---
title: "Cell Differentiation Study"
subtitle: "Poisson GLM with Interaction Effects"
date: today
nocite: |
  @bartlett_deviance_2014
  @conflicted_package
  @dplyr_package
  @fahrmeir_package
  @gee_package
  @geepack_package_2006
  @noauthor_poisson_2026
  @venables_modern_2002
---

## TODO

- Add reference to MASS book - effectively the user manual for a lot of R functions
- Standardise the quasi Poisson notation.
- Add more direct tables from the book.
- Fix the working residuals file
- Fix titles and subtitles throughout

## Overview

This script demonstrates:

- **Poisson regression** for count data (cell differentiation)
- Testing **synergistic vs. additive effects** of two treatment factors
- Model diagnostics using **deviance and Pearson statistics**
- **Dispersion estimation** for overdispersion assessment

The study investigates whether TNF (Tumor Necrosis Factor) and IFN (Interferon)
agents stimulate cell differentiation **synergistically or independently**.

## Package Setup and Data

As in the previous script, for safety we import the `conflicted` package first
to manage any function name conflicts.

```{r}
library(conflicted)
```

Then we import the other packages we need to complete the analysis.

```{r}
library(dplyr)
library(Fahrmeir)
library(gee)
library(geepack)
library(MASS)
```

Note that the MASS package is documented in the book @venables_modern_2002
[and is available for free for unimelb students at this link](https://go.openathens.net/redirector/unimelb.edu.au?url=https%3A%2F%2Flink.springer.com%2F10.1007%2F978-0-387-21706-2).

We can import and preview the `cells` data from the `Fahrmeir` package.
As always, it is good practice to inspect the dataframe to get an idea of the structure,
and types of the data.

```{r}
data(cells)
cells_tbl <- as_tibble(cells)

cells_tbl
```

All the data are encoded as integers.

## Example 2.3

The purpose of this data is to determine whether the two agents stimulate cell differentiation synergistically (there is an interaction between the agents) or independently.

We initially just fit the Poisson GLM as in Ex 2.3.
Note that `y ~ TNF * IFN` automatically includes the intercept terms and individual covariates for `TNF` and `IFN`.
It is equivalent to `y ~ 1 + TNF + IFN + TNF * IFN`.
The default link is the log link, meaning that we assume $\log(\texttt{y})$ depends linearly on the covariates.

```{r}
glm_poisson <- glm(formula = y ~ TNF * IFN, family = poisson, data = cells_tbl)

glm_poisson_summary <- summary(glm_poisson)
```

We can inspect the whole summary object directly with `glm_summary`,
or just the coefficients (appropriately rounded) to compare directly to @fahrmeir_multivariate_2001 page 37.
```{r}
round(glm_poisson_summary$coeff, digits = 4)
```

We see that these estimates line up appropriately.
The interactive effect has a small coefficient, suggesting that perhaps there may be no true effect,
however the $p$-value is very small.
We continue to explore this in @fahrmeir_multivariate_2001 Example 2.6.

## Example 2.6 (page 54)

We must check whether the model is appropriate for the data.
If not, this could explain why the $p$-value is small.

We explore and test the deviance using an asymptotic $\chi^2$ test.

::: {.callout-tip collapse="true"}
## What should the degrees of freedom be for this test? Expand for answer.

There must be $16 - 4 = 12$ degrees of freedom, since the saturated model has 16 parameters,
and the model we are testing has 4.
:::

We directly access the deviance with

```{r}
glm_poisson$deviance
```

which is identical to the book.

We are also able to access the scaled covariance and the unscaled covariance matrices.

```{r}
glm_poisson_summary$cov.scaled
glm_poisson_summary$cov.unscaled
```

::: {.callout-tip collapse="true"}
## Why are these covariance matrices identical? Expand for answer.

The scaled covariance matrix is scaled by the dispersion parameter $\phi$,
but the Poisson model has a fixed dispersion parameter $\phi= 1$.
:::

## Example 2.6

We calculate the Pearson $\chi^2$ statistic:

```{r}
pearson_chisq_stat <-
    sum(
        (glm_poisson$y - glm_poisson$fitted.values)^2 /
            glm_poisson$fitted.values
    )
pearson_chisq_stat
```

which is the same as in @fahrmeir_multivariate_2001 page 54.

When we try to form an estimate $\hat{\phi}$ for the dispersion parameter $\phi$,
we divide the Pearson $\chi^2$ statistic by the degrees of freedom.

```{r}
est_disperson <- pearson_chisq_stat / 12
round(est_disperson, 3)
```

::: {.callout-note}
This is not exactly the same as the book (which has 11.734), but it is unknown why.
:::

This gives us an idea that the data may in fact be very overdispersed,
and hence model may not be a good fit.

```{r}
1 - pchisq(glm_poisson$deviance, df = 12)
1 - pchisq(pearson_chisq_stat, df = 12)
```

This is confirmed by both the deviance test and the Pearson $\chi^2$ tests,
which reject the model.

:::{.callout-warning}
There is an important caveat to this result.
The sample size of the data is $16$, since the data isn't actually grouped.
Since this is equal to the degrees of freedom in the saturated model,
the dimensionality of the model grows with more samples
(which must necessarily be new unique combinations of the covariates).
Therefore the asymptotics are not fully valid, since we don't have a stabilising null distribution.
For more information see what the biostatistician Jonathan Bartlett says in
[his blog post](https://thestatsgeek.com/2014/04/26/deviance-goodness-of-fit-test-for-poisson-regression/).

There is a case to redeem the tests when the mean of the Poisson random variables are large.
Heuristically, this is because the Poisson distribution looks normal
(see the [Wikipedia page on the Poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution)
for a visual demonstration), and for the normal distribution, the chi-squared distribution is exact.
For more information see also @venables_modern_2002 page 187.
:::

```{r}
est_disperson <- pearson_chisq_stat / glm_poisson$df.residual
est_disperson
```

So now we perform quasi "model-based" (also known as "naive") approach
by assuming the variance structure ($\sigma^2(\mu) = \phi\mu$) is correctly specified:

```{r}
glm_quasi_poisson <- glm(
    formula = y ~ TNF * IFN,
    family = quasipoisson,
    data = cells_tbl
)

glm_quasi_poisson_summary <- summary(glm_quasi_poisson)
glm_quasi_poisson_summary
```

Note that since the variance structure is assumed to be correct,
the estimates are not derived using the sandwich estimator.

We can compare the coefficients between the from the full Poisson model.

```{r}
round(glm_poisson_summary$coefficients, 5)
round(glm_quasi_poisson_summary$coefficients, 5)
```

We see that the coefficients are the exact same, but the standard errors are significantly different.

The estimating equations in both the standard and quasi Poisson models are the same,
giving us identical coefficient estimates, but the dispersion parameter effects our standard errors.

The following functions still work with the quasi Poisson model.

```{r}
#| eval: false
anova(glm_quasi_poisson)
resid(glm_quasi_poisson)
resid(glm_quasi_poisson, type = "deviance")
resid(glm_quasi_poisson, type = "working")
resid(glm_quasi_poisson, type = "pearson")
```

These two are different now, since the scaled covariance should be $\phi$ times the unscaled covariance.

```{r}
glm_quasi_poisson_summary$cov.scaled
glm_quasi_poisson_summary$cov.unscaled
```

We can do a sanity check in order to determine if we can recover the scaled deviance.

```{r}
glm_poisson_summary$cov.unscaled * glm_quasi_poisson_summary$dispersion
```

We check also that the dispersion estimate reported by `glm()` matches our own calculation based on Pearson residuals.

```{r}
est_disperson
glm_quasi_poisson_summary$dispersion
```

It is close but not exactly the same. We can recover the exact number reported by `glm()` as follows:

```{r}
sum(glm_quasi_poisson$weights * glm_quasi_poisson$residuals^2) /
    12
```

:::{.callout-warning}
The dispersion estimate provided by `glm()` actually leverages the "working residuals" and the IWLS weights! So it is not the same as our standard estimate based on Pearson residuals! Note that the weights here are same as `glm_poisson$weights`, the weights by assuming $\phi = 1$ under the full likelihood model.

For a detailed mathematical explanation of why this difference exists, see [Appendix: Estimating the Dispersion with Working Residuals](A1-dispersion-working-residuals.qmd).
:::


## Example 2.7

Plotting variance against mean by TNF. Note that `var()` computes the unbiased variance estimate by default; to reproduce the results on p.59 of @fahrmeir_multivariate_2001, we have to use the biased variance estimate.

```{r}
cells_summary <- cells_tbl |>
    group_by(TNF) |>
    summarize(
        est_var = var(y) * 3 / 4,
        est_mean = mean(y),
        .groups = "drop"
    )
plot(
    cells_summary$est_mean,
    cells_summary$est_var,
    xlab = "mean",
    ylab = "variance"
)
```

Use the gee function for the "robust inference" in this example instead. This should get the second column of table 2.7 corresponding to the linear working variance structure. The dispersion provided by gee is the same as our own estimated dispersion based on Pearson residuals!

```{r}
gee_mu <- gee(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = quasi(variance = "mu", link = "log"),
    data = cells_tbl
)

summary(gee_mu)

summary(gee_mu)$coeff
```

The following should be the robust z-values, i.e. estimates rescaled by sandwich variance estimate.

```{r}
robust_z <- gee_mu$coeff / sqrt(diag((gee_mu)$robust.variance))
round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
```

The following should be the estimates rescaled by model-based variance estimate.

```{r}
naivez <- gee_mu$coeff / sqrt(diag((gee_mu)$naive.variance))
round(2 * pnorm(abs(naivez), lower.tail = FALSE), 3)
```

Try to fit with geeglm from the geepack instead.

```{r}
gee_mu_geepack <- geeglm(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = poisson(link = "log"),
    data = cells_tbl,
    scale.fix = FALSE
)

gee_mu_geepack_fix <- geeglm(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = poisson(link = "log"),
    data = cells_tbl,
    scale.fix = TRUE
)
```

I think the second column in Table 2.7 is not correctly printed; they have printed the naive $p$-values, which should have been the robust $p$-values.

Now the third column of table 2.7 corresponding to the purely quadratic working variance structure. The robust $p$-values match the ones in @fahrmeir_multivariate_2001, but not the naive ones, so all the $p$-values reported in parentheses of Table 2.7 in @fahrmeir_multivariate_2001 should have been robust $p$-values.

```{r}
glm_gee_mu2 <- gee(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = quasi(variance = "mu^2", link = "log"),
    data = cells_tbl
)

summary(glm_gee_mu2)$scale
robust_z <- glm_gee_mu2$coeff / sqrt(diag(glm_gee_mu2$robust.variance))
round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
naivez <- glm_gee_mu2$coeff / sqrt(diag(glm_gee_mu2$naive.variance))
round(2 * pnorm(abs(naivez), lower.tail = FALSE), 3)
```

Bibliography

@fahrmeir_multivariate_2001
