---
title: "Cell Differentiation Study"
subtitle: "Poisson GLM with Interaction Effects"
date: today
nocite: |
  @conflicted_package
  @dplyr_package
  @fahrmeir_package
  @gee_package
  @geepack_package_2006
  @venables_modern_2002
---

## Overview

This script demonstrates:

- **Poisson regression** for count data (cell differentiation)
- Testing **synergistic vs. additive effects** of two treatment factors
- Model diagnostics using **deviance and Pearson statistics**
- **Dispersion estimation** for overdispersion assessment

The study investigates whether TNF (Tumor Necrosis Factor) and IFN (Interferon)
agents stimulate cell differentiation **synergistically or independently**.

## Package Setup and Data

As in the previous script, for safety we import the `conflicted` package first
to manage any function name conflicts.

```{r}
library(conflicted)
```

Then we import the other packages we need to complete the analysis.

```{r}
library(dplyr)
library(Fahrmeir)
library(gee)
library(geepack)
library(MASS)
```

We can import and preview the `cells` data from the `Fahrmeir` package.
As always, it is good practice to inspect the dataframe to get an idea of the structure,
and types of the data.

```{r}
data(cells)
cells_tbl <- as_tibble(cells)

cells_tbl
```

All the data are encoded as integers.

## Example 2.3

Aims to study whether the two agents stimulate cell differentiation synergistically or independently.

We initially just fit the Poisson GLM as in Ex 2.3.
Note that `y ~ TNF * IFN` automatically includes the intercept terms and individual covariates for `TNF` and `IFN`.
It is equivalent to `y ~ 1 + TNF + IFN + TNF * IFN`.

```{r}
glm_poisson <- glm(formula = y ~ TNF * IFN, family = poisson, data = cells_tbl)

glm_poisson_summary <- summary(glm_poisson)
```

We can inspect the whole summary object directly with `glm_summary`,
or just the coefficients (appropriately rounded) to compare directly to @fahrmeir_multivariate_2001 page 37.
```{r}
round(glm_poisson_summary$coeff, digits = 4)
```

We see that these estimates line up appropriately.

## Example 2.6

We now continue with this dataset as in Example 2.6.
We can see from the data above that the $p$-value for the interaction term is very small.
However before we jump to the conclusion that therefore it must be significant,
we consider the possibility that the model may not be appropriate for the data.

```{r}
anova(glm_poisson)
round(glm_poisson$coeff, digits = 4)
glm_poisson$deviance
```

## Covariance Matrices of Beta

These two are the same, which makes sense under the exponential family assumption is true.

```{r}
glm_poisson_summary$cov.scaled
glm_poisson_summary$cov.unscaled
```

## Covariance Matrices of Beta

These two are the same, which makes sense under the exponential family assumption is true.

```{r}
glm_poisson_summary$cov.scaled
glm_poisson_summary$cov.unscaled
```

## Example 2.6

Estimated Pearson statistic, same as in Eg 2.6. Both reject the model, but we should take these results with a grain of salt.

```{r}
pearson_chisq_stat <-
    sum(
        (glm_poisson$y - glm_poisson$fitted.values)^2 /
            glm_poisson$fitted.values
    )

1 - pchisq(glm_poisson$deviance, df = 12)
1 - pchisq(pearson_chisq_stat, df = 12)
```

The estimated dispersion in Example 2.6, very close to the book.

```{r}
est_disperson <- pearson_chisq_stat / glm_poisson$df.residual
est_disperson
```

Perform quasi "model-based" (also known as "naive") approach assuming the variance structure is correctly specified:

```{r}
glm_quasi_poisson <- glm(
    formula = y ~ TNF * IFN,
    family = quasipoisson,
    data = cells_tbl
)

summary(glm_quasi_poisson)

summary(glm_quasi_poisson)$coefficients
anova(glm_quasi_poisson)
resid(glm_quasi_poisson)
resid(glm_quasi_poisson, type = "deviance")
resid(glm_quasi_poisson, type = "working")
resid(glm_quasi_poisson, type = "pearson")
```

These two are different now by the scale/dispersion.

```{r}
summary(glm_quasi_poisson)$cov.scaled
summary(glm_quasi_poisson)$cov.unscaled

glm_poisson_summary$cov.unscaled * summary(glm_quasi_poisson)$dispersion
```

**WARNING**: the dispersion estimate provided by glm() actually leverages the "working residuals" and the IWLS weights! So it is not the same as our standard estimate based on Pearson residuals! Note that the weights here are same as `glm_poisson$weights`, the weights by assuming phi = 1 under the full likelihood model.

```{r}
summary(glm_quasi_poisson)$dispersion

with(glm_quasi_poisson, sum(weights * residuals^2)) /
    glm_quasi_poisson$df.residual
```

**WARNING**: the dispersion estimate provided by glm() actually leverages the "working residuals" and the IWLS weights! So it is not the same as our standard estimate based on Pearson residuals! Note that the weights here are same as `glm_poisson$weights`, the weights by assuming phi = 1 under the full likelihood model.

```{r}
summary(glm_quasi_poisson)$dispersion

with(glm_quasi_poisson, sum(weights * residuals^2)) /
    glm_quasi_poisson$df.residual
```

## Example 2.7

Plotting variance against mean by TNF. Note that `var()` computes the unbiased variance estimate by default; to reproduce the results on p.59 of @fahrmeir_multivariate_2001, we have to use the biased variance estimate.

```{r}
cells_tbl_grouped_tnf <- group_by(cells_tbl, TNF)
cells_summary <- summarize(
    cells_tbl_grouped_tnf,
    est_var = var(y) * 3 / 4,
    est_mean = mean(y),
    .groups = "drop"
)
plot(
    cells_summary$est_mean,
    cells_summary$est_var,
    xlab = "mean",
    ylab = "variance"
)
```

Use the gee function for the "robust inference" in this example instead. This should get the second column of table 2.7 corresponding to the linear working variance structure. The dispersion provided by gee is the same as our own estimated dispersion based on Pearson residuals!

```{r}
gee_mu <- gee(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = quasi(variance = "mu", link = "log"),
    data = cells_tbl
)

summary(gee_mu)

summary(gee_mu)$coeff
```

The following should be the robust z-values, i.e. estimates rescaled by sandwich variance estimate.

```{r}
robust_z <- gee_mu$coeff / sqrt(diag((gee_mu)$robust.variance))
round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
```

The following should be the estimates rescaled by model-based variance estimate.

```{r}
naivez <- gee_mu$coeff / sqrt(diag((gee_mu)$naive.variance))
round(2 * pnorm(abs(naivez), lower.tail = FALSE), 3)
```

Try to fit with geeglm from the geepack instead.

```{r}
gee_mu_geepack <- geeglm(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = poisson(link = "log"),
    data = cells_tbl,
    scale.fix = FALSE
)

gee_mu_geepack_fix <- geeglm(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = poisson(link = "log"),
    data = cells_tbl,
    scale.fix = TRUE
)
```

I think the second column in Table 2.7 is not correctly printed; they have printed the naive $p$-values, which should have been the robust $p$-values.

Now the third column of table 2.7 corresponding to the purely quadratic working variance structure. The robust $p$-values match the ones in @fahrmeir_multivariate_2001, but not the naive ones, so all the $p$-values reported in parentheses of Table 2.7 in @fahrmeir_multivariate_2001 should have been robust $p$-values.

```{r}
glm_gee_mu2 <- gee(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells_tbl)),
    family = quasi(variance = "mu^2", link = "log"),
    data = cells_tbl
)

summary(glm_gee_mu2)$scale
robust_z <- glm_gee_mu2$coeff / sqrt(diag(glm_gee_mu2$robust.variance))
round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
naivez <- glm_gee_mu2$coeff / sqrt(diag(glm_gee_mu2$naive.variance))
round(2 * pnorm(abs(naivez), lower.tail = FALSE), 3)
```

Bibliography

@fahrmeir_multivariate_2001
