---
title: "(Quasi-)Poisson GLM: Cell Differentiation Study"
subtitle: "@fahrmeir_multivariate_2001 Examples 2.3, 2.6, and 2.7"
date: today
nocite: |
  @bartlett_deviance_2014
  @noauthor_poisson_2026
  @venables_modern_2002
---

## Package Setup and Data

As in the previous script, for safety we import the `conflicted` package first
to manage any function name conflicts.

```{r}
#| label: 02-celldiff-chunk-1
library(conflicted)
```

We import the other packages needed to complete the analysis.

```{r}
#| label: 02-celldiff-chunk-2
library(dplyr)
library(Fahrmeir)
library(gee)
library(geepack)
library(ggplot2)
library(MASS)
```

Thorough MASS documentation is available in the book
[Modern Applied Statistics with S (MASS)](https://go.openathens.net/redirector/unimelb.edu.au?url=https%3A%2F%2Flink.springer.com%2F10.1007%2F978-0-387-21706-2){target="_blank"}
 @venables_modern_2002
which is available to you for free.

## Example 2.3 (p. 37): Cell Differentiation Data

### Load and Inspect Data

We now attempt to replicate the results in @fahrmeir_multivariate_2001 Example 2.3 (p. 37).
We import and preview the `cells` data from the `Fahrmeir` package.
The purpose of this dataset is to explore whether the two agents
stimulate cell differentiation synergistically (is an interaction between the agents)
or independently.

```{r}
#| label: 02-celldiff-chunk-3
data(cells)
cells_tbl <- as_tibble(cells)
cells_tbl
```

All the variables are encoded as integers
and so we do not need to manipulate the data types.
It is also good practice in RStudio to run `?cells` to inspect
[the documentation for the dataset](https://search.r-project.org/CRAN/refmans/Fahrmeir/html/cells.html){target="_blank"}.

### Fit Poisson GLM Model

We initially just fit the Poisson GLM as in Ex 2.3.

::: {.callout-note collapse="true"}
## What is the canonical link function for Poisson regression? Expand for answer.

The default link is the log link,
meaning that we assume $\log(\mu)$ depends linearly on the covariates.
:::

```{r}
#| label: 02-celldiff-chunk-4
glm_poisson <- glm(formula = y ~ TNF * IFN, family = poisson, data = cells_tbl)
```

::: {.callout-tip collapse="true"}
## glm() Syntax Explanation

- `formula = y ~ TNF * IFN`:
Response variable `y` modeled with main effects and interaction for `TNF` and `IFN`.
This is equivalent to `y ~ 1 + TNF + IFN + TNF * IFN`.
- `family = poisson`: Specifies Poisson distribution with default log link function
- `data = cells_tbl`: Data frame containing the variables
:::

### Inspect Results

We inspect the returned coefficients.

```{r}
#| label: 02-celldiff-chunk-5
glm_poisson_summary <- summary(glm_poisson)
round(glm_poisson_summary$coefficients, digits = 4)
```

We see that these estimates line up with those reported
in @fahrmeir_multivariate_2001 p. 38.
Two things should catch our eye as requiring further investigation:

1. the interactive effect has a small coefficient
suggesting that perhaps **there may be no true effect**,
2. the $p$-value for the interaction is very small,
suggesting that **there is a true effect**.

We continue to explore this in @fahrmeir_multivariate_2001 Example 2.6.

## Example 2.6 (p. 54): Goodness-of-Fit and Overdispersion

A simple explanation for the apparent contradiction in the previous example is that
the data is not well explained by the Poisson model.
We test the goodness of fit with the deviance and the Pearson $\chi^2$ statistic
using an asymptotic $\chi^2$ test.

::: {.callout-note collapse="true"}
## What will the degrees of freedom be for this test? Expand for answer.

There must be $16 - 4 = 12$ degrees of freedom,
since the saturated model has 16 parameters,
and the model we are testing has 4.
:::

### Test Goodness of Fit

We can directly access the deviance
and we can calculate the Pearson chi-squared statistic with
```{r}
#| label: 02-celldiff-chunk-6
poisson_deviance <- glm_poisson$deviance
round(poisson_deviance, 1)
pearson_chisq_stat <-
  sum(
    (glm_poisson$y - glm_poisson$fitted.values)^2 /
      glm_poisson$fitted.values
  )
round(pearson_chisq_stat, 1)
```

which are identical to the values reported on page 54 of @fahrmeir_multivariate_2001.

We inspect the covariance matrices of the coefficient estimates
```{r}
#| label: 02-celldiff-chunk-7
glm_poisson_summary$cov.scaled
glm_poisson_summary$cov.unscaled
```

::: {.callout-note collapse="true"}
## Why are these covariance matrices identical? Expand for answer.

The scaled covariance matrix is scaled by the dispersion parameter $\phi$,
but the Poisson model has a fixed dispersion parameter $\phi= 1$.
:::

We can look for overdispersion as evidence of poor fit.
The typical way of creating an estimate $\hat{\phi}$ for the dispersion parameter $\phi$,
is by dividing the Pearson $\chi^2$ statistic by the degrees of freedom.

```{r}
#| label: 02-celldiff-chunk-8
est_dispersion <- pearson_chisq_stat / 12
round(est_dispersion, 3)
```

Heuristically, this is not close to a value of 1,
which is what we would expect if the Poisson model was a good fit.

::: {.callout-caution}
For an unknown reason our $\hat{\phi}$
is not exactly the same as the book (which has 11.734),
but it is very close.
:::

We now formally test the goodness of fit
by looking at the $p$-values of the deviance and Pearson $\chi^2$ statistics.

```{r}
#| label: 02-celldiff-chunk-9
1 - pchisq(poisson_deviance, df = 12)
1 - pchisq(pearson_chisq_stat, df = 12)
```

::: {.callout-important collapse="true"}
## What is the correct interpretation of these results? Expand for answer.

Both the deviance test and the Pearson $\chi^2$ tests reject the null hypothesis
that the data are produced from the Poisson model
at any reasonable level of significance.
:::

:::{.callout-warning}
There is an important caveat to this result.
The sample size of the data is $16$, since the data isn't actually grouped.
Since this is equal to the degrees of freedom in the saturated model,
the dimensionality of the model grows with more samples.
Therefore the asymptotics are not fully valid, since we don't have a stabilising null distribution.
For more information see what the biostatistician Jonathan Bartlett says in
[his blog post](https://thestatsgeek.com/2014/04/26/deviance-goodness-of-fit-test-for-poisson-regression/){target="_blank"}.

There is a case to redeem the tests when the mean of the Poisson random variables are large
(given by @fahrmeir_multivariate_2001 p. 54 in this example).
Heuristically, this is because the Poisson distribution looks normal
(see the [Wikipedia page on the Poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution){target="_blank"}
for a visual demonstration), and for the normal distribution, the chi-squared distribution is exact.
For more information see also @venables_modern_2002 page 187.
:::

### Fit a Quasi-Poisson GLM Model

Having rejected the Poisson model, we move to a quasi "model-based"
(also known as "naive") approach
by loosening the variance structure, but still assuming we can correctly specify it.
We start with a linear and overdispersed quasi-Poisson variance structure
$\sigma^2(\mu) = \phi\mu$,
and we will try to replicate the results of Table 2.6 on page 54 of @fahrmeir_multivariate_2001 copied below (we have corrected the coefficient estimate -0.001 reported for `TNF*IFN`in the book to -0.0001 according to our own fits),
showing the comparison between the two models:

| Parameter | Poisson, $\phi = 1$ | Poisson, $\hat{\phi} = 11.734$  |
|:---------:|--------------------:|--------------------------------:|
| `1`       | 3.436 (.0)          | 3.436 (.0)                      |
| `TNF`     | .016 (.0)           | .016 (.0)                       |
| `IFN`     | .009 (.0)           | .009 (.0)                       |
| `TNF*IFN` | -.0001 (.0)          | -.0001 (.22)                     |

: Log-linear model fits to cellular differentiation data based on Poisson-likelihoods
[Adapted from @fahrmeir_multivariate_2001, Table 2.6, p. 54] {#tbl-poisson-compare}

```{r}
#| label: 02-celldiff-chunk-10
glm_quasi_poisson <- glm(
  formula = y ~ TNF * IFN,
  family = quasipoisson,
  data = cells_tbl
)

glm_quasi_poisson_summary <- summary(glm_quasi_poisson)
glm_quasi_poisson_summary
```

Note that since the variance structure is assumed to be correct,
the standard errors are not derived using the more robust sandwich covariance matrix.

### Inspect Quasi-Poisson Results

We can compare the coefficients between the from the full Poisson model.

```{r}
#| label: 02-celldiff-chunk-11
round(glm_poisson_summary$coefficients, 5)
round(glm_quasi_poisson_summary$coefficients, 5)
```

We see that the coefficients are the exact same and match @tbl-poisson-compare,
but the standard errors between the two models are significantly different.
This is because the estimating equations
in both the standard and quasi-Poisson models are identical,
giving us identical coefficient estimates,
but the dispersion parameter effects our standard errors.

Moreover, the $p$-values reported for the coefficients in `glm_quasi_poisson_summary` are calibrated based on the $t_{n-p}= t_{16-4}$ distribution, which are a bit different from those in @tbl-poisson-compare calibrated only based on a standard normal distribution. For example, taking the rescaled coefficient estimate $-1.22748$ for `TNF:IFN`, calibrating based on the standard normal gives `2*(pnorm(1.22748, lower.tail = F)) = 0.2196422`, while calibrating based on $t_{12}$ gives `2*(pt(1.22748, df = 12,  lower.tail = F)) = 0.2431777`.

As with the typical `glm()` and `lm()` analysis,
we can still use the standard `anova()` function,
and extract the various residuals:

```{r}
#| label: 02-celldiff-chunk-12
residuals_tbl <- tibble(
  default = resid(glm_quasi_poisson),
  deviance = resid(glm_quasi_poisson, type = "deviance"),
  working = resid(glm_quasi_poisson, type = "working"),
  pearson = resid(glm_quasi_poisson, type = "pearson")
)
residuals_tbl
```

We now also expect the scaled covariance matrix
to be different from the unscaled covariance matrix,
since the dispersion parameter is no longer fixed to 1.

```{r}
#| label: 02-celldiff-chunk-13
glm_quasi_poisson_summary$cov.unscaled
glm_quasi_poisson_summary$cov.scaled
```

We can do a sanity check in order to determine if we can recover the scaled covariance matrix

```{r}
#| label: 02-celldiff-chunk-14
glm_poisson_summary$cov.unscaled * glm_quasi_poisson_summary$dispersion
```

This looks good!


:::{.callout-caution}

## Dispersion estimation with `glm()` does not match our Pearson residuals estimate

We can see that the dispersion estimate reported by `glm()`
does not exactly match our own calculation based on Pearson residuals.

```{r}
#| label: 02-celldiff-chunk-15
est_dispersion
glm_quasi_poisson_summary$dispersion
```

We can recover the exact number reported by `glm()` as follows:

```{r}
#| label: 02-celldiff-chunk-16
sum(glm_quasi_poisson$weights * glm_quasi_poisson$residuals^2) / 12
```

The dispersion estimate provided by `glm()` actually leverages the
"working residuals" and the IWLS weights.
So it is not the same as our standard estimate based on Pearson residuals!
Note that the weights here are same as `glm_poisson$weights`,
the weights by assuming $\phi = 1$ under the full likelihood model.

For a detailed mathematical explanation of why this difference exists, see [Appendix: Estimating the Dispersion with Working Residuals](A1-dispersion-working-residuals.qmd).
:::

## Example 2.7: Variance Functions and GEE

### Assess Variance-Mean Relationship

We can further explore different variance structures for the data,
beyond just the linear variance structure $\sigma^2(\mu) = \phi \mu$.

We can get some visual ideas by plotting the variance against mean per covariate.

:::{.callout-caution}
`var()` computes the unbiased variance estimate by default.
To visually reproduce the results on p. 59 of @fahrmeir_multivariate_2001,
we have to use the biased variance estimate.
:::

```{r}
#| label: 02-celldiff-chunk-17
variance_summary_by_TNF <- cells_tbl |>
  group_by(TNF) |>
  summarize(
    est_var = var(y) * 3 / 4,
    est_mean = mean(y),
    .groups = "drop"
  )
ggplot(variance_summary_by_TNF, aes(x = est_mean, y = est_var, color = factor(TNF))) +
  geom_point(size = 3) +
  labs(
    x = "Mean",
    y = "Biased Variance",
    color = "TNF Level"
  ) +
  theme_minimal()
```

We can do a similar plot by `IFN` instead of `TNF`:

```{r}
#| label: 02-celldiff-chunk-18
variance_summary_by_IFN <- cells_tbl |>
  group_by(IFN) |>
  summarize(
    est_var = var(y) * 3 / 4,
    est_mean = mean(y),
    .groups = "drop"
  )
ggplot(variance_summary_by_IFN, aes(x = est_mean, y = est_var, color = factor(IFN))) +
  geom_point(size = 3) +
  labs(
    x = "Mean",
    y = "Biased Variance",
    color = "IFN Level"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
## ggplot2 Syntax Explanation

- `ggplot(data, aes(...))`: Initialize plot with aesthetic mappings (x, y, color, size, etc.)
  - `color = factor(TNF)`: Convert numeric to factor for discrete coloring
- `geom_point()`: Add scatter plot layer
- `labs()`: Set axis and legend labels (`color = "TNF Level"` customizes legend title)
- `theme_minimal()`: Apply clean theme
- Layers combined with `+` (similar to piping)

The equivalent base R code to create the same scatter plot would be:
```{r}
#| label: 02-celldiff-chunk-19
#| eval: false
plot(
  variance_summary_by_TFN$est_mean,
  variance_summary_by_TFN$est_var,
  xlab = "Mean",
  ylab = "Biased Variance"
)
```
:::

The non-linear relationship between the variance and the means suggests that we should consider some non-linear variance structures.
In particular, we compare:

- $\sigma^2(\mu) = \phi \mu$ (linear)
- $\sigma^2(\mu) = \phi \mu^2$ (quadratic)
- $\sigma^2(\mu) = \phi \mu + {\phi}_2 \mu^2$ (linear + quadratic---not explored here).

We leave the linear + quadratic case for further exploration.

### Fit GEE with Linear Variance

We will now use the `gee()` function for "robust inference".
This means that we do not assume that the variance structure is correctly specified,
but instead use the sandwich variance estimator for inference.
We now try to partially replicate Table 2.7 from @fahrmeir_multivariate_2001,
adapted below:

| Parameter     | $\sigma^2(\mu) = \phi\mu$ | $\sigma^2(\mu) = \phi\mu^2$   |
|:-------------:|--------------------------:|------------------------------:|
| `1`           | 3.436 (0.0)               | 3.394 (0.0)                   |
| `TNF`         | 0.016 (0.0)               | 0.016 (0.0)                   |
| `IFN`         | 0.009 (0.0)               | 0.009 (0.003)                 |
| `TNF*IFN`     | -0.001 (0.22)             | -0.001 (0.099)                |
| $\hat{\phi}$  | 11.734                    | 0.243                         |

: Log-linear model fits to cellular differentiation data based on quasi-likelihoods
(Table 2.7, p. 60)

We start by considering the linear variance structure $\sigma^2(\mu) = \phi \mu$.

```{r}
#| label: 02-celldiff-chunk-20
gee_mu <- gee(
  formula = y ~ TNF * IFN,
  id = seq_len(nrow(cells_tbl)),
  data = cells_tbl,
  family = quasi(variance = "mu", link = "log")
)
```

::: {.callout-tip collapse="true"}
## `gee()` Syntax Explanation

- `id`: Cluster identifier for which observations are correlated.
Here `seq_len(nrow(cells_tbl))` creates unique IDs (1, 2, ..., n) for each row,
treating observations as independent
- `family = quasi(variance = "...", link = "...")`: Specifies the assumed variance function and link function
  - `variance = "mu"`: Linear variance structure $\sigma^2(\mu) = \phi \mu$
  - `link = "log"`: Log link function
:::

```{r}
#| label: 02-celldiff-chunk-21
gee_mu_summary <- summary(gee_mu)
gee_mu_summary
```

We can see now that the dispersion estimate is the same
as the one we manually calculated based on Pearson residuals.
We compare the coefficients and standard errors to the previous models:

```{r}
#| label: 02-celldiff-chunk-22
round(gee_mu_summary$coefficients, 4)
round(glm_quasi_poisson_summary$coefficients, 4)
```

The coefficients are be identical to the previous models,
since the same score function is being solved.
We can see that the naive standard errors
are those assuming the variance structure is correct,
while the robust standard errors are those using the sandwich variance estimator.

::: {.callout-note collapse="true"}
## Why use the sandwich variance estimator? Expand for answer.

The sandwich variance estimator is "robust"
because it is guaranteed to be asymptotically correct
even if the variance structure is misspecified.
If we incorrectly assume $\sigma^2(\mu) = \phi\mu$
when the true structure is different,
the naive standard errors will be biased,
but the sandwich estimator will still provide valid inference.
:::

We can then calculate the $p$-values
corresponding to the naive and robust standard errors.

:::{.callout-tip collapse="true"}
## Understanding the z-value Calculation

`diag(...)` extracts the diagonal elements from a matrix (variance-covariance matrix in this case).
Standard errors are obtained by taking the square root of these diagonal elements:
`sqrt(diag(covariance_matrix))`.
Dividing the coefficient by its standard error gives the z-value.
:::

Assuming we have correctly specified the variance structure (naive approach):

```{r}
#| label: 02-celldiff-chunk-23
naive_z_values <- tibble(
  coefficient = names(gee_mu$coefficients),
  z_value = gee_mu$coefficients / sqrt(diag((gee_mu)$naive.variance)),
  p_value = round(2 * pnorm(abs(z_value), lower.tail = FALSE), 3)
)
naive_z_values
```

Without assuming the variance structure is correct (robust approach):

```{r}
#| label: 02-celldiff-chunk-24
robust_z_values <- tibble(
  coefficient = names(gee_mu$coefficients),
  z_value = gee_mu$coefficients / sqrt(diag((gee_mu)$robust.variance)),
  p_value = round(2 * pnorm(abs(z_value), lower.tail = FALSE), 3)
)
robust_z_values
```

:::{.callout-warning}
Note that in @fahrmeir_multivariate_2001 Table 2.7,
the $p$-values of the $\sigma^2(\mu) = \phi \mu$ column
corresponds to the naive $p$-values,
whereas they should actually correspond to the robust $p$-values (read ahead below).
:::

Expand the following note for an alternative implementation of GEE models
using the `geepack` package.

:::{.callout-note collapse="true"}
## Alternative `geepack` Implementation

There is an alternative package `geepack` which also implements GEE models.
Under the hood, there are some small differences in implementation between `gee` and `geepack`.

```{r}
#| label: 02-celldiff-chunk-25
geepack_mu <- geeglm(
  formula = y ~ TNF * IFN,
  family = poisson(link = "log"),
  data = cells_tbl,
  id = seq_len(nrow(cells_tbl))
)
geepack_mu
geepack_mu_summary <- summary(geepack_mu)
geepack_mu_summary
```

:::

### Fit GEE with Quadratic Variance

Finally, we try the quadratic variance structure $\sigma^2(\mu) = \phi \mu^2$.

```{r}
#| label: 02-celldiff-chunk-26
gee_mu2 <- gee(
  formula = y ~ TNF * IFN,
  id = seq_len(nrow(cells_tbl)),
  family = quasi(variance = "mu^2", link = "log"),
  data = cells_tbl
)
gee_mu2_summary <- summary(gee_mu2)
gee_mu2_summary
round(gee_mu2_summary$scale, 3)
```

The coefficients almost match those in @fahrmeir_multivariate_2001 Table 2.7 on page 60.

For a final sanity check, we manually compute both the naive and robust $Z$ and $p$-values.

```{r}
#| label: 02-celldiff-chunk-27
gee_mu2_inference <- tibble(
  coefficient = names(gee_mu2$coefficients),
  naive_z = gee_mu2$coefficients / sqrt(diag(gee_mu2$naive.variance)),
  naive_p = round(2 * pnorm(abs(naive_z), lower.tail = FALSE), 3),
  robust_z = gee_mu2$coefficients / sqrt(diag(gee_mu2$robust.variance)),
  robust_p = round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
)
gee_mu2_inference
```

Unlike for the linear variance structure,
for the quadratic variance structure,
the robust $p$-values match the ones in @fahrmeir_multivariate_2001.
This confirms that all the $p$-values
reported in parentheses of Table 2.7 of @fahrmeir_multivariate_2001
should have been robust $p$-values.

```{r}
#| label: 02-celldiff-chunk-28
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "Fahrmeir", "gee", "geepack", "ggplot2", "MASS")
for (pkg in packages) {
  cit <- citation(pkg)

  # Check if multiple citations
  if (length(cit) > 1) {
    cat("- **", pkg, "**:\n", sep = "")
    for (i in seq_along(cit)) {
      cit_text <- format(cit[i], style = "text")
      cat("    - ", cit_text, "\n\n", sep = "")
    }
  } else {
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
  }
}
```
