---
title: "Example 3.1 & 3.11: Caesar Dataset Analysis"
subtitle: "Multinomial Logistic Regression and Category-Specific Effects"
date: today
nocite: |
    @fahrmeir_multivariate_2001
    @venables_modern_2002
---

## TODO

- Go through and add appropriate documentation to functions.

## Overview

This script replicates the analyses
from Examples 3.1 and 3.11 in @fahrmeir_multivariate_2001.

## Packages and Data

Load all required packages and the prepared Caesar dataset.
As always, we load `conflicted` to manage any function name conflicts across packages.

```{r}
#| label: load-conflicted
library(conflicted)
```

Then we load the necessary packages for data manipulation and modelling.

```{r}
#| label: load-modelling-packages
library(dplyr)
library(Fahrmeir)
library(glmnet)
library(nnet)
library(tidyr)
library(readr)
library(VGAM)
```

## Example 3.1: Caesarian Birth Study (p. 73)

This data is the same as the data we examined in [the first script](01-ANODEV.qmd.qmd),
but now we distinguish between the two infection types `Inf1` and `Inf2`
instead of just a binary `Inf` variable. We import the ready made dataset.

We do this using a basic multinomial logit model
with no infection as the reference category.

:::{.callout-important collapse="true"}
## What is the Model Specification: Expand for Answer
$$
\log\frac{P(\texttt{Inf1})}{P(\texttt{Noinf})}
= \beta_{10} + \beta_{1N}\mathbb{1}_{\texttt{NoPlan}}
+ \beta_{1F}\mathbb{1}_{\texttt{RiskF}} + \beta_{1A}\mathbb{1}_{\texttt{Antib}}
$$
$$
\log\frac{P(\texttt{Inf2})}{P(\texttt{Noinf})}
= \beta_{20} + \beta_{2N}\mathbb{1}_{\texttt{NoPlan}}
+ \beta_{2F}\mathbb{1}_{\texttt{RiskF}} + \beta_{2A}\mathbb{1}_{\texttt{Antib}}
$$
:::

```{r}
#| label: read-caes-data
caes_dat <- read_csv("data/Example3-1Caes.csv")
caes_dat
```

:::{.callout-note collapse="true"}
## Alternative Data Preparation (from `Fahrmeir` Package)

Instead of using the pre-prepared CSV file,
we could also start from the original `caesar` dataset in the `Fahrmeir` package,
and manipulate it into the required format for modeling.

```{r}
#| label: fahrmeir-caesar
head(caesar)
```

The code below:

1. Converts the caesar data to a tibble and removes unused columns
2. Removes any rows with zero counts
3. Pivots from long format (one row per infection type) to wide format
(one row per covariate combination)
4. Renames infection type codes `1`, `2`, and `3`
to `Inf1`, `Inf2`, and `Noinf` respectively
5. Computes total counts and recodes factors to integers matching the textbook coding

```{r}
#| label: prepare-caesar-from-package
caesar_clean <- caesar |>
    as_tibble() |>
    select(-patco, -yl) |>
    dplyr::filter(w > 0) |>
    pivot_wider(
        names_from = y,
        values_from = w,
        values_fill = 0
    ) |>
    rename(Noinf = `3`, Inf1 = `1`, Inf2 = `2`) |>
    mutate(
        size = Noinf + Inf1 + Inf2,
        factor = as.integer(factor == "risk factors"),
        noplan = as.integer(noplan == "not"),
        antib = as.integer(antib == "antibiotics")
    ) |>
    arrange(noplan, antib, factor) |>
    relocate(size, Noinf, Inf1, Inf2, noplan, antib, factor)

caesar_clean
```

:::

### Fit with `multinom()` from `nnet`

To fit the a multinomial log-linear model
we use the `multinom()` function from the `nnet` package.
The `nnet` package is written by the authors of
and well documented in @venables_modern_2002.
To see more information about the `multinom()` function,
you can
[refer to the documentation](https://search.r-project.org/CRAN/refmans/nnet/html/multinom.html)
by running `?multinom` in your R environment.
It is especially useful to check the arguments controlling optimisation
(`maxit`, `trace`, `Hess`) when reproducing textbook results.

:::{.callout-note collapse="true"}
## `multinom` Optimization Method

The `multinom()` function uses the BFGS optimization method,
which is a quasi-Newton method for finding
the maximum likelihood estimates of the model parameters.
This is quicker than the Fisher scoring algorithm
which requires computing the Hessian matrix at each iteration.
:::

We extract the response variables as a matrix format for the `multinom` function,
```{r}
#| label: extract-response-matrix
caes_response <- caes_dat |>
    select(noInf, Inf1, Inf2) |>
    as.matrix()

caes_response
```

and fit the model.

```{r}
#| label: fit-multinom-caes
caes_nnet <- multinom(
    caes_response ~ NoPlan + Antib + RiskF,
    data = caes_dat
)
caes_nnet_summary <- summary(caes_nnet)
caes_nnet_summary
```

The first column (`noInf`) of the response is treated as reference category as desired.
We can print the coefficients in a format that matches the
table on @fahrmeir_multivariate_2001 page 75.

```{r}
#| label: coef-table-caes-nnet
round(t(coef(caes_nnet_summary)), 3)
```

We can also have a look at the deviance of the fitted model.

```{r}
#| label: deviance-caes-nnet
caes_nnet$deviance
```

:::{.callout-caution}
## Deviance Interpretation

Looking at the documentation of the `multinom()` function by running `?multinom`,
we see that the deviance reported by `multinom()` is
$-2 \times \text{log-likelihood}$
and has **not been adjusted** by the saturated model (or the scale parameter).
This differs from deviance in `glm()` and our usual deviance.
Always check the documentation for how deviance is calculated.
:::

:::{.callout-tip collapse="true"}
## Exploring Model Objects

Use `attributes(caes_nnet)` to see all available components of a model object.
This reveals what elements you can access
(like `$deviance`, `$fitted.values`, `$residuals`)
and helps you understand the object's structure.
Similarly, `str()` shows the internal structure,
and `names()` lists just the accessible components.

```{r}
#| label: attributes-caes-nnet
attributes(caes_nnet)
```

:::

:::{.callout-note collapse="true"}
## Fit with `glmnet` Package

This section is not examinable, but just for fun.
`glmnet` is written by Jerome Friedman and Trevor Hastie,
and is a popular package for regularized regression
(that is, Lasso and Ridge regression).
The codebase is in Fortran, and is very fast.

We can use `glmnet` to fit the same multinomial model
by setting `lambda = 0` (giving no weight to the regularisation step)
to get the unregularized estimates.
If you look up `?glmnet`, pay attention to the multinomial parameterization,
since the returned coefficients are not on the same baseline coding as `multinom()`.

```{r}
#| label: glmnet-multinomial-fit
caes_predictors <- caes_dat |>
    select(NoPlan, Antib, RiskF) |>
    as.matrix()

caes_glmnet <- glmnet(
    x = caes_predictors,
    y = caes_response,
    family = "multinomial",
    lambda = 0
)

# Extract and display coefficients
coef(caes_glmnet, s = 0)
```

These coefficients correspond to the alphas in

$$
\tilde{u}_{ir}
= u_{ir} - u_{ik}
= \underbrace{(\alpha_{r0} - \alpha_{k0})}_{\beta_{r0}}
+ \mathbf{z}_i^T\underbrace{
    (\boldsymbol{\alpha}_r - \boldsymbol{\alpha}_k)
}_{\boldsymbol{\beta}_r}
= \beta_{r0} + \mathbf{z}_i^T\boldsymbol{\beta}_r.
$$

Therefore, we can retrieve the `multinom()` coefficients
by taking differences of the `glmnet` alphas:

1. Extract slope coefficients for each category
and compute differences relative to the reference category
2. Extract intercept coefficients and compute differences
relative to the reference category intercept

```{r}
#| label: glmnet-to-multinom-coefs
beta_matrix <- cbind(
    caes_glmnet$beta[[2]] - caes_glmnet$beta[[1]],
    caes_glmnet$beta[[3]] - caes_glmnet$beta[[1]]
)
beta_matrix <- rbind(
    caes_glmnet$a0[-1] - caes_glmnet$a0[1],
    beta_matrix
)
round(beta_matrix, 3)
```

These also matches the coefficients from @fahrmeir_multivariate_2001 p. 75.

:::

::: {.callout-note collapse="true"}
## Alternative Using `Fahrmeir` Package Data

We can fit instead using the `caesar` data from the Fahrmeir package,
setting preferred reference levels before modeling.

```{r}
#| label: fit-multinom-caesar-weighted
caesar_prepared <- caesar |>
    as_tibble() |>
    mutate(
        y = relevel(y, ref = "3"),
        noplan = relevel(noplan, ref = "planned"),
        antib = relevel(antib, ref = "without"),
        factor = relevel(factor, ref = "without")
    )

caes_nnet_prepared <- multinom(
    y ~ noplan + antib + factor,
    data = caesar_prepared,
    weights = w
)

summary(caes_nnet_prepared)
round(t(coef(caes_nnet_prepared)), 3)
```

This again matches the textbook.

We can also check the deviance of this fit,
which should be the same as before since it's the same model.

```{r}
#| label: compare-deviance-multinom-fits
caes_nnet_prepared$deviance
caes_nnet$deviance
```

:::

### Explore the Fitted Values and Residuals

We now inspect the fitted values and residuals of the `caes_nnet` model.
The fitted values are the predicted probabilities for each category.

```{r}
#| label: print-fitted-values
caes_nnet$fitted.values
```

The residuals are the differences
between the observed proportions and the fitted probabilities.

```{r}
#| label: print-residuals
caes_nnet$residuals
```

We can manually validate the residuals by computing the observed proportions
and subtracting the fitted probabilities.

```{r}
#| label: manual-residual-check
caes_nnet_prop <- caes_dat |>
    mutate(
        noInf = noInf / size,
        Inf1 = Inf1 / size,
        Inf2 = Inf2 / size
    ) |>
    select(noInf, Inf1, Inf2)

caes_nnet_prop - caes_nnet$fitted
```

Given a new observation, we can either predict the most likely category
(with `type = "class"`) or the probabilities for each category
(with `type = "probs"`).
Checking `?predict.multinom` is recommended here,
because these two `type` options return different output scales and structures.

```{r}
#| label: predict-new-case
new_case <- tibble(NoPlan = 1, Antib = 1, RiskF = 0)

predict(caes_nnet, newdata = new_case, type = "probs")
predict(caes_nnet, newdata = new_case, type = "class")
```

For a sanity check, we can also compute the predicted probabilities manually
using the coefficients.

```{r}
#| label: manual-probability-check
coef_matrix <- coef(caes_nnet)
ratios <- exp(coef_matrix %*% c(1, 1, 1, 0))
prob_no_inf <- 1 / (sum(ratios) + 1)

prob_no_inf
```

:::{.callout-caution}
We see that the manually calculated probability for `noInf`
is slightly different from the one returned by `predict()`.
The exact reason for this discrepancy is not clear.
:::

:::{.callout-note collapse="true"}
## Alternative Fit with `VGAM` Package

The `VGAM` package will be used a lot more later in the course,
because it fits a wide variety of multivariate response regression models.
It is not as fast as `glmnet` or `nnet` for large datasets,
because it uses the Fisher scoring algorithm (like the book).

Is is good practice to
[read the documentation](https://search.r-project.org/CRAN/refmans/VGAM/html/vglm.html)
for the `vglm()` function
with `?vglm` to understand the arguments and output structure.
Also check `?multinomial`, especially `refLevel` and `parallel`,
since these choices directly affect model interpretation.

We use the raw data `caes_dat` extracted from the CSV file.
`multinomial()` has a `refLevel` argument to specify the reference category.

```{r}
#| label: fit-vgam-caes
caes_vgam <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    family = multinomial(refLevel = "noInf"),
    data = caes_dat
)
```

:::{.callout-tip collapse="true"}
## `vglm` Default Reference Level
[Looking at the documentation](https://search.r-project.org/CRAN/refmans/VGAM/html/multinomial.html)
for `multinomial()`,
we can see that the default parameter for `refLevel` is `"(Last)"`.
This means that the last category in the response matrix is treated as the reference.
In our case, this would be `Inf2`, which is not what we want.
we ensure that `noInf` is treated as the reference category
by setting `refLevel = "noInf"`.
:::

```{r}
#| label: summary-vgam-caes
summary(caes_vgam)
```

:::{.callout-warning}
## Log-Likelihood Differences Between Packages

Remember that the deviance reported by `nnet::multinom()`
is twice the negative log-likelihood.
Therefore the log-likelihood from `nnet` can be computed as:
```{r}
#| label: nnet-loglik-from-deviance
-caes_nnet$deviance / 2
```

This is significantly different from the log-likelihood reported by
`VGAM::vglm()` for the same data and model.
This is because different packages
treat grouped vs. ungrouped multinomial data differently.
See [Appendix A2](A2-likelihood-multinomial.qmd)
for a detailed explanation of how this arises
and why you must be careful when comparing results across packages.
:::

Extracting the coefficients from the `vglm` fit
we see they also those in @fahrmeir_multivariate_2001 p. 75,
and the fitted probabilities match those from `multinom()`.

```{r}
#| label: coef-and-fitted-vgam
round(coef(caes_vgam, matrix.out = TRUE), 3)

fitted(caes_vgam)
```

:::

## Example 3.11

We now move on to Example 3.11.
Looking at the coefficients from the multinomial logit model,
we can see that the coefficients for `NoPlan` and `RiskF`
are very similar across the two infection types.
Therefore we can constrain the coefficients for `NoPlan` and `RiskF` to be equal
across infection types using the `parallel` argument.

:::{.callout-important collapse="true"}
## What is the Model Specification: Expand for Answer
$$
\log\frac{P(\texttt{Inf1})}{P(\texttt{Noinf})}
= \beta_{10} + \beta_{N}\mathbb{1}_{\texttt{NoPlan}}
+ \beta_{F}\mathbb{1}_{\texttt{RiskF}} + \beta_{1A}\mathbb{1}_{\texttt{Antib}}
$$
$$
\log\frac{P(\texttt{Inf2})}{P(\texttt{Noinf})}
= \beta_{20} + \beta_{N}\mathbb{1}_{\texttt{NoPlan}}
+ \beta_{F}\mathbb{1}_{\texttt{RiskF}} + \beta_{2A}\mathbb{1}_{\texttt{Antib}}
$$
:::

We run the regression, and extract the coefficients.

```{r}
#| label: fit-vgam-parallel
caes_vgam_parallel <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = caes_dat,
    family = multinomial(
        refLevel = "noInf",
        parallel = TRUE ~ NoPlan + RiskF - 1
    )
)

coefvlm(caes_vgam_parallel, matrix.out = TRUE)
```

:::{.callout-note collapse="true"}
## `VGAM` Using `Fahrmeir` Package Data

We can try running the `vglm()` function on the original `caesar` data
[we manipulated earlier](#alternative-original-fahrmeir-data)
from the `Fahrmeir` package.
We don't fit the parallel model here.

```{r}
#| label: fit-vgam-fahrmeir
#| warning: false
#| error: true

vglm(
    formula = y ~ noplan + antib + factor,
    data = caesar_prepared,
    weights = w,
    family = multinomial(refLevel = "3")
)
```

This error is due to the zero-count rows. We remove them and try again.

```{r}
#| label: fit-vgam-filtered-fahrmeir

caesar_no_zero_w <- caesar_prepared |> dplyr::filter(w > 0)
caesar_no_zero_w

levels(caesar_no_zero_w$noplan)
levels(caesar_no_zero_w$factor)
levels(caesar_no_zero_w$antib)

caes_vgam2 <- vglm(
    formula = y ~ noplan + antib + factor,
    data = caesar_no_zero_w,
    weights = w,
    family = multinomial(refLevel = "3")
)

summary(caes_vgam2)
```

We can compare the coefficients from this fit to the earlier `vglm` fit on the wide-format data.

```{r}
#| label: coefs-vgam2

coefvlm(caes_vgam2, matrix.out = TRUE)
coefvlm(caes_vgam, matrix.out = TRUE)
```

```{r}
#| label: compare-criteria-vgam-fits

caes_vgam@criterion
caes_vgam2@criterion
```

We can see that the deviance and log-likelihoods are different between the fits.
The deviance of `caes_vgam` matches that of in the book,
however the deviance of `caes_vgam2` is just two times the negative log-likelihood of the fitted model.

:::{.callout-tip collapse="true"}
## Why `@` Here Instead of `$`?

In R, `@` accesses **slots** of S4 objects, while `$` accesses named elements
in lists, data frames, and many S3 objects.

`vglm()` returns an S4 object, so `criterion` is stored as a slot,
and we use `caes_vgam@criterion` rather than `caes_vgam$criterion`.
:::
:::

:::{.callout-note collapse="true"}
## Using `dplyr` and `tidyr` to Reshape Data

When fitting multinomial models, the data can be structured in either long or wide format.
We can convert the data back into long format and confirm we still get the same fit.

The manipulation below:

1. Removes the total-count column (`size`)
because the long-form weight column will carry counts
2. Pivots infection outcomes (`noInf`, `Inf1`, `Inf2`) into two columns:
category name (`inf_type`) and count (`w`)
3. Sets an explicit factor order for `inf_type`
so the reference and comparisons are reproducible

```{r}
#| label: fit-nnet-long-form

caesar_long <- caes_dat |>
    select(-size) |>
    pivot_longer(
        cols = noInf:Inf2,
        names_to = "inf_type",
        values_to = "w"
    ) |>
    mutate(inf_type = factor(inf_type, levels = c("noInf", "Inf1", "Inf2")))

caesar_long

caes_nnet_long <- multinom(
    inf_type ~ NoPlan + Antib + RiskF,
    data = caesar_long,
    weights = w
)

coef(caes_nnet_long)
coef(caes_nnet_prepared)
```

:::

## Package citations and references

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c(
    "conflicted",
    "dplyr",
    "Fahrmeir",
    "glmnet",
    "nnet",
    "tidyr",
    "readr",
    "VGAM"
)
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```