---
title: "Example 3.1 & 3.11: Caesar Dataset Analysis"
subtitle: "Multinomial Logistic Regression and Category-Specific Effects"
date: today
nocite: |
    @fahrmeir_multivariate_2001
    @venables_modern_2002
---

## TODO

- Go through and add apppropriate documentation to functions.

## Overview

This document mirrors the R script for Examples 3.1 and 3.11 (Fahrmeir & Tutz), covering:

- Multinomial logistic regression on three infection categories (noInf, Inf1, Inf2)
- Fits via `nnet::multinom()` (BFGS) and `VGAM::vglm()` (Fisher scoring)
- A constrained parallel-coefficients model (Example 3.11)
- Additional fits with `glmnet` and alternative wide-format data

## Packages and Data

Load all required packages and the prepared Caesar dataset.
As always, we load `conflicted` to manage any function name conflicts across packages.

```{r}
#| label: load-conflicted
library(conflicted)
```

Then we load the necessary packages for data manipulation and modelling.

```{r}
#| label: load-modelling-packages
library(dplyr)
library(Fahrmeir)
library(glmnet)
library(nnet)
library(tidyr)
library(readr)
library(VGAM)
```

## Example 3.1: Caesarian Birth Study (p. 73)

```{r}
#| label: read-caes-data
caes_dat <- read_csv("data/Example3-1Caes.csv")
caes_dat
```

:::{.callout-note collapse="true"}
## Alternative Data Preparation (from `Fahrmeir` Package)

Alternatively, one can manipulate the dataset directly from the Fahrmeir package using `dplyr`. The code below:

1. Converts the caesar data to a tibble and removes unused columns
2. Filters to non-zero weights
3. Pivots from long format (one row per infection type) to wide format (one row per covariate combination)
4. Renames infection type codes to interpretable labels (3→Noinf, 1→Inf1, 2→Inf2)
5. Computes total counts and recodes factors to integers matching the textbook coding

```{r}
#| label: prepare-caesar-from-package
caesar_clean <- caesar |>
    as_tibble() |>
    select(-patco, -yl) |>
    dplyr::filter(w > 0) |>
    pivot_wider(
        names_from = y,
        values_from = w,
        values_fill = 0
    ) |>
    rename(Noinf = `3`, Inf1 = `1`, Inf2 = `2`) |>
    mutate(
        size = Noinf + Inf1 + Inf2,
        factor = as.integer(factor == "risk factors"),
        noplan = as.integer(noplan == "not"),
        antib = as.integer(antib == "antibiotics")
    ) |>
    arrange(noplan, antib, factor) |>
    relocate(size, Noinf, Inf1, Inf2, noplan, antib, factor)

caesar_clean
```

:::

### Fit with `multinom()` from `nnet`

To fit the a multinomial log-linear model
we use the `multinom()` function from the `nnet` package.
The `nnet` package is largely documented in @venables_modern_2002,
and is also written by the authors of that book.
To see more information about the `multinom()` function,
you can
[refer to the documentation](https://search.r-project.org/CRAN/refmans/nnet/html/multinom.html)
by running `?multinom` in your R environment.

:::{.callout-note}
The `multinom()` function uses the BFGS optimization method,
which is a quasi-Newton method for finding
the maximum likelihood estimates of the model parameters.
This is quicker than the Fisher scoring algorithm
which requires computing the Hessian matrix at each iteration.
:::

We convert the response variables to a matrix format suitable for `multinom()`,
```{r}
#| label: fit-multinom-caes
caes_response <- caes_dat |>
    select(noInf, Inf1, Inf2) |>
    as.matrix()

caes_response

caes_nnet <- multinom(
    caes_response ~ NoPlan + Antib + RiskF,
    data = caes_dat
)
caes_nnet_summary <- summary(caes_nnet)
caes_nnet_summary
```

The first column (`noInf`) of the response is treated as reference category.
We can print the coefficients in a format that matches the
table on @fahrmeir_multivariate_2001 page 75.

```{r}
#| label: coef-table-caes-nnet
round(t(coef(caes_nnet_summary)), 3)
```

This matches the textbook results.
We can also have a look at the deviance of the fitted model.

```{r}
#| label: deviance-caes-nnet
caes_nnet$deviance
```

:::{.callout-caution}
## Deviance Interpretation

The deviance reported by `multinom()` is simply
$-2 \times \text{log-likelihood}$
and has **not been adjusted** by the saturated model.
This differs from deviance in `glm()` and our usual deviance.
Always check the documentation for how deviance is calculated.
:::

These values match the textbook results

:::{.callout-tip collapse="true"}
## Exploring Model Objects

Use `attributes(caes_nnet)` to see all available components of a model object.
This reveals what elements you can access
(like `$deviance`, `$fitted.values`, `$residuals`)
and helps you understand the object's structure.
Similarly, `str()` shows the internal structure,
and `names()` lists just the accessible components.

```{r}
#| label: attributes-caes-nnet
attributes(caes_nnet)
```

:::


## `glmnet` Comparison

:::{.callout-note}

This section is not examinable, but just for fun.
`glmnet` is written by Jerome Friedman and Trevor Hastie,
and is a popular package for regularized regression
(that is, Lasso and Ridge regression).
The codebase is in Fortran, and is very fast.

We can use `glmnet` to fit the same multinomial model
by setting `lambda = 0` to get the unregularized estimates.

```{r}
#| label: glmnet-multinomial-fit
caes_predictors <- caes_dat |>
    select(NoPlan, Antib, RiskF) |>
    as.matrix()

caes_glmnet <- glmnet(
    x = caes_predictors,
    y = caes_response,
    family = "multinomial",
    lambda = 0
)

# Extract and display coefficients
coef(caes_glmnet, s = 0)
```

These coefficients are different from the `multinom()` fit,
but rather correspond to the alphas in

$$
\tilde{u}_{ir} = u_{ir} - u_{ik} = \underbrace{(\alpha_{r0} - \alpha_{k0})}_{\beta_{r0}} + \mathbf{z}_i^T\underbrace{(\boldsymbol{\alpha}_r - \boldsymbol{\alpha}_k)}_{\boldsymbol{\beta}_r} = \beta_{r0} + \mathbf{z}_i^T\boldsymbol{\beta}_r.
$$

We can therefore retrieve the `multinom()` coefficients by taking differences of the `glmnet` alphas:

1. Extract slope coefficients for each category and compute differences relative to the reference category
2. Extract intercept coefficients and compute differences relative to the reference category intercept

```{r}
#| label: glmnet-to-multinom-coefs
beta_matrix <- cbind(
    caes_glmnet$beta[[2]] - caes_glmnet$beta[[1]],
    caes_glmnet$beta[[3]] - caes_glmnet$beta[[1]]
)
beta_matrix <- rbind(
    caes_glmnet$a0[-1] - caes_glmnet$a0[1],
    beta_matrix
)
round(beta_matrix, 3)
```

This also matches the coefficients from @fahrmeir_multivariate_2001 p. 75.

:::

::: {.callout-note}
## Alternative: Original Fahrmeir Data

We can fit instead using the `caesar` data from the Fahrmeir package, setting preferred reference levels before modeling.

```{r}
#| label: fit-multinom-caesar-weighted
caesar_prepared <- caesar |>
    as_tibble() |>
    mutate(
        y = relevel(y, ref = "3"),
        noplan = relevel(noplan, ref = "planned"),
        antib = relevel(antib, ref = "without"),
        factor = relevel(factor, ref = "without")
    )

caes_nnet2 <- multinom(
    y ~ noplan + antib + factor,
    data = caesar_prepared,
    weights = w
)

summary(caes_nnet2)
round(t(coef(caes_nnet2)), 3)
```

This again matches the textbook.

We can also check the deviance of this fit,
which should be the same as before since it's the same model.

```{r}
#| label: compare-deviance-multinom-fits
caes_nnet2$deviance
caes_nnet$deviance
```

:::

### Explore the Fitted Values and Residuals

We now inspect the fitted values and residuals of the `caes_nnet` model.
The fitted values are the predicted probabilities for each category.

```{r}
#| label: print-fitted-values
caes_nnet$fitted.values
```

The residuals are the differences
between the observed proportions and the fitted probabilities.

```{r}
#| label: print-residuals
caes_nnet$residuals
```

We can manually validate the residuals by computing the observed proportions
and subtracting the fitted probabilities.

```{r}
#| label: manual-residual-check
caes_nnet_prop <- caes_dat |>
    mutate(
        noInf = noInf / size,
        Inf1 = Inf1 / size,
        Inf2 = Inf2 / size
    ) |>
    select(noInf, Inf1, Inf2)

caes_nnet_prop - caes_nnet$fitted
```

Given a new observation, we can either predict the most likely category
(with `type = "class"`) or the probabilities for each category
(with `type = "probs"`).

```{r}
#| label: predict-new-case
new_case <- tibble(NoPlan = 1, Antib = 1, RiskF = 0)

predict(caes_nnet, newdata = new_case, type = "probs")
predict(caes_nnet, newdata = new_case, type = "class")
```

For a sanity check, we can also compute the predicted probabilities manually
using the coefficients.

```{r}
#| label: manual-probability-check
coef_matrix <- coef(caes_nnet)
ratios <- exp(coef_matrix %*% c(1, 1, 1, 0))
prob_no_inf <- 1 / (sum(ratios) + 1)

prob_no_inf
```

:::{.callout-caution}
For some unknown reason,
the `predict()` function returns a slightly different probability
for the `noInf` category than the manual calculation.
:::

### Fit with `VGAM`

The `VGAM` package will be used a lot more later in the course,
because it fits a wide variety of multivariate response regression models.
It is not as fast as `glmnet` or `nnet` for large datasets,
because it uses the Fisher scoring algorithm.

Is is good practice to
[read the documentation](https://search.r-project.org/CRAN/refmans/VGAM/html/vglm.html)
for the `vglm()` function
with `?vglm` to understand the arguments and output structure.

```{r}
#| label: fit-vgam-caes
caes_vgam <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    family = multinomial(refLevel = "noInf"),
    data = caes_dat
)
```

:::{.callout-tip collapse="true"}
## `vglm` Syntax
[Looking at the documentation](https://search.r-project.org/CRAN/refmans/VGAM/html/multinomial.html)
for `multinomial()`,
we can see that the default parameter for `refLevel` is `"(Last)"`.
This means that the last category in the response matrix is treated as the reference.
In our case, this is `Inf2`, which is not what we want.
By setting `refLevel = "noInf"`, we ensure that `noInf` is treated as the reference category,
and the coefficients are interpreted relative to `noInf`.
:::

```{r}
#| label: summary-vgam-caes
summary(caes_vgam)
```

:::{.callout-warning}
## Log-Likelihood Differences Between Packages

Remember that the deviance reported by `nnet::multinom()`
is twice the negative log-likelihood.
Therefore the log-likelihood from `nnet` can be computed as:
```{r}
#| label: nnet-loglik-from-deviance
-caes_nnet$deviance / 2
```

This is significantly different from the log-likelihood reported by
`VGAM::vglm()` for the same data and model.
This is because different packages
treat grouped vs. ungrouped multinomial data differently.
See [Appendix A2](A2-likelihood-multinomial.qmd)
for a detailed explanation of how this arises
and why you must be careful when comparing results across packages.
:::

Extracting the coefficients from the `vglm` fit we see they also those in @fahrmeir_multivariate_2001 p. 75,
and the fitted probabilities match those from `multinom()`.

```{r}
#| label: coef-and-fitted-vgam
round(coef(caes_vgam, matrix.out = TRUE), 3)

fitted(caes_vgam)
```

## Example 3.11

Constrain `NoPlan` and `RiskF` to be equal across categories using the `parallel` argument.

```{r}
#| label: fit-vgam-parallel
caes_vgam_parallel <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = caes_dat,
    family = multinomial(
        refLevel = "noInf",
        parallel = TRUE ~ NoPlan + RiskF - 1
    )
)

coefvlm(caes_vgam_parallel, matrix.out = TRUE)
```

## VGAM on Filtered Fahrmeir Data

Filter out zero-count rows, reset reference levels, and fit with `vglm`; compare criteria and log-likelihoods.

```{r}
#| label: fit-vgam-filtered-fahrmeir
caesar_no_zero_w <- dplyr::filter(caesar, w > 0)
levels(caesar_no_zero_w$noplan)
levels(caesar_no_zero_w$antib)
levels(caesar_no_zero_w$factor)

caesar_no_zero_w$noplan <- relevel(caesar_no_zero_w$noplan, ref = "planned")
caesar_no_zero_w$antib <- relevel(caesar_no_zero_w$antib, ref = "without")
caesar_no_zero_w$factor <- relevel(caesar_no_zero_w$factor, ref = "without")

caes_vgam2 <- vglm(
    formula = y ~ noplan + antib + factor,
    data = caesar_no_zero_w,
    weights = w,
    family = VGAM::multinomial(refLevel = "3")
)

summary(caes_vgam2)
coefvlm(caes_vgam2, matrix.out = TRUE)

attributes(caes_vgam)$criterion
attributes(caes_vgam2)$criterion
logLik(caes_vgam)
logLik(caes_vgam2)
```

## Long-Form Data (Wide-to-Long)

Reshape the prepared data to long format and fit the multinomial model with weights; coefficients should match earlier fits.

```{r}
#| label: fit-nnet-long-form
caesar_new <- pivot_longer(
    data = caes_dat[, -1],
    cols = noInf:Inf2,
    names_to = "inf_type",
    values_to = "w"
)
caesar_new$inf_type <- factor(
    caesar_new$inf_type,
    levels = unique(caesar_new$inf_type)
)
# Extract coefficient matrix

caes_nnet_tibble <- multinom(
    inf_type ~ NoPlan + Antib + RiskF,
    data = caesar_new,
    weights = w
)
summary(caes_nnet_tibble)$coeff
summary(caes_nnet2)$coeff
```

## Package citations and references

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "Fahrmeir", "glmnet", "nnet", "tidyr", "readr", "VGAM")
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```