---
title: "Example 3.2 & 3.4: BTR Analysis"
subtitle: "Ordinal Regression on Breathing Test Results"
date: today
---

## Overview

This script reproduces **Examples 3.2 and 3.4** from Fahrmeir & Tutz, analyzing **breathing test results (BTR)** with ordinal outcomes.

Topics covered:
- **Ordinal logistic regression** with `polr()` (proportional odds/cumulative logit)
- Alternative link functions: probit, complementary log-log, log-log
- Contrast coding systems: dummy vs. effect coding
- Model comparison and proportional odds assumption testing
- Saturated model deviance calculations

## Data Loading and Setup

```{r}
library(conflicted)
library(MASS)
library(dplyr)
```

Load the BTR dataset. It contains breathing test results (ordinal outcome) with covariates age and smoking status, along with frequency counts for each combination.

```{r}
btr_dat <- read.csv(
    "data/Example3-2BTR.csv",
    colClasses = c("factor", "factor", "factor", "numeric")
)

str(btr_dat)
btr_dat
```

## Factor Coding and Contrast Systems

The breathing test result (BTR) must be converted to an **ordered factor** since we have three ordered categories. By default, R uses **dummy coding** with `contr.treatment`.

First, check and set the appropriate factor structure:

```{r}
is.factor(btr_dat$BTR)
is.ordered(btr_dat$BTR)

# Set BTR to ordinal (ordered categorical)
btr_dat$BTR <- as.ordered(btr_dat$BTR)

is.factor(btr_dat$Age)
is.factor(btr_dat$Smoking)
is.factor(btr_dat$Freq)
```

Examine the current contrast coding scheme:

```{r}
getOption(x = "contrasts")

contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

The dummy coding creates indicator variables for each level relative to a reference category. This is the default but leads to coefficient estimates that differ from Fahrmeir & Tutz due to their use of **effect coding**.

## Model Fitting: Proportional Odds Models

The `polr()` function fits cumulative logit (and other) models. Key note: **the sign of coefficients in `polr()` is opposite to the textbook formula**—compare with the documentation to reconcile.

### Main Model with Interaction (Logistic Link)

Fit the cumulative logistic/proportional odds model with age, smoking, and their interaction:

```{r}
btr_logit <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit)
```

Extract model components:

```{r}
# Linear predictors (eta values)
btr_logit$lp

# Threshold parameters (cutpoints/intercepts)
btr_logit$zeta
```

Verify coefficients by hand-computing the linear predictor:

```{r}
Z <- cbind(
    (btr_dat$Age == "40to59"),
    (btr_dat$Smoking == "2Former"),
    (btr_dat$Smoking == "3Current"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "2Former"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "3Current")
)

# Hand-computed eta should match btr_logit$lp
head(Z %*% btr_logit$coeff)
head(btr_logit$lp)
```

### Model Comparison (with and without interaction)

Fit a main-effects-only model and compare using the likelihood ratio test:

```{r}
btr_logit1 <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit1)
```

::: {.callout-note}
The LR statistic in the `anova()` output is the difference between the deviances of the two models. Each deviance reported is `-2 * log(likelihood)` for that model.
:::

```{r}
anova(btr_logit, btr_logit1)
```

## Alternative Link Functions

The proportional odds assumption can be tested by fitting models with different link functions. The `polr()` function supports:
- **logistic**: cumulative logit model (default)
- **probit**: cumulative normal model
- **cloglog**: complementary log-log (grouped Cox/proportional hazards)
- **loglog**: extreme maximal-value distribution

### Grouped Cox / Complementary Log-Log

```{r}
btr_ph <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

summary(btr_ph)
```

### Extreme Maximal-Value Distribution

```{r}
btr_extm <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

summary(btr_extm)
```

## Effect Coding (Fahrmeir & Tutz Parameterization)

By default, R uses dummy coding. To match the textbook and get effect coding (deviation contrasts), switch the global contrast option and refit:

```{r}
options(contrasts = c("contr.sum", "contr.poly"))

# Verify the contrast change
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

Refit both models with effect coding:

```{r}
btr_logit_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

btr_logit1_eff <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit_eff)
summary(btr_logit1_eff)
```

::: {.callout-tip}
With effect coding, the coefficients now match Table 3.3 in Fahrmeir & Tutz. The deviances remain the same (model fit is unchanged), only the parameterization differs.
:::

```{r}
anova(btr_logit_eff, btr_logit1_eff)
```

## Saturated Model and Deviance Calculation

The deviance reported by `polr()` is `-2 * log(likelihood)` for the fitted model. To compute the **residual deviance** (difference from saturated model), we manually compute the saturated model log-likelihood.

```{r}
# Reshape frequency counts into a matrix (rows = covariate combinations, cols = BTR categories)
data_mat <- t(matrix(btr_dat$Freq, nr = 3))

# Compute MLE probabilities for the saturated model (empirical proportions)
probs <- data_mat / rowSums(data_mat)

# Compute saturated log-likelihood via multinomial
loglike_saturated <- 0
for (row in seq_len(nrow(data_mat))) {
    for (col in 1:3) {
        eval_vec <- rep(0, 3)
        eval_vec[col] <- 1
        increment <- data_mat[row, col] *
            dmultinom(x = eval_vec, prob = probs[row, ], log = TRUE)
        if (!is.na(increment)) loglike_saturated <- loglike_saturated + increment
    }
}

cat("Saturated model log-likelihood:", loglike_saturated, "\n")
```

Compute residual deviances (deviance of fitted model minus deviance of saturated model):

```{r}
residual_dev_logit <- btr_logit_eff$deviance - abs(2 * loglike_saturated)
residual_dev_logit1 <- btr_logit1_eff$deviance - abs(2 * loglike_saturated)

cat("Residual deviance (interaction):", residual_dev_logit, "\n")
cat("Residual deviance (main effects):", residual_dev_logit1, "\n")
cat("These should match Table 3.3 in Fahrmeir & Tutz\n")
```

Compute residual deviances for alternative link functions:

```{r}
# Refit with effect coding
btr_ph_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

btr_extm_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

residual_dev_ph <- btr_ph_eff$deviance - abs(2 * loglike_saturated)
residual_dev_extm <- btr_extm_eff$deviance - abs(2 * loglike_saturated)

cat("Residual deviance (cloglog):", residual_dev_ph, "\n")
cat("Residual deviance (loglog):", residual_dev_extm, "\n")
```

## Model Diagnostics

### Fitted Values and Predictions

```{r}
# Fitted probabilities (first few rows)
head(predict(btr_logit_eff, btr_dat, type = "probs"))
```

### Model Selection with AIC

```{r}
# Automatic stepwise selection based on AIC
btr_step <- step(btr_logit_eff)

summary(btr_step)
```

### Variance-Covariance Matrix and Confidence Intervals

```{r}
# Variance-covariance matrix of parameter estimates
vcov(btr_logit_eff)

# Standard errors from summary
sqrt(diag(vcov(btr_logit_eff)))
```

Compute confidence intervals using the profile likelihood (more accurate than Wald intervals):

```{r}
confint(btr_logit_eff)
```

### Profile Likelihood

For detailed inference, examine the profile likelihood for individual parameters:

```{r}
profile(btr_logit_eff)
```

## Key Learning Points

1. **Ordinal outcomes**: BTR has three ordered categories; use ordered factors
2. **Proportional odds**: Assumes coefficients are constant across all cutpoints
3. **Link functions**: Different links (logit, probit, cloglog, loglog) test proportional odds
4. **Contrast coding**: Dummy vs. effect coding changes parameterization, not model fit
5. **Deviance interpretation**: Always distinguish between `-2*log(L)` and residual deviance
6. **Saturated model**: Needed to compute residual deviance for GOF

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Venables, W. N., & Ripley, B. D. (2002). Modern applied statistics with S (4th ed.)
- [polr() documentation](https://www.rdocumentation.org/packages/MASS/versions/7.3-51.6/topics/polr)
