---
title: "Ordinal Regression: Breathing Test Results."
subtitle: "@fahrmeir_multivariate_2001 Examples 3.2 & 3.4"
date: last-modified
editor:
  markdown:
    wrap: sentence
    canonical: true
---

## Data Loading and Setup

```{r}
#| label: libs
library(conflicted)
library(dplyr)
library(MASS)
library(purrr)
library(readr)
library(tidyr)
```

We load the BTR dataset.
It contains breathing test results with 3 levels: normal, borderline, and abnormal (which we can consider as ordered).
We want to analyze the effects of the covariates age and smoking status.

```{r}
#| label: load-btr
btr_dat <- read_csv(
  "data/Example3-2BTR.csv"
)
```

As it is, we can see that the variables `BTR`, `Age`, and `Smoking` are all read as characters.
This is not what we want, and we will use the `col_types` argument to specify that we want them as factors.

```{r}
#| label: load-btr-factor
btr_dat <- read_csv(
  "data/Example3-2BTR.csv",
  col_types = cols(
    BTR = col_factor(),
    Age = col_factor(),
    Smoking = col_factor(),
    Freq = col_double()
  )
)

btr_dat
```

`BTR`, `Age`, and `Smoking` are all factors now.
We next check if the breathing test result (`BTR`) is encoded as an ordered factor, and if not, we convert it to an ordered factor:

```{r}
#| label: order-btr
is.ordered(btr_dat$BTR)
btr_dat$BTR <- as.ordered(btr_dat$BTR)
```

We can now confirm the conversion was successful.

```{r}
#| label: check-btr-ordered
class(btr_dat$BTR)
btr_dat$BTR
```

The default ordering happens to be good for us, but if we needed to change the order, we could specify it with `factor(..., levels = c(...), ordered = TRUE)`.

We now examine the current contrast coding scheme:

```{r}
#| label: get-contrasts-opt
getOption("contrasts")
```

`"contr.treatment"` corresponds to dummy coding for unordered factors.
For ordered factors, the default is `"contr.poly"` (orthogonal polynomials); we won't cover polynomial contrasts in this subject but refer to p.33 of @chambers_hastie_sms_1992 for more details.
We consider `Age` and `Smoking` as unordered factors, and we will use dummy coding.

```{r}
#| label: show-contrasts
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

The default reference categories are `never-smoked` for `Smoking`, and the `<40` for Age.

::: callout-note
-   **Dummy coding**: Reference category is 0, other categories are 1; coefficients represent difference from reference category
-   **Effect coding**: Reference category is -1, other categories are 0 or 1; coefficients represent difference from overall mean
:::

## Model Fitting: Proportional Odds Models

The `MASS::polr()` function fits cumulative logit (and other) models.
[Looking at the documentation](https://search.r-project.org/CRAN/refmans/MASS/html/polr.html){target="_blank"} using `?polr`, particularly how the linear predictor $\eta$ is defined therein; the signs of beta coefficients in `MASS::polr()` is **opposite** to the corresponding coefficients in the textbook (@fahrmeir_multivariate_2001)'s formula.

We first fit a cumulative logistic (proportional odds) model with an intercept, age, smoking.
We also include an interaction effect between age and smoking.

```{r}
#| label: fit-logit-int
btr_logit_inter <- polr(
  BTR ~ Age * Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "logistic"
)
```

::: {.callout-tip collapse="true"}
## `MASS::polr()` Function Syntax

-   **`formula`**: Model structure `response ~ predictors`; response must be an ordered factor
-   **`data`**: Data frame containing the variables
-   **`weights`**: Case weights (here, frequency counts for each covariate combination)
-   **`Hess = TRUE`**: Return the Hessian matrix; needed for `summary()` and confidence intervals
-   **`method`**: Link function---`"logistic"` (default, proportional odds), `"probit"`, `"cloglog"`, `"loglog"`, or `"cauchit"`
:::

The `MASS::polr()` function is limited.
It can model standard cumulative models (see section 3.3.1 of @fahrmeir_multivariate_2001), but doesn't allow for extended cumulative model with threshold variables (see section 3.3.2 of @fahrmeir_multivariate_2001).

```{r}
#| label: summary-logit-int
summary(btr_logit_inter)
```

::: {.callout-important collapse="true"}
## SELF TEST: Provided that a worker is in the 40–59 age group, what is his/her (cumulative) odds of having a normal test result if s/he is former smoker, compared to if s/he has never smoked?

The fitted model has cumulative logit link of the form \begin{multline*}
\log\frac{P(\texttt{BTR}=\texttt{Normal})}{P(\texttt{BTR}>\texttt{Normal})} = \theta_1
+ \gamma_1\texttt{Age40to59}+\gamma_2\texttt{Smoking2Former}+\gamma_3\texttt{Smoking3Current} +\\
\gamma_4\texttt{Age40to59}\cdot \texttt{Smoking2Former}
+ \gamma_5\texttt{Age40to59}\cdot \texttt{Smoking3Current}
\end{multline*} and \begin{multline*}
\log\frac{P(\texttt{BTR} \leq \texttt{Borderline})}{P(\texttt{BTR}>\texttt{Borderline})} = \theta_2
+ \gamma_1\texttt{Age40to59}+\gamma_2\texttt{Smoking2Former}+\gamma_3\texttt{Smoking3Current} +\\
\gamma_4\texttt{Age40to59}\cdot \texttt{Smoking2Former}
+ \gamma_5\texttt{Age40to59}\cdot \texttt{Smoking3Current},
\end{multline*} where the MLE of the parameters are

$$
\hat{\theta}_1=2.8334,  
\hat{\theta}_2=4.3078,
\hat{\gamma}_1=0.8857,  
\hat{\gamma}_2=-0.6973 , 
\hat{\gamma}_3=-0.3472 ,
\hat{\gamma}_4=-1.1458 ,
\hat{\gamma}_5=-2.2007.
$$

Given that a worker is in the 40–59 age group, the cumulative odds of having a normal test result for a former smoker is $\exp(-0.6973 - 1.1458)= 0.158$ times that for a person who has never smoked.

As it is a "proportional odds" model, the same is true for the cumulative odds of having "$\leq$ Borderline" testing result.
:::

There is no residual readily available, but we can observe the linear predictor $\eta$ **as defined in the documentation page for the `MASS::polr()` function without the intercept**.

```{r}
#| label: linear-predictor
btr_logit_inter$lp
```

The $\zeta$'s are the theshold parameters $\theta$'s as defined in our textbook (@fahrmeir_multivariate_2001) and slides.

```{r}
#| label: threshold-zeta
btr_logit_inter$zeta
```

We verify the linear predictors by hand-computing the linear predictor.

```{r}
#| label: manual-lp-check
Z <- cbind(
  (btr_dat$Age == "40to59"),
  (btr_dat$Smoking == "2Former"),
  (btr_dat$Smoking == "3Current"),
  (btr_dat$Age == "40to59") * (btr_dat$Smoking == "2Former"),
  (btr_dat$Age == "40to59") * (btr_dat$Smoking == "3Current")
)

tibble(
  manual = c(Z %*% coef(btr_logit_inter)),
  polr = btr_logit_inter$lp
)
```

We see that they line up.

### Fit Model without Interaction (Main Effects Only)

We fit a main-effects-only model without interaction.

```{r}
#| label: fit-logit-main
btr_logit_main <- polr(
  BTR ~ Age + Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "logistic"
)

summary(btr_logit_main)
```

Again the signs of the covariate coefficients are the opposite to our model defined in lectures.
Since the main effects model is nested within the interaction model, we can compare them with a likelihood ratio test.

::: {.callout-important collapse="true"}
## SELF TEST: How Many Degrees of Freedom will the Likelihood Ratio Test have?

The test has 2 degrees of freedom since we are adding two interaction terms (`Age40to59:Smoking2Former` and `Age40to59:Smoking3Current`).
:::

We can use `anova()`:

```{r}
#| label: anova-logit-models
anova(btr_logit_inter, btr_logit_main)
```

::: {.callout-important collapse="true"}
## SELF TEST: Interpret the Likelihood Ratio Test Results

We reject the null hypothesis that there is no interaction, at all reasonable significance levels.
:::

::: callout-note
The LR statistic in the `anova()` output is the difference between the deviances of the two models.
Each residual deviance reported is $-2 \times \log(\text{likelihood})$ for that model (more on this to follow).
:::

## Alternative Link Functions

The proportional odds assumption can be tested by fitting models with different link functions.
The `MASS::polr()` function supports:

-   **logistic**: cumulative logit model (default)
-   **probit**: cumulative normal model
-   **cloglog**: complementary log-log (grouped Cox/proportional hazards)
-   **loglog**: extreme maximal-value distribution.

As discussed in lecture, these can be motivated by the so-called "threshold approach" where the response is a categorized version of an underlying latent continuous variables.
Figure 3.1 of @fahrmeir_multivariate_2001 displays the shapes of the error densities of the latent variable with respect to these models .

### Grouped Cox / Proportional Hazards (Complementary Log-Log Link)

```{r}
#| label: fit-cloglog-int
btr_ph <- polr(
  BTR ~ Age + Smoking + Age:Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "cloglog"
)

summary(btr_ph)
```

### Extreme Maximal-Value Distribution

```{r}
#| label: fit-loglog-int
btr_extm <- polr(
  BTR ~ Age + Smoking + Age:Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "loglog"
)

summary(btr_extm)
```

## Matching our Results to @fahrmeir_multivariate_2001 Example 3.4

By default, R uses dummy coding.
To match to the numerical results in Table 3.3 of @fahrmeir_multivariate_2001 Example 3.4, we need to use effect coding.
We switch the global contrast option and refit.

```{r}
#| label: set-effect-coding
options(contrasts = c("contr.sum", "contr.poly"))

# Verify the contrast change
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

We now have the last categories coded as -1, as also done in the textbook.
Refitting both models:

```{r}
#| label: fit-logit-eff
btr_logit_inter_eff <- polr(
  BTR ~ Age + Smoking + Age:Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "logistic"
)

btr_logit_main_eff <- polr(
  BTR ~ Age + Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "logistic"
)

btr_logit_inter_eff_summary <- summary(btr_logit_inter_eff)
btr_logit_inter_eff_summary
btr_logit_main_eff_summary <- summary(btr_logit_main_eff)
btr_logit_main_eff_summary
```

The values reported in the summaries now match the textbook (up to rounding errors and sign change of the covariate coefficients).
In addition, we can redo the likelihood ratio test with the new models to confirm we get the same result.

```{r}
#| label: anova-logit-eff
anova(btr_logit_inter_eff, btr_logit_main_eff)
```

As expected, the likelihood ratio test results are the same as before, because using effect coding, as opposed to dummy coding, amounts only to a reparametrization of the model.

## Saturated Model and Deviance Calculation

In each of the regressed model summaries, there is a value "Residual Deviance" reported.
We need to be careful about what this exactly means; as usual, this can be highly package dependent.

The `deviance` component (reported as "residual deviance" by the `summary()` function) contained in a `polr` object is $-2 \times \log(\text{likelihood})$ for the fitted model.
We can manually compute the (scaled) deviance as defined in class , i.e. twice the difference between the maximum log-likelihoods of the saturated model and the GLM.

We aim to compute the maximum log-likelihood of the saturated model.
We first create a matrix with each row corresponding to a unique covariate combination, with three columns counting the frequencies of each response category for that covariate combination.
This can be accomplished elegantly by the function `pivot_wider()` in the `tidyr` package.

```{r}
#| label: make-count-matrix
btr_counts <- btr_dat |>
  pivot_wider(
    id_cols = c(Age, Smoking),
    names_from = BTR,
    values_from = Freq
  )
btr_counts

data_mat <- btr_counts |>
  dplyr::select(-Age, -Smoking) |>
  as.matrix()
data_mat
```

We can also represent the data as a matrix of empirical proportions (dividing each row by it's row sum).

```{r}
#| label: count-proportions
probs <- data_mat / rowSums(data_mat)
probs
```

This is the also the matrix of maximum likelihood estimates under the saturated model (**Why?**).
We can manually calculate the maximum log-likelihood of the saturated model using the multinomial distribution by first calculating the log-likelihood for each cell, (treating $0 \log(0) := 0$, as one of the cells in `data_mat` is a zero entry), and summing over all cells to get the total log-likelihood for the saturated model.

```{r}
#| label: loglike-matrix
loglike_matrix <- ifelse(data_mat > 0, data_mat * log(probs), 0)
loglike_matrix
```

```{r}
#| label: loglike-saturated
loglike_saturated <- sum(loglike_matrix)
loglike_saturated
```

::: {.callout-note collapse="true"}
## SELF TEST: Explain Why We Compute the Saturated Log-Likelihood with this Sum

The above computation is under the ungrouped view of the multinomial data as in [Appendix A2](A2-likelihood-multinomial.qmd){target="_blank"}.
The log-likelihood is then: $$
\sum_{i = 1}^6 \sum_{j = 1}^3 \Bigl(\log(p_{ij}) \tilde{y}_{ij} \Bigr),
$$ where $\tilde{y}_{ij}$ (the entries in `data_mat`) is the count of workers in response category $j$ (one of "Normal", "Border" or "Abnormal") within covariate group $i$ (defined by "Age" and "Smoking"), and we treat $0 \log(0) = 0$.
:::

We can recover the deviance reported in Table 3.3 of @fahrmeir_multivariate_2001 by computing twice the difference between the reported `deviance` of the fitted model and the log-likelihood of the saturated model (just manually computed).

```{r}
#| label: residual-dev-logit
residual_dev_logit <- 2 * loglike_saturated + deviance(btr_logit_inter_eff)
residual_dev_logit
```

This confirms that the so called `deviance` component of a `polr` object is actually just twice the negative of the log-likelihood of the fitted model (under the ungrouped view in [Appendix A2](A2-likelihood-multinomial.qmd){target="_blank"}).
This is very different from what we usually consider the deviance.

### Returning the Deviance for Alternative Link Functions

We also fit the models with complementary log-log and loglog link functions using effect coding, then summarize results in a comparison table similar to Table 3.3 of @fahrmeir_multivariate_2001, p.90:

::: callout-tip
The table is built in three steps:

1.  Fit one model per link and extract coefficient estimates with `coef(summary(...))[ ,"Value"]`.
2.  Build two small tibbles (one for coefficients, one for residual deviance) and stack them with `dplyr::bind_rows()`.
3.  Use `dplyr::mutate(across(where(is.numeric), ~ round(., 3)))` to round all numeric columns to 3 decimal places for cleaner display.

`rownames(coef(...))` supplies the parameter labels so each estimate aligns across link functions.
:::

```{r}
#| label: fit-alt-links-eff

# Fit complementary log-log (proportional hazards)
btr_ph_eff <- polr(
  BTR ~ Age + Smoking + Age:Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "cloglog"
)
btr_ph_eff_summary <- summary(btr_ph_eff)

# Fit loglog (extreme maximal-value)
btr_extm_eff <- polr(
  BTR ~ Age + Smoking + Age:Smoking,
  data = btr_dat,
  weights = Freq,
  Hess = TRUE,
  method = "loglog"
)
btr_extm_eff_summary <- summary(btr_extm_eff)

# Create a list of models for easy iteration
models <- list(
  Logistic = btr_logit_inter_eff_summary,
  `Complementary Log-Log` = btr_ph_eff_summary,
  Loglog = btr_extm_eff_summary
)

# Extract variable names
coef_names <- rownames(coef(models[[1]]))

results_table <- bind_rows(
  # Coefficients
  tibble(
    Parameter = coef_names,
    `Cum Log` = coef(btr_logit_inter_eff_summary)[, "Value"],
    `Prop Haz` = coef(btr_ph_eff_summary)[, "Value"],
    `Ext Max Val` = coef(btr_extm_eff_summary)[, "Value"]
  ),
  # Residual Deviance
  tibble(
    Parameter = "Deviance",
    `Cum Log` = 2 * loglike_saturated + deviance(btr_logit_inter_eff),
    `Prop Haz` = 2 * loglike_saturated + deviance(btr_ph_eff),
    `Ext Max Val` = 2 * loglike_saturated + deviance(btr_extm_eff)
  )
) |>
  mutate(across(where(is.numeric), ~ round(., 3)))

results_table
```

::: {.callout-tip collapse="true"}
## Alternative (completely optional!): Nested Data + Functional Mapping Workflow

This can be written in a nested/list-column style using functional programming.
The idea is to store one row per link function, fit each model via `purrr::map()`, then map again to extract coefficients and residual deviance.

```{r}
#| label: fit-alt-links-eff-functional

link_specs <- tibble(
  label = c("Cum Log", "Prop Haz", "Ext Max Val"),
  method = c("logistic", "cloglog", "loglog")
)

# Fit one model per row (list-column of models)
model_tbl <- link_specs |>
  mutate(
    model = map(
      method,
      ~ polr(
        BTR ~ Age + Smoking + Age:Smoking,
        data = btr_dat,
        weights = Freq,
        Hess = TRUE,
        method = .x
      )
    ),
    coef_value = map(model, ~ coef(summary(.x))[, "Value"]),
    residual_dev = map_dbl(model, ~ 2 * loglike_saturated + deviance(.x))
  )

# Expand mapped coefficients to long format
coef_long <- model_tbl |>
  mutate(
    Parameter = map(coef_value, names),
    estimate = coef_value
  ) |>
  dplyr::select(label, Parameter, estimate) |>
  tidyr::unnest(cols = c(Parameter, estimate))

# Add residual deviance row in matching long format
dev_long <- model_tbl |>
  transmute(
    label,
    Parameter = "Residual Deviance",
    estimate = residual_dev
  )

results_table_map <- bind_rows(coef_long, dev_long) |>
  pivot_wider(
    names_from = label,
    values_from = estimate
  ) |>
  mutate(across(where(is.numeric), ~ round(., 4)))

results_table_map
```
:::

We have now matched Table 3.3 of @fahrmeir_multivariate_2001 (with reverse signs for the non-threshold coefficients).

It is **highly recommended** that you read through Example 3.4 in the textbook to get an idea of intepreting the results.
It is worth considering here that the proportional hazard has the lowest deviance and hence the best fit.

## Other Tricks and Commands

### Residuals

The `MASS::polr()` function does not readily provide residuals.

```{r}
#| label: residuals-check
residuals(btr_logit_inter_eff)
```

### Fitted Values and Predictions

Fitted probabilities can be extracted directly or using the generic `predict()` function:

```{r}
#| label: fitted-probs

head(fitted(btr_logit_inter_eff))
head(predict(btr_logit_inter_eff, type = "probs"))
```

Both approaches yield identical results.

### Model Selection with AIC

The `step()` function performs automatic model selection using the AIC criterion.
(not directly covered in this course unless time allows)

```{r}
#| label: model-selection-aic
step(btr_logit_inter_eff)
```

### Variance-Covariance Matrix and Standard Errors

We can extract the estimated variance-covariance matrix of the parameter estimates using `vcov()`.
This should roughly line up with the inverse of the Fisher information matrix.

```{r}
#| label: vcov-extraction
logit_vcov <- vcov(btr_logit_inter_eff)
logit_vcov
```

From this we can match the standard errors reported in the summary output, since they are the square root of the diagonal elements of the variance-covariance matrix:

```{r}
#| label: vcov-comparison

btr_logit_inter_eff_coefs <- coef(btr_logit_inter_eff_summary)
btr_logit_inter_eff_coefs

se_from_vcov <- sqrt(diag(logit_vcov))
se_from_summary <- btr_logit_inter_eff_coefs[, "Std. Error"]

se_from_vcov == se_from_summary
```

We should be able to recover the same confidence intervals by using the variance-covariance matrix and a normal approximation (Wald intervals).

We can check the covariance matrix against the inverse of the Hessian matrix (Fisher information) returned by `MASS::polr()`; note that I can simply refer to it as `btr_logit_inter_eff$Hess` instead of `btr_logit_inter_eff$Hessian` because of [partial matching in R](https://humdrumr.ccml.gtcmt.gatech.edu/reference/partialMatching.html#:~:text=Partial%20matching%20explained,as%20do%20many%20humdrumR%20functions.):

```{r}
#| label: hessian-comparison
hessian_inv <- solve(btr_logit_inter_eff$Hess)
hessian_inv
```

We can visually see that the two matrices `hessian_inv` and `logit_vcov` show slight differences in certain smaller entries.
(The exact reason isn't known to us yet; mostly numerical.)

```{r}
#| label: 04-example3-2btr-chunk-32
all.equal(hessian_inv, logit_vcov)
```

### Confidence Intervals

R provides the generic method `confint()` to compute confidence intervals for model parameters.

```{r}
#| label: confint-profile
confint(btr_logit_inter_eff) |> round(3)
```

We can attempt to recovery the same confidence intervals using a Wald approximation based on the coefficient estimates and their standard errors:

```{r}
#| label: confint-wald-manual-upper
# Manual Wald CI (upper limit)
coeff_est <- btr_logit_inter_eff_coefs[, 1]
wald_upper <- coeff_est + qnorm(0.975) * sqrt(diag(vcov(btr_logit_inter_eff)))
wald_lower <- coeff_est + qnorm(0.025) * sqrt(diag(vcov(btr_logit_inter_eff)))
tibble(wald_lower, wald_upper)
```

These *Wald* confidence intervals are different from those computed by `confint()`; in particular, the Wald ones are symmetric around the parameter estimates, but those by `confint()` aren't.

This is because the `confint()` function uses the *profile likelihood* confidence intervals contructed by inverting the likelihood ratio tests, which is more suitable for some non-linear regression problems.
See @agresti_cda_2013 section 1.3.4 for constructing CIs by inverting tests, and also @agresti_cda_2013 section 3.2.6 for constructing profile likelihood confidence intervals.

Run `profile(btr_logit_inter_eff)` and see @venables_modern_2002, p. 220 for more details on the `profile()` function.

## Package Citations and Session Information

::: {.callout-tip collapse="true"}
## Session Information

```{r}
sessionInfo()
```
:::

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "MASS", "purrr", "readr", "tidyr")
for (pkg in packages) {
  cit <- citation(pkg)

  # Check if multiple citations
  if (length(cit) > 1) {
    cat("- **", pkg, "**:\n", sep = "")
    for (i in seq_along(cit)) {
      cit_text <- format(cit[i], style = "text")
      cat("    - ", cit_text, "\n\n", sep = "")
    }
  } else {
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
  }
}
```
