---
title: "Example 3.2 & 3.4: Breathing Test Results Analysis"
subtitle: "Ordinal Regression on Breathing Test Results"
date: today
---

## TODO

- Look at 3.3.3 pg 31/40 for variable information on thetas and gammas.
- Write up more about the different link functions (2025-04-11 lecture 23 min)

## Data Loading and Setup

```{r}
#| label: libs
library(conflicted)
library(dplyr)
library(MASS)
library(readr)
library(tidyr)
```

We load the BTR dataset.
It contains breathing test results (ordinal outcome) with covariates age and smoking status,
along with frequency counts for each combination.

```{r}
#| label: load-btr
btr_dat <- read_csv(
    "data/Example3-2BTR.csv",
    col_types = cols(
        BTR = col_factor(),
        Age = col_factor(),
        Smoking = col_factor(),
        Freq = col_double()
    )
)

btr_dat
```

The breathing test result (BTR) must be converted to an **ordered factor** since we have three ordered categories. By default, R uses **dummy coding** with `contr.treatment`.

First, we check if the BTR variable is currently an ordered factor:

```{r}
#| label: check-btr-ordered
is.ordered(btr_dat$BTR)
```

Currently it is not ordered, so we convert it to an ordered factor:
```{r}
#| label: order-btr
btr_dat$BTR <- as.ordered(btr_dat$BTR)
class(btr_dat$BTR)
btr_dat$BTR
```

The default ordering happens to be good for us,
but if we needed to change the order, we could specify it with `factor(..., levels = c(...), ordered = TRUE)`.

```{r}
#| label: check-factor-types
is.factor(btr_dat$Age)
is.factor(btr_dat$Smoking)
is.factor(btr_dat$Freq)
```

Examine the current contrast coding scheme:

```{r}
#| label: get-contrasts-opt
getOption("contrasts")
```

When you see "contr.treatment" for unordered factors, it means dummy coding is being used. For ordered factors, the default is "contr.poly" (orthogonal polynomials).
Age and Smoking are unordered factors, so they will use dummy coding.

```{r}
#| label: show-contrasts
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

Here it is choosing the never-smoked category as the reference for Smoking, and the <40 category as the reference for Age.
These are what we want to match @fahrmeir_multivariate_2001.

The dummy coding creates indicator variables for each level relative to a reference category. This is the default but leads to coefficient estimates that differ from Fahrmeir & Tutz due to their use of **effect coding**.

## Model Fitting: Proportional Odds Models

The `polr()` function fits cumulative logit (and other) models. Key note: **the sign of coefficients in `polr()` is opposite to the textbook formula**—[compare with the documentation](https://search.r-project.org/CRAN/refmans/MASS/html/polr.html) using `?polr` to reconcile.

Fit the cumulative logistic/proportional odds model with and intercept, age, smoking, and their interaction:

```{r}
#| label: fit-logit-int
btr_logit <- polr(
    BTR ~ Age * Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)
```

::: {.callout-tip collapse="true"}
## `polr()` Function Syntax

- **`formula`**: Model structure `response ~ predictors`; response must be an ordered factor
- **`data`**: Data frame containing the variables
- **`weights`**: Case weights (here, frequency counts for each covariate combination)
- **`Hess = TRUE`**: Return the Hessian matrix; needed for `summary()` and confidence intervals
- **`method`**: Link function—`"logistic"` (default, proportional odds), `"probit"`, `"cloglog"`, `"loglog"`, or `"cauchit"`
:::

The `polr()` function is limited.
It doesn't allowed for extended cumulative model where we can have threshold variables.
We can see that the only coefficients that change with the thresholds are the thresholds themselves shown below.

```{r}
#| label: summary-logit-int
summary(btr_logit)
```

There is only one coefficient for `Age`, since the other age category is the reference.
Similarly for `Smoking`, we only have two coefficients reported.
The interaction terms are also present as expected.

We need to be able to interpret the coefficients.
If you have a smoker who is Age 40 to 59 who is a former smokers, then the cumulative odds of having a normal test compared to not having a normal test is $\exp()$ the sum of those two estimates (0.158) than a person who has never smoked.

If smoking has no association with the breathing test results, then we would expect this number to be 1.


There is no residual readily available, but we can observe the linear predictor (represented as $\eta$ in the documentation).
```{r}
#| label: linear-predictor
# Linear predictors (eta values)
btr_logit$lp
```

For our notation, if we are trying to recover $x*\hat{\gamma}$, which is our version of the linear predictor, we need to invert the signs.

The $\zeta$'s are our theshold parameters $\theta$

```{r}
#| label: threshold-zeta
# Threshold parameters (cutpoints/intercepts)
btr_logit$zeta
```

We can verify the linear predictors by hand-computing the linear predictor:

```{r}
#| label: manual-lp-check
Z <- cbind(
    (btr_dat$Age == "40to59"),
    (btr_dat$Smoking == "2Former"),
    (btr_dat$Smoking == "3Current"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "2Former"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "3Current")
)

# Hand-computed eta should match btr_logit$lp
tibble(
    manual = c(Z %*% btr_logit$coeff),
    polr = btr_logit$lp
)
```

We see that they line up.

### Fit Model without Interaction

Here we fit a main-effects-only model without interaction:

```{r}
#| label: fit-logit-main
btr_logit1 <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit1)
```

Again the signs of the coefficients are the opposite to our model.

Since the interaction model is nested within the main effects model, we can compare them with a likelihood ratio test.
The test should have 2 degrees of freedom since we are adding two interaction terms (`Age40to59:Smoking2Former` and `Age40to59:Smoking3Current`).

We can use `anova()`:

```{r}
#| label: anova-logit-models
anova(btr_logit, btr_logit1)
```

We reject the null hypothesis that there is no interaction,
at all reasonable significance levels.

::: {.callout-note}
The LR statistic in the `anova()` output is the difference between the deviances of the two models. Each deviance reported is `-2 * log(likelihood)` for that model.
:::

## Alternative Link Functions

The proportional odds assumption can be tested by fitting models with different link functions.
The `polr()` function supports:

- **logistic**: cumulative logit model (default)
- **probit**: cumulative normal model
- **cloglog**: complementary log-log (grouped Cox/proportional hazards)
- **loglog**: extreme maximal-value distribution

### Grouped Cox / Proportional Hazards (Complementary Log-Log Link)

The grouped Cox model has

```{r}
#| label: fit-cloglog-int
btr_ph <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

summary(btr_ph)
```

### Extreme Maximal-Value Distribution

This model is just a reparameterisation of the proportional hazard model.

```{r}
#| label: fit-loglog-int
btr_extm <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

summary(btr_extm)
```

## Matching our Results to @fahrmeir_multivariate_2001 Example 3.4

By default, R uses dummy coding.
To match to the textbook, we need to use effect coding.
We switch the global contrast option and refit:

```{r}
#| label: set-effect-coding
options(contrasts = c("contr.sum", "contr.poly"))

# Verify the contrast change
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

We now have the reference category coded as -1, as required.
Therefore we refit both models:

```{r}
#| label: fit-logit-eff
btr_logit_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

btr_logit1_eff <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit_eff)
summary(btr_logit1_eff)
```

The values reported in the summaries now match the textbook (excluding rounding errors).

```{r}
#| label: anova-logit-eff
anova(btr_logit_eff, btr_logit1_eff)
```

As expected, the likelihood ratio test results are the same as before since the models are the same, just with different parameterisations.

## Saturated Model and Deviance Calculation

Is each of the regressed model summaries,
there is a value "Residual Deviance" reported.
We need to be careful to understand what this exactly means, because as we have previously seen,
this can be highly package dependent.

The deviance reported by `polr()` is `-2 * log(likelihood)` for the fitted model. To compute the **residual deviance** (difference from saturated model), we manually compute the saturated model log-likelihood.

To explore, we first compute the log-likelihood of the saturated model.
We create a matrix with each row corresponding to a unique covariate combination,
and each column counting the frequency of each response per category for that covariate combination.

```{r}
#| label: make-count-matrix
btr_counts <- btr_dat |>
    pivot_wider(
        id_cols = c(Age, Smoking),
        names_from = BTR,
        values_from = Freq
    )
btr_counts

data_mat <- btr_counts |>
    dplyr::select(-Age, -Smoking) |>
    as.matrix()
data_mat
```

We can also represent the data as a matrix of empirical proportions (dividing each row by it's row sum).

```{r}
#| label: count-proportions
probs <- data_mat / rowSums(data_mat)
probs
```

This is the also the matrix of maximum likelihood estimates
for the saturated model.
Therefore we manually calculate the log-likelihood of the saturated model
using the multinomial distribution.
We first calculate the likelihood for each cell, treating $0 \log(0) = 0$,
and then we sum over all cells to get the total log-likelihood for the saturated model.

```{r}
#| label: loglike-matrix
loglike_matrix <- ifelse(data_mat > 0, data_mat * log(probs), 0)
loglike_matrix
```

```{r}
#| label: loglike-saturated
loglike_saturated <- sum(loglike_matrix)
loglike_saturated
```

::: {.callout-note collapse="true"}
# Explain why we can compute the saturated log-likelihood with this sum
The above computation is under the ungrouped view of the data.
The probability mass function for the multinomial distribution is hence:
$$P(Y_1 = y_1, \ldots, Y_k = y_k) = \prod_{i = 1}^{6} p_1^{y_{i1}} p_2^{y_{i2}} p_3^{y_{i3}},$$
where $y_i$ are the observed counts and $p_i$ are the probabilities for each category.
The log-likelihood is then:
$$\sum_{i,j} y_{ij} \log(p_{ij}),$$
where we treat $0 \log(0) = 0$.
:::

We can recover the deviance reported in Table 3.3 of @fahrmeir_multivariate_2001
by computing the residual deviance as the difference
between the reported 'deviance' of the fitted model and the deviance of the saturated model.

```{r}
#| label: residual-dev-logit
residual_dev_logit <- 2 * loglike_saturated + btr_logit_eff$deviance
residual_dev_logit
```

This tells us that the so called 'deviance' reported by `polr()` is actually just
2 times negative of the log-likelihood of the fitted model.
This is very different from what we usually consider the deviance.

### Returning the Deviance for Alternative Link Functions

Fit models with complementary log-log and loglog link functions using effect coding, then summarize results in a comparison table:

```{r}
#| label: fit-alt-links-eff

# Fit complementary log-log (proportional hazards)
btr_ph_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

# Fit loglog (extreme maximal-value)
btr_extm_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

# Create a list of models for easy iteration
models <- list(
    Logistic = summary(btr_logit_eff),
    `Complementary Log-Log` = summary(btr_ph_eff),
    Loglog = summary(btr_extm_eff)
)

# Extract variable names
coef_names <- rownames(coef(models[[1]]))

results_table <- bind_rows(
    # Coefficients
    tibble(
        Parameter = coef_names,
        `Cum Log` = coef(summary(btr_logit_eff))[,"Value"],
        `Prop Haz` = coef(summary(btr_ph_eff))[,"Value"],
        `Ext Max Val` = coef(summary(btr_extm_eff))[,"Value"]
    ),
    # Residual Deviance
    tibble(
        Parameter = "Residual Deviance",
        `Cum Log` = 2 * loglike_saturated + btr_logit_eff$deviance,
        `Prop Haz` = 2 * loglike_saturated + btr_ph_eff$deviance,
        `Ext Max Val` = 2 * loglike_saturated + btr_extm_eff$deviance
    )
) |>
    mutate(across(where(is.numeric), ~ round(., 4)))

results_table

```

We have now matched Table 3.3 of @fahrmeir_multivariate_2001
(with inverse signs for the coefficients).

At this point, it is a good idea to read through Example 3.4 in the textbook
to get an idea of the results interpretation.

The proportional hazard has the lowest deviance, and gives a much smaller threshold parameter.

## Other Tricks and Commands

### Residuals

The `polr()` function does not readily provide residuals.

```{r}
#| label: residuals-check
residuals(btr_logit_eff)
```

### Fitted Values and Predictions

Fitted probabilities can be extracted directly or using the generic `predict()` function:

```{r}
#| label: fitted-probs
# Extracting fitted probabilities directly
head(btr_logit_eff$fitted)
head(predict(btr_logit_eff, type = "probs"))
```

Both approaches yield identical results.

### Model Selection with AIC

The `step()` function performs automatic model selection using the AIC criterion.
(not directly covered in this course unless time allows)

```{r}
#| label: model-selection-aic
step(btr_logit_eff)
```

### Variance-Covariance Matrix and Standard Errors

We can extract the estimated variance-covariance matrix of the parameter estimates using `vcov()`.
This should line up with the inverse of the Fisher information matrix.

```{r}
#| label: vcov-extraction
logit_vcov <- vcov(btr_logit_eff)
logit_vcov
```

From this we can match the standard errors reported in the summary output,
since they are the square root of the diagonal elements of the variance-covariance matrix:

```{r}
#| label: vcov-comparison
se_from_vcov <- sqrt(diag(logit_vcov))
se_from_summary <- summary(btr_logit_eff)$coefficients[, "Std. Error"]

se_from_vcov == se_from_summary
```
We should be able to recover the same confidence intervals by using the variance-covariance matrix and a normal approximation (Wald intervals).

We can check the covariance matrix against the inverse of the Hessian matrix (Fisher information) returned by `polr()`:

```{r}
#| label: hessian-comparison
hessian_inv <- solve(btr_logit_eff$Hess)
```

We can see that the matrices are slightly different.
```{r}
all.equal(hessian_inv, logit_vcov)
```

### Confidence Intervals

R provides the generic method `confint()` to compute confidence intervals for model parameters.

```{r}
#| label: confint-profile
confint(btr_logit_eff) |> round(3)
```

We can attempt to recovery the same confidence intervals using a Wald approximation based on the coefficient estimates and their standard errors:

```{r}
#| label: confint-wald-manual-upper
# Manual Wald CI (upper limit)
coeff_est <- summary(btr_logit_eff)$coefficients[, 1]
wald_upper <- coeff_est + qnorm(0.975) * sqrt(diag(vcov(btr_logit_eff)))
wald_lower <- coeff_est + qnorm(0.025) * sqrt(diag(vcov(btr_logit_eff)))
tibble(wald_lower, wald_upper)
```

These values are different from the confidence intervals computed by `confint()`.
This is because the `confint()` function uses the profile likelihood method,
which is more suitable for some non-linear regression problems.

For more details see @venables_modern_2002, p. 220,
and run `profile(btr_logit_eff)`.

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "MASS", "readr")
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```