---
title: "Ordinal Regression on Breathing Test Results."
subtitle: "@fahrmeir_multivariate_2001 Examples 3.2 & 3.4: Breathing Test Results"
date: today
---

## TODO

- Look at 3.3.3 pg 31/40 for variable information on thetas and gammas.
- Write up more about the different link functions (2025-04-11 lecture 23 min)

## Data Loading and Setup

```{r}
#| label: libs
library(conflicted)
library(dplyr)
library(MASS)
library(purrr)
library(readr)
library(tidyr)
```

We load the BTR dataset.
It contains breathing test results normal, borderline, and abnormal
(which we can consider ordered categories),
and considered the effects of the covariates age and smoking status.

```{r}
#| label: load-btr
btr_dat <- read_csv(
    "data/Example3-2BTR.csv",
    col_types = cols(
        BTR = col_factor(),
        Age = col_factor(),
        Smoking = col_factor(),
        Freq = col_double()
    )
)

btr_dat
```

We can see that `BTR`, `Age`, and `Smoking` are all factors.
We check if the breathing test result (`BTR`) is encoded as an ordered factor,
and if not, we convert it to an ordered factor.

```{r}
#| label: order-btr
is.ordered(btr_dat$BTR)
btr_dat$BTR <- as.ordered(btr_dat$BTR)
```

We can now confirm the conversion was successful.
```{r}
#| label: check-btr-ordered
class(btr_dat$BTR)
btr_dat$BTR
```

The default ordering happens to be good for us,
but if we needed to change the order, we could specify it with
`factor(..., levels = c(...), ordered = TRUE)`.

We now examine the current contrast coding scheme:
```{r}
#| label: get-contrasts-opt
getOption("contrasts")
```

`"contr.treatment"` corresponds to dummy coding for unordered factors.
For ordered factors, the default is `"contr.poly"` (orthogonal polynomials).
We consider `Age` and `Smoking` as unordered factors, and we will use dummy coding.

```{r}
#| label: show-contrasts
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

The default reference categories are `never-smoked` for `Smoking`,
and the `<40` for Age,
which matches @fahrmeir_multivariate_2001,
however they use **effect coding** rather than **dummy coding.**

::: {.callout-note}
- **Dummy coding**: Reference category is 0, other categories are 1;
coefficients represent difference from reference category
- **Effect coding**: Reference category is -1, other categories are 0 or 1;
coefficients represent difference from overall mean
:::

## Model Fitting: Proportional Odds Models

The `polr()` function fits cumulative logit (and other) models.
[Looking at the documentation](https://search.r-project.org/CRAN/refmans/MASS/html/polr.html)
using `?polr`,
the sign of coefficients in `polr()` is **opposite** to the textbook formula.

We first fit a cumulative logistic/proportional odds model
with an intercept, age, smoking.
We also include an interaction effect between age and smoking.

```{r}
#| label: fit-logit-int
btr_logit_inter <- polr(
    BTR ~ Age * Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)
```

::: {.callout-tip collapse="true"}
## `polr()` Function Syntax

- **`formula`**: Model structure `response ~ predictors`;
response must be an ordered factor
- **`data`**: Data frame containing the variables
- **`weights`**: Case weights (here, frequency counts for each covariate combination)
- **`Hess = TRUE`**: Return the Hessian matrix;
needed for `summary()` and confidence intervals
- **`method`**: Link function---`"logistic"`
(default, proportional odds), `"probit"`, `"cloglog"`, `"loglog"`, or `"cauchit"`
:::

The `polr()` function is limited.
It can model standard cumulative models
(see section 3.3.1 of @fahrmeir_multivariate_2001),
but doesn't allow for extended cumulative model with threshold variables
(see section 3.3.2 of @fahrmeir_multivariate_2001).

```{r}
#| label: summary-logit-int
summary(btr_logit_inter)
```

::: {.callout-important collapse="true"}
## SELF TEST: Interpret Coefficients in `polr()` for a Former Smoker age 40 to 59

A former smoker who is Age 40 to 59 then the cumulative odds of having a normal test
compared to not having a normal test is $\exp(2.8334)$
times the odds for a person who has never smoked.

If smoking has no association with the breathing test results,
then we would expect this number to be 1.
:::


There is no residual readily available, but we can observe the linear predictor
(represented as $\eta$ in the documentation,
but referred to in our subject with $\mathbf{x} \hat{\boldsymbol{\gamma}}$).
```{r}
#| label: linear-predictor
btr_logit_inter$lp
```

The $\zeta$'s are our theshold parameters $\theta$

```{r}
#| label: threshold-zeta
btr_logit_inter$zeta
```

We verify the linear predictors by hand-computing the linear predictor.

```{r}
#| label: manual-lp-check
Z <- cbind(
    (btr_dat$Age == "40to59"),
    (btr_dat$Smoking == "2Former"),
    (btr_dat$Smoking == "3Current"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "2Former"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "3Current")
)

tibble(
    manual = c(Z %*% btr_logit_inter$coeff),
    polr = btr_logit_inter$lp
)
```

We see that they line up.

### Fit Model without Interaction

We fit a main-effects-only model without interaction.

```{r}
#| label: fit-logit-main
btr_logit_main <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit_main)
```

Again the signs of the coefficients are the opposite to our model.

Since the interaction model is nested within the main effects model,
we can compare them with a likelihood ratio test.

::: {.callout-important collapse="true"}
## SELF TEST: How Many Degrees of Freedom will the Likelihood Ratio Test have?
The test has 2 degrees of freedom since we are adding two interaction terms (`Age40to59:Smoking2Former` and `Age40to59:Smoking3Current`).
:::

We can use `anova()`:

```{r}
#| label: anova-logit-models
anova(btr_logit_inter, btr_logit_main)
```

::: {.callout-important collapse="true"}
## SELF TEST: Interpret the Likelihood Ratio Test Results

We reject the null hypothesis that there is no interaction,
at all reasonable significance levels.
:::

::: {.callout-note}
The LR statistic in the `anova()` output is the difference
between the deviances of the two models.
Each deviance reported is $-2 \times \log(\text{likelihood})$ for that model.
:::

## Alternative Link Functions

The proportional odds assumption can be tested
by fitting models with different link functions.
The `polr()` function supports:

- **logistic**: cumulative logit model (default)
- **probit**: cumulative normal model
- **cloglog**: complementary log-log (grouped Cox/proportional hazards)
- **loglog**: extreme maximal-value distribution.

See Figure 3.1 of @fahrmeir_multivariate_2001 for some good diagrams
demonstrating the shapes of these link functions.

### Grouped Cox / Proportional Hazards (Complementary Log-Log Link)

We run a grouped Cox model with the complementary log-log link function.

```{r}
#| label: fit-cloglog-int
btr_ph <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

summary(btr_ph)
```

### Extreme Maximal-Value Distribution

This model is just a reparameterisation of the proportional hazard model.

```{r}
#| label: fit-loglog-int
btr_extm <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

summary(btr_extm)
```

## Matching our Results to @fahrmeir_multivariate_2001 Example 3.4

By default, R uses dummy coding.
To match to the textbook, we need to use effect coding.
We switch the global contrast option and refit.

```{r}
#| label: set-effect-coding
options(contrasts = c("contr.sum", "contr.poly"))

# Verify the contrast change
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

We now have the final categories coded as -1, as required.
Therefore we refit both models:

```{r}
#| label: fit-logit-eff
btr_logit_inter_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

btr_logit_main_eff <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit_inter_eff)
summary(btr_logit_main_eff)
```

The values reported in the summaries now match the textbook (excluding rounding errors).

In addition, we can redo the likelihood ratio test with the new models
to confirm we get the same result.

```{r}
#| label: anova-logit-eff
anova(btr_logit_inter_eff, btr_logit_main_eff)
```

As expected, the likelihood ratio test results are the same as before.

## Saturated Model and Deviance Calculation

In each of the regressed model summaries,
there is a value "Residual Deviance" reported.
We need to be careful to understand what this exactly means,
because as we have previously seen,
this can be highly package dependent.

The deviance reported by `polr()` is
$-2 \times \log(\text{likelihood})$ for the fitted model.
We can manually compute the **residual deviance** (difference from saturated model).

We first compute the log-likelihood of the saturated model.
We create a matrix with each row corresponding to a unique covariate combination,
and each column counting the frequency of each response per category
for that covariate combination.

```{r}
#| label: make-count-matrix
btr_counts <- btr_dat |>
    pivot_wider(
        id_cols = c(Age, Smoking),
        names_from = BTR,
        values_from = Freq
    )
btr_counts

data_mat <- btr_counts |>
    dplyr::select(-Age, -Smoking) |>
    as.matrix()
data_mat
```

We can also represent the data as a matrix of empirical proportions
(dividing each row by it's row sum).

```{r}
#| label: count-proportions
probs <- data_mat / rowSums(data_mat)
probs
```

This is the also the matrix of maximum likelihood estimates
for the saturated model.
We can manually calculate the log-likelihood of the saturated model
using the multinomial distribution
by first calculating the likelihood for each cell, (treating $0 \log(0) := 0$),
and summing over all cells to get the total log-likelihood for the saturated model.

```{r}
#| label: loglike-matrix
loglike_matrix <- ifelse(data_mat > 0, data_mat * log(probs), 0)
loglike_matrix
```

```{r}
#| label: loglike-saturated
loglike_saturated <- sum(loglike_matrix)
loglike_saturated
```

::: {.callout-note collapse="true"}
## SELF TEST: Explain Why We Compute the Saturated Log-Likelihood with this Sum

The above computation is under the ungrouped view of the data.
The probability mass function for the multinomial distribution is hence:
$$
    \Pr(Y_1 = y_1, \ldots, Y_k = y_k)
    = \prod_{i = 1}^{6} p_1^{y_{i1}} p_2^{y_{i2}} p_3^{y_{i3}},
$$
where $y_i$ are the observed counts and $p_i$ are the probabilities for each category.
The log-likelihood is then:
$$\sum_{i,j} y_{ij} \log(p_{ij}),$$
where we treat $0 \log(0) = 0$.
:::

We can recover the deviance reported in Table 3.3 of @fahrmeir_multivariate_2001
by computing the residual deviance as the difference
between the reported 'deviance' of the fitted model
and the deviance of the saturated model.

```{r}
#| label: residual-dev-logit
residual_dev_logit <- 2 * loglike_saturated + btr_logit_inter_eff$deviance
residual_dev_logit
```

This confirms that the so called 'deviance' reported by `polr()`
is actually just twice the negative of the log-likelihood of the fitted model.
This is very different from what we usually consider the deviance.

### Returning the Deviance for Alternative Link Functions

We fit models with complementary log-log and loglog link functions using effect coding,
then summarize results in a comparison table:

::: {.callout-tip}
The table is built in three steps:

1. Fit one model per link and extract coefficient estimates
with `coef(summary(...))[ ,"Value"]`.
2. Build two small tibbles (one for coefficients, one for residual deviance)
and stack them with `bind_rows()`.
3. Use `mutate(across(where(is.numeric), ~ round(., 4)))`
to round all numeric columns for cleaner display.

`rownames(coef(...))` supplies the parameter labels
so each estimate aligns across link functions.
:::

```{r}
#| label: fit-alt-links-eff

# Fit complementary log-log (proportional hazards)
btr_ph_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

# Fit loglog (extreme maximal-value)
btr_extm_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

# Create a list of models for easy iteration
models <- list(
    Logistic = summary(btr_logit_inter_eff),
    `Complementary Log-Log` = summary(btr_ph_eff),
    Loglog = summary(btr_extm_eff)
)

# Extract variable names
coef_names <- rownames(coef(models[[1]]))

results_table <- bind_rows(
    # Coefficients
    tibble(
        Parameter = coef_names,
        `Cum Log` = coef(summary(btr_logit_inter_eff))[,"Value"],
        `Prop Haz` = coef(summary(btr_ph_eff))[,"Value"],
        `Ext Max Val` = coef(summary(btr_extm_eff))[,"Value"]
    ),
    # Residual Deviance
    tibble(
        Parameter = "Residual Deviance",
        `Cum Log` = 2 * loglike_saturated + btr_logit_inter_eff$deviance,
        `Prop Haz` = 2 * loglike_saturated + btr_ph_eff$deviance,
        `Ext Max Val` = 2 * loglike_saturated + btr_extm_eff$deviance
    )
) |>
    mutate(across(where(is.numeric), ~ round(., 4)))

results_table
```

::: {.callout-tip collapse="true"}
## Alternative: Nested Data + Functional Mapping Workflow

This can be written in a nested/list-column style using functional programming.
The idea is to store one row per link function, fit each model via `map()`,
then map again to extract coefficients and residual deviance.

```{r}
#| label: fit-alt-links-eff-functional

link_specs <- tibble(
    label = c("Cum Log", "Prop Haz", "Ext Max Val"),
    method = c("logistic", "cloglog", "loglog")
)

# Fit one model per row (list-column of models)
model_tbl <- link_specs |>
    mutate(
        model = map(
            method,
            ~ polr(
                BTR ~ Age + Smoking + Age:Smoking,
                data = btr_dat,
                weights = Freq,
                Hess = TRUE,
                method = .x
            )
        ),
        coef_value = map(model, ~ coef(summary(.x))[ , "Value"]),
        residual_dev = map_dbl(model, ~ 2 * loglike_saturated + .x$deviance)
    )

# Expand mapped coefficients to long format
coef_long <- model_tbl |>
    mutate(
        Parameter = map(coef_value, names),
        estimate = coef_value
    ) |>
    dplyr::select(label, Parameter, estimate) |>
    tidyr::unnest(cols = c(Parameter, estimate))

# Add residual deviance row in matching long format
dev_long <- model_tbl |>
    transmute(
        label,
        Parameter = "Residual Deviance",
        estimate = residual_dev
    )

results_table_map <- bind_rows(coef_long, dev_long) |>
    pivot_wider(
        names_from = label,
        values_from = estimate
    ) |>
    mutate(across(where(is.numeric), ~ round(., 4)))

results_table_map
```

:::

We have now matched Table 3.3 of @fahrmeir_multivariate_2001
(with inverse signs for the coefficients).

It is **highly recommended** that you read through Example 3.4 in the textbook
to get an idea of intepreting the results. It is worth considering here
that the proportional hazard has the lowest deviance,
and gives a much smaller threshold parameter.

## Other Tricks and Commands

### Residuals

The `polr()` function does not readily provide residuals.

```{r}
#| label: residuals-check
residuals(btr_logit_inter_eff)
```

### Fitted Values and Predictions

Fitted probabilities can be extracted directly
or using the generic `predict()` function:

```{r}
#| label: fitted-probs

head(btr_logit_inter_eff$fitted)
head(predict(btr_logit_inter_eff, type = "probs"))
```

Both approaches yield identical results.

### Model Selection with AIC

The `step()` function performs automatic model selection using the AIC criterion.
(not directly covered in this course unless time allows)

```{r}
#| label: model-selection-aic
step(btr_logit_inter_eff)
```

### Variance-Covariance Matrix and Standard Errors

We can extract the estimated variance-covariance matrix of the parameter estimates
using `vcov()`.
This should line up with the inverse of the Fisher information matrix.

```{r}
#| label: vcov-extraction
logit_vcov <- vcov(btr_logit_inter_eff)
logit_vcov
```

From this we can match the standard errors reported in the summary output,
since they are the square root of the diagonal elements
of the variance-covariance matrix:

```{r}
#| label: vcov-comparison
se_from_vcov <- sqrt(diag(logit_vcov))
se_from_summary <- summary(btr_logit_inter_eff)$coefficients[, "Std. Error"]

se_from_vcov == se_from_summary
```

We should be able to recover the same confidence intervals
by using the variance-covariance matrix and a normal approximation (Wald intervals).

We can check the covariance matrix
against the inverse of the Hessian matrix (Fisher information) returned by `polr()`:

```{r}
#| label: hessian-comparison
hessian_inv <- solve(btr_logit_inter_eff$Hess)
hessian_inv
```

We can see that the matrices are slightly different.
```{r}
#| label: 04-example3-2btr-chunk-32
all.equal(hessian_inv, logit_vcov)
```

### Confidence Intervals

R provides the generic method `confint()`
to compute confidence intervals for model parameters.

```{r}
#| label: confint-profile
confint(btr_logit_inter_eff) |> round(3)
```

We can attempt to recovery the same confidence intervals using a Wald approximation
based on the coefficient estimates and their standard errors:

```{r}
#| label: confint-wald-manual-upper
# Manual Wald CI (upper limit)
coeff_est <- summary(btr_logit_inter_eff)$coefficients[, 1]
wald_upper <- coeff_est + qnorm(0.975) * sqrt(diag(vcov(btr_logit_inter_eff)))
wald_lower <- coeff_est + qnorm(0.025) * sqrt(diag(vcov(btr_logit_inter_eff)))
tibble(wald_lower, wald_upper)
```

These values are different from the confidence intervals computed by `confint()`.
This is because the `confint()` function uses the profile likelihood method,
which is more suitable for some non-linear regression problems.

For more details see @venables_modern_2002, p. 220,
and run `profile(btr_logit_inter_eff)`.

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "MASS", "purrr", "readr", "tidyr")
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```