---
title: "Example 3.2 & 3.4: Breathing Test Results Analysis"
subtitle: "Ordinal Regression on Breathing Test Results"
date: today
---

## TODO

- Look at 3.3.3 pg 31/40 for variable information on thetas and gammas.
- Write up more about the different link functions (2025-04-11 lecture 23 min)

## Data Loading and Setup

```{r}
library(conflicted)
library(dplyr)
library(MASS)
library(readr)
```

We load the BTR dataset.
It contains breathing test results (ordinal outcome) with covariates age and smoking status,
along with frequency counts for each combination.

```{r}
btr_dat <- read_csv(
    "data/Example3-2BTR.csv",
    col_types = cols(
        BTR = col_factor(),
        Age = col_factor(),
        Smoking = col_factor(),
        Freq = col_double()
    )
)

btr_dat
```

The breathing test result (BTR) must be converted to an **ordered factor** since we have three ordered categories. By default, R uses **dummy coding** with `contr.treatment`.

First, we check if the BTR variable is currently an ordered factor:

```{r}
is.ordered(btr_dat$BTR)
```

Currently it is not ordered, so we convert it to an ordered factor:
```{r}
btr_dat$BTR <- as.ordered(btr_dat$BTR)
class(btr_dat$BTR)
btr_dat$BTR
```

The default ordering happens to be good for us,
but if we needed to change the order, we could specify it with `factor(..., levels = c(...), ordered = TRUE)`.

```{r}
is.factor(btr_dat$Age)
is.factor(btr_dat$Smoking)
is.factor(btr_dat$Freq)
```

Examine the current contrast coding scheme:

```{r}
getOption("contrasts")
```

When you see "contr.treatment" for unordered factors, it means dummy coding is being used. For ordered factors, the default is "contr.poly" (orthogonal polynomials).
Age and Smoking are unordered factors, so they will use dummy coding.

```{r}
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

Here it is choosing the never-smoked category as the reference for Smoking, and the <40 category as the reference for Age.
These are what we want to match @fahrmeir_multivariate_2001.

The dummy coding creates indicator variables for each level relative to a reference category. This is the default but leads to coefficient estimates that differ from Fahrmeir & Tutz due to their use of **effect coding**.

## Model Fitting: Proportional Odds Models

The `polr()` function fits cumulative logit (and other) models. Key note: **the sign of coefficients in `polr()` is opposite to the textbook formula**—[compare with the documentation](https://search.r-project.org/CRAN/refmans/MASS/html/polr.html) using `?polr` to reconcile.

Fit the cumulative logistic/proportional odds model with and intercept, age, smoking, and their interaction:

```{r}
btr_logit <- polr(
    BTR ~ Age * Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)
```

::: {.callout-tip collapse="true"}
## `polr()` Function Syntax

- **`formula`**: Model structure `response ~ predictors`; response must be an ordered factor
- **`data`**: Data frame containing the variables
- **`weights`**: Case weights (here, frequency counts for each covariate combination)
- **`Hess = TRUE`**: Return the Hessian matrix; needed for `summary()` and confidence intervals
- **`method`**: Link function—`"logistic"` (default, proportional odds), `"probit"`, `"cloglog"`, `"loglog"`, or `"cauchit"`
:::

The `polr()` function is limited.
It doesn't allowed for extended cumulative model where we can have threshold variables.
We can see that the only coefficients that change with the thresholds are the thresholds themselves shown below.

```{r}
summary(btr_logit)
```

There is only one coefficient for `Age`, since the other age category is the reference.
Similarly for `Smoking`, we only have two coefficients reported.
The interaction terms are also present as expected.

We need to be able to interpret the coefficients.
If you have a smoker who is Age 40 to 59 who is a former smokers, then the cumulative odds of having a normal test compared to not having a normal test is $\exp()$ the sum of those two estimates (0.158) than a person who has never smoked.

If smoking has no association with the breathing test results, then we would expect this number to be 1.


There is no residual readily available, but we can observe the linear predictor (represented as $\eta$ in the documentation).
```{r}
# Linear predictors (eta values)
btr_logit$lp
```

For our notation, if we are trying to recover $x*\hat{\gamma}$, which is our version of the linear predictor, we need to invert the signs.

The $\zeta$'s are our theshold parameters $\theta$

```{r}
# Threshold parameters (cutpoints/intercepts)
btr_logit$zeta
```

We can verify the linear predictors by hand-computing the linear predictor:

```{r}
Z <- cbind(
    (btr_dat$Age == "40to59"),
    (btr_dat$Smoking == "2Former"),
    (btr_dat$Smoking == "3Current"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "2Former"),
    (btr_dat$Age == "40to59") * (btr_dat$Smoking == "3Current")
)

# Hand-computed eta should match btr_logit$lp
tibble(
    manual = c(Z %*% btr_logit$coeff),
    polr = btr_logit$lp
)
```

We see that they line up.

### Fit Model without Interaction

Here we fit a main-effects-only model without interaction:

```{r}
btr_logit1 <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit1)
```

Again the signs of the coefficients are the opposite to our model.

Since the interaction model is nested within the main effects model, we can compare them with a likelihood ratio test.
The test should have 2 degrees of freedom since we are adding two interaction terms (`Age40to59:Smoking2Former` and `Age40to59:Smoking3Current`).

We can use `anova()`:

```{r}
anova(btr_logit, btr_logit1)
```

We reject the null hypothesis that there is no interaction,
at all reasonable significance levels.

::: {.callout-note}
The LR statistic in the `anova()` output is the difference between the deviances of the two models. Each deviance reported is `-2 * log(likelihood)` for that model.
:::

## Alternative Link Functions

The proportional odds assumption can be tested by fitting models with different link functions.
The `polr()` function supports:

- **logistic**: cumulative logit model (default)
- **probit**: cumulative normal model
- **cloglog**: complementary log-log (grouped Cox/proportional hazards)
- **loglog**: extreme maximal-value distribution

### Grouped Cox / Proportional Hazards (Complementary Log-Log Link)

The grouped Cox model has

```{r}
btr_ph <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

summary(btr_ph)
```

### Extreme Maximal-Value Distribution

This model is just a reparameterisation of the proportional hazard model.

```{r}
btr_extm <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

summary(btr_extm)
```

## Matching our Results to @fahrmeir_multivariate_2001 Example 3.4

By default, R uses dummy coding.
To match to the textbook, we need to use effect coding.
We switch the global contrast option and refit:

```{r}
options(contrasts = c("contr.sum", "contr.poly"))

# Verify the contrast change
contrasts(btr_dat$Smoking)
contrasts(btr_dat$Age)
```

We now have the reference category coded as -1, as required.
Therefore we refit both models:

```{r}
btr_logit_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

btr_logit1_eff <- polr(
    BTR ~ Age + Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(btr_logit_eff)
summary(btr_logit1_eff)
```

The values reported in the summaries now match the textbook (excluding rounding errors).

```{r}
anova(btr_logit_eff, btr_logit1_eff)
```

As expected, the likelihood ratio test results are the same as before since the models are the same, just with different parameterisations.

## Saturated Model and Deviance Calculation

The deviance reported by `polr()` is `-2 * log(likelihood)` for the fitted model. To compute the **residual deviance** (difference from saturated model), we manually compute the saturated model log-likelihood.

```{r}
# Reshape frequency counts into a matrix (rows = covariate combinations, cols = BTR categories)
data_mat <- t(matrix(btr_dat$Freq, nr = 3))

# Compute MLE probabilities for the saturated model (empirical proportions)
probs <- data_mat / rowSums(data_mat)

# Compute saturated log-likelihood via multinomial
loglike_saturated <- 0
for (row in seq_len(nrow(data_mat))) {
    for (col in 1:3) {
        eval_vec <- rep(0, 3)
        eval_vec[col] <- 1
        increment <- data_mat[row, col] *
            dmultinom(x = eval_vec, prob = probs[row, ], log = TRUE)
        if (!is.na(increment)) {
            loglike_saturated <- loglike_saturated + increment
        }
    }
}

cat("Saturated model log-likelihood:", loglike_saturated, "\n")
```

Compute residual deviances (deviance of fitted model minus deviance of saturated model):

```{r}
residual_dev_logit <- btr_logit_eff$deviance - abs(2 * loglike_saturated)
residual_dev_logit1 <- btr_logit1_eff$deviance - abs(2 * loglike_saturated)

cat("Residual deviance (interaction):", residual_dev_logit, "\n")
cat("Residual deviance (main effects):", residual_dev_logit1, "\n")
cat("These should match Table 3.3 in Fahrmeir & Tutz\n")
```

Compute residual deviances for alternative link functions:

```{r}
# Refit with effect coding
btr_ph_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "cloglog"
)

btr_extm_eff <- polr(
    BTR ~ Age + Smoking + Age:Smoking,
    data = btr_dat,
    weights = Freq,
    Hess = TRUE,
    method = "loglog"
)

residual_dev_ph <- btr_ph_eff$deviance - abs(2 * loglike_saturated)
residual_dev_extm <- btr_extm_eff$deviance - abs(2 * loglike_saturated)

cat("Residual deviance (cloglog):", residual_dev_ph, "\n")
cat("Residual deviance (loglog):", residual_dev_extm, "\n")
```

## Model Diagnostics

### Fitted Values and Predictions

```{r}
# Fitted probabilities (first few rows)
head(predict(btr_logit_eff, btr_dat, type = "probs"))
```

### Model Selection with AIC

```{r}
# Automatic stepwise selection based on AIC
btr_step <- step(btr_logit_eff)

summary(btr_step)
```

### Variance-Covariance Matrix and Confidence Intervals

```{r}
# Variance-covariance matrix of parameter estimates
vcov(btr_logit_eff)

# Standard errors from summary
sqrt(diag(vcov(btr_logit_eff)))
```

Compute confidence intervals using the profile likelihood (more accurate than Wald intervals):

```{r}
confint(btr_logit_eff)
```

### Profile Likelihood

For detailed inference, examine the profile likelihood for individual parameters:

```{r}
profile(btr_logit_eff)
```

```{r}
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "MASS", "readr")
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```