---
title: "Sequential Regression: Tonsil Size and Streptococcus"
subtitle: "@fahrmeir_multivariate_2001 Examples 3.6 and 3.7"
date: today
nocite: |
    @fahrmeir_multivariate_2001
---

## Overview

This script reproduces Examples 3.6 and 3.7 from @fahrmeir_multivariate_2001.
We analyse the relationship between tonsil size of children and their
*Streptococcus pyogenes* carrier status.

The response has three ordered categories:

1. `present`
2. `enlarged`
3. `great_enlarged`.

There is a good physical interpretation of the ordering,
where a normal tonsil must first become enlarged before it can become greatly enlarged,
and so we are motivated to try a sequential regression model.

## Data and Packages

```{r}
#| label: 05-tonsil-load-packages
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(MASS)
library(tidyr)
library(VGAM)
```

Run `?tonsil` for information about the data.
This dataset is very simple, since it only contains one predictor.
We load it and preview the contents to see how it's structured.

```{r}
#| label: 05-tonsil-load-and-preview-data
data(tonsil)
tonsil_tbl <- as_tibble(tonsil)
tonsil_tbl
```

### Data Reshaping

The `tonsil` data are in long format: one row per carrier-status and size category.
For `vglm`, we want one row per carrier-status and one column per size count,
referred to as wide format.
We can reshape using `tidyr::pivot_wider()`.
We also rename the size category to be more descriptive
and convert it to a factor with labels.

```{r}
#| label: 05-tonsil-reshape-to-wide
tonsil_wide <- tonsil_tbl |>
    mutate(
        Size = factor(
            Size,
            levels = c(1, 2, 3),
            labels = c("present", "enlarged", "great_enlarged")
        )
    ) |>
    dplyr::select(Streptococcus.p, Size, n) |>
    pivot_wider(names_from = Size, values_from = n)

tonsil_wide
```

::: {.callout-tip collapse="true"}
## What each step is doing

- `mutate(...)`:
creates or updates columns; here it rewrites `Size` as a labelled factor.
- `factor(...)`: converts numeric codes (`1, 2, 3`)
into ordered category labels used in output.
- `dplyr::select(Streptococcus.p, Size, n)`:
keeps only the columns needed for reshaping.
- `pivot_wider(names_from = Size, values_from = n)`:
converts long data into wide data, so each size category becomes its own count column.
- `|>`:
passes the result of one step to the next, making the workflow readable top-to-bottom.
:::

### Understanding Factor Coding

As in the other scripts, we want to ensure effect coding (`contr.sum`)
is used for factors.
This gives coefficients that are directly comparable to the textbook output.

```{r}
#| label: 05-tonsil-set-factor-contrasts
contrasts(tonsil_wide$Streptococcus.p)
options(contrasts = c("contr.sum", "contr.poly"))
contrasts(tonsil_wide$Streptococcus.p)
```

## Model 1: Cumulative Link Model (Proportional Odds)

We first fit a cumulative logit model.

```{r}
#| label: 05-tonsil-fit-cumulative-parallel
fit_cumulative <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = cumulative(parallel = TRUE ~ Streptococcus.p - 1),
    data = tonsil_wide
)

summary(fit_cumulative)
```

::: {.callout-caution}
Here, the `parallel = TRUE ~ Streptococcus.p - 1`
argument specifies that the effect of `Streptococcus.p` is the same across all levels,
and the `- 1` removes the intercept from the model,
which sets the intercept to be different,
since the intercepts here are exactly the $\theta$s.
:::

Note that the $p$-values do not line up with those
in Table 3.6 of @fahrmeir_multivariate_2001 for a completely unknown reason.
If you manage to figure out why, please let me know.

## Model 2: Extended Cumulative Model

Here we don't set `parallel = TRUE` and allow category-specific effects.

```{r}
#| label: 05-tonsil-fit-cumulative-nonparallel
fit_cumulative_extended <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = cumulative(),
    data = tonsil_wide
)

summary(fit_cumulative_extended)
```

## Model 3: Sequential Model

Here we fit a sequential model using `VGAM::sratio()`.

::: {.callout-important collapse="true"}
## Make sure you can write down the mathematical form of the model. Expand for hint.

HINT: it should have the form:
\begin{equation*}
    \Pr(Y = r | Y \geq r, x) = \ldots
\end{equation*}

Check @fahrmeir_multivariate_2001 page 94 for answer.
:::

```{r}
#| label: 05-tonsil-fit-sequential-parallel
fit_sequential <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = sratio(parallel = TRUE ~ Streptococcus.p - 1),
    data = tonsil_wide
)

summary(fit_sequential)
```

Again, the $p$-values do not line up with those in Table 3.6 for an unknown reason,
but the deviances do match.
We can see that the deviance for the sequential model is much smaller
than that of the cumulative model which suggests a much better fit to the data.

::: {.callout-tip collapse="true"}
## Alternative: MASS::polr() Implementation

Here we fit the same cumulative model using `MASS::polr()` to verify our $p$-values.
We convert our data back to long format.

```{r}
#| label: 05-tonsil-prepare-polr-data
polr_dat <- tonsil_wide |>
    pivot_longer(
        cols = c(present, enlarged, great_enlarged),
        names_to = "size",
        values_to = "weight"
    ) |>
    mutate(
        size = ordered(size, levels = c("present", "enlarged", "great_enlarged"))
    )

polr_dat
```

::: {.callout-tip collapse="true"}
## What each step is doing

- `pivot_longer(...)`: turns the wide count columns back into long format (one row per category).
- `cols = c(present, enlarged, great_enlarged)`: chooses which columns to stack into one variable.
- `names_to = "size"`: stores the former column names as category labels in a new `size` column.
- `values_to = "weight"`: stores the corresponding counts in a new `weight` column used as case weights.
- `ordered(...)`: makes `size` an ordered factor so `polr()` respects the category order.
- `|>`: pipes each intermediate result to the next transformation step.
:::

```{r}
#| label: 05-tonsil-fit-polr-model
polr_obj <- polr(
    formula = size ~ Streptococcus.p,
    data = polr_dat,
    weights = weight,
    Hess = TRUE
)

summary(polr_obj)
```

The `polr()` standard errors agree with the cumulative `vglm()` fit.
:::

## Reproducibility

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true
packages <- c("conflicted", "dplyr", "Fahrmeir", "MASS", "tidyr", "VGAM")
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```

```{r}
#| label: 05-tonsil-session-info
sessionInfo()
```
