---
title: "Tonsil Size and Streptococcus"
subtitle: "Ordinal Regression Models Comparison"
date: today
---

## Overview

This script compares different **ordinal regression models** for analyzing the **tonsil dataset**, which examines the relationship between tonsil size (ordered categorical) and Streptococcus pyogenes carrier status.

## Data and Packages

```{r}
#| label: 05-tonsil-chunk-1
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(VGAM)
```

```{r}
#| label: 05-tonsil-chunk-2
?tonsil
```

```{r}
#| label: 05-tonsil-chunk-3
data(tonsil)
head(tonsil)
```

### Data Reshaping

Change the dataset into wide format for the vglm function. The commented-out code below shows an alternative approach using tidyverse:

```{r}
#| label: 05-tonsil-chunk-4
# Alternative tidyverse approach (commented out):
# tonsil_df <- dplyr::tbl_df(tonsil)
# tonsil_grouped <- dplyr::group_by(tonsil_df, Streptococcus.p)
# tonsil_summary <- dplyr::summarize(
#   tonsil_grouped,
#   present = n[which(Size == 1)],
#   enlarged = n[which(Size == 2)],
#   great_enlarged = n[which(Size == 3)]
# )
# tonsil <- as.data.frame(tonsil_summary)
```

Using reshape instead:

```{r}
#| label: 05-tonsil-chunk-5
tonsil <- reshape(
    tonsil,
    idvar = "Streptococcus.p",
    timevar = "Size",
    direction = "wide"
)

tonsil
```

Change the column names:

```{r}
#| label: 05-tonsil-chunk-6
colnames(tonsil) <- c(
    "Streptococcus.p",
    "present",
    "enlarged",
    "great_enlarged"
)
```

### Setting Contrasts

Change to effect coding for Streptococcus.p:

```{r}
#| label: 05-tonsil-chunk-7
options(contrasts = c("contr.sum", "contr.poly"))
contrasts(tonsil$Streptococcus.p)
```

## Model 1: Cumulative Link Model (Proportional Odds)

Fit with proportional odds assumption (parallel constraint):

```{r}
#| label: 05-tonsil-chunk-8
fit_cumulative <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = cumulative(parallel = TRUE ~ Streptococcus.p - 1),
    data = tonsil
)

summary(fit_cumulative)
```

These numbers check out with Table 3.6 in Example 3.7 of F & T.

## Model 2: Extended (Partial Proportional Odds)

Fit without the parallel constraint to allow category-specific effects:

```{r}
#| label: 05-tonsil-chunk-9
fit_cumulative_extended <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = cumulative(),
    data = tonsil
)

summary(fit_cumulative_extended)
```

## Model 3: Sequential (Continuation Ratio) Model

Fit sequential model with proportionality assumption:

```{r}
#| label: 05-tonsil-chunk-10
fit_sequential <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = sratio(parallel = TRUE ~ Streptococcus.p - 1),
    data = tonsil
)

summary(fit_sequential)
```

Note: The p-values reported in F&T appear to be incorrect.

## Alternative: MASS::polr() Function

Try using `polr()` from MASS package for comparison:

```{r}
#| label: 05-tonsil-chunk-11
response <- c(
    "present",
    "enlarged",
    "great_enlarged",
    "present",
    "enlarged",
    "great_enlarged"
)

response <- ordered(
    response,
    levels = c("present", "enlarged", "great_enlarged")
)

x <- c(
    "carriers",
    "carriers",
    "carriers",
    "noncarriers",
    "noncarriers",
    "noncarriers"
)

polr_obj <- MASS::polr(
    formula = response ~ x,
    weights = c(
        as.numeric(tonsil[1, 2:4]),
        as.numeric(tonsil[2, 2:4])
    )
)

summary(polr_obj)
```

The standard errors are the same as vglm. These numbers check out with Table 3.6 in Example 3.7 of F & T.
