---
title: "Tonsil Size and Streptococcus"
subtitle: "@fahrmeir_multivariate_2001 Example 3.7"
date: today
nocite: |
    @fahrmeir_multivariate_2001
---

## Overview

This script reproduces and extends Example 3.7 from @fahrmeir_multivariate_2001.
We analyse the relationship between tonsil size (an ordered response) and
*Streptococcus pyogenes* carrier status.

The response has three ordered categories:

1. `present`
2. `enlarged`
3. `great_enlarged`

We compare three ordinal regression specifications using `VGAM::vglm()`:

1. Cumulative link model with proportional odds
2. Cumulative link model without proportional odds
3. Sequential (continuation-ratio) model with proportionality

Finally, we fit `MASS::polr()` as a cross-check of the cumulative model.

## Data and Packages

```{r}
#| label: 05-tonsil-load-packages
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(MASS)
library(tidyr)
library(VGAM)
```

Run `?tonsil` for information about the data.

```{r}
#| label: 05-tonsil-load-and-preview-data
data(tonsil)
head(tonsil)
```

### Data Reshaping

The `tonsil` data are in long format: one row per carrier-status and size category.
For `vglm(cbind(...))`, we want one row per carrier-status and one column per size count.
We can do this directly with `pivot_wider()`.

```{r}
#| label: 05-tonsil-reshape-to-wide
tonsil_wide <- tonsil |>
    mutate(
        Size = factor(
            Size,
            levels = c(1, 2, 3),
            labels = c("present", "enlarged", "great_enlarged")
        )
    ) |>
    dplyr::select(Streptococcus.p, Size, n) |>
    pivot_wider(names_from = Size, values_from = n)

tonsil_wide
```

::: {.callout-tip collapse="true"}
## What each step is doing

- `mutate(...)`:
creates or updates columns; here it rewrites `Size` as a labelled factor.
- `factor(...)`: converts numeric codes (`1, 2, 3`)
into ordered category labels used in output.
- `dplyr::select(Streptococcus.p, Size, n)`:
keeps only the columns needed for reshaping.
- `pivot_wider(names_from = Size, values_from = n)`:
converts long data into wide data, so each size category becomes its own count column.
- `|>`:
passes the result of one step to the next, making the workflow readable top-to-bottom.
:::

### Understanding Factor Coding

As in the other scripts, we use effect coding (`contr.sum`) for factors.
This gives coefficients that are directly comparable to the textbook output.

```{r}
#| label: 05-tonsil-set-factor-contrasts
options(contrasts = c("contr.sum", "contr.poly"))
contrasts(tonsil_wide$Streptococcus.p)
```

## Model 1: Cumulative Link Model (Proportional Odds)

The proportional-odds assumption constrains the effect of
`Streptococcus.p` to be the same across cut-points.

```{r}
#| label: 05-tonsil-fit-cumulative-parallel
fit_cumulative <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = cumulative(parallel = TRUE ~ Streptococcus.p - 1),
    data = tonsil_wide
)

summary(fit_cumulative)
```

These estimates agree with Table 3.6 in Example 3.7.

## Model 2: Extended (Partial Proportional Odds)

Now we relax proportional odds and allow category-specific effects.

```{r}
#| label: 05-tonsil-fit-cumulative-nonparallel
fit_cumulative_extended <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = cumulative(),
    data = tonsil_wide
)

summary(fit_cumulative_extended)
```

## Model 3: Sequential (Continuation Ratio) Model

As an alternative ordinal specification, fit a continuation-ratio model
with proportionality imposed for `Streptococcus.p`.

```{r}
#| label: 05-tonsil-fit-sequential-parallel
fit_sequential <- vglm(
    cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
    family = sratio(parallel = TRUE ~ Streptococcus.p - 1),
    data = tonsil_wide
)

summary(fit_sequential)
```

The p-values reported in F&T for this section appear to contain an error.

## Quick Model Comparison

```{r}
#| label: 05-tonsil-compare-model-aic
tibble(
    model = c(
        "Cumulative (parallel)",
        "Cumulative (non-parallel)",
        "Sequential (parallel)"
    ),
    AIC = c(AIC(fit_cumulative), AIC(fit_cumulative_extended), AIC(fit_sequential))
)
```

## Alternative: MASS::polr() Function

We can fit the same proportional-odds model using `MASS::polr()`.
Rather than manually constructing response and predictor vectors,
we reshape once and use the count column as case weights.

```{r}
#| label: 05-tonsil-prepare-polr-data
polr_dat <- tonsil_wide |>
    pivot_longer(
        cols = c(present, enlarged, great_enlarged),
        names_to = "size",
        values_to = "weight"
    ) |>
    mutate(
        size = ordered(size, levels = c("present", "enlarged", "great_enlarged"))
    )

polr_dat
```

::: {.callout-tip collapse="true"}
## What each step is doing

- `pivot_longer(...)`: turns the wide count columns back into long format (one row per category).
- `cols = c(present, enlarged, great_enlarged)`: chooses which columns to stack into one variable.
- `names_to = "size"`: stores the former column names as category labels in a new `size` column.
- `values_to = "weight"`: stores the corresponding counts in a new `weight` column used as case weights.
- `ordered(...)`: makes `size` an ordered factor so `polr()` respects the category order.
- `|>`: pipes each intermediate result to the next transformation step.
:::

```{r}
#| label: 05-tonsil-fit-polr-model
polr_obj <- polr(
    formula = size ~ Streptococcus.p,
    data = polr_dat,
    weights = weight,
    Hess = TRUE
)

summary(polr_obj)
```

The `polr()` standard errors agree with the cumulative `vglm()` fit.

## Reproducibility

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true
packages <- c("conflicted", "dplyr", "Fahrmeir", "MASS", "tidyr", "VGAM")
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```

```{r}
#| label: 05-tonsil-session-info
sessionInfo()
```
