---
title: "Conditional Models: Happiness Data"
subtitle: "@fahrmeir_multivariate_2001 Example 3.12"
date: today
nocite: |
  @fahrmeir_multivariate_2001
editor:
  markdown:
    wrap: 72
---

## Overview

This script demonstrates **ordinal (cumulative link) regression** on the **happiness dataset**, analyzing **ordered categorical responses**.

## Data Loading

```{r}
#| label: 06-happiness-chunk-1
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(tidyr)
library(VGAM)
```

We load the `happy` dataset from the `Fahrmeir` package, which contains data on happiness levels.
Since it is a small dataset, we print it in full.
As usual, we will convert it to a tibble first.

```{r}
#| label: 06-happiness-chunk-2
data(happy)
happy_tbl <- as_tibble(happy)
happy_tbl
```

## Part 1: School Education Levels by Sex

Arrange the data into wide form for the first regression part in
Section 3.5.3 of @fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-3
happy_wide <- happy_tbl |>
    mutate(
        School = factor(
            School,
            levels = c("<12", "12", "13-16", ">16"),
            labels = c("school1", "school2", "school3", "school4")
        )
    ) |>
    summarize(n = sum(n), .by = c(Sex, School)) |>
    pivot_wider(
        names_from = School,
        values_from = n,
        values_fill = 0
    )

happy_wide
```

::: {.callout-tip collapse="true"}
## Understanding the tidying

- `mutate()` recodes factor levels so the column names match the model inputs.
- `summarize(..., .by = ...)` aggregates counts within each group.
- `pivot_wider()` spreads groups into columns and fills missing combinations with 0.
:::

### Setting Effect Coding

Effect coding is needed to reproduce the numbers in the book:

```{r}
#| label: 06-happiness-chunk-4a
options("contrasts")
```

`contr.treatment` is the default (corresponding to dummy coding),
but we need `contr.sum` for effect coding:

```{r}
#| label: 06-happiness-chunk-4b
contrasts(happy_wide$Sex)

options(contrasts = c("contr.sum", "contr.poly"))
contrasts(happy_wide$Sex)
```

We now have effect coding, however we need to switch the reference group to female,
in order to match the book.

```{r}
#| label: 06-happiness-chunk-4c
contrasts(happy_wide$Sex) <- as.matrix(c(-1, 1))
contrasts(happy_wide$Sex)
```

### Fit First Model

Now we can fit the model with `school1` as the reference level.

```{r}
#| label: 06-happiness-chunk-5
happy_logit1_vglm <- vglm(
    cbind(school1, school2, school3, school4) ~ Sex,
    data = happy_wide,
    family = cumulative()
)

summary(happy_logit1_vglm)
```

The coefficient values are the same as those reported in the first six
rows of Table 3.11 in @fahrmeir_multivariate_2001.

::: {.callout-warning}
The reported standard errors (also reproduced below)
and p-values presented here should be treated with caution.

**I'm not sure why these ones would be wrong.
Wouldn't it just be the standard errors and p-values for the second part of the table that are wrong? Dennis to clarify**

```{r}
#| label: 06-happiness-chunk-7
happy_logit1_vglm |>
    vcov() |>
    diag() |>
    sqrt()
```
:::

We can access the linear predictors

```{r}
happy_logit1_vglm@predictors
```

## Part 2: Happiness Levels by Education

Here we arrange the data into wide form for the second regression.
Arrange the data into wide form for the second regression part in
Section 3.5.3 of @fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-8
happy2 <- happy_tbl |>
    mutate(
        School = factor(School, levels = c("<12", "12", "13-16", ">16")),
        Rep.happiness = recode(
            Rep.happiness,
            "Not to happy" = "happy_low",
            "Pretty happy" = "happy_medium",
            "Very happy" = "happy_high"
        )
    ) |>
    summarize(n = sum(n), .by = c(Sex, School, Rep.happiness)) |>
    pivot_wider(
        names_from = Rep.happiness,
        values_from = n,
        values_fill = 0
    )

happy2
```

### Setting Contrasts for Part 2

We check the coding scheme for `School`:

```{r}
#| label: 06-happiness-chunk-9a
contrasts(happy2$School)
```

Here we actually want to switch from effect to dummy coding with `<12` as reference level:

```{r}
#| label: 06-happiness-chunk-9b
options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(happy2$School)
```

`<12` is the reference level.

### Fit Second Model

Regression for the second line of display (3.5.3) in
@fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-10
happy_logit2_vglm <- vglm(
    cbind(happy_low, happy_medium, happy_high) ~ School,
    data = happy2,
    family = cumulative(link = "logitlink")
)

summary(happy_logit2_vglm)
```

When we compare the coefficients here to those in Table 3.11,
only the two intercept estimates (corresponding to reference level
`<12` of School) match the numbers in the book.
This is because under the dummy coding scheme,
we are picking less than 12 years of school as our reference level.
Therefore all the school intercepts are to be compared relative to less than 12 years of school.
In @fahrmeir_multivariate_2001, the numbers are the absolute levels.

### Recovering Coefficients

To recover the original estimates, we add the intercepts to each
pair of coefficients for the other schooling levels:

```{r}
#| label: 06-happiness-chunk-11
happy_logit2_vglm_coeff <- coef(happy_logit2_vglm)
happy_logit2_vglm_coeff

rep(happy_logit2_vglm_coeff[1:2], 3) + happy_logit2_vglm_coeff[3:8]
```

These recover the values reported in the last six rows of Table 3.11 in
@fahrmeir_multivariate_2001.

## Reproducing the Deviance

We now attempt to reproduce the devaince.

To retrieve the likelihood of the full GLM model,
we sum the log likelihood resulting from the first and second parts of
the simple regressive cumulative model.

:::{callout-note}
Make sure you know why this is the case.
:::

```{r}
#| label: 06-happiness-chunk-12
log_lik_glm <- logLik(happy_logit1_vglm) + logLik(happy_logit2_vglm)
log_lik_glm
```

Alternatively we can extract the log-likelihood of each of individual groups separately:

```{r}
#| label: 06-happiness-chunk-13
logLik(happy_logit1_vglm, sum = FALSE)
logLik(happy_logit2_vglm, sum = FALSE)
```

### Computing Saturated Model Likelihood

Now we just need to compute the log-likelihood of the saturated model.
We split the dataset into a male group and female group.
Within each group we have 12 observations
corresponding to the 12 combinations of schooling and happiness levels
(meaning we have 11 degrees of freedom for the male and female groups separately).
The maximum likelihood estimators for each group
are exactly the observed proportions of each happiness level within each group.

Therefore we can calculate the log-likelihood of the saturated model by summing the log-likelihoods as follows:

```{r}
#| label: 06-happiness-chunk-14
happy_male <- dplyr::filter(happy, Sex == "Males")
happy_female <- dplyr::filter(happy, Sex == "Females")

log_lik_saturated <- dmultinom(
    happy_male$n,
    prob = happy_male$n / sum(happy_male$n),
    log = TRUE
) +
    dmultinom(
        happy_female$n,
        prob = happy_female$n / sum(happy_female$n),
        log = TRUE
    )

log_lik_saturated
```

### Deviance

Finally, we can compute the deviance as twice the difference between the log-likelihood of the saturated model and the log-likelihood of the fitted model:

```{r}
#| label: 06-happiness-chunk-15
2 * (log_lik_saturated - log_lik_glm)
```

## Problems with the Reported Standard Errors and $p$-values

**This bit was slightly unclear to me, Dennis to clarify if I have misunderstood.**

The second part of the the table are treating the model as
by treating the years at school as fixed, and fitting a GLM to the data.
The standard errors for are hence derived from the Fisher information matrix
under the assumption that years at school are fixed.
If we really want to understand the model as a whole,
we need to treat both $Y_1$ and $Y_2$ as random.
In practice this is probably very difficult to do,
as we need to tease out the explicit form for the main function and so on.

## Package Citations and References

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c(
    "conflicted",
    "dplyr",
    "Fahrmeir",
    "tidyr",
    "VGAM"
)
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("  - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```
