---
title: "Ordinal Regression: Happiness Data"
subtitle: "Cumulative Link Models for Ordered Outcomes"
date: today
---

## Overview

This script demonstrates **ordinal (cumulative link) regression** on the **happiness dataset**, analyzing **ordered categorical responses**.

Note: To reproduce the first regression in section 3.5.3 of F&T, we cannot use `polr()` since it doesn't allow for category-specific beta coefficients.

## Data Loading

```{r}
library(conflicted)
library(Fahrmeir)
library(VGAM)
library(dplyr)
```

```{r}
data(happy)
happy
?happy
```

## Part 1: School Education Levels by Sex

Arrange the data into wide form for the first regression part in (3.5.3):

```{r}
happy_tibble <- dplyr::tbl_df(happy)

happy_grouped1 <- dplyr::group_by(happy_tibble, Sex)

happy_summary1 <- dplyr::summarize(
    happy_grouped1,
    school1 = as.numeric(sum(n[which(School == "<12")])),
    school2 = as.numeric(sum(n[which(School == "12")])),
    school3 = as.numeric(sum(n[which(School == "13-16")])),
    school4 = as.numeric(sum(n[which(School == ">16")]))
)

happy1 <- as.data.frame(happy_summary1)
happy1
```

### Setting Contrasts for Part 1

Effect coding is needed to reproduce the numbers in the book:

```{r}
options("contrasts")
contrasts(happy1$Sex)

options(contrasts = c("contr.sum", "contr.poly"))
contrasts(happy1$Sex)

contrasts(happy1$Sex) <- as.matrix(c(-1, 1))
```

Adjust the contrast matrix so female serves as reference.

### Fit First Model

Regression for the first line of display (3.5.3) in F&T:

```{r}
happy_logit1_vglm <- vglm(
    cbind(school1, school2, school3, school4) ~ Sex,
    data = happy1,
    family = cumulative()
)

summary(happy_logit1_vglm)
```

The coefficient values are the same as those reported in the first six rows of Table 3.11 in F&T.

```{r}
attributes(happy_logit1_vglm)$predictors
```

We can also look at the predictors.

```{r}
sqrt(diag(vcov(happy_logit1_vglm)))
```

Same as the std. error reported by `summary(happy_logit1_vglm)`.

## Part 2: Happiness Levels by Education

Arrange the data into wide form for the second regression part in (3.5.3):

```{r}
happy_grouped2 <- dplyr::group_by(happy_tibble, Sex, School)

happy_summary2 <- dplyr::summarize(
    happy_grouped2,
    happy_low = as.numeric(sum(n[which(Rep.happiness == "Not to happy")])),
    happy_medium = as.numeric(sum(n[which(Rep.happiness == "Pretty happy")])),
    happy_high = as.numeric(sum(n[which(Rep.happiness == "Very happy")]))
)

happy2 <- as.data.frame(happy_summary2)
happy2
```

### Setting Contrasts for Part 2

Switch to dummy coding with `<12` as reference level:

```{r}
contrasts(happy2$School)

options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(happy2$School)
```

`<12` is the reference level.

### Fit Second Model

Regression for the second line of display (3.5.3) in F&T:

```{r}
happy_logit2_vglm <- vglm(
    cbind(happy_low, happy_medium, happy_high) ~ School,
    data = happy2,
    family = cumulative(link = "logitlink")
)

summary(happy_logit2_vglm)
```

`cumulative()` must be regressed with an intercept.

### Recovering Coefficients

Only the two intercept estimates (corresponding to reference level `<12` of School) match the numbers in the book. To recover the original estimates in F&T, we need to add the intercepts to each pair of coefficients for each other Schooling levels:

```{r}
happy_logit2_vglm_coeff <- attributes(happy_logit2_vglm)$coeff
happy_logit2_vglm_coeff

rep(happy_logit2_vglm_coeff[1:2], 3) + happy_logit2_vglm_coeff[3:8]
```

The numbers in the books are now recovered, as reported in the last 6 rows of F&T Table 3.11.

## Log-Likelihood and Model Fit

Add the log likelihood resulting from the first and second parts of (3.5.3), to give the likelihood of the full GLM model:

```{r}
log_lik_glm <- logLik(happy_logit1_vglm) + logLik(happy_logit2_vglm)
log_lik_glm
```

Alternatively:

```{r}
logLik(happy_logit1_vglm, sum = FALSE)
logLik(happy_logit2_vglm, sum = FALSE)
```

These get the log likelihood of individual groups.

### Computing Saturated Model Likelihood

```{r}
happy_male <- happy[which(happy$Sex == "Males"), ]
happy_female <- happy[which(happy$Sex == "Females"), ]

logLik_saturated <- dmultinom(
    happy_male$n,
    prob = happy_male$n / sum(happy_male$n),
    log = TRUE
) +
    dmultinom(
        happy_female$n,
        prob = happy_female$n / sum(happy_female$n),
        log = TRUE
    )

logLik_saturated
```

### Deviance

This is the deviance 13.27, as reported in the book:

```{r}
2 * (logLik_saturated - log_lik_glm)
```
