---
title: "Ordinal Regression: Happiness Data"
subtitle: "@fahrmeir_multivariate_2001 Section 3.5.3"
date: today
nocite: |
  @fahrmeir_multivariate_2001
editor:
  markdown:
    wrap: 72
---

## Overview

This script demonstrates **ordinal (cumulative link) regression** on the **happiness dataset**, analyzing **ordered categorical responses**.

## Data Loading

```{r}
#| label: 06-happiness-chunk-1
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(tidyr)
library(VGAM)
```

We load the `happy` dataset from the `Fahrmeir` package, which contains data on happiness levels.
Since it is a small dataset, we print it in full.
As usual, we will convert it to a tibble first.

```{r}
#| label: 06-happiness-chunk-2
data(happy)
happy_tbl <- as_tibble(happy)
happy_tbl
```

## Part 1: School Education Levels by Sex

Arrange the data into wide form for the first regression part in
Section 3.5.3 of @fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-3
happy_wide <- happy_tbl |>
    mutate(
        School = factor(
            School,
            levels = c("<12", "12", "13-16", ">16"),
            labels = c("school1", "school2", "school3", "school4")
        )
    ) |>
    summarize(n = sum(n), .by = c(Sex, School)) |>
    pivot_wider(
        names_from = School,
        values_from = n,
        values_fill = 0
    )

happy_wide
```

::: {.callout-tip collapse="true"}
## Understanding the tidying

- `mutate()` recodes factor levels so the column names match the model inputs.
- `summarize(..., .by = ...)` aggregates counts within each group.
- `pivot_wider()` spreads groups into columns and fills missing combinations with 0.
:::

### Setting Effect Coding

Effect coding is needed to reproduce the numbers in the book:

```{r}
#| label: 06-happiness-chunk-4a
options("contrasts")
```

`contr.treatment` is the default (corresponding to dummy coding),
but we need `contr.sum` for effect coding:

```{r}
#| label: 06-happiness-chunk-4b
contrasts(happy_wide$Sex)

options(contrasts = c("contr.sum", "contr.poly"))
contrasts(happy_wide$Sex)
```

We now have effect coding, however we need to switch the reference group to female,
in order to match the book.

```{r}
#| label: 06-happiness-chunk-4c
contrasts(happy_wide$Sex) <- as.matrix(c(-1, 1))
contrasts(happy_wide$Sex)
```

### Fit First Model

Now we can fit the model with `school1` as the reference level.

```{r}
#| label: 06-happiness-chunk-5
happy_logit1_vglm <- vglm(
    cbind(school1, school2, school3, school4) ~ Sex,
    data = happy_wide,
    family = cumulative()
)

summary(happy_logit1_vglm)
```

The coefficient values are the same as those reported in the first six
rows of Table 3.11 in @fahrmeir_multivariate_2001.

::: {.callout-warning}
The reported standard errors (also reproduced below)
and p-values presented here should be treated with caution.

```{r}
#| label: 06-happiness-chunk-7
happy_logit1_vglm |>
    vcov() |>
    diag() |>
    sqrt()
```
:::

We can access the linear predictors

```{r}
happy_logit1_vglm@predictors
```

## Part 2: Happiness Levels by Education

Here we arrange the data into wide form for the second regression.
Arrange the data into wide form for the second regression part in
Section 3.5.3 of @fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-8
happy2 <- happy_tbl |>
    mutate(
        School = factor(School, levels = c("<12", "12", "13-16", ">16")),
        Rep.happiness = recode(
            Rep.happiness,
            "Not to happy" = "happy_low",
            "Pretty happy" = "happy_medium",
            "Very happy" = "happy_high"
        )
    ) |>
    summarize(n = sum(n), .by = c(Sex, School, Rep.happiness)) |>
    pivot_wider(
        names_from = Rep.happiness,
        values_from = n,
        values_fill = 0
    )

happy2
```

### Setting Contrasts for Part 2

We check the coding scheme for `School`:

```{r}
#| label: 06-happiness-chunk-9a
contrasts(happy2$School)
```

Here we actually want to switch from effect to dummy coding with `<12` as reference level:

```{r}
#| label: 06-happiness-chunk-9b
options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(happy2$School)
```

`<12` is the reference level.

### Fit Second Model

Regression for the second line of display (3.5.3) in
@fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-10
happy_logit2_vglm <- vglm(
    cbind(happy_low, happy_medium, happy_high) ~ School,
    data = happy2,
    family = cumulative(link = "logitlink")
)

summary(happy_logit2_vglm)
```

`cumulative()` must be regressed with an intercept.

### Recovering Coefficients

Only the two intercept estimates (corresponding to reference level
`<12` of School) match the numbers in the book. To recover the original
estimates in @fahrmeir_multivariate_2001, add the intercepts to each
pair of coefficients for the other schooling levels:

```{r}
#| label: 06-happiness-chunk-11
happy_logit2_vglm_coeff <- coef(happy_logit2_vglm)
happy_logit2_vglm_coeff

rep(happy_logit2_vglm_coeff[1:2], 3) + happy_logit2_vglm_coeff[3:8]
```

These recover the values reported in the last six rows of Table 3.11 in
@fahrmeir_multivariate_2001.

## Log-Likelihood and Model Fit

Add the log likelihood resulting from the first and second parts of (3.5.3), to give the likelihood of the full GLM model:

```{r}
#| label: 06-happiness-chunk-12
log_lik_glm <- logLik(happy_logit1_vglm) + logLik(happy_logit2_vglm)
log_lik_glm
```

Alternatively:

```{r}
#| label: 06-happiness-chunk-13
logLik(happy_logit1_vglm, sum = FALSE)
logLik(happy_logit2_vglm, sum = FALSE)
```

These get the log likelihood of individual groups.

### Computing Saturated Model Likelihood

```{r}
#| label: 06-happiness-chunk-14
happy_male <- dplyr::filter(happy, Sex == "Males")
happy_female <- dplyr::filter(happy, Sex == "Females")

log_lik_saturated <- dmultinom(
    happy_male$n,
    prob = happy_male$n / sum(happy_male$n),
    log = TRUE
) +
    dmultinom(
        happy_female$n,
        prob = happy_female$n / sum(happy_female$n),
        log = TRUE
    )

log_lik_saturated
```

### Deviance

This gives deviance $13.27$, as reported in the book:

```{r}
#| label: 06-happiness-chunk-15
2 * (log_lik_saturated - log_lik_glm)
```


## Package Citations and References

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c(
    "conflicted",
    "dplyr",
    "Fahrmeir",
    "tidyr",
    "VGAM"
)
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("  - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```
