---
title: "Ordinal Regression: Happiness Data"
subtitle: "@fahrmeir_multivariate_2001 Section 3.5.3"
date: today
nocite: |
  @fahrmeir_multivariate_2001
editor:
  markdown:
    wrap: 72
---

## Overview

This script demonstrates **ordinal (cumulative link) regression** on the **happiness dataset**, analyzing **ordered categorical responses**.

## Data Loading

```{r}
#| label: 06-happiness-chunk-1
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(tidyr)
library(VGAM)
```

We load the `happy` dataset from the `Fahrmeir` package, which contains data on happiness levels.
Since it is a small dataset, we print it in full.

```{r}
#| label: 06-happiness-chunk-2
data(happy)
happy
```

## Part 1: School Education Levels by Sex

Arrange the data into wide form for the first regression part in
Section 3.5.3 of @fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-3
happy_wide <- happy |>
    as_tibble() |>
    mutate(
        School = factor(
            School,
            levels = c("<12", "12", "13-16", ">16"),
            labels = c("school1", "school2", "school3", "school4")
        )
    ) |>
    summarize(n = sum(n), .by = c(Sex, School)) |>
    pivot_wider(
        names_from = School,
        values_from = n,
        values_fill = 0
    )

happy_wide
```

::: {.callout-tip collapse="true"}
## Understanding the tidying

- `as_tibble()` makes the data print-friendly and consistent with tidyverse tools.
- `mutate()` recodes factor levels so the column names match the model inputs.
- `summarize(..., .by = ...)` aggregates counts within each group.
- `pivot_wider()` spreads groups into columns and fills missing combinations with 0.
:::

### Setting Contrasts for Part 1

Effect coding is needed to reproduce the numbers in the book:

```{r}
#| label: 06-happiness-chunk-4
options("contrasts")
contrasts(happy_wide$Sex)

options(contrasts = c("contr.sum", "contr.poly"))
contrasts(happy_wide$Sex)

contrasts(happy_wide$Sex) <- as.matrix(c(-1, 1))
```

Adjust the contrast matrix so females serve as the reference group.

### Fit First Model

Regression for the first line of display (3.5.3) in
@fahrmeir_multivariate_2001:
To reproduce the first regression in Section 3.5.3 of
@fahrmeir_multivariate_2001, we cannot use `polr()` since it does not
allow for category-specific beta coefficients.

```{r}
#| label: 06-happiness-chunk-5
happy_logit1_vglm <- vglm(
    cbind(school1, school2, school3, school4) ~ Sex,
    data = happy_wide,
    family = cumulative()
)

summary(happy_logit1_vglm)
```

The coefficient values are the same as those reported in the first six
rows of Table 3.11 in @fahrmeir_multivariate_2001.

```{r}
#| label: 06-happiness-chunk-6
model.matrix(happy_logit1_vglm)
```

We can also inspect the design matrix.

```{r}
#| label: 06-happiness-chunk-7
sqrt(diag(vcov(happy_logit1_vglm)))
```

These match the standard errors reported by `summary(happy_logit1_vglm)`.

## Part 2: Happiness Levels by Education

Arrange the data into wide form for the second regression part in
Section 3.5.3 of @fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-8
happy2 <- happy |>
    as_tibble() |>
    mutate(
        School = factor(School, levels = c("<12", "12", "13-16", ">16")),
        Rep.happiness = recode(
            Rep.happiness,
            "Not too happy" = "happy_low",
            "Not to happy" = "happy_low",
            "Pretty happy" = "happy_medium",
            "Very happy" = "happy_high"
        )
    ) |>
    summarize(n = sum(n), .by = c(Sex, School, Rep.happiness)) |>
    tidyr::pivot_wider(
        names_from = Rep.happiness,
        values_from = n,
        values_fill = 0
    )

happy2
```

### Setting Contrasts for Part 2

Switch to dummy coding with `<12` as reference level:

```{r}
#| label: 06-happiness-chunk-9
contrasts(happy2$School)

options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(happy2$School)
```

`<12` is the reference level.

### Fit Second Model

Regression for the second line of display (3.5.3) in
@fahrmeir_multivariate_2001:

```{r}
#| label: 06-happiness-chunk-10
happy_logit2_vglm <- vglm(
    cbind(happy_low, happy_medium, happy_high) ~ School,
    data = happy2,
    family = cumulative(link = "logitlink")
)

summary(happy_logit2_vglm)
```

`cumulative()` must be regressed with an intercept.

### Recovering Coefficients

Only the two intercept estimates (corresponding to reference level
`<12` of School) match the numbers in the book. To recover the original
estimates in @fahrmeir_multivariate_2001, add the intercepts to each
pair of coefficients for the other schooling levels:

```{r}
#| label: 06-happiness-chunk-11
happy_logit2_vglm_coeff <- coef(happy_logit2_vglm)
happy_logit2_vglm_coeff

rep(happy_logit2_vglm_coeff[1:2], 3) + happy_logit2_vglm_coeff[3:8]
```

These recover the values reported in the last six rows of Table 3.11 in
@fahrmeir_multivariate_2001.

## Log-Likelihood and Model Fit

Add the log likelihood resulting from the first and second parts of (3.5.3), to give the likelihood of the full GLM model:

```{r}
#| label: 06-happiness-chunk-12
log_lik_glm <- logLik(happy_logit1_vglm) + logLik(happy_logit2_vglm)
log_lik_glm
```

Alternatively:

```{r}
#| label: 06-happiness-chunk-13
logLik(happy_logit1_vglm, sum = FALSE)
logLik(happy_logit2_vglm, sum = FALSE)
```

These get the log likelihood of individual groups.

### Computing Saturated Model Likelihood

```{r}
#| label: 06-happiness-chunk-14
happy_male <- dplyr::filter(happy, Sex == "Males")
happy_female <- dplyr::filter(happy, Sex == "Females")

log_lik_saturated <- dmultinom(
    happy_male$n,
    prob = happy_male$n / sum(happy_male$n),
    log = TRUE
) +
    dmultinom(
        happy_female$n,
        prob = happy_female$n / sum(happy_female$n),
        log = TRUE
    )

log_lik_saturated
```

### Deviance

This gives deviance $13.27$, as reported in the book:

```{r}
#| label: 06-happiness-chunk-15
2 * (log_lik_saturated - log_lik_glm)
```


## Package Citations and References

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c(
    "conflicted",
    "dplyr",
    "Fahrmeir",
    "tidyr",
    "VGAM"
)
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("  - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```
