---
title: "Conditional Models: Happiness Data"
subtitle: "@fahrmeir_multivariate_2001 Example 3.12"
date: today
nocite: |
  @fahrmeir_multivariate_2001
---

## Overview

We will fit a simple asymmetric model to a dataset on happiness levels from Table 3.10 of p.115 in @fahrmeir_multivariate_2001.
We want to determine the association between gender, level of education, and happiness levels.
We will do this in two parts, first by setting gender as a predictor for years of schooling, and then by setting years of schooling as a predictor for happiness levels.

The model is specified more formally in the equation (3.5.3) in @fahrmeir_multivariate_2001.

## Data Loading

As always, we start by loading the necessary packages (and installing any we don't already have).

```{r}
#| label: 06-happiness-chunk-1
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(tidyr)
library(VGAM)
```

We load the `happy` dataset from the `Fahrmeir` package, which contains data on happiness levels.
As usual, we will convert it to a tibble first.

```{r}
#| label: 06-happiness-chunk-2
data(happy)
happy_tbl <- as_tibble(happy)
happy_tbl
```

## Part 1: Marginal Distribution of School Education Levels by Sex

We want to fit a cumulative logit model to the marginal distribution of schooling, predicted by gender.
To do this, we first need to arrange the data into wide form.

```{r}
#| label: 06-happiness-chunk-3
happy_wide <- happy_tbl |>
  mutate(
    School = factor(
      School,
      levels = c("<12", "12", "13-16", ">16"),
      labels = c("school1", "school2", "school3", "school4")
    )
  ) |>
  summarize(n = sum(n), .by = c(Sex, School)) |>
  pivot_wider(
    names_from = School,
    values_from = n,
    values_fill = 0
  )

happy_wide
```

::: {.callout-tip collapse="true"}
## Understanding the tidying

-   `mutate()` recodes factor levels so the column names match the model inputs.
-   `summarize(..., .by = ...)` aggregates counts within each group.
-   `pivot_wider()` spreads groups into columns and fills missing combinations with 0 (if there is any).
:::

### Setting Effect Coding

Effect coding is needed to reproduce the numbers in the book:

```{r}
#| label: 06-happiness-chunk-4a
options("contrasts")
```

`contr.treatment` is the default (corresponding to dummy coding), but we need `contr.sum` for effect coding:

```{r}
#| label: 06-happiness-chunk-4b
contrasts(happy_wide$Sex)

options(contrasts = c("contr.sum", "contr.poly"))
contrasts(happy_wide$Sex)
```

We now have effect coding, however we need to switch the reference group to female, in order to match the book.

```{r}
#| label: 06-happiness-chunk-4c
contrasts(happy_wide$Sex) <- as.matrix(c(-1, 1))
contrasts(happy_wide$Sex)
```

### Fit First Model

Now we can fit the model with `school1` as the reference level.

```{r}
#| label: 06-happiness-chunk-5
happy_logit1_vglm <- vglm(
  cbind(school1, school2, school3, school4) ~ Sex,
  data = happy_wide,
  family = cumulative()
)

summary(happy_logit1_vglm)
```

The coefficient values are the same as those reported in the first six rows of Table 3.11 on p.117 of  @fahrmeir_multivariate_2001.

::: callout-warning

We can reproduce the standard errors (and hence the $p$-values) reported in  the first six rows of Table 3.11 by using the code below:

```{r}
#| label: 06-happiness-chunk-7
happy_logit1_vglm |>
  vcov() |>
  diag() |>
  sqrt()
```

However, we probably should  treat these standard errors and $p$-values in Table 3.11 with a grain of salt. See the end for more explanation.

:::

We can access the linear predictors

```{r}
happy_logit1_vglm@predictors
```

## Part 2: Happiness Levels by Education

Now we are interested in predicting happiness levels by years of schooling.
We again arrange our data into a wide form with the happiness levels as columns.

```{r}
#| label: 06-happiness-chunk-8
happy2 <- happy_tbl |>
  mutate(
    School = factor(School, levels = c("<12", "12", "13-16", ">16")),
    Rep.happiness = recode(
      Rep.happiness,
      "Not to happy" = "happy_low",
      "Pretty happy" = "happy_medium",
      "Very happy" = "happy_high"
    )
  ) |>
  summarize(n = sum(n), .by = c(Sex, School, Rep.happiness)) |>
  pivot_wider(
    names_from = Rep.happiness,
    values_from = n,
    values_fill = 0
  )

happy2
```

### Setting Contrasts for Part 2

We check the coding scheme for `School`:

```{r}
#| label: 06-happiness-chunk-9a
contrasts(happy2$School)
```

Here we actually want to switch from effect to dummy coding with `<12` as reference level:

```{r}
#| label: 06-happiness-chunk-9b
options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(happy2$School)
```

`<12` is the reference level.

### Fit Second Model

We now fit the model.

::: callout-note
This regression ignores the `Sex` column (which is what we want).
:::

```{r}
#| label: 06-happiness-chunk-10
happy_logit2_vglm <- vglm(
  cbind(happy_low, happy_medium, happy_high) ~ School,
  data = happy2,
  family = cumulative
)

summary(happy_logit2_vglm)
```

When we compare the coefficients here to those in Table 3.11, only the two intercept estimates (corresponding to reference level `<12` of School) match the numbers in the book.
This is because under the dummy coding scheme, we are picking less than 12 years of school as our reference level.
In contrast, the book treats each school intercept as it's own absolute level (see second line of (3.5.3) in @fahrmeir_multivariate_2001).

### Recovering Coefficients

Therefore to recover the original estimates, we add the intercepts to each pair of coefficients for the other schooling levels:

```{r}
#| label: 06-happiness-chunk-11
happy_logit2_vglm_coeff <- coef(happy_logit2_vglm)
happy_logit2_vglm_coeff

rep(happy_logit2_vglm_coeff[1:2], 3) + happy_logit2_vglm_coeff[3:8]
```

These recover the values reported in the last six rows of Table 3.11 in @fahrmeir_multivariate_2001 for the other three schooling levels, i.e. 12 years, 13-16 years, and $\geq$ 17 years.

## Reproducing the Deviance

We now attempt to reproduce the deviance.

To retrieve the likelihood of the full GLM model, we sum the log likelihoods resulting from the two cumulative models in the first and second parts of display $(3.5.3)$ in @fahrmeir_multivariate_2001.

::: callout-note
Make sure you know why the log likelihood of the full GLM can be computed by simply summing the log likelihoods of the two components.
:::

```{r}
#| label: 06-happiness-chunk-12
log_lik_glm <- logLik(happy_logit1_vglm) + logLik(happy_logit2_vglm)
log_lik_glm
```

Alternatively we can extract the log-likelihood of each of the individual groups separately:

```{r}
#| label: 06-happiness-chunk-13
logLik(happy_logit1_vglm, sum = FALSE)
logLik(happy_logit2_vglm, sum = FALSE)
```

### Computing Saturated Model Likelihood

Now we just need to compute the log-likelihood of the saturated model.
We split the dataset into a male group and female group.
Within each group we have 12 observations corresponding to the 12 combinations of schooling and happiness levels (meaning we have 11 degrees of freedom for the male and female groups separately).
The maximum likelihood estimators for each group are exactly the observed proportions for the different combinations of schooling and happiness levels within that group.

Therefore we can calculate the log-likelihood of the saturated model by summing the log-likelihoods as follows:

```{r}
#| label: 06-happiness-chunk-14
happy_male <- dplyr::filter(happy, Sex == "Males")
happy_female <- dplyr::filter(happy, Sex == "Females")

happy_male
happy_female

log_lik_saturated <- dmultinom(
  happy_male$n,
  prob = happy_male$n / sum(happy_male$n),
  log = TRUE
) +
  dmultinom(
    happy_female$n,
    prob = happy_female$n / sum(happy_female$n),
    log = TRUE
  )

log_lik_saturated
```

### Deviance

Finally, we can compute the deviance as twice the difference between the log-likelihood of the saturated model and the log-likelihood of the fitted model:

```{r}
#| label: 06-happiness-chunk-15
2 * (log_lik_saturated - log_lik_glm)
```

This is same as the deviance reported on p.116 in @fahrmeir_multivariate_2001.

## Potential problems with the reported Standard Errors and $p$-values in Table 3.11 of @fahrmeir_multivariate_2001

The model prescribed by display $(3.5.3)$ on p.116 of @fahrmeir_multivariate_2001 is by nature one single GLM with $(Y_1, Y_2)$ (years of school and happiness levels) as response (precisely, a multinomial response with 1 trial and 12 categories) and $x$ (sex) as the covariate. To properly get the standard errors for all the 14 parameter estimates in Table 3.11, strictly speaking, one should first get an estimate of  the $14 \times 14$ Fisher information of this GLM (e.g. use the observed information matrix), and take the 14 diagonal entries of the inverse of it as the estimated variance for these  parameter estimates. 

Apparently, the standard errors presented in Table 3.11 are not obtained this way: From  `summary(happy_logit1_vglm)` (which reproduces all the standard deviations for $\theta_1, \theta_2, \theta_3, \beta_1^{(1)}, \beta_2^{(1)}, \beta_3^{(1)}$ in Table 3.11) and  `summary(happy_logit2_vglm)`(which also reproduces the same standard deviations $0.109$ and $0.092$ for $\theta_{11}$ and $\theta_{12}$ in Table 3.11), we can see that these standard errors are only computed based on treating the two conditional models in $(3.5.3)$ as two separate GLMs. These may not  be the same standard errors we could derive by using the (proper) "one single GLM" approach described in the prior paragraph; I leave it to the interested student to check if the two approaches actually coincide.

## Package Citations and References

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c(
  "conflicted",
  "dplyr",
  "Fahrmeir",
  "tidyr",
  "VGAM"
)
for (pkg in packages) {
  cit <- citation(pkg)

  # Check if multiple citations
  if (length(cit) > 1) {
    cat("- **", pkg, "**:\n", sep = "")
    for (i in seq_along(cit)) {
      cit_text <- format(cit[i], style = "text")
      cat("  - ", cit_text, "\n\n", sep = "")
    }
  } else {
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
  }
}
```
