---
title: "Ohio Respiratory Data: GEE Analysis"
subtitle: "Generalized Estimating Equations with Repeated Measures"
date: today
---

## Overview

This script aims to reproduce the results in Example 6.4 of Fahrmeir & Tutz using the Ohio respiratory dataset.

## Packages and Setup

```{r}
library(conflicted)
library(gee)
library(geepack)
```

We aim to reproduce the results in Example 6.4 of F & T.

```{r}
?ohio
```

```{r}
head(ohio, 20)
```

## Understanding the Data

```{r}
str(ohio)
```

First set the right classes for the variables in the dataset:

```{r}
class(ohio$resp)
class(ohio$age)
class(ohio$smoke)

ohio$age <- as.factor(ohio$age)
ohio$smoke <- as.factor(ohio$smoke)
```

Age coding:
- age = -2 is 7 years old
- age = -1 is 8 years old
- age = 0 is 9 years old
- age = 1 is 10 years old

## Contrast Coding

Check default contrasts:

```{r}
contrasts(ohio$age)
contrasts(ohio$smoke)
```

F and T use effect coding in the example so we need to set the contrasts:

```{r}
options(contrasts = c("contr.sum", "contr.poly"))

contrasts(ohio$age)
contrasts(ohio$smoke)
```

Contrast for smoke status needs to be further adjusted:

```{r}
contrasts(ohio$smoke) <- matrix(c(-1, 1), 2, 1)
contrasts(ohio$smoke)
```

## GEE Models

### Independence Structure

```{r}
geefit_indep <- gee(
    formula = resp ~ age * smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "independence"
)

summary(geefit_indep)$coeff
```

### Exchangeable Structure

```{r}
geefit_exch <- gee(
    formula = resp ~ age * smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "exchangeable"
)

summary(geefit_exch)$coeff
```

### Unstructured Correlation

```{r}
geefit_unstruct <- gee(
    formula = resp ~ age * smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "unstructured"
)

summary(geefit_unstruct)$coeff
```

Same numbers as Table 6.8 in F & T:

```{r}
summary(geefit_indep)$coeff
summary(geefit_exch)$coeff
summary(geefit_unstruct)$coeff
```

## Robust Variance Matrices

```{r}
geefit_indep$robust.variance
geefit_exch$robust.variance
geefit_unstruct$robust.variance
```

The three robust variances happen to look exactly the same here, probably just a numerical coincidence. However, if we take a difference, R says they differ by very small numbers:

```{r}
geefit_indep$robust.variance - geefit_exch$robust.variance
```

## Scale Parameter

```{r}
summary(geefit_indep)$scale
```

Now we tease out the Pearson residuals and compute the method of moment dispersion estimator by hand:

```{r}
pearson_resid <- (ohio$resp - geefit_indep$fitted) /
    sqrt(
        geefit_indep$fitted.values * (1 - geefit_indep$fitted.values)
    )

scale_by_hand <- sum(pearson_resid^2) /
    (geefit_indep$nobs - length(geefit_indep$coeff))

scale_by_hand
```

Same number as the one reported in summary.

### Verifying Naive vs Robust Standard Errors

We first regress by fixing the scale to be 1:

```{r}
geefit_indep_scale_fix <- gee(
    formula = resp ~ age * smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "independence",
    scale.fix = TRUE
)
```

Verify relationship: Naive SE × sqrt(scale) = Model SE:

```{r}
summary(geefit_indep_scale_fix)$coeff[, "Naive S.E."] *
    sqrt(summary(geefit_indep)$scale)
```

Is same as:

```{r}
summary(geefit_indep)$coeff[, "Naive S.E."]
```

So it only differs by the dispersion parameter, which makes sense.

## Beta Estimate for Age = 10

With effect coding, age=10 is minus the sum of the other three age coefficients:

```{r}
beta_age_10 <- -sum(summary(geefit_indep)$coeff[
    c("age1", "age2", "age3"),
    1
])

beta_age_10
```

Same as the book.

### Standard Error for Combined Effect

Robust SE for combined age=10 coefficient (using contrast matrix):

```{r}
robust_se_age_10 <- sqrt(
    rep(1, 3) %*%
        geefit_indep$robust.variance[
            c("age1", "age2", "age3"),
            c("age1", "age2", "age3")
        ] %*%
        rep(1, 3)
)

robust_se_age_10
```

Same as reported on p.273 in the book.

## Table 6.9: Main Effects Models

Fit without interactions:

```{r}
geefit_indep_no_interaction <- gee(
    formula = resp ~ age + smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "independence"
)

geefit_exch_no_interaction <- gee(
    formula = resp ~ age + smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "exchangeable"
)

geefit_unstruct_no_interaction <- gee(
    formula = resp ~ age + smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "unstructured"
)
```

Results (same as the numbers given in Table 6.9):

```{r}
summary(geefit_indep_no_interaction)$coeff
summary(geefit_exch_no_interaction)$coeff
summary(geefit_unstruct_no_interaction)$coeff
```

The robust variances are not the same in general:

```{r}
geefit_indep_no_interaction$robust.variance
geefit_exch_no_interaction$robust.variance
geefit_unstruct_no_interaction$robust.variance

summary(geefit_exch)$coeff
```

## GEE with Unstructured Correlation

Fit GEE model with unstructured correlation (unique correlation for each pair of time points):

```{r}
geefit_unstruct <- gee(
    formula = resp ~ age * smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "unstructured"
)

summary(geefit_unstruct)$coeff
```

## Comparing Coefficient Estimates

Compare coefficients across the three correlation structures - same numbers as Table 6.8 in Fahrmeir & Tutz:

```{r}
summary(geefit_indep)$coeff
summary(geefit_exch)$coeff
summary(geefit_unstruct)$coeff
```

## Robust Variance Matrices

Extract robust (sandwich) variance matrices from each model:

```{r}
geefit_indep$robust.variance
geefit_exch$robust.variance
geefit_unstruct$robust.variance
```

The three robust variances happen to look exactly the same here, probably just a numerical coincidence. However, when computing differences, R shows they differ by very small numbers:

```{r}
geefit_indep$robust.variance - geefit_exch$robust.variance
```

## Scale Parameter Estimation

Extract the scale (dispersion) parameter from the summary:

```{r}
summary(geefit_indep)$scale
```

Now tease out the Pearson residuals and compute the method of moments dispersion estimator by hand:

```{r}
pearson_resid <- (ohio$resp - geefit_indep$fitted) /
    sqrt(
        geefit_indep$fitted.values * (1 - geefit_indep$fitted.values)
    )

scale_by_hand <- sum(pearson_resid^2) /
    (geefit_indep$nobs - length(geefit_indep$coeff))
scale_by_hand
```

This gives the same number as reported in summary.

## Naive vs Robust Standard Errors

To verify how they compute the naive SE, fit the model with scale fixed to 1:

```{r}
geefit_indep_scale_fix <- gee(
    formula = resp ~ age * smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "independence",
    scale.fix = TRUE
)

summary(geefit_indep_scale_fix)$coeff[, "Naive S.E."] *
    sqrt(summary(geefit_indep)$scale)
```

This is the same as:

```{r}
summary(geefit_indep)$coeff[, "Naive S.E."]
```

So it only differs by the dispersion parameter, which makes sense.

## Computing Beta for Age 10

With effect coding, the coefficient for age = 10 (not in the design matrix) is computed as:

```{r}
beta_age_10 <- -sum(summary(geefit_indep)$coeff[
    c("age1", "age2", "age3"),
    1
])
beta_age_10
```

This is the same as the book.

## Robust Standard Error for Age 10

Compute the robust SE for the combined age=10 coefficient:

```{r}
robust_se_age_10 <- sqrt(
    rep(1, 3) %*%
        geefit_indep$robust.variance[
            c("age1", "age2", "age3"),
            c("age1", "age2", "age3")
        ] %*%
        rep(1, 3)
)

robust_se_age_10
```

This is the same as reported on p.273 in the book.

## Table 6.9: Models Without Interactions

Now fit the simpler main effects models (without age:smoke interaction):

```{r}
geefit_indep_no_interaction <- gee(
    formula = resp ~ age + smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "independence"
)

geefit_exch_no_interaction <- gee(
    formula = resp ~ age + smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "exchangeable"
)

geefit_unstruct_no_interaction <- gee(
    formula = resp ~ age + smoke,
    data = ohio,
    family = binomial,
    id = id,
    corstr = "unstructured"
)
```

Results from Table 6.9:

```{r}
summary(geefit_indep_no_interaction)$coeff
summary(geefit_exch_no_interaction)$coeff
summary(geefit_unstruct_no_interaction)$coeff
```

These are the same as the numbers given in Table 6.9.

## Robust Variances for Main Effects Models

Compare robust variances across structures - they are not the same in general:

```{r}
geefit_indep_no_interaction$robust.variance
geefit_exch_no_interaction$robust.variance
geefit_unstruct_no_interaction$robust.variance
```

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Liang, K. Y., & Zeger, S. L. (1986). Longitudinal data analysis using generalized linear models. *Biometrika*.
