---
title: "Ohio Respiratory Data: GEE Analysis"
subtitle: "@fahrmeir_multivariate_2001 Example 6.4"
---

## Data Loading and Setup

```{r}
#| label: libs
library(conflicted)
library(gee)
library(geepack)
library(dplyr)
```

::: callout-note
Previously we have mentioned that the `geepack` implements its `geeglm` function differently to the `gee` function in the `gee` package.
Here we load it simply to access the `ohio` dataset.
:::

We reproduce the key GEE results for the Ohio respiratory data in @fahrmeir_multivariate_2001.
The response variable is binary (whether a respiratory condition is present), repeatedly measured over time within child.

The package `geepack` provides the `ohio` dataset, which we use here.
As always it is a good practice to inspect the help page for the dataset before proceeding by running `?ohio`.

We convert the dataset to a tibble, and inspect it.

```{r}
#| label: data-peek
ohio_tbl <- as_tibble(ohio)
ohio_tbl
```

The `age` and `smoke` variables are currently integers, but should instead be factors.
We convert them to factors before fitting models.

```{r}
#| label: preprocess-classes
ohio_tbl <- ohio_tbl %>%
  mutate(
    age = as.factor(age),
    smoke = as.factor(smoke)
  )
ohio_tbl
```

The age coding in the dataset is:

-   $\texttt{age} = -2$ corresponds to 7 years old
-   $\texttt{age} = -1$ corresponds to 8 years old
-   $\texttt{age} = 0$ corresponds to 9 years old
-   $\texttt{age} = 1$ corresponds to 10 years old

## Contrast Coding

As in the textbook, we need effect coding for unordered factors.

```{r}
#| label: show-default-contrasts
getOption("contrasts")
contrasts(ohio_tbl$age)
contrasts(ohio_tbl$smoke)
```

It is currently set to dummy coding, so we change it to effect coding and verify:

```{r}
#| label: set-effect-coding
options(contrasts = c("contr.sum", "contr.poly"))

contrasts(ohio_tbl$age)
contrasts(ohio_tbl$smoke)
```

10 years old is already the reference level for the age factor, but we need to explicitly set the reference level for `smoke` to match the textbook's.

```{r}
#| label: set-smoking-contrast
contrasts(ohio_tbl$smoke) <- matrix(c(-1, 1), 2, 1)
contrasts(ohio_tbl$smoke)
```

## Model Fitting: GEE with Interaction

We fit the interaction effect model: $\texttt{resp} \sim \texttt{smoke} * \texttt{age}$ under three working correlation structures.

::: callout-tip
## What `id` and `corstr` Do in `gee()`

-   `id` identifies which rows belong to the same subject/cluster. In this dataset, repeated observations from the same child share the same `id`.
-   `corstr` specifies the working correlation pattern within each `id` cluster.

Allowed `corstr` values in `gee` include:

-   `"independence"`: assumes within-subject correlation is 0
-   `"fixed"`: user-supplied fixed working correlation
-   `"stat_M_dep"` / `"non_stat_M_dep"`: not covered in this course
-   `"exchangeable"`: one common correlation for all within subject pairs
-   `"AR-M"`: autoregressive structure **Dennis can delete explaination, don't think this is used**
-   `"unstructured"`: each pairwise correlation can differ

For argument details and optional tuning parameters, run `?gee`.
:::

### Independence Structure

We first fit the model under the independence working correlation.

```{r}
#| label: fit-independence
geefit_indep <- gee(
  formula = resp ~ smoke * age,
  id = id,
  data = ohio_tbl,
  family = binomial,
  corstr = "independence"
)

summary_indep <- summary(geefit_indep)

coefs_indep <- coef(summary_indep)

coefs_indep |>
  round(3)
```

::: {.callout-note collapse="true"}
## Reading `coefs_indep` Notation

`coefs_indep` is a matrix, with model terms in rows and summary statistics in columns.

-   `coefs_indep["age1", "Estimate"]` gets one cell
-   `coefs_indep[c("age1", "age2", "age3"), "Estimate"]` gets a vector of selected rows from one column
-   `coefs_indep[, "Naive S.E."]` gets the full `Naive S.E.` column

In matrix indexing, the format is always `[rows, columns]`.
:::

These values replicate the values in Table 6.8 of @fahrmeir_multivariate_2001 (up to numerical rounding).

### Exchangeable and Unstructured Correlation Structures

We also try the exchangeable and unstructured working correlations.

```{r}
#| label: fit-exchangeable
geefit_exch <- gee(
  formula = resp ~ smoke * age,
  id = id,
  data = ohio_tbl,
  family = binomial,
  corstr = "exchangeable"
)

summary(geefit_exch) |>
  coef() |>
  round(3)
```

```{r}
#| label: fit-unstructured
geefit_unstruct <- gee(
  formula = resp ~ smoke * age,
  id = id,
  data = ohio_tbl,
  family = binomial,
  corstr = "unstructured"
)

summary(geefit_unstruct) |>
  coef() |>
  round(3)
```

::: {.callout-important collapse="true"}
## SELF TEST: Compare Coefficients Across Working Correlations

Fill in:

-   Which coefficients look most sensitive to the working correlation?
-   Which are nearly unchanged?
-   Why is this behavior expected under the GEE framework?
:::

::: callout-note
## Robust (Sandwich) Variance Matrices

**Dennis: Without looking into this, I find this behaviour very weird**

We can also extract the robust variance matrices under each working correlation structure.

```{r}
#| label: robust-vcov-int
geefit_indep$robust.variance |> round(6)
geefit_exch$robust.variance |> round(6)
geefit_unstruct$robust.variance |> round(6)
```

Eyeballing the three matrices, we can see that they are all almost identical.
We can confirm this by computing the difference between the independence and exchangeable robust variance matrices:

```{r}
#| label: robust-vcov-diff
geefit_indep$robust.variance - geefit_exch$robust.variance
```

In theory all the robust variance matrices should differ, but here, perhaps due to the nature of the data, they are very close to each other.
:::

## Scale Parameter and Pearson-Based Check

We can also access the scale parameter estimate from the model:

```{r}
#| label: scale-from-summary
geefit_indep$scale
```

We can see that there is very minimal overdispersion here.

We can derive the scale estimate ourselves.
Compute the method-of-moments dispersion estimate by hand from Pearson residuals:

```{r}
#| label: scale-by-hand
pearson_resid <- (ohio_tbl$resp - geefit_indep$fitted) /
  sqrt(geefit_indep$fitted.values * (1 - geefit_indep$fitted.values))

length(pearson_resid)
```

::: callout-note
The number of Pearson residuals should match the number of observations used in the model fit, which is the same as the number of rows in `ohio_tbl`.
Each row is an observation of a child, where each child is observed multiple times.
:::

```{r}
scale_by_hand <- sum(pearson_resid^2) /
  (geefit_indep$nobs - length(coef(geefit_indep)))

scale_by_hand
```

The hand-computed value matches the model summary scale.

## Naive Standard Errors with Scale Fixed at 1

To verify the naive standard errors reported by `gee`, we refit with `scale.fix = TRUE`, which fixes the scale parameter at 1.

```{r}
#| label: fit-indep-scale-fix
geefit_indep_scale_fix <- gee(
  formula = resp ~ smoke * age,
  id = id,
  data = ohio_tbl,
  family = binomial,
  corstr = "independence",
  scale.fix = TRUE
)

geefit_indep_scale_fix$scale
```

We can see that the scale is fixed at 1.

Check relationship:

$$
    \text{Naive SE (scale fixed)} \times \sqrt{\hat\phi}
=
    \text{Naive SE (reported in original fit)}.
$$

```{r}
#| label: se-rescaling-comparison
naive_se_scale_fix <- coef(summary(geefit_indep_scale_fix))[, "Naive S.E."]

tibble(
  term = rownames(coefs_indep),
  se_rescaled = naive_se_scale_fix *
    sqrt(geefit_indep$scale),
  se_original = coefs_indep[, "Naive S.E."]
)
```

## Derived Parameter: Coefficient for Age = 10

With effect coding, we can extract the coefficient for for age 10.
It is the negative sum of the three age coefficients, since the age 10 group is the reference group for the age factor.

```{r}
#| label: beta-age-10
coefs_indep |> round(3)

beta_age_10 <- -sum(coefs_indep[
  c("age1", "age2", "age3"),
  "Estimate"
])

beta_age_10
```

This value matches page 273 of @fahrmeir_multivariate_2001.

::: {.callout-important collapse="true"}
## SELF TEST: Interpretation of Age-10 Coefficient being Negative

The probability of infection for a child in the age-10 group is lower than the average probability across all age groups, holding smoking status fixed.
:::

@fahrmeir_multivariate_2001 also reports a standard deviation for this coefficient, which we can also reproduce.
It will simple The robust standard error for this linear combination is:

```{r}
#| label: robust-se-age-10
age_coef_names <- c("age1", "age2", "age3")

age_robust_vcov <- geefit_indep$robust.variance[
  age_coef_names,
  age_coef_names
]

age_robust_vcov |> round(4)

robust_se_age_10 <- sqrt(
  rep(1, 3) %*%
    age_robust_vcov %*%
    rep(1, 3)
)

robust_se_age_10
```

## Table 6.9: Main-Effects Models (No Interaction)

The robust z-values (or alternatively the estimates and robust standard errors) reported in the model output tell us that the interaction terms are not statistically significant.

Therefore we fit models without the interaction terms, and try to recover the main effect coefficients in @fahrmeir_multivariate_2001 Table 6.9.

```{r}
#| label: fit-main-effects
geefit_indep_no_interaction <- gee(
  formula = resp ~ smoke + age,
  id = id,
  data = ohio_tbl,
  family = binomial,
  corstr = "independence"
)

summary(geefit_indep_no_interaction) |>
  coef() |>
  round(3)
```

These values line up with Table 6.9 of @fahrmeir_multivariate_2001.

We can do the same for the exchangeable and unstructured working correlations:

```{r}
geefit_exch_no_interaction <- gee(
  formula = resp ~ smoke + age,
  id = id,
  data = ohio_tbl,
  family = binomial,
  corstr = "exchangeable"
)

geefit_unstruct_no_interaction <- gee(
  formula = resp ~ smoke + age,
  id = id,
  data = ohio_tbl,
  family = binomial,
  corstr = "unstructured"
)

summary(geefit_exch_no_interaction) |>
  coef() |>
  round(3)
summary(geefit_unstruct_no_interaction) |>
  coef() |>
  round(3)
```

Here we can see that the robust variances differ more across working correlation structures than in the interaction model.:

```{r}
#| label: robust-vcov-main
geefit_indep_no_interaction$robust.variance |> round(6)
geefit_exch_no_interaction$robust.variance |> round(6)
geefit_unstruct_no_interaction$robust.variance |> round(6)
```

## Package Citations

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "gee", "geepack", "dplyr")
for (pkg in packages) {
  cit <- citation(pkg)

  # Check if multiple citations
  if (length(cit) > 1) {
    cat("- **", pkg, "**:\n", sep = "")
    for (i in seq_along(cit)) {
      cit_text <- format(cit[i], style = "text")
      cat("    - ", cit_text, "\n\n", sep = "")
    }
  } else {
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
  }
}
```
