---
title: "GEE Analysis: CTSIB Balance Test Data"
subtitle: "Generalized Estimating Equations for Repeated Measures"
date: today
---

## Overview

This script reproduces the simulations on p.292 of Faraway, 2nd edition, 2016, or p.225 in the first edition.

## Data Setup

Set default contrasts and load the data:

```{r}
library(conflicted)
library(gee)
library(geepack)
library(faraway)

options(contrasts = c("contr.treatment", "contr.poly"))

data(ctsib)
```

Read the description of the data:

```{r}
?ctsib
```

Take a look at the data:

```{r}
head(ctsib, 20)
str(ctsib)
```

### Creating Binary Response

Create binary responses: 1 represents completely stable, otherwise not:

```{r}
ctsib$stable <- ifelse(ctsib$CTSIB == 1, 1, 0)
```

## GEE with Exchangeable Correlation

Fit GEE with scale fixed at 1:

```{r}
gg <- gee(
    stable ~ Sex + Age + Height + Weight + Surface + Vision,
    id = Subject,
    family = binomial,
    data = ctsib,
    corstr = "exchangeable",
    scale.fix = TRUE
)

gg$coefficients
summary(gg)
```

Faraway didn't fit with scale.fix = FALSE, but we try it:

```{r}
gg_scale <- gee(
    stable ~ Sex + Age + Height + Weight + Surface + Vision,
    id = Subject,
    family = binomial,
    data = ctsib,
    corstr = "exchangeable",
    scale.fix = FALSE
)

gg_scale$coefficients
gg_scale$scale
```

## Alternative: geepack Package

Use the geepack package. Note: its estimate of dispersion is NOT based on our typical Pearson residual method of moment estimator. See the original paper by Yan and Fine:

```{r}
# modgeep <- geeglm(
#   stable ~ Sex + Age + Height + Weight + Surface + Vision,
#   id = Subject,
#   corstr = "exchangeable",
#   scale.fix = TRUE,
#   data = ctsib,
#   family = binomial
# )
# modgeep$coefficients
# summary(modgeep)
# summary(modgeep)$scale.value
# summary(modgeep)$dispersion
#
# modgeep.scale <- geeglm(
#   stable ~ Sex + Age + Height + Weight + Surface + Vision,
#   id = Subject,
#   corstr = "exchangeable",
#   scale.fix = FALSE,
#   data = ctsib,
#   family = binomial
# )
#
# modgeep.scale$coefficients
# summary(modgeep.scale)$dispersion
```
