---
title: "Generalized Estimating Equations: CTSIB Balance Test Data"
subtitle: "@faraway_extending_2016 Section 13.5"
date: last-modified
editor:
  markdown:
    wrap: sentence
    canonical: true
---

## Overview

We now look at the CTSIB balance test dataset from the very applied book by @faraway_extending_2016, which can be [accessed online](https://research.ebsco.com/plink/3806c0de-f79d-330d-9c4e-7bb136efaa53){target="_blank"}.
In the second edition of this book, it is in section 13.5, page 292.

## Data Setup

First, we load the data and set the default contrasts to effect coding:

```{r}
#| label: 08-gee-ctsib-chunk-1
library(conflicted)
library(dplyr)
library(faraway)
library(gee)

options(contrasts = c("contr.treatment", "contr.poly"))

data(ctsib)
```

As always, we should spend some time [reading the documentation](https://search.r-project.org/CRAN/refmans/faraway/html/ctsib.html){target="_blank"} to understand the data set.

```{r}
#| label: 08-gee-ctsib-chunk-2
#| #| eval: false
?ctsib
```

We can also inspect the data types by converting the data to a tibble:

```{r}
#| label: 08-gee-ctsib-chunk-3
ctsib_tbl <- as_tibble(ctsib)
head(ctsib_tbl, 20)
```

### Creating Binary Response

We convert the ordinal CTSIB outcome into a binary variable indicating "stable" (when `CTSIB == 1`) vs. "not stable" balance, just like what was done on p.277 of @faraway_extending_2016.

```{r}
#| label: 08-gee-ctsib-chunk-4

ctsib_tbl <- ctsib_tbl |>
  mutate(stable = (CTSIB == 1))

ctsib_tbl
```

## Fitting the Model with `gee::gee()`

We fit a model using the generalised estimating equations assuming **exchangeable correlation structure** among repeated measurements within each subject.

::: {.callout-important collapse="true"}
## SELF TEST: What is the Exchangeable Correlation Structure?

The exchangeable correlation structure assumes that any pair of observations within the same cluster (subject) has the same correlation.
:::

```{r}
#| label: 08-gee-ctsib-chunk-5
gg <- gee(
  stable ~ Sex + Age + Height + Weight + Surface + Vision,
  id = Subject,
  data = ctsib_tbl,
  family = binomial,
  corstr = "exchangeable",
  scale.fix = TRUE
)

coef(gg)
summary(gg)
```

::: {.callout-note collapse="true"}
## Revision on `gee::gee()` syntax

-   `stable ~ Sex + Age + Height + Weight + Surface + Vision`: a main effects model, with `stable` as the response.
-   `id = Subject`: identifies the subject-level clusters for repeated measures.
-   `data = ctsib_tbl`: the data frame used to evaluate the variables in the formula.
-   `family = binomial`: a binomial mean-variance model for the binary response.
-   `corstr = "exchangeable"`: assumes a common within-subject correlation for all repeated observations.
-   `scale.fix = TRUE`: fixes the scale parameter fixed to 1.
:::

### Empirical Scale Estimation

Alternatively, we can also estimate the scale from the data (which @faraway_extending_2016 chose not to do for reasons explained on their p.292. But they also recognize generally there wouldn't be a strong reason to fix it).

```{r}
#| label: 08-gee-ctsib-chunk-6
gg_scale <- gee(
  stable ~ Sex + Age + Height + Weight + Surface + Vision,
  id = Subject,
  family = binomial,
  data = ctsib_tbl,
  corstr = "exchangeable",
  scale.fix = FALSE
)
summary(gg_scale)
gg_scale$scale
```

We can compare the coefficient estimates from the fixed scale vs. empirical scale models:

```{r}
#| label: 08-gee-ctsib-chunk-6b
coef_compare <- tibble(
  term = names(coef(gg)),
  fixed_scale = coef(gg),
  empirical_scale = coef(gg_scale),
)

coef_compare
```

The coefficient estimates actually appear numerically identical, which we cannot fully explain at the moment: Under the exchangeable correlation working model, the equicorrelation $\alpha$'s estimation (embedded in the working covariance matrix $\Sigma_i$ for each multivariate ${\bf y}_i$) should also depend on an estimated dispersion $\phi$ (if it is not fixed).
Therefore, the dispersion shouldn't just simply factor out as a scalar in the estimating equations and the two approaches (fixed scale vs unfixed scale) should be solving non-identical equations.

::: {.callout-note collapse="true"}
## Alternative: geepack Package (optional and non-examinable)

The **geepack** package provides an alternative implementation of GEE with some methodological differences from the paper by @yan2004estimating.
This is the same package used in @faraway_extending_2016.
The code is provided here without commentary.

### Fitting with geepack (Fixed Scale)

```{r}
#| label: 08-gee-ctsib-chunk-7
library(geepack)

modgeep <- geeglm(
  stable ~ Sex + Age + Height + Weight + Surface + Vision,
  id = Subject,
  corstr = "exchangeable",
  scale.fix = TRUE,
  data = ctsib_tbl,
  family = binomial
)

coef(modgeep)
summary(modgeep)
summary(modgeep)$scale.value
summary(modgeep)$dispersion
```

### Fitting with geepack (Empirical Scale)

```{r}
#| label: 08-gee-ctsib-chunk-8
modgeep_scale <- geeglm(
  stable ~ Sex + Age + Height + Weight + Surface + Vision,
  id = Subject,
  corstr = "exchangeable",
  scale.fix = FALSE,
  data = ctsib_tbl,
  family = binomial
)

coef(modgeep_scale)
summary(modgeep_scale)$scale.value
summary(modgeep_scale)$dispersion
```

### Final Comparison: All Four Models

```{r}
#| label: 08-gee-ctsib-chunk-9
all_terms <- sort(unique(c(
  names(coef(gg)),
  names(coef(gg_scale)),
  names(coef(modgeep)),
  names(coef(modgeep_scale))
)))

coef_compare_all <- tibble(
  term = all_terms,
  gee_fixed = unname(coef(gg)[all_terms]),
  gee_empirical = unname(coef(gg_scale)[all_terms]),
  geepack_fixed = unname(coef(modgeep)[all_terms]),
  geepack_empirical = unname(coef(modgeep_scale)[all_terms])
)

coef_compare_all
```
:::

## Package Citations and Session Information

::: {.callout-tip collapse="true"}
## Session Information

```{r}
sessionInfo()
```
:::

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "gee", "geepack", "dplyr", "faraway")
for (pkg in packages) {
  cit <- citation(pkg)

  # Check if multiple citations
  if (length(cit) > 1) {
    cat("- **", pkg, "**:\n", sep = "")
    for (i in seq_along(cit)) {
      cit_text <- format(cit[i], style = "text")
      cat("    - ", cit_text, "\n\n", sep = "")
    }
  } else {
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
  }
}
```
