---
title: "Linear Mixed Models: Pulp Brightness and PSID Longitudinal Data"
subtitle: "@faraway_extending_2016 Sections 10.1 and 11.1"
date: last-modified
editor:
  markdown:
    wrap: sentence
    canonical: true
---

## @faraway_extending_2016 Section 10.1: Pulp Brightness

### Data Loading

This script demonstrates **linear mixed effects models** using the `lme4` package, by @bates_fitting_2015, using the example in @faraway_extending_2016 section 10.1, page 200.
This package is the standard package for fitting mixed effects models.

The first dataset we use measures paper brightness (continuous response) depending on which shift operator prepared the pulp.

```{r}
#| label: 09-lm-mixed-effect-chunk-1
library(conflicted)
library(dplyr)
library(faraway)
library(ggplot2)
library(lme4)
```

As always we can run `?pulp` to get information about the dataset:

```{r}
#| label: 09-lm-mixed-effect-chunk-2
#| eval: false
?pulp
```

```{r}
#| label: 09-lm-mixed-effect-chunk-3
pulp_tbl <- as_tibble(pulp)
pulp_tbl
```

We can see that there are 20 observations of brightness, with 5 measurements for each of 4 operators.

We want to model brightness using an intercept only model with random operator effects.

::: {.callout-important collapse="true"}
## SELF TEST: What is the mathematical formulation of this model?

The model is: $$
y_{ij} = \beta_0 + b_i + \varepsilon_{ij}
$$ where $y_{ij}$ is the brightness for the $j$-th measurement of the $i$-th operator, $\beta_0$ is the overall mean brightness (fixed effect), $b_i \sim N(0, \sigma_b^2)$ is the random effect for operator $i$, and $\varepsilon_{ij} \sim N(0, \sigma_{\varepsilon}^2)$ is the observation-level error.

We estimate $\beta_0$ as the fixed effect, and $\sigma_b^2$ and $\sigma_{\varepsilon}^2$ as the variance components for the random effects and residuals, respectively.
:::

We fit a model using a REML criterion, and a model using a MLE criterion, to compare the results.

::: {.callout-important collapse="true"}
## SELF TEST: REML vs. MLE in linear mixed models

In mixed effects models we need to estimate the variance components (random effects variance and residual variance) in addition to the fixed effects.
The standard MLE estimate has a strong downward bias for variance components (it will underestimate them).
REML corrects for this bias.
:::

```{r}
#| label: 09-lm-mixed-effect-chunk-4
m_pulp_reml <- lmer(bright ~ 1 + (1 | operator), pulp_tbl)
m_pulp_mle <- lmer(bright ~ 1 + (1 | operator), pulp_tbl, REML = FALSE)
```

::: {.callout-tip collapse="true"}
## `lme4::lmer()` Syntax

`lme4::lmer()` uses the pattern `response ~ fixed effects + (random effects | group), data = ...`.
The second argument is the data frame, and `REML = FALSE` switches from REML to MLE.
Here `(1 | operator)` gives a random intercept for each operator.

Note that the `1 +` in the fixed effects is optional since the intercept is included by default, but here it is included explicitly for clarity.
:::

The `m_pulp_reml` object is of class lmerMod.
Let's examine both fits:

```{r}
#| label: 09-lm-mixed-effect-chunk-5
sum_pulp_reml <- summary(m_pulp_reml)
sum_pulp_reml
```

We get one fixed effect (the overall mean brightness) and one random effect (the operator-level variation).

**DENNIS TO ADD SOME EXPOSITION ON THE STANDARD ERROR OF FIXED EFFECTS HERE**

For the random effect we get two variance parameters.
The `Residual` variance is the estimated variance $\sigma_{\varepsilon}^2$ of the obervation-level errors $\varepsilon_{ij}$, while the `operator` variance is the estimated variance $\sigma_b^2$ of the operator-level random effects $b_i$.
Here the value reported by `Std.Dev.` is just the square root of the variance (not a standard error of the variance estimate).

We can also see the estimates for the MLE fit:

```{r}
#| label: 09-lm-mixed-effect-chunk-6
summary(m_pulp_mle)
```

We can see that (as expected) the MLE variance estimates are smaller than the REML estimates, while the fixed effect estimate is the same.

We can directly extract the fixed effect estimates:

```{r}
#| label: 09-lm-mixed-effect-chunk-7
tibble(
  REML = fixef(m_pulp_reml),
  MLE = fixef(m_pulp_mle)
)
```

::: callout-caution
The fixed effect estimates here are the same between REML and MLE, **only because the design is balanced**.
Generally they will differ for unbalanced designs.

**DENNIS: TO ADD MORE CONTEXT IF NEEDED, CLAIM IS THAT REML VARIANCE ESTIMATES MUST BE SCALAR MULTIPLE OF MLE VARIANCE ESTIMATES IN BALANCED DESIGNS**
:::

#### Extracting Model Components

Usually we don't extract the random effects estimates directly (since there is only a posterior distribution), but we can extract the conditional modes using the generic method `lme4::ranef()`:

```{r}
#| label: 09-lm-mixed-effect-chunk-8
as_tibble(ranef(m_pulp_reml))
```

Since the conditional distribution is normal, the conditional modes are also the conditional means.

Variance-covariance matrix for the fixed effects:

```{r}
#| label: 09-lm-mixed-effect-chunk-9
sum_pulp_reml$vcov
```

This is a constant in this case since we are only estimating one fixed effect (the intercept).

We can directly access the estimated residual standard deviation $\sigma_{\varepsilon}$:

```{r}
#| label: 09-lm-mixed-effect-chunk-10
sum_pulp_reml$sigma
```

We can access both variance-covariance components ($\sigma_b^2$ and $\sigma_{\varepsilon}^2$):

```{r}
#| label: 09-lm-mixed-effect-chunk-11
sum_pulp_reml$varcor
```

And the usual generic method `coef()` function gives us the fixed effect coefficient table:

```{r}
#| label: 09-lm-mixed-effect-chunk-12
coef(sum_pulp_reml)
```

### Hand-Computed Fixed Effect Estimate

Since fixed effect estimates for linear mixed models are a special case of solving a generalised least squares (GLS) estimator, we can verify the fixed effect estimate by hand-computing the GLS solution from the estimated variance structure:

```{r}
#| label: 09-lm-mixed-effect-chunk-13

components <- as_tibble(VarCorr(m_pulp_reml))$vcov
components
Vi <- components[2] * diag(5) + components[1] * matrix(1, nrow = 5, ncol = 5)
V <- bdiag(Vi, Vi, Vi, Vi)
Z <- rep(1, 20)
```

::: {.callout-tip collapse="true"}
## What This Code Does and Syntax Notes

1.  `components <- tibble::as_tibble(lme4::VarCorr(m_pulp_reml))$vcov` extracts estimated variance components from the fitted model. `lme4::VarCorr(...)` returns random-effect and residual variance information, `tibble::as_tibble(...)` turns it into a tidy table, and `$vcov` selects the variance column.
2.  `Vi <- components[2] * ...` builds the $5 \times 5$ covariance matrix for one operator. `diag(5)` creates an identity matrix, and `matrix(1, nrow = 5, ncol = 5)` creates a $5 \times 5$ matrix of ones.
3.  `V <- Matrix::bdiag(Vi, Vi, Vi, Vi)` creates a block-diagonal covariance matrix for all 4 operators. `Matrix::bdiag(...)` stacks the operator-level covariance blocks on the diagonal.
4.  `Z <- rep(1, 20)` creates the intercept-only fixed-effect design vector for all 20 observations.
:::

Hand-computed GLS estimate of the fixed effect (intercept):

```{r}
#| label: 09-lm-mixed-effect-chunk-14
(1 / (t(Z) %*% solve(V) %*% Z)) * t(Z) %*% solve(V) %*% pulp$bright
```

Hand-computed standard error (GLS covariance):

```{r}
#| label: 09-lm-mixed-effect-chunk-15
sqrt(1 / (t(Z) %*% solve(V) %*% Z))
```

### Model Diagnostics

This course does not focus on model diagnostics, but the general idea behind these plots is to evaluate whether our model assumptions are met.
The default `residuals` function accounts for the estimated random effects.
These residuals can be regarded as estimates of the observation-level errors $\varepsilon_{ij}$, which is usually what we want for diagnostics:

```{r}
#| label: 09-lm-mixed-effect-chunk-16
resid_m_pulp_reml <- residuals(m_pulp_reml)
fitted_m_pulp_reml <- fitted(m_pulp_reml)
diagnostics_tbl <- tibble(residual = resid_m_pulp_reml, fitted = fitted_m_pulp_reml)

ggplot(diagnostics_tbl, aes(sample = residual)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical Quantiles", y = "Residuals") +
  theme_minimal()

ggplot(diagnostics_tbl, aes(x = fitted, y = residual)) +
  geom_point() +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  labs(x = "Fitted", y = "Residuals") +
  theme_minimal()
```

These plots show that it is a reasonable assumption that the residuals are approximately normally distributed and have constant variance across fitted values.

::: callout-note
We can compute residuals by hand to verify:

```{r}
#| label: 09-lm-mixed-effect-chunk-17
operator_intercept <- ranef(m_pulp_reml)[["operator"]][["(Intercept)"]]
residual_manual <- pulp$bright - rep(operator_intercept, each = 5) - fixef(m_pulp_reml)[["(Intercept)"]]
```

Verify they match the model output:

```{r}
#| label: 09-lm-mixed-effect-chunk-18
all(residual_manual == residuals(m_pulp_reml))
```
:::

## @faraway_extending_2016 Section 11.1: PSID Longitudinal Data

Now we fit a model with **random slopes**.
The Panel Study of Income Dynamics (PSID) dataset tracks income over time for individuals.
We use `log(income)` as the response.

Load and examine data:

```{r}
#| label: 09-lm-mixed-effect-chunk-19
#| eval: false
?psid
```

So this dataset has 1,661 observations on the following 6 variables:

-   `age`: age in 1968
-   `educ`: years of education
-   `sex`: sex of individual (`F` or `M`)
-   `income`: annual income in dollars
-   `year`: calendar year
-   `person`: ID number for individual

```{r}
#| label: 09-lm-mixed-effect-chunk-20

psid_tbl <- as_tibble(psid)
psid_tbl
```

We create a new variable `cyear`, the time relative to 1978

```{r}
#| label: 09-lm-mixed-effect-chunk-21
psid_tbl <- mutate(psid_tbl, cyear = year - 78)
```

::: callout-note
**Centering predictors**: By centering year at 1978, the intercept in our model will represent log-income in 1978.
:::

### REML vs. MLE with Random Slopes

We fit models with both random intercept and random slope (person growth rate varies) using both the REML (default) and the biased MLE (`REML = FALSE`) estimation methods:

```{r}
#| label: 09-lm-mixed-effect-chunk-22
m_psid_reml <- lmer(
  log(income) ~ cyear * sex + age + educ + (cyear | person),
  data = psid_tbl
)

m_psid_mle <- lmer(
  log(income) ~ cyear * sex + age + educ + (cyear | person),
  data = psid_tbl,
  REML = FALSE
)
```

::: callout-note
`(cyear | person)` gives **both** a random intercept and slope for each person.
:::

::: {.callout-important collapse="true"}
## SELF TEST: What is the mathematical formulation of this model?

The model is: $$
\begin{split}
\log(\texttt{income})_{ij} &= \beta_0 + \beta_1 \texttt{cyear}_{j} + \beta_2 \mathbb{1}(\texttt{sex}_{i}=\texttt{M}) + \beta_3 \texttt{age}_{i} + \beta_4 \texttt{educ}_{i}\\
&\quad + \beta_5 \texttt{cyear}_{j}\mathbb{1}(\texttt{sex}_{i}=\texttt{M}) + b_{0i} + b_{1i}\texttt{cyear}_{j} + \varepsilon_{ij}
\end{split}
$$ where $\log(\texttt{income}_{ij})$ corresponds to person $i$ at time $j$, and $\mathbb{1}(\texttt{sex}_{i}=\texttt{M})$ corresponds to the `sexM` coefficient in the output.
The random effects are correlated: $$
\begin{pmatrix} b_{0i} \\ b_{1i} \end{pmatrix}
\sim N\Biggl(
\begin{pmatrix}0\\0\end{pmatrix},
\begin{pmatrix}
\sigma_{b_0}^2 & \sigma_{b_{01}}\\
\sigma_{b_{01}} & \sigma_{b_1}^2
\end{pmatrix}
\Biggr),
\quad
\varepsilon_{ij}\sim N(0,\sigma_\varepsilon^2).
$$
:::

Examine REML fit:

```{r}
#| label: 09-lm-mixed-effect-chunk-23
s_psid_reml <- summary(m_psid_reml)
s_psid_reml
```

The above summary is more complex than the previous example because we have more fixed effects and a more complex random effects structure.
The `Correlation of Fixed Effects` table shows the correlation between the fixed effect estimates.
The covariance matrix of the fixed effects estimates can easily be derived from this.

The `Random effects` table shows the variance components $\sigma_{b_0}^2$ and $\sigma_{b_1}^2$ for the random intercept and random slope $b_{1i}$, as well as their correlation (from which $\sigma_{b_{01}}$ can be calculated), and the residual variance $\sigma_\varepsilon^2$.

We can directly access the variance-covariance matrix of fixed effects:

```{r}
#| label: 09-lm-mixed-effect-chunk-24
s_psid_reml$vcov
```

We can compare the fixed effect estimates from the REML and MLE fits:

```{r}
#| label: 09-lm-mixed-effect-chunk-25

tibble(
  names = names(fixef(m_psid_reml)),
  REML = fixef(m_psid_reml),
  MLE = fixef(m_psid_mle)
)
```

Note that here the fixed effect estimates differ slightly between REML and MLE, which is to be expected.

Finally we can access at the standard errors of the fixed effects:

```{r}
#| label: 09-lm-mixed-effect-chunk-26
sqrt(diag(s_psid_reml$vcov))
```

As above, we can examine the variance-covariance of random effects:

```{r}
#| label: 09-lm-mixed-effect-chunk-27
VarCorr(m_psid_reml)
```

::: callout-note
## Interpreting the Random Effects Output

The standard deviation of the random intercept tells us how much variability there is in initial log-income across individuals.
Since income is on a log-scale, this values is multiplicative---95% of incomes are between $\exp(-2\sigma_{b_0})$= `r {round(exp(-2*as_tibble(s_psid_reml$varcor)$sdcor[1]), 2)}` and $\exp(2\sigma_{b_0})$= `r {round(exp(2*as_tibble(s_psid_reml$varcor)$sdcor[1]), 2)}` times the overall mean income in 1978.

The standard deviation of the random slope tells us how much variability there is in income growth rates across individuals.

A low correlation value tells us that the slope of income growth over time, is minimally effected by the initial salary (intercept).
:::

## Package Citations and Session Information

::: {.callout-tip collapse="true"}
## Session Information

```{r}
sessionInfo()
```
:::

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "faraway", "ggplot2", "lme4")
for (pkg in packages) {
  cit <- citation(pkg)

  # Check if multiple citations
  if (length(cit) > 1) {
    cat("- **", pkg, "**:\n", sep = "")
    for (i in seq_along(cit)) {
      cit_text <- format(cit[i], style = "text")
      cat("    - ", cit_text, "\n\n", sep = "")
    }
  } else {
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
  }
}
```
