---
title: "Generalised Linear Mixed Models: Ohio Respiratory Data and Wine Bitterness"
subtitle: "@fahrmeir_multivariate_2001 Examples 7.1--7.4"
date: last-modified
editor:
  markdown:
    wrap: sentence
    canonical: true
---

::: {.callout-tip collapse="true"}
## GLMM Software Options

Many R packages exist for fitting generalised linear mixed models, using different estimation methods:

-   **lme4::glmer** (most widely used): Frequentist GLMMs via Laplace approximation (`nAGQ = 1`) or adaptive Gauss-Hermite quadrature (`nAGQ > 1`)
-   **MASS::glmmPQL**: Penalized quasi-likelihood (PQL; approximate, fast, but can be biased for binary/sparse outcomes)
-   **MCMCglmm**: Bayesian GLMMs fitted by Markov chain Monte Carlo (MCMC)
-   **glmmTMB** (very popular modern alternative): Frequentist GLMMs via Template Model Builder/Laplace; especially strong for count models, zero inflation, and overdispersion
-   **ordinal::clmm**: Cumulative link mixed models for ordinal outcomes (an ordinal GLMM family)
-   **glmmML**: Classical random-intercept GLMMs using adaptive Gauss-Hermite quadrature (more limited model scope)
-   **ASREML-R**: Commercial software with broad mixed-model/GLMM capability, widely used in biostatistics/agricultural genetics
-   **glmmADMB**: GLMMs via AD Model Builder/Laplace (historically important, but largely superseded and no longer actively maintained)
-   **glmm** (from Jim Lindsey's `repeated` package): Older AGHQ-based approach for repeated-measures GLMMs (now uncommon)
-   **sabreR**, **mcemGLM**, **gamlss.mx**: Specialized/less commonly used frameworks that can fit mixed or latent/random-effect extensions in narrower settings
:::

## Ohio Respiratory Data @fahrmeir_multivariate_2001 Example 7.1 and 7.3

### Setup

The Ohio respiratory dataset contains repeated respiratory status measurements (yes/no) for children in a longitudinal study, with covariates including age and smoking status.
We'll fit GLMM models to account for within-child correlation using random intercepts.
Our goal is to reproduce some of Table 7.2 from @fahrmeir_multivariate_2001.

Load required packages and do the standard preparation of the Ohio respiratory dataset:

```{r}
#| label: 10-glm-mixed-effect-chunk-1
library(conflicted)
library(dplyr)
library(lme4)
library(geepack)
library(MASS)
library(ordinal)

ohio_tbl <- as_tibble(ohio)
ohio_tbl

ohio_tbl <- mutate(ohio_tbl, across(c(age, smoke), factor))

# Check default contrasts
contrasts(ohio_tbl$age)
contrasts(ohio_tbl$smoke)

options(contrasts = c("contr.sum", "contr.poly"))

contrasts(ohio_tbl$age)
contrasts(ohio_tbl$smoke)

contrasts(ohio_tbl$smoke) <- matrix(c(-1, 1), 2, 1)
contrasts(ohio_tbl$smoke)
```

### `lme4::glmer()` Fitting with Laplace Approximation

When fitting a generalised linear mixed model (GLMM), the likelihood involves integrating over the random effects, which typically has no closed-form solution.
Therefore we need to choose a numerical integration method to approximate the likelihood.
The `lme4::glmer()` function provides two main methods for this: the Laplace approximation and adaptive Gauss-Hermite quadrature (AGHQ).
The default `glmer()` approach uses the **Laplace approximation**.
Once we have chosen the integration method, we also need to specify the optimization algorithm; the Nelder-Mead optimizer is recommended by @faraway_extending_2016.

```{r}
glmer_fit_nelder_mead <- glmer(
  resp ~ age * smoke + (1 | id),
  data = ohio_tbl,
  family = binomial,
  control = glmerControl(
    optimizer = "Nelder_Mead"
  )
)
```

::: {.callout-important collapse="true"}
## SELF TEST: What does the `(1 | id)` formula mean?

This specifies a random intercept model: `(1 | id)` means each subject (`id`) has its own random intercept drawn from a normal distribution with mean 0 and estimated standard deviation.
The `1` represents the intercept term; `| id` conditions on subject ID.
This allows each subject to have a different baseline probability of respiratory status while maintaining a common effect of the fixed covariates.
:::

We can see that there is an error message about convergence.
Checking `?lme4::troubleshooting` suggests that this type of error may be fixed by trying a different optimiser.
The other optimiser available is Bound Optimization BY Quadratic Approximation (BOBYQA).
Let's try that:

```{r}
#| label: 10-glm-mixed-effect-chunk-5
glmer_fit <- glmer(
  resp ~ age * smoke + (1 | id),
  data = ohio_tbl,
  family = binomial,
  control = glmerControl(
    optimizer = "bobyqa"
  )
)

logLik(glmer_fit) - logLik(glmer_fit_nelder_mead)
as_tibble(coef(summary(glmer_fit)) - coef(summary(glmer_fit_nelder_mead)))
```

There are no longer convergence issues, and comparing the log-likelihoods, we see that the log-likelihood is slightly higher with the BOBYQA optimizer.
The coefficient estimates are slightly different, but the overall pattern of results is the same.

```{r}
summary(glmer_fit)
```

The output is very similar to that in the linear mixed model fit, with fixed effects estimates, as well as a correlation matrix for the fixed effects, and an estimate of the random effect variance.

The coefficient estimates are in the right ballpark, but ultimately don't match very closely with those in the book.
Instead we follow the book's approach and try using adaptive Gauss-Hermite quadrature (AGHQ) to estimate the integration.

::: callout-caution
## Standard Errors in Table 7.2

Although strictly speaking the mixed effects model is not in the exponential family, it is still a likelihood-based model, and so in theory, we should be able to approximate a Fisher information matrix to compute standard errors.
Unfortunately the `lme4::glmer()` function is not well documented enough for us to reproduce the standard errors in the book.
:::

### `lme4::glmer()` Fitting with Adaptive Gauss-Hermite Quadrature

In order to get more accurate estimates, we can improve the numerical integration by using adaptive Gauss-Hermite quadrature (AGHQ) instead of the Laplace approximation.
The `nAGQ` argument in `glmer()` controls the number of quadrature points used for the integration.
The default `nAGQ = 1` corresponds to the Laplace approximation, while `nAGQ > 1` uses multiple quadrature points for a more accurate approximation of the likelihood.

Here we try `nAGQ = 10`, which is what @fahrmeir_multivariate_2001 claim to use in their analysis.

```{r}
#| label: 10-glm-mixed-effect-chunk-6
glmer_fit_n_agq_10 <- glmer(
  resp ~ age * smoke + (1 | id),
  data = ohio_tbl,
  family = binomial,
  control = glmerControl(
    optimizer = "bobyqa"
  ),
  nAGQ = 10
)

summary(glmer_fit_n_agq_10)
```

Comparing to Table 7.2, we are significantly closer to the book's reported estimates.

### Random Effects Estimates

As with the LMMs, we can use the `ranef()` function to extract the conditional modes of the random effects for each subject:

```{r}
#| label: 10-glm-mixed-effect-chunk-7
ranef(glmer_fit)$id |> tail(20)
```

Note that now we are no longer dealing with normally distributed errors, and so the modes are not necessarily the same as the means of the random effects distribution.

::: {.callout-tip collapse="true"}
## Alternative: Penalized Quasi-Likelihood (not examinable)

Penalized quasi-likelihood (PQL) is an alternative (and much faster) method that doesn't require numerical integration.
See Chapter 13 in @faraway_extending_2016 for more details.

```{r}
#| label: 10-glm-mixed-effect-chunk-8
glmm_pql_fit <- glmmPQL(
  resp ~ age * smoke,
  random = ~ 1 | id,
  family = binomial,
  data = ohio_tbl
)

summary(glmm_pql_fit)
```

```{r}
#| label: 10-glm-mixed-effect-chunk-9
as_tibble(ranef(glmm_pql_fit))
```
:::

## Wine Bitterness: @fahrmeir_multivariate_2001 Example 7.2 and 7.4

Now we demonstrate **ordinal regression with random effects**.
For this we consider the `ordinal::wine` dataset, described in Example 7.2 of @fahrmeir_multivariate_2001, which contains ratings of wine bitterness by different judges under different conditions.

As always, we should inspect the data help page

```{r}
#| label: 10-glm-mixed-effect-chunk-9.5
?wine
```

The covariates are - `response`: a continuous measure of bitterness - `rating`: an ordered factor rating of bitterness (1-5) - `temp`: temperature (cold/warm) - `contact`: contact with skin (yes/no) - `judge`: judge ID

We will regress the ordinal `rating` on the covariates `temp` and `contact`, while including a random intercept for `judge` to account for judge-specific differences in rating tendencies, in order to reproduce some of the results in Table 7.4 of @fahrmeir_multivariate_2001.

### Data Setup and Contrast Coding

Even though the book says that no contact is the reference level, the analysis appears to be done with contact as the reference level.

```{r}
#| label: 10-glm-mixed-effect-chunk-10
wine_tbl <- as_tibble(wine)
wine_tbl

wine_tbl <- wine_tbl |>
  mutate(
    temp = factor(temp, levels = c("warm", "cold")),
    contact = factor(contact, levels = c("yes", "no"))
  )
contrasts(wine_tbl$contact)
contrasts(wine_tbl$temp)
```

Reset contrasts to dummy coding:

```{r}
#| label: 10-glm-mixed-effect-chunk-11
options(contrasts = c("contr.treatment", "contr.poly"))

# Check default contrasts
contrasts(wine_tbl$contact)
contrasts(wine_tbl$temp)
```

### Fixed Effects Ordinal Model: `MASS::polr()`

To replicate the first column of Table 7.4, we fit using `MASS::polr()`:

```{r}
#| label: 10-glm-mixed-effect-chunk-13
wine_fixed_fit_polr <- polr(
  rating ~ temp + contact,
  data = wine_tbl,
  Hess = TRUE,
  method = "logistic"
)

summary(wine_fixed_fit_polr)
```

Our coefficient estimates for the non-threshold coefficients are relatively close to the book's (up to the signs as we expect, explained in [the first ordinal regression script](04-Example3-2BTR.qmd#model-fitting-proportional-odds-models)).

The book reports thresholds differently to how we do, but we can easily convert our results to the book's formulation (see the equation (7.5.1) on page 320 to see why this is the case).
Let's check:

```{r}
#| label: 10-glm-mixed-effect-chunk-15

thresholds <- wine_fixed_fit_polr$zeta
alphas <- c(thresholds[1], log(diff(thresholds)))
alphas
```

::: {.callout-tip collapse="true"}
## What does `c(thresholds[1], log(diff(thresholds)))` do?

This line re-parameterises the ordered thresholds into the book's unconstrained parameterisation (`alphas`):

-   `thresholds[1]`: keeps the first cutpoint unchanged.
-   `diff(thresholds)`: computes successive gaps\
    $(\theta_2-\theta_1,\theta_3-\theta_2,\ldots)$.
-   `log(diff(thresholds))`: log-transforms those positive gaps so they can be modelled on the real line.
-   `c(...)`: concatenates the first cutpoint and the logged gaps into one vector.

So if `thresholds = (theta_1, theta_2, theta_3, theta_4)`, then this creates $$
\alpha = \left(\theta_1,\log(\theta_2-\theta_1),\log(\theta_3-\theta_2),\log(\theta_4-\theta_3)\right).
$$

This guarantees ordered thresholds when converting back, because exponentiating the gap terms makes each gap strictly positive.
:::

These are also close but not exact.
For a final sanity check, we can compare our log-likelihood with the book's:

```{r}
#| label: 10-glm-mixed-effect-chunk-12
logLik(wine_fixed_fit_polr)
```

The log-likelihood is close to the reported log-likehood in Table 7.4, but not exact.
In fact, ours is marginally higher than the book's.
The reason for all the small mismatches is not clear.

::: {.callout-tip collapse="true"}
## Alternative Fit with `ordinal::clm()`

In order to make sure that the mismatches aren't due to the specific implementation of `polr()`, we can try a different function `ordinal::clm()`:

```{r}
#| label: 10-glm-mixed-effect-chunk-14
wine_fixed_fit_clm <- clm(rating ~ temp + contact, data = wine_tbl)

summary(wine_fixed_fit_clm)
coef(wine_fixed_fit_clm)
logLik(wine_fixed_fit_clm)

thresholds_clm <- wine_fixed_fit_clm$alpha
alphas_clm <- c(thresholds_clm[1], log(diff(thresholds_clm)))
alphas_clm
```

The values retrieved are almost the same as those obtained with `polr()`.
:::

### Random Effects Model: `ordinal::clmm()`

Now we fit a random effects model using `ordinal::clmm()`, which is designed for cumulative link mixed models (CLMMs).

```{r}
#| label: 10-glm-mixed-effect-chunk-16
wine_mixed_fit <- clmm(
  rating ~ temp + contact + (1 | judge),
  data = wine_tbl,
  nAGQ = 10
)

summary(wine_mixed_fit)
```

Again these results are close to, but not exactly the same as, those in Table 7.4.

We can compare threshold estimates again as above:

```{r}
thresholds_mixed <- wine_mixed_fit$alpha
alphas_mixed <- c(thresholds_mixed[1], log(diff(thresholds_mixed)))
alphas_mixed
```

We can extract fitted values and contrast structure from the fitted model.
Note that the fitted values are evaluated at the conditional modes of the random effects distribution (see `?clmm`).

```{r}
#| label: 10-glm-mixed-effect-chunk-17
# Fitted values for first few observations
wine_mixed_fit$fitted.values
```

As always, to get the estimated conditional modes of the random effects, we use the generic `ranef()` function.

```{r}
#| label: 10-glm-mixed-effect-chunk-18
ranef(wine_mixed_fit)
```

## Package Citations

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "geepack", "lme4", "MASS", "ordinal")
for (pkg in packages) {
  cit <- citation(pkg)

  # Check if multiple citations
  if (length(cit) > 1) {
    cat("- **", pkg, "**:\n", sep = "")
    for (i in seq_along(cit)) {
      cit_text <- format(cit[i], style = "text")
      cat("    - ", cit_text, "\n\n", sep = "")
    }
  } else {
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
  }
}
```
