---
title: "Survival Analysis"
subtitle: "Breast Cancer Data: Kaplan-Meier, Exponential and Weibull Models"
date: last-modified
editor:
  markdown:
    wrap: sentence
    canonical: true
---

**DENNIS: You may want to add more exposition to this code. I'm not exactly sure what happens in the lectures but when you are explaining it you are quite succinct (particularly in the sections on parametric models).**

## Data Loading and Setup

We load packages and breast cancer survival data, and inspect the stucture:

```{r}
#| label: 11-survival-chunk-1
library(conflicted)
library(dplyr)
library(ggplot2)
library(readr)
library(scales)
library(survival)

bcancer <- read_csv("data/survival.csv")

bcancer

tail(bcancer)
```

::: callout-note
**Data structure for survival analysis**:

- `time`: Time in days until event (death) or censoring (last follow-up)
- `status`: Censoring indicator (0 = censored, 1 = event observed)
- `group`: Grouping variable (here: staining status: Positive (P) or Negative (N)) this indicates whether metastasis was observed in the lymph nodes, which is a prognostic factor for breast cancer survival.

The `Surv()` function combines time and status into a survival object for modeling.
:::

## Diagnostics

### Kaplan-Meier Estimation

We fit the **Kaplan-Meier** (non-parametric) estimator for the survival function by group:

```{r}
#| label: 11-survival-chunk-2
cancer_km <- survfit(Surv(time, status) ~ group, data = bcancer)

cancer_km

summary(cancer_km)
```

Here the `survfit` object contains the estimated survival probabilities at each event time, along with confidence intervals and the number at risk.
So for `group=N` at time 23, the estimated survival probability is 0.923.

`n.risk` is the number of patients still at risk just before that time, and `n.event` is the number of events (deaths) that occur at that time. `n.risk` can decrease due to events or censoring, which is why it doesn't always decrease by the number of events.

The generic `summary()` method for `survfit` objects also has a `data.frame = TRUE` option which we can use and convert to a tibble for easier manipulation:

```{r}
km_summary_tbl <- summary(cancer_km, data.frame = TRUE) |> as_tibble()
str(km_summary_tbl)
```

We use `str()` to inspect the structure of the summary data frame, because our tibble has many columns, and doesn't display neatly.

### Plotting Setup

Build a tidy data frame from the Kaplan-Meier fit so we can reuse it across the plots below:

```{r}
#| label: 11-survival-chunk-3

km_plot_data <- km_summary_tbl |>
    mutate(
        group = factor(
            strata,
            levels = c("group=N", "group=P"),
            labels = c("Negative Stain", "Positive Stain")
        )
    ) |>
    select(group, time, surv, lower, upper) |>
    arrange(group, time)

km_plot_data
```

Since we'll be making multiple survival plots, we can define a common theme and colour palette.
We use colourblind-friendly colours [as suggested by Paul Tol](https://sronpersonalpages.nl/~pault/):

```{r}
#| label: 11-survival-chunk-3b

stain_colours <- c(
    "Negative Stain" = "#EE6677",
    "Positive Stain" = "#4477AA"
)

survival_theme <- theme_minimal(base_size = 14) +
    theme(
        legend.position = "top",
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank()
    )
```

::: {.callout-tip collapse="true"}
## Syntax walkthrough (shared plot settings):

- `stain_colours <- c(...)` creates a **named vector**: the names (`"Negative Stain"`, `"Positive Stain"`) must match the group labels in the data, and the values are the colours used in the legend and lines.
- `theme_minimal(base_size = 14)` sets a clean base style and readable default text size.
- `+ theme(...)` modifies specific theme elements on top of `theme_minimal()`.
- `legend.position = "top"` moves the legend above the plot, which is usually easier to scan for two-group comparisons.
- `legend.title = element_text(face = "bold")` and `plot.title = element_text(face = "bold")` make key text elements stand out.
- `panel.grid.minor = element_blank()` removes minor gridlines to reduce visual clutter.
:::

Plot Kaplan-Meier survival curves for the two groups:

```{r}
#| label: 11-survival-chunk-4
ggplot(km_plot_data, aes(x = time, y = surv, colour = group)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = group), alpha = 0.18, colour = NA) +
    geom_step(linewidth = 0.9) +
    scale_colour_manual(values = stain_colours, name = "Group") +
    scale_fill_manual(values = stain_colours, name = "Group") +
    scale_y_continuous(
        limits = c(0, 1),
        breaks = seq(0, 1, 0.2)
    ) +
    labs(
        title = "Kaplan-Meier survival curves",
        subtitle = "Step drops show the estimated survival probability over time;\nshaded bands show 95% confidence intervals",
        x = "Time",
        y = "Survival probability",
        colour = NULL
    ) +
    survival_theme
```

::: {.callout-tip collapse="true"}
## Plot syntax walkthrough (Kaplan-Meier)

Read this plot from top to bottom as a sequence of layers:

1. `ggplot(km_plot_data, aes(...))` picks the data and default mappings (`time` on x-axis, survival `surv` on y-axis, colour by `group`).
2. `geom_ribbon(aes(ymin = lower, ymax = upper, fill = group), ...)` draws the 95% confidence interval band between the lower and upper survival limits.
3. `geom_step(...)` draws Kaplan-Meier style step curves.
4. `scale_colour_manual(...)` applies the named palette so the step-line legend labels match the colours.
5. `scale_fill_manual(...)` uses the same palette for the confidence-band fill.
6. `scale_y_continuous(...)` constrains and labels the y-axis to survival-probability values between 0 and 1.
7. `labs(...)` adds title, subtitle, and axis labels.
8. `survival_theme` applies shared styling (legend placement, text emphasis, and cleaner gridlines).
:::

### Weibull Suitability Check

Before fitting parametric models, check if the Weibull distribution is suitable.
The **log-log transformation** of a Weibull survival curve is linear: $$\log(-\log(S(t))) = \alpha \log(\lambda t),$$ and so plotting $\log(-\log(S(t)))$ vs. $\log(t)$ should yield a straight line if the Weibull model is appropriate.

```{r}
#| label: 11-survival-chunk-5
weibull_plot_data <- km_plot_data |>
    dplyr::filter(time > 0, surv > 0) |>
    mutate(
        log_time = log(time),
        log_log_surv = log(-log(surv))
    )

ggplot(weibull_plot_data, aes(x = log_time, y = log_log_surv, colour = group)) +
    geom_point(size = 2.2, alpha = 0.8) +
    geom_line(linewidth = 0.9) +
    scale_colour_manual(values = stain_colours, name = "Group") +
    labs(
        title = "Weibull diagnostic plot",
        subtitle = "A roughly straight pattern suggests a Weibull model is reasonable",
        x = "log(Time)",
        y = "log(-log(Survival))",
        colour = NULL
    ) +
    survival_theme
```

::: {.callout-tip collapse="true"}
## Plot syntax walkthrough (Weibull diagnostic)

1. `ggplot(weibull_plot_data, aes(x = log_time, y = log_log_surv, colour = group))` sets the data source and maps transformed variables to the axes.
2. `geom_point(size = 2.2, alpha = 0.8)` adds plotted observations with size and transparency controls.
3. `geom_line(linewidth = 0.9)` connects points within each colour group.
4. `scale_colour_manual(values = stain_colours, name = "Group")` applies a named colour vector and legend title.
5. `labs(...)` sets the plot title, subtitle, and axis labels.
6. `survival_theme` applies the shared theme object.
:::

Here we have a roughly linear pattern for both groups, suggesting the Weibull model may be a reasonable parametric choice for modeling survival in this dataset.

### Proportional Hazards Assumption

Many survival models (e.g., Cox) assume **proportional hazards**: the hazard ratio between groups is constant over time.
We can check this by plotting the complementary log-log transformation of the survival curves: $\log(-\log(S(t)))$ vs. time, and looking for a parallel shift between groups.

```{r}
#| label: 11-survival-chunk-6
cloglog_plot_data <- km_plot_data |>
    dplyr::filter(time > 0, surv > 0) |>
    mutate(cloglog = log(-log(surv)))

ggplot(cloglog_plot_data, aes(x = time, y = cloglog, colour = group)) +
    geom_step(linewidth = 0.9) +
    scale_colour_manual(values = stain_colours, name = "Group") +
    labs(
        title = "Proportional hazards check via cloglog plot",
        subtitle = "A parallel shift supports the proportional hazards assumption",
        x = "Time",
        y = "log(-log(Survival))",
        colour = NULL
    ) +
    survival_theme
```

::: {.callout-tip collapse="true"}
## Plot syntax walkthrough (cloglog PH check)

1. `ggplot(cloglog_plot_data, aes(x = time, y = cloglog, colour = group))` maps time and the transformed `cloglog` response.
2. `geom_step(linewidth = 0.9)` draws the observed transformed values as step functions.
3. `scale_colour_manual(values = stain_colours, name = "Group")` uses the predefined palette and legend title.
4. `labs(...)` defines the title, subtitle, and axis labels.
5. `survival_theme` adds the common visual styling.
:::

Here is it plausible that the curves are roughly parallel, which supports the proportional hazards assumption.

## Exponential Regression

Now we fit parametric survival models to estimate the effect of group on survival time.
The first model we fit is the exponential survival model:

```{r}
#| label: 11-survival-chunk-7
cancer_exp <- survreg(
    Surv(time, status) ~ group,
    data = bcancer,
    dist = "exponential"
)

summary(cancer_exp)
```

::: {.callout-important collapse="true"}
## SELF TEST: What are the model assumptions for the exponential survival model?

- The time-to-event follows an exponential distribution, which implies a constant hazard rate over time.
- The effect of the group is multiplicative on the hazard scale (i.e., it shifts the rate parameter $\lambda$ by a factor of $e^{\beta}$).
- Censoring is non-informative (the reason for censoring is unrelated to the survival time).
:::

We can recover the p-value for the group effect using the Wald test.

**Wald Test**:

```{r}
#| label: 11-survival-chunk-8

exp_coefs <- coef(cancer_exp)
exp_coefs
exp_cov <- vcov(cancer_exp)
exp_cov

beta <- exp_coefs[2]
var_beta <- exp_cov[2, 2]

wald_stat <- beta / sqrt(var_beta)
2 * pnorm(-abs(wald_stat)) |> round(3)
```

We can also do a likelihood ratio test comparing the exponential model with group to a null model without group:

```{r}
#| label: 11-survival-chunk-9

exp_loglik <- cancer_exp$loglik
lr_stat <- -2 * (exp_loglik[1] - exp_loglik[2])
pchisq(q = lr_stat, df = 1, lower.tail = FALSE) |> round(3)
```

### Extracting Rate Parameters

For the exponential model, the rate parameter (hazard) is $$\lambda = e^{(\beta_0 + \beta_1 \cdot \text{group})}$$:

```{r}
#| label: 11-survival-chunk-10
lambda_n <- exp(-exp_coefs[1])
lambda_p <- exp(-sum(exp_coefs))

lambda_n
lambda_p
```

::: callout-note
Negative stain group: $\lambda_N =$ `r round(lambda_n, 4)` events/time\
Positive stain group: $\lambda_P =$ `r round(lambda_p, 4)` events/time

Higher $\lambda$ means shorter survival (higher hazard rate).
:::

### Verification: MLE via Method of Moments

For the exponential distribution, the MLE is:
$$\hat{\lambda} = \frac{\text{\# events}}{\text{total time at risk}}.$$
Therefore, we can verify our MLE estimates by calculating the event rates directly from the data:

```{r}
#| label: 11-survival-chunk-11
bcancer |>
    group_by(group) |>
    summarise(rate = sum(status) / sum(time), .groups = "drop")
```

These match our model-based estimates, confirming that the MLE for the exponential model is consistent with the method of moments calculation.

### Delta Method Confidence Interval

Recall there are two methods we use to construct confidence intervals. Here, we use the **delta method** (linearization).
$$\text{var}(g(X)) \approx (g'(X))^2 \cdot \text{var}(X)$$

```{r}
#| label: 11-survival-chunk-12

var_lambda <- (exp(-beta))^2 * var_beta

CI <- exp(-beta) + qnorm(0.975) * sqrt(var_lambda) * c(-1, 1)
CI |> round(3)
```

::: callout-note
The lower bound is positive (doesn't cross zero), which is good since $\lambda$ must be positive.
However, the delta method doesn't always respect parameter constraints; a safer approach is to transform:
:::

**Better approach**: Transform to enforce positivity:

```{r}
#| label: 11-survival-chunk-13
# Build CI on log scale, then exponentiate
exp(-(beta - qnorm(0.975) * sqrt(var_beta) * c(-1, 1))) |>
    round(3)
```

## Weibull Regression

The **Weibull model** generalizes the exponential by allowing a shape parameter $\alpha > 0$.
When $\alpha = 1$, it reduces to exponential; $\alpha > 1$ means hazard increases with time; $\alpha < 1$ means hazard decreases (realistic in diseases that primarily effect infants such as malaria).

We fit the Weibull survival model, and extract some of its parameters:

```{r}
#| label: 11-survival-chunk-14
cancer_wbl <- survreg(
    Surv(time, status) ~ group,
    data = bcancer,
    dist = "weibull"
)

summary(cancer_wbl)
wbl_coef <- coef(cancer_wbl)
wbl_scale <- cancer_wbl$scale
wbl_cov <- vcov(cancer_wbl)

wbl_coef
wbl_scale
wbl_cov
```

::: callout-caution
Note: `survreg` parameterizes the scale as $\sigma = 1/\alpha$, so the shape parameter is $\alpha = 1/\text{scale}$.
:::

### Rate and Shape Parameters

Extract and interpret the Weibull parameters:

```{r}
#| label: 11-survival-chunk-15
lambda_n <- exp(-wbl_coef[1])
lambda_p <- exp(-sum(wbl_coef))

alpha <- 1 / wbl_scale

s_n_100 <- exp(-(lambda_n * 100)^alpha)
s_p_100 <- exp(-(lambda_p * 100)^alpha)

lambda_n
lambda_p
alpha
s_n_100
s_p_100
```

::: {.callout-important collapse="true"}
## SELF TEST: What do these survival probabilities mean?

`s_n_100` is the estimated survival probability at time 100 for negative-stain patients.
`s_p_100` is the same for positive-stain patients.
:::

### Hazard Ratio

We can also estimate the hazard ratio (positive stain vs. negative stain):

```{r}
#| label: 11-survival-chunk-16
hazard_ratio <- (lambda_p / lambda_n)^alpha
hazard_ratio
```

::: callout-note
A hazard ratio \> 1 means positive-stain patients have higher hazard (die sooner).
\< 1 means better survival for positive-stain.
:::

### Testing the Shape Parameter

We can do a Wald test or likelihood ratio test to check if the shape parameter $\alpha$ is significantly different from 1, which would indicate that the Weibull model provides a better fit than the exponential model.

#### Wald Test

```{r}
#| label: 11-survival-chunk-17
wald_z <- log(wbl_scale) / sqrt(wbl_cov[3, 3])

# p-value for H0: log(scale) = 0, i.e., scale = 1, i.e., alpha = 1
2 * pnorm(-abs(wald_z))
```

Here we don't reject the null hypothesis that $\alpha = 1$, which suggests the exponential model may be sufficient.

#### Likelihood Ratio Test

We also check this with a likelihood ratio test comparing the Weibull model to the exponential model.

```{r}
#| label: 11-survival-chunk-18

wbl_loglik <- logLik(cancer_wbl)
wbl_loglik
lr_stat <- 2 * (wbl_loglik - exp_loglik[2])

pchisq(
    lr_stat,
    df = 1,
    lower.tail = FALSE
)
```

Again, we don't reject the null hypothesis that $\alpha = 1$.

## Package Citations

```{r}
#| label: package-citations
#| results: asis
#| code-fold: true

packages <- c("conflicted", "dplyr", "ggplot2", "readr", "scales", "survival")
for (pkg in packages) {
    cit <- citation(pkg)

    # Check if multiple citations
    if (length(cit) > 1) {
        cat("- **", pkg, "**:\n", sep = "")
        for (i in seq_along(cit)) {
            cit_text <- format(cit[i], style = "text")
            cat("    - ", cit_text, "\n\n", sep = "")
        }
    } else {
        cit_text <- format(cit, style = "text")
        cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
    }
}
```
