---
title: "Survival Analysis"
subtitle: "Breast Cancer Data: Kaplan-Meier, Exponential and Weibull Models"
date: last-modified
editor:
  markdown:
    wrap: sentence
    canonical: true
---

## Data Loading and Setup

We load packages and breast cancer survival data, and inspect the stucture:

```{r}
#| label: 11-survival-chunk-1
library(conflicted)
library(dplyr)
library(ggplot2)
library(readr)
library(scales)
library(survival)

bcancer <- read_csv("data/survival.csv")

bcancer

tail(bcancer)
```

::: callout-note
**Data structure for survival analysis**:

- `time`: Time until event (in some unit, here likely days or months)
- `status`: Censoring indicator (0 = censored, 1 = event observed)
- `group`: Grouping variable (here: staining status: Positive (P) or Negative (N)) this indicates whether metastasis was observed in the lymph nodes, which is a prognostic factor for breast cancer survival.

The `Surv()` function combines time and status into a survival object for modeling.
:::

## Diagnostics

### Kaplan-Meier Estimation

We fit the **Kaplan-Meier** (non-parametric) estimator for the survival function by group:

```{r}
#| label: 11-survival-chunk-2
cancer_km <- survfit(Surv(time, status) ~ group, data = bcancer)

cancer_km

summary(cancer_km)
```

Here the `survfit` object contains the estimated survival probabilities at each event time, along with confidence intervals and the number at risk.
So for `group=N` at time 23, the estimated survival probability is 0.923.

`n.risk` is the number of patients still at risk just before that time, and `n.event` is the number of events (deaths) that occur at that time. `n.risk` can decrease due to events or censoring, which is why it doesn't always decrease by the number of events.

The generic `summary()` method for `survfit` objects also has a `data.frame = TRUE` option which we can use and convert to a tibble for easier manipulation:

```{r}
km_summary_tbl <- summary(cancer_km, data.frame = TRUE) |> as_tibble()
str(km_summary_tbl)
```

We use `str()` to inspect the structure of the summary data frame, because our tibble has many columns, and doesn't display neatly.

### Plotting Setup

Build a tidy data frame from the Kaplan-Meier fit so we can reuse it across the plots below:

```{r}
#| label: 11-survival-chunk-3

km_plot_data <- km_summary_tbl |>
    mutate(
        group = factor(
            strata,
            levels = c("group=N", "group=P"),
            labels = c("Negative Stain", "Positive Stain")
        )
    ) |>
    select(group, time, surv, lower, upper) |>
    arrange(group, time)

km_plot_data
```

Since we'll be making multiple survival plots, we can define a common theme and colour palette.
We use colourblind-friendly colours [as suggested by Paul Tol](https://sronpersonalpages.nl/~pault/):

```{r}
#| label: 11-survival-chunk-3b

stain_colours <- c(
    "Negative Stain" = "#EE6677",
    "Positive Stain" = "#4477AA"
)

survival_theme <- theme_minimal(base_size = 14) +
    theme(
        legend.position = "top",
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank()
    )
```

::: {.callout-tip collapse="true"}
## Syntax walkthrough (shared plot settings):

- `stain_colours <- c(...)` creates a **named vector**: the names (`"Negative Stain"`, `"Positive Stain"`) must match the group labels in the data, and the values are the colours used in the legend and lines.
- `theme_minimal(base_size = 14)` sets a clean base style and readable default text size.
- `+ theme(...)` modifies specific theme elements on top of `theme_minimal()`.
- `legend.position = "top"` moves the legend above the plot, which is usually easier to scan for two-group comparisons.
- `legend.title = element_text(face = "bold")` and `plot.title = element_text(face = "bold")` make key text elements stand out.
- `panel.grid.minor = element_blank()` removes minor gridlines to reduce visual clutter.
:::

Plot Kaplan-Meier survival curves for the two groups:

```{r}
#| label: 11-survival-chunk-4
ggplot(km_plot_data, aes(x = time, y = surv, colour = group)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = group), alpha = 0.18, colour = NA) +
    geom_step(linewidth = 0.9) +
    scale_colour_manual(values = stain_colours, name = "Group") +
    scale_fill_manual(values = stain_colours, name = "Group") +
    scale_y_continuous(
        limits = c(0, 1),
        breaks = seq(0, 1, 0.2)
    ) +
    labs(
        title = "Kaplan-Meier survival curves",
        subtitle = "Step drops show the estimated survival probability over time;\nshaded bands show 95% confidence intervals",
        x = "Time",
        y = "Survival probability",
        colour = NULL
    ) +
    survival_theme
```

::: {.callout-tip collapse="true"}
## Plot syntax walkthrough (Kaplan-Meier)

Read this plot from top to bottom as a sequence of layers:

1. `ggplot(km_plot_data, aes(...))` picks the data and default mappings (`time` on x-axis, survival `surv` on y-axis, colour by `group`).
2. `geom_ribbon(aes(ymin = lower, ymax = upper, fill = group), ...)` draws the 95% confidence interval band between the lower and upper survival limits.
3. `geom_step(...)` draws Kaplan-Meier style step curves.
4. `scale_colour_manual(...)` applies the named palette so the step-line legend labels match the colours.
5. `scale_fill_manual(...)` uses the same palette for the confidence-band fill.
6. `scale_y_continuous(...)` constrains and labels the y-axis to survival-probability values between 0 and 1.
7. `labs(...)` adds title, subtitle, and axis labels.
8. `survival_theme` applies shared styling (legend placement, text emphasis, and cleaner gridlines).
:::

### Weibull Suitability Check

Before fitting parametric models, check if the Weibull distribution is suitable.
The **log-log transformation** of a Weibull survival curve is linear: $$\log(-\log(S(t))) = \alpha \log(\lambda t),$$ and so plotting $\log(-\log(S(t)))$ vs. $\log(t)$ should yield a straight line if the Weibull model is appropriate.

```{r}
#| label: 11-survival-chunk-5
weibull_plot_data <- km_plot_data |>
    dplyr::filter(time > 0, surv > 0) |>
    mutate(
        log_time = log(time),
        log_log_surv = log(-log(surv))
    )

ggplot(weibull_plot_data, aes(x = log_time, y = log_log_surv, colour = group)) +
    geom_point(size = 2.2, alpha = 0.8) +
    geom_line(linewidth = 0.9) +
    scale_colour_manual(values = stain_colours, name = "Group") +
    labs(
        title = "Weibull diagnostic plot",
        subtitle = "A roughly straight pattern suggests a Weibull model is reasonable",
        x = "log(Time)",
        y = "log(-log(Survival))",
        colour = NULL
    ) +
    survival_theme
```

::: {.callout-tip collapse="true"}
## Plot syntax walkthrough (Weibull diagnostic)

1. `ggplot(weibull_plot_data, aes(x = log_time, y = log_log_surv, colour = group))` sets the data source and maps transformed variables to the axes.
2. `geom_point(size = 2.2, alpha = 0.8)` adds plotted observations with size and transparency controls.
3. `geom_line(linewidth = 0.9)` connects points within each colour group.
4. `scale_colour_manual(values = stain_colours, name = "Group")` applies a named colour vector and legend title.
5. `labs(...)` sets the plot title, subtitle, and axis labels.
6. `survival_theme` applies the shared theme object.
:::

Here we have a roughly linear pattern for both groups, suggesting the Weibull model may be a reasonable parametric choice for modeling survival in this dataset.

### Proportional Hazards Assumption

Many survival models (e.g., Cox) assume **proportional hazards**: the hazard ratio between groups is constant over time.
We can check this by plotting the complementary log-log transformation of the survival curves: $\log(-\log(S(t)))$ vs. time, and looking for a parallel shift between groups.

```{r}
#| label: 11-survival-chunk-6
cloglog_plot_data <- km_plot_data |>
    dplyr::filter(time > 0, surv > 0) |>
    mutate(cloglog = log(-log(surv)))

ggplot(cloglog_plot_data, aes(x = time, y = cloglog, colour = group)) +
    geom_step(linewidth = 0.9) +
    scale_colour_manual(values = stain_colours, name = "Group") +
    labs(
        title = "Proportional hazards check via cloglog plot",
        subtitle = "A parallel shift supports the proportional hazards assumption",
        x = "Time",
        y = "log(-log(Survival))",
        colour = NULL
    ) +
    survival_theme
```

::: {.callout-tip collapse="true"}
## Plot syntax walkthrough (cloglog PH check)

1. `ggplot(cloglog_plot_data, aes(x = time, y = cloglog, colour = group))` maps time and the transformed `cloglog` response.
2. `geom_step(linewidth = 0.9)` draws the observed transformed values as step functions.
3. `scale_colour_manual(values = stain_colours, name = "Group")` uses the predefined palette and legend title.
4. `labs(...)` defines the title, subtitle, and axis labels.
5. `survival_theme` adds the common visual styling.
:::

Here is it plausible that the curves are roughly parallel, which supports the proportional hazards assumption.

## Exponential Regression

The **exponential model** is a parametric survival model with constant hazard rate.
It assumes failures follow a Poisson process.

Fit exponential survival model:

```{r}
#| label: 11-survival-chunk-7
cancer_exp <- survreg(
    Surv(time, status) ~ group,
    data = bcancer,
    dist = "exponential"
)

summary(cancer_exp)
```

::: callout-note
**Exponential model assumptions**: Constant hazard rate (hazard doesn't change over time).
This is restrictive and often violated in practice, but is a useful baseline and computationally simple.
:::

### Testing the Group Effect

**Wald Test**:

```{r}
#| label: 11-survival-chunk-8
wald_stat <- cancer_exp$coeff[2] / sqrt(cancer_exp$var[2, 2])
2 * pnorm(-abs(wald_stat))
```

This gives a p-value for whether group significantly affects survival time.

**Likelihood Ratio Test**:

```{r}
#| label: 11-survival-chunk-9
lr_stat <- -2 * (cancer_exp$loglik[1] - cancer_exp$loglik[2])
pchisq(q = lr_stat, df = 1, lower.tail = FALSE)
```

An alternative p-value based on the likelihood ratio statistic.

::: {.callout-important collapse="true"}
## SELF TEST: When would Wald and LR tests disagree?

Both test the same hypothesis but use different asymptotics.
For moderate sample sizes, they usually agree.
With small samples or if the coefficient is near a boundary (e.g., near no effect), they can diverge.
LR tests are generally more robust in finite samples.
:::

### Extracting Rate Parameters

For the exponential model, the rate parameter (hazard) is $\lambda = e^{(intercept + \beta \cdot \text{group})}$:

```{r}
#| label: 11-survival-chunk-10
lambda_n <- exp(-cancer_exp$coef[1])
lambda_p <- exp(-sum(cancer_exp$coef))

lambda_n
lambda_p
```

::: callout-note
Negative stain group: $\lambda_N =$ `r lambda_n` events/time\
Positive stain group: $\lambda_P =$ `r lambda_p` events/time

Higher $\lambda$ means shorter survival (higher hazard rate).
:::

### Verification: MLE via Method of Moments

For the exponential distribution, the MLE is:
$$\hat{\lambda} = \frac{\text{\# events}}{\text{total time at risk}}$$

```{r}
#| label: 11-survival-chunk-11
bcancer |>
    group_by(group) |>
    summarise(rate = sum(status) / sum(time), .groups = "drop")
```

::: {.callout-important collapse="true"}
## SELF TEST: Do these match the model rates above?

These hand-computed values should match (to numerical precision) the rates extracted above.
This verifies that `survreg` is fitting via maximum likelihood and our interpretation of the rates is correct.
:::

### Delta Method Confidence Interval

Construct a 95% CI for $\lambda_P$ using the **delta method** (linearization):

```{r}
#| label: 11-survival-chunk-12
beta <- cancer_exp$coeff[2]
var_beta <- (cancer_exp$var)[2, 2]

# Delta method: var(e^X) ≈ (e^X)^2 * var(X)
var_lambda <- (exp(-beta))^2 * var_beta

# CI
exp(-beta) + qnorm(0.975) * sqrt(var_lambda)
exp(-beta) - qnorm(0.975) * sqrt(var_lambda)
```

::: callout-note
The lower bound is positive (doesn't cross zero), which is good since $\lambda$ must be positive.
However, the delta method doesn't always respect parameter constraints; a safer approach is to transform:
:::

**Better approach**: Transform to enforce positivity:

```{r}
#| label: 11-survival-chunk-13
# Build CI on log scale, then exponentiate
exp(-(beta - qnorm(0.975) * sqrt(var_beta)))
exp(-(beta + qnorm(0.975) * sqrt(var_beta)))
```

::: callout-note
This guarantees both bounds are positive because exponentiation preserves the ordering and always produces positive values.
:::

## Weibull Regression

The **Weibull model** generalizes the exponential by allowing a shape parameter $\alpha > 0$.
When $\alpha = 1$, it reduces to exponential; $\alpha > 1$ means hazard increases with time (wear-out); $\alpha < 1$ means hazard decreases (infant mortality).

Fit Weibull survival model:

```{r}
#| label: 11-survival-chunk-14
cancer_wbl <- survreg(
    Surv(time, status) ~ group,
    data = bcancer,
    dist = "weibull"
)

summary(cancer_wbl)
cancer_wbl$coef
cancer_wbl$scale
cancer_wbl$var
```

::: callout-note
Note: `survreg` parameterizes the scale as $\sigma = 1/\alpha$, so the shape parameter is $\alpha = 1/\text{scale}$.
A scale close to 1 means $\alpha \approx 1$ (nearly exponential); scale \>\> 1 means $\alpha << 1$ (rapidly decreasing hazard).
:::

### Rate and Shape Parameters

Extract and interpret the Weibull parameters:

```{r}
#| label: 11-survival-chunk-15
lambda_n <- exp(-cancer_wbl$coef[1])
lambda_p <- exp(-sum(cancer_wbl$coef))

alpha <- 1 / cancer_wbl$scale

s_n_100 <- exp(-(lambda_n * 100)^alpha)
s_p_100 <- exp(-(lambda_p * 100)^alpha)

lambda_n
lambda_p
alpha
s_n_100
s_p_100
```

::: {.callout-important collapse="true"}
## SELF TEST: What do these survival probabilities mean?

`s_n_100` is the estimated survival probability at time 100 for negative-stain patients.
`s_p_100` is the same for positive-stain patients.
Compare these to the Kaplan-Meier estimates for similar times to see how well the parametric model fits the data.
:::

### Hazard Ratio

Compute the hazard ratio (positive stain vs. negative stain):

```{r}
#| label: 11-survival-chunk-16
hazard_ratio <- (lambda_p / lambda_n)^alpha
hazard_ratio
```

::: callout-note
A hazard ratio \> 1 means positive-stain patients have higher hazard (die sooner).
\< 1 means better survival for positive-stain.
:::

### Testing the Shape Parameter

**Wald test** for whether shape = 1 (exponential):

```{r}
#| label: 11-survival-chunk-17
wald_z <- log(cancer_wbl$scale) / sqrt(cancer_wbl$var[3, 3])

# p-value for H0: log(scale) = 0, i.e., scale = 1, i.e., alpha = 1
2 * pnorm(-abs(wald_z))
```

**Likelihood ratio test** comparing Weibull to exponential:

```{r}
#| label: 11-survival-chunk-18
lr_stat <- 2 * (cancer_wbl$loglik[2] - cancer_exp$loglik[2])

qchisq(.95, df = 1)

pchisq(
    lr_stat,
    df = 1,
    lower.tail = FALSE
)
```

::: {.callout-important collapse="true"}
## SELF TEST: Does the Weibull model fit better than exponential?

Compare the LR test p-value: if p \< 0.05, the shape parameter is significantly different from 1, meaning Weibull provides better fit than exponential.
Also compare AIC/BIC or look at the scaled deviances to assess relative model quality.
:::

## Key Learning Points

1.  **Survival data structure**: Time-to-event ($t$) and censoring indicator (0/1).
    Censoring indicates incomplete observation—we know $T > t$ but not the exact value.

2.  **Non-parametric vs. parametric**:

    -   **Kaplan-Meier**: Non-parametric, makes no distributional assumption, handles censoring via risk sets
    -   **Parametric models**: Assume specific distribution (exponential, Weibull), more efficient if assumption holds

3.  **Diagnostic plots**:

    -   **Log-log plot**: Linearity suggests Weibull fit is appropriate
    -   **cloglog plot**: Parallelism checks proportional hazards assumption

4.  **Model parameters**:

    -   **Exponential**: Single rate parameter $\lambda$ (constant hazard)
    -   **Weibull**: Rate $\lambda$ and shape $\alpha$ (hazard can increase/decrease over time)

5.  **Interpretation**:

    -   Hazard ratio: relative risk between groups
    -   Survival probability at time $t$: probability of not having event by time $t$
    -   Rates are on the inverse scale: higher rate = shorter survival time

6.  **Inference**:

    -   **Wald tests**: Test individual parameters (less robust in finite samples)
    -   **LR tests**: Compare nested models (more robust)
    -   Always verify with bootstrapping or sensitivity analyses when possible

