---
title: "Likelihood for Grouped and Ungrouped Multinomial Data"
subtitle: "Differences Between Packages"
date: today
author:
  - name: Dennis Leung
---

## Overview

Let
$$
\mathbf{y}_i = (y_{i1}, \dots, y_{ik}), \qquad i = 1, \dots, n
$$
be i.i.d random $k$-vectors distributed as
$$
\mathbf{y}_i \sim \operatorname{Multinom}(1, \boldsymbol{\pi}).
$$
That is, each $\mathbf{y}_i$ is a multinomial distributed random vector of 1 trial
with class probability $k$-vector
$$
\boldsymbol{\pi} = ({\pi}_1, \dots, {\pi}_k).
$$
For any given $\mathbf{y}_i$, only one of $y_{i1}, \ldots, y_{ik}$ can equal 1.

There are two ways to view the likelihood of multinomial data:

1. Ungrouped View:
If we compute the likelihood of the data as is, we get:
$$
L_{\text{u}}(\boldsymbol{\pi}) = \prod_{i=1}^{n} {\pi}_1^{y_{i1}} \cdots {\pi}_k^{y_{ik}}
$$
with log-likelihood:
$$
\ell_{\text{u}}(\boldsymbol{\pi}) = \sum_{j=1}^{k} \log({\pi}_j) \sum_{i=1}^{n} y_{ij}
$$ {#eq-ungrouped}

2. Grouped View
Alternatively, we can group the $\mathbf{y}_i$'s as:
$$
\mathbf{x} = \sum_{i=1}^{n} \mathbf{y}_i
$$
so
$$
\mathbf{x} = (x_1, \ldots, x_k)
= \biggl(\sum_i y_{i1}, \ldots, \sum_i y_{ik}\biggr)
\sim \operatorname{Multinom}(n, \boldsymbol{\pi}),
$$
and treat $\mathbf{x}$ as our raw data. In this view, the likelihood is:
$$
L_{\text{g}}(\boldsymbol{\pi}) = \frac{n!}{x_1! \cdots x_k!} {\pi}_1^{x_1} \cdots {\pi}_k^{x_k}
$$
with log-likelihood:
$$
\ell_{\text{g}}(\boldsymbol{\pi})
= {
    \log \biggl(\frac{n!}{x_1! \cdots x_k!} \biggr)
    + \sum_{j=1}^{k} \log({\pi}_j)\sum_{i=1}^{n} y_{ij}
}
$$ {#eq-grouped}

Comparing log-likelihood equations (-@eq-ungrouped) and (-@eq-grouped),
we see that the grouped and ungrouped log-likelihoods differ by:

$$
\log\biggl(\frac{n!}{x_1! \cdots x_k!}\biggr)
$$

This term adjusts for the different permutations
that can give rise to the same grouped multinomial data.

## Example: C-Section Dataset

```{r}
#| code-fold: true
#| message: false
library(conflicted)
library(readr)
library(VGAM)
library(nnet)
```

Consider the original wide format of the C-section dataset:

```{r}
caes_dat <- read_csv("data/Example3-1Caes.csv")
caes_dat
```

### Using nnet::multinom()

When we fit a multinomial model using `nnet::multinom()`,
treating the data as ungrouped:

```{r}
#| message: false
caes_nnet <- nnet::multinom(
    cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = caes_dat
)
nnet_logLik <- caes_nnet$deviance / (-2)
nnet_logLik

```

The reported log-likelihood is `r round(nnet_logLik, 3)`.

### Using VGAM::vglm()

When we fit using `VGAM::vglm()` with the data in grouped format:

```{r}
Caes_vgam <- VGAM::vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = caes_dat,
    family = VGAM::multinomial(refLevel = "noInf")
)
vgam_logLik <- logLik(Caes_vgam)
vgam_logLik
```

The reported log-likelihood is now a different number!

### Explanation

To understand the difference,
we compute the sum of the "permutation factors"
$$
\log\biggl(\frac{n!}{x_1! \cdots x_k!}\biggr)
$$
in (-@eq-grouped) for the seven covariate groups:

```{r}
perm_fact_sum <- sum(log(
    factorial(caes_dat$size) /
        (factorial(caes_dat$noInf) *
            factorial(caes_dat$Inf1) *
            factorial(caes_dat$Inf2))
))
perm_fact_sum

```

This is precisely the difference between the two log-likelihoods:

```{r}
vgam_logLik - nnet_logLik
```

## Conclusion

For the original "wide format data":

- `nnet::multinom()` treats it as **ungrouped data**
- `VGAM::vglm()` treats it as **grouped data**

**Moral:** One must be very careful
when interpreting numbers reported by different packages.
The same data can produce different values for the what seems to be the same quantity.
