---
title: "Likelihood for Grouped and Ungrouped Multinomial Data"
subtitle: "Differences Between Packages"
date: today
author:
  - name: Dennis Leung
---

## Overview

Let $\mathbf{y}_i = (y_{i1}, \ldots, y_{ik})$, $i = 1, \ldots, n$ be i.i.d random $k$-vectors distributed as $\text{Multinom}(1, \boldsymbol{\pi})$. That is, each $\mathbf{y}_i$ is a multinomial distributed random vector of 1 trial with class probability $k$-vector $\boldsymbol{\pi} = (\pi_1, \ldots, \pi_k)$. For any given $\mathbf{y}_i$, only one of $y_{i1}, \ldots, y_{ik}$ can equal 1.

There are two ways to view the likelihood of multinomial data:

## Ungrouped View

If we compute the likelihood of the data as is, we get:

$$
L(\boldsymbol{\pi}) = \prod_{i=1}^{n} \pi_1^{y_{i1}} \cdots \pi_k^{y_{ik}}
$$

with log-likelihood:

$$
\ell(\boldsymbol{\pi}) = \sum_{j=1}^{k} \log(\pi_j) \sum_{i=1}^{n} y_{ij}
$$ {#eq-ungrouped}

## Grouped View

Alternatively, we can group the $\mathbf{y}_i$'s as:

$$
\mathbf{x} = \sum_{i=1}^{n} \mathbf{y}_i
$$

so $\mathbf{x} = (x_1, \ldots, x_k) = \left(\sum_i y_{i1}, \ldots, \sum_i y_{ik}\right) \sim \text{Multinom}(n, \boldsymbol{\pi})$, and treat $\mathbf{x}$ as our raw data. In this view, the likelihood is:

$$
L(\boldsymbol{\pi}) = \frac{n!}{x_1! \cdots x_k!} \pi_1^{x_1} \cdots \pi_k^{x_k}
$$

with log-likelihood:

$$
\ell(\boldsymbol{\pi}) = \log\left(\frac{n!}{x_1! \cdots x_k!}\right) + \sum_{j=1}^{k} \log(\pi_j) \sum_{i=1}^{n} y_{ij}
$$ {#eq-grouped}

## Comparison

Comparing equations @eq-ungrouped and @eq-grouped, we see that the grouped and ungrouped log-likelihoods differ by:

$$
\log\left(\frac{n!}{x_1! \cdots x_k!}\right)
$$

This term adjusts for the different permutations that can give rise to the same grouped multinomial data.

## Practical Example: C-Section Dataset

Consider the original wide format of the C-section dataset:

```{r}
Caes.dat <- read.csv("data/Example3-1Caes.csv")
Caes.dat
```

### Using nnet::multinom()

When we fit a multinomial model using `nnet::multinom()`, treating the data as ungrouped:

```{r}
Caes_nnet <- nnet::multinom(as.matrix(Caes.dat[, 2:4]) ~ NoPlan + Antib + RiskF,
    data = Caes.dat
)
Caes_nnet$deviance / (-2)
```

The reported log-likelihood is $-160.9371$.

### Using VGAM::vglm()

When we fit using `VGAM::vglm()` with the data in grouped format:

```{r}
Caes_vgam <- VGAM::vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = Caes.dat,
    family = VGAM::multinomial(refLevel = "noInf")
)
logLik(Caes_vgam)
```

The reported log-likelihood is now $-20.88715$, a different number!

### Explanation

To understand the difference, we compute the sum of the multinomial coefficients for the seven covariate groups:

```{r}
sum(log(factorial(Caes.dat$size) /
    (factorial(Caes.dat$noInf) * factorial(Caes.dat$Inf1) * factorial(Caes.dat$Inf2))))
```

This equals $140.05$, which is precisely the difference between the two log-likelihoods:

```{r}
-20.88715 - (-160.9371)
```

## Conclusion

For the original "wide format data":

- `nnet::multinom()` treats it as **ungrouped data**
- `VGAM::vglm()` treats it as **grouped data**

**Moral:** One must be very careful when interpreting numbers reported by different packages. The same data can produce different log-likelihood values depending on whether the package treats it as grouped or ungrouped multinomial data.
