---
title: "ANOVA and Deviance Analysis"
subtitle: "Binomial GLM with Logit and Probit Links"
date: today
---

## Overview

The [Fahrmeir & Tutz book is available to download for free from the University of Melbourne Library](https://go.openathens.net/redirector/unimelb.edu.au?url=https%3A%2F%2Flink.springer.com%2F10.1007%2F978-1-4757-3454-6)

This script reproduces **Examples 2.1, 2.4** from Fahrmeir & Tutz (pages 30 and 51).
We consider the Caesarian birth study with 251 births.
We consider the binary response variable $y$,
which is 1 for infection, and 0 for no infection.

There are 3 categorical variables:

1. `NOPLAN`: whether or not the Caesarian was planned or not
2. `FACTOR`: whether there are risk factors present or not
3. `ANTIB`: whether antibiotics were given to prevent infection (rather than treat infection)

We use fit the logistic regression model
$$
\ln\biggl(\frac{\Pr(\text{infection})}{\Pr(\text{no infection})}\biggr)
= \beta_{0} + \beta_{1}\texttt{NOPLAN}
+ \beta_{2}\texttt{FACTOR} +\beta_{3}\texttt{ANTIB}
$$
to the data.

We want to replicate the following table from page 30:

| Covariate | Effect    |
| :-------- | --------: |
| 1         | -1.89     |
| `NOPLAN`  | 1.07      |
| `FACTOR`  | 2.03      |
| `ANTIB`   | -3.25     |

## Data Setup

We start by loading the Caesar dataset from the Fahrmeir package.
If you don't already have the packages installed, run
```{r}
#| eval: false
install.packages(c("conflicted", "dplyr", "Fahrmeir"))
```

The `conflicted` library forces you to choose a which library
you want to call a function from if it is in two or more loaded libraries.
This is very important, particularly if you are looking at legacy code (old code)
from someone else who may have had particular functions loaded.

```{r}
library(conflicted) # avoid calling functions in multiple libraries
library(dplyr) # manipulate dataframes
library(forcats) # deal with categorical data (factors)
library(Fahrmeir) # load datasets from the book
```

We load the Caeser dataset in (from the Fahrmeir package),
and inspect the top few rows using `head`.

```{r}
data(caesar)
head(caesar)
```

You can inspect information about the data set
by running the following command in RStudio

```{r}
#| eval: false
help(caeser)
```

## Data Transformation

The raw `caesar` dataset has one row per birth (251 rows total).
For binomial regression, we need to aggregate the data:
one row per unique combination of covariates (`noplan`, `factor`, `antib`),
with columns for total cases and infection counts in each group.

First, convert to a tibble for clearer printing, and easier data manipulation.
Note that tibble print the data types of each column.

```{r}
caesar_tbl <- as_tibble(caesar)
caesar_tbl
```

### Grouping and Summarizing

Now we use a **pipe** (`|>`) to chain operations together.
Think of the pipe as saying "and then...":

- Take `caesar_tbl` **and then**
- Group rows by covariate combinations **and then**
- Count totals and infections per group

```{r}
caesar_tbl <- caesar_tbl |>
    group_by(noplan, factor, antib) |>
    summarize(
        size = sum(w),
        Infect_count = sum(yl * w),
        .groups = "drop"
    )
caesar_tbl
```

Setting `.groups = "drop"` removes the grouping after `summarize()`,
giving us a normal tibble, and avoiding confusing behaviour in later operations.

The third row has `size` zero, and so we remove it
(and any other empty variable combinations)

```{r}
caesar_tbl <- caesar_tbl |>
    dplyr::filter(size != 0)

caesar_tbl
```

Unfold the following code to see the equivalent unpiped version.

```{r}
#| eval: false
#| code-fold: true
# Step 1: Group by covariates
caesar_tbl <- group_by(caesar_tbl, noplan, factor, antib)

# Step 2: Compute totals per group
caesar_tbl <- summarize(
    caesar_tbl,
    size = sum(w),
    Infect_count = sum(yl * w),
    .groups = "drop"
)

# Step 3: Remove empty cells
caesar_tbl <- dplyr::filter(caesar_tbl, size != 0)
```

## Understanding Factor Coding

It is good practice to understand the data structure we have.
From the above tibble we can see that we have 7 observations of 5 variables.
The benefit of tibbles is that we can see what class each variable has.
`noplan`, `factor`, and `antib` have the class factor.
`size` and `Infect_count` have the class double (numeric).

Look at a description of the structure of caesar and the coding scheme of each
variable of type "factor" using the function `contrasts()`.

```{r}
contrasts(caesar_tbl$noplan)
contrasts(caesar_tbl$factor)
contrasts(caesar_tbl$antib)
```

We can see that these are coded in the exact opposite way that F & T describes,
and so we need to redefine the reference levels of these variables.
Set the reference level for each factor. The reference level (first level) becomes
the baseline category in regression coefficients.

```{r}
caesar_tbl_fct <- caesar_tbl |>
    mutate(
        noplan = fct_relevel(noplan, "planned"),
        factor = fct_relevel(factor, "without"),
        antib = fct_relevel(antib, "without")
    )

contrasts(caesar_tbl$noplan)
contrasts(caesar_tbl$factor)
contrasts(caesar_tbl$antib)
```

## Main Effects Model

Now that our dataset looks good, we are ready to go ahead with the regression.
`Infect_count / size` as our $y$ gives us the proportion of women
with an infection per group of covariates.
When we use grouped data, the weight should always be the same
as the number of counts of that combination of covariates, hence `weights = size`.

Fit the regression with the logit link (default).
The unscaled deviance gives the same scaled deviance for binomial and poisson data.
(Note:
[The documentation of `glm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html)
about what it is reporting for the deviance may be misleading,
because it ignores the fact that deviance only becomes
twice the log-likelihood difference when rescaled by the dispersion.)

```{r}
caesar_glm_fct <- glm(
    Infect_count / size ~ antib + factor + noplan,
    family = binomial,
    data = caesar_tbl,
    weights = size
)

caesar_glm_fct
```

These results line up with the ones in the book (yay).

Alternatively, we can assign integer values to the factors.
This makes the dataframe look exactly like it matches the books notation exactly.

```{r}
caesar_tbl_int <- caesar_tbl |>
    mutate(
        noplan = as.integer(noplan == "not"),
        factor = as.integer(factor == "risk factors"),
        antib = as.integer(antib == "antibiotics")
    )

caesar_tbl_int
caesar_tbl
```

We can see that we get the same results for the integer version

```{r}
caesar_int_glm <- glm(
    Infect_count / size ~ antib + factor + noplan,
    family = binomial,
    data = caesar_tbl_int,
    weights = size
)

caesar_int_glm
```

We can see the results more fully by using `summary()`

```{r}
summary(caesar_int_glm)
```

It is important to understand how `glm()` reports the summary.
We can see that the reported deviance is `r round(deviance(caesar_int_glm), 3)`.
This reported deviance is what we call the **unscaled deviance**.
In the binomial and Poisson cases,
the scaled deviance will be the same as the unscaled deviance.

::: {.callout-note}
The documentation of `glm()` about what it is reporting for the deviance
may be misleading, because it ignores the fact that deviance only becomes
twice the log-likelihood difference when rescaled by the dispersion.
:::

We can now consider the significance of each of the variables

```{r}
anova(caesar_int_glm, test = "Chisq")
```

This table closely matches the one in the notes.
All three effects are highly significant.

```{r}
anova(caesar_int_glm, test = "F")
```

::: {.callout-tip collapse="true"}
## Why is an F test inappropriate here? Expand for answer

For binomial GLMs, the natural test for nested models
is based on differences in deviance,
which asymptotically follows a chi-square distribution.
The F test is appropriate for Gaussian linear models with unknown dispersion
(and certain quasi-likelihood cases), but not for binomial GLMs with fixed dispersion.
:::

## Model Diagnostics

We can extract residuals of different types; here we collect them
into a single tibble and round to three decimals for readability.

```{r}
tibble(
    deviance = resid(caesar_int_glm, type = "deviance"),
    pearson = resid(caesar_int_glm, type = "pearson"),
    working = resid(caesar_int_glm, type = "working"),
    response = resid(caesar_int_glm, type = "response")
)
```

As a sanity check, if we sum the squares of the deviance residuals,
we should recover the deviance.

Perform Pearson and deviance tests for model goodness-of-fit.
Compare fitted and realized proportions as on the top of p.52.

```{r}
sum(resid(caesar_int_glm, type = "deviance")^2)
deviance(caesar_int_glm)
```

We can also test the goodness of fit using the deviance, and the Pearson statistic.
Both of these converge to the same asymptotic chi-squared distribution.
If we assume that the number of observations in each group is sufficiently large,
the degrees of freedom in the chi-squared distribution is
the difference between the number of
parameters in the saturated model (7, make sure you can explain why),
and the degrees of freedom in the glm model (4).

Both the deviance test
```{r}
1 - pchisq(caesar_int_glm$deviance, df = 3)
```

and the Pearson test
```{r}
1 - pchisq(sum(resid(caesar_int_glm, type = "pearson")^2), df = 3)
```

return values under 0.05,
and so under both tests (assuming sufficient asymptotics apply),
we reject null hypothesis that the data can be explained by these models,
with significant level of 0.05.

We can investigate why this might be by comparing the observed
and fitted proportions side by side:

```{r}
tibble(
    observed = round(caesar_tbl_int$Infect_count / caesar_tbl_int$size, 2),
    fitted = round(caesar_int_glm$fitted.values, 2)
)
```

We can see that this lines up with what we have in the book (Example 2.4),
with a significant discrepancy in the 3rd row
(`unplanned, no risk factors, without antibiotics`).

## Alternative Link Function: Probit

We can also fit with a probit link as was done on p. 31 (towards the end of
Example 2.1). The coefficient estimates should be the same as those reported by the
book.

```{r}
caesar_glm_probit <- glm(
    Infect_count / size ~ antib + factor + noplan,
    data = caesar_tbl_int,
    weights = size,
    family = binomial(link = "probit")
)
summary(caesar_glm_probit)$coeff
```

## Interaction Effects

We now follow **Example 2.4** and add in interaction effects.

### Adding `noplan:antib` Interaction

We test the interaction between `noplan` (planned/emergency surgery) and `antib`
(antibiotic use).
We are aiming to replicate the results in book

::: {.callout-tip collapse="true"}
## What do we expect to happen to the deviance? Expand for answer.

The deviance must decrease, since it is a larger, nested model.
:::

```{r}
caesar_glm_noplan_antib <- glm(
    Infect_count / size ~ factor + noplan * antib,
    data = caesar_tbl_int,
    weights = size,
    family = binomial
)

anova(caesar_glm_noplan_antib)
```

The change in deviance agrees with the numbers in **Example 2.4**.

To test the significance of this interaction effect,
we use the limiting chi-squared distribution with degree of freedom one.

::: {.callout-tip}
Make sure you know why the degree of freedom is one.
:::

We extract the deviance for the interaction term and compute the $p$-value manually:

```{r}
anova_result <- anova(caesar_glm_noplan_antib)
interaction_deviance <- tail(anova_result$Deviance, 1)
interaction_deviance

1 - pchisq(interaction_deviance, df = 1)
```

We do not reject the null hypothesis that the interaction effect is zero.

### Adding `noplan:factor` Interaction

This is not the only possible interaction effect.
We now try the interaction effect `noplan:factor`.

```{r}
caesar_glm_noplan_factor <- glm(
    Infect_count / size ~ noplan + factor + antib + noplan:factor,
    data = caesar_tbl_int,
    weights = size,
    family = binomial
)

# eqiuvalently, you can do
# caesar_glm_noplan_factor <- glm(Infect_count/size ~  antib + noplan*factor,
#                    data=caesar_tbl , weight=size, family=binomial)

summary(caesar_glm_noplan_factor)
```

We see some enormous estimates and standard errors for some coefficients.
The standard errors are the diagonal of the inverse of the Fisher information matrix.
Therefore the values of the Fisher information are very small.
Therefore, when we plot the log-likelihood it will have a very small second derivative,
and will be very flat in these dimensions.

This means that the estimates are likely to be very unreliable.
Read the book in detail to understand why, but in summary,
there are two rows with no infection counts,
which means that certain conditions are not satisfied.
See the section **Uniqueness and existence of MLE's** for more information.

:::{.callout-warning collapse="true"}
## Book typo

The book says
*'The reason is that data are too sparse with responses for the cell
`NOPLAN = 0`, `FACTOR = 0` and `ANTIB = 1` all in the "yes" category,
and for the cell `NOPLAN = 1`, `FACTOR= 0` and `ANTIB = 0`
all in the "no" category.'*
however, there is a typo. The first "yes" should instead say "no".
:::

If we ignore the problem for the time being,
we can continue by testing for goodness-of-fit.
Compare fitted and realized proportions as on p.53.

```{r}
1 - pchisq(sum(resid(caesar_glm_noplan_factor, type = "pearson")^2), df = 2)
1 - pchisq(sum(resid(caesar_glm_noplan_factor, type = "deviance")^2), df = 2)
```

Both statistics give a large $p$-value, and so we don't reject the interaction.

```{r}
tibble(
    observed = round(caesar_tbl_int$Infect_count / caesar_tbl_int$size, 2),
    fitted = round(caesar_glm_noplan_factor$fitted.values, 2)
)
```

We can see that the observed and fitted values are very similar.

## Handling Separation Issues

The book recommends that to make the analysis more robust,
we should add fictitious data to the two zero `Infect_count` cells
to avoid existence of MLE issue.
We do this by adding a 'count' of 0.5 infections and 1 total case to problematic row.

```{r}
caesar_tbl_int
caesar_augment <- caesar_tbl_int
caesar_augment$Infect_count[3] <- 0.5
caesar_augment$size[3] <- caesar_augment$size[3] + 1

caesar_augment

caesar_glm_noplan_factor_augment <- glm(
    Infect_count / size ~ noplan + factor + antib + noplan:factor,
    data = caesar_augment,
    weights = size,
    family = binomial
)

summary(caesar_glm_noplan_factor_augment)
```

Compare observed and fitted proportions:

```{r}
tibble(
    observed = round(caesar_augment$Infect_count / caesar_augment$size, 2),
    fitted = round(caesar_glm_noplan_factor_augment$fitted.values, 2)
)
```

Test goodness-of-fit using Pearson and deviance statistics:

```{r}
pearson_pval <- 1 -
    pchisq(
        sum(resid(caesar_glm_noplan_factor_augment, type = "pearson")^2),
        df = 2
    )
deviance_pval <- 1 -
    pchisq(
        sum(resid(caesar_glm_noplan_factor_augment, type = "deviance")^2),
        df = 2
    )

tibble(
    test = c("Pearson", "Deviance"),
    p_value = c(pearson_pval, deviance_pval)
)
```

We do not reject the new model with `noplan:factor` interaction
at usual levels such as 0.1 or 0.05.

## Reproducibility

### Session Information

```{r}
sessionInfo()
```

### Package Citations

Below are the formal citations for the packages used in this analysis:

```{r}
#| results: asis
#| code-fold: true
packages <- c("conflicted", "dplyr", "forcats", "Fahrmeir")
for (pkg in packages) {
    cit <- citation(pkg)
    # Extract only the text citation, not BibTeX
    cit_text <- format(cit, style = "text")
    cat("- **", pkg, "**: ", cit_text, "\n\n", sep = "")
}
```

## References

- Fahrmeir, L., & Tutz, G. (2001). *Multivariate Statistical Modelling Based on Generalized Linear Models (Second Edition)*. Springer. https://doi.org/10.1007/978-1-4757-3454-6
- [R documentation on `glm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html)
