---
title: "ANOVA and Deviance Analysis"
subtitle: "Binomial GLM with Logit and Probit Links"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script reproduces **Examples 2.1, 2.4, and 2.6** from Fahrmeir & Tutz, demonstrating:

- Binomial logistic regression on the Caesar dataset
- Hypothesis testing using deviance and Pearson statistics
- Interaction effects in GLMs
- Model goodness-of-fit assessment

## Data Setup

We start by loading the Caesar dataset from the Fahrmeir package and transforming it from contingency table format to individual-level data.

```{r}
#| message: false
library(Fahrmeir)
library(dplyr)

# Load and inspect the dataset
data(caesar)
head(caesar)
```

The dataset contains information about surgical infections with factors including:
- **Antibiotic use** (yes/no)
- **Risk factors** (yes/no)
- **Surgery type** (planned/emergency)

## Data Transformation

We aggregate the data to compute the number of infections and total cases for each combination of factors:

```{r}
# Convert to tibble format for easier manipulation
caesar_tibble <- as_tibble(caesar)

# Group by combinations of covariates
caesar_grouped <- caesar_tibble |>
  group_by(noplan, factor, antib) |>
  summarize(
    size = sum(w),
    Infect_count = sum(yl * w),
    .groups = 'drop'
  )

# Remove rows with zero cases
caesar <- caesar_grouped |>
  filter(size != 0) |>
  as.data.frame()

rownames(caesar) <- 1:nrow(caesar)
caesar
```

## Understanding Factor Coding

By default, R uses **dummy coding** for factors. Let's inspect and adjust the reference levels to match the textbook conventions:

```{r}
# Check current contrast coding
contrasts(caesar$noplan)
contrasts(caesar$factor)
contrasts(caesar$antib)

# Reset reference levels
caesar$noplan <- relevel(caesar$noplan, ref = "planned")
caesar$factor <- relevel(caesar$factor, ref = "without")
caesar$antib <- relevel(caesar$antib, ref = "without")

# Verify new contrasts
contrasts(caesar$noplan)
```

## Main Effects Model

Fit a binomial logistic regression with the three main effects:

```{r}
# Fit GLM with binomial family (logit link by default)
caesar_glm <- glm(
  Infect_count/size ~ antib + factor + noplan,
  data = caesar,
  weights = size,
  family = binomial()
)

summary(caesar_glm)
```

### Model Diagnostics

**Deviance and Goodness-of-Fit:**

```{r}
# Extract residuals of different types
deviance_resid <- resid(caesar_glm, type = "deviance")
pearson_resid <- resid(caesar_glm, type = "pearson")

# Goodness-of-fit tests
deviance_stat <- sum(deviance_resid^2)
pearson_stat <- sum(pearson_resid^2)

df_resid <- caesar_glm$df.residual
p_deviance <- 1 - pchisq(deviance_stat, df = df_resid)
p_pearson <- 1 - pchisq(pearson_stat, df = df_resid)

cat("Deviance statistic:", deviance_stat, "\n")
cat("Pearson statistic:", pearson_stat, "\n")
cat("p-value (deviance):", p_deviance, "\n")
cat("p-value (Pearson):", p_pearson, "\n")
```

**Comparison of fitted and observed proportions:**

```{r}
comparison <- data.frame(
  Actual = caesar$Infect_count / caesar$size,
  Fitted = caesar_glm$fitted.values
)
round(comparison, 3)
```

## Alternative Link Function: Probit

R allows easy comparison with different link functions. Here's the same model with a probit link:

```{r}
caesar_glm_probit <- glm(
  Infect_count/size ~ antib + factor + noplan,
  data = caesar,
  weights = size,
  family = binomial(link = "probit")
)

summary(caesar_glm_probit)
```

## Interaction Effects

### Model with noplan:antib Interaction

First, test the interaction between noplan (planned/emergency surgery) and antib (antibiotic use):

```{r}
# Fit model with noplan:antib interaction
caesar_glm_noplan_antib <- glm(
  Infect_count/size ~ factor + noplan * antib,
  data = caesar,
  weights = size,
  family = binomial()
)

summary(caesar_glm_noplan_antib)
```

**ANOVA Table:**

```{r}
anova(caesar_glm_noplan_antib, test = "Chisq")

# Extract the interaction deviance difference
anova_table <- anova(caesar_glm_noplan_antib, test = "Chisq")
interaction_deviance <- anova_table$Deviance[4]  # Last row in ANOVA table
p_noplan_antib <- 1 - pchisq(interaction_deviance, df = 1)

cat("Deviance change for noplan:antib interaction:", round(interaction_deviance, 4), "\n")
cat("p-value:", round(p_noplan_antib, 4), "\n")
cat("Conclusion: Interaction is", ifelse(p_noplan_antib < 0.05, "significant", "not significant"), "\n")
```

### Model with noplan:factor Interaction

```{r}
# Fit model with noplan:factor interaction
caesar_glm_int <- glm(
  Infect_count/size ~ noplan + factor + antib + noplan:factor,
  data = caesar,
  weights = size,
  family = binomial()
)

summary(caesar_glm_int)
```

**Model Diagnostics:**

```{r}
# Extract residuals
pearson_resid_int <- resid(caesar_glm_int, type = "pearson")
deviance_resid_int <- resid(caesar_glm_int, type = "deviance")

# Goodness-of-fit tests
pearson_stat_int <- sum(pearson_resid_int^2)
deviance_stat_int <- sum(deviance_resid_int^2)
df_int <- caesar_glm_int$df.residual

p_pearson_int <- 1 - pchisq(pearson_stat_int, df = df_int)
p_deviance_int <- 1 - pchisq(deviance_stat_int, df = df_int)

cat("Goodness-of-fit tests (noplan:factor model):\n")
cat("Pearson statistic:", round(pearson_stat_int, 4), "on", df_int, "df\n")
cat("p-value:", round(p_pearson_int, 4), "\n\n")
cat("Deviance statistic:", round(deviance_stat_int, 4), "on", df_int, "df\n")
cat("p-value:", round(p_deviance_int, 4), "\n")
```

**Comparison of fitted vs actual proportions:**

```{r}
comparison_int <- data.frame(
  Actual = caesar$Infect_count / caesar$size,
  Fitted = caesar_glm_int$fitted.values
)
round(comparison_int, 2)
```

## Handling Separation Issues

Some cells in the contingency table may be empty (zero infections), which can cause **complete separation** and infinite MLE estimates. A practical solution is to add fictitious data to stabilize estimation.

**Before augmentation:**

```{r}
cat("Original data (rows with 0 infections):\n")
print(caesar[which(caesar$Infect_count == 0), ])
```

**Adding fictitious data:**

```{r}
# Augment data: add 0.5 infection and 1 total case to problematic cell
caesar_aug <- caesar
caesar_aug$Infect_count[3] <- 0.5
caesar_aug$size[3] <- caesar_aug$size[3] + 1

cat("\nAugmented data:\n")
print(caesar_aug)
```

**Fit with augmented data:**

```{r}
# Refit model with augmented data (noplan:factor interaction)
caesar_glm_aug <- glm(
  Infect_count/size ~ noplan + factor + antib + noplan:factor,
  data = caesar_aug,
  weights = size,
  family = binomial()
)

summary(caesar_glm_aug)
```

**Comparison of actual vs fitted proportions:**

```{r}
comparison_aug <- data.frame(
  Actual = caesar_aug$Infect_count / caesar_aug$size,
  Fitted = caesar_glm_aug$fitted.values
)
round(comparison_aug, 2)
```

**Goodness-of-fit with augmented data:**

```{r}
# Pearson and deviance tests
pearson_aug <- sum(resid(caesar_glm_aug, type = "pearson")^2)
deviance_aug <- sum(resid(caesar_glm_aug, type = "deviance")^2)
df_aug <- caesar_glm_aug$df.residual

p_pearson_aug <- 1 - pchisq(pearson_aug, df = df_aug)
p_deviance_aug <- 1 - pchisq(deviance_aug, df = df_aug)

cat("Goodness-of-fit tests (augmented data, noplan:factor model):\n")
cat("Pearson chi-square:", round(pearson_aug, 4), "on", df_aug, "df, p =", round(p_pearson_aug, 4), "\n")
cat("Deviance:", round(deviance_aug, 4), "on", df_aug, "df, p =", round(p_deviance_aug, 4), "\n")
cat("\nConclusion: Model", ifelse(p_deviance_aug < 0.05, "is rejected", "is not rejected"), "at 0.05 level\n")
```

## Key Takeaways

1. **Data transformation**: Converting contingency tables to analysis format requires careful aggregation
2. **Contrast coding**: Choose reference levels strategically for interpretability
3. **Model comparison**: Use deviance and Pearson statistics for goodness-of-fit
4. **Interaction testing**: Compare nested models to test interaction significance
5. **Link functions**: Both logit and probit are commonly used; results often similar

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling Based on Generalized Linear Models*
- [R documentation on `glm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html)
