---
title: "ANOVA and Deviance Analysis"
subtitle: "Binomial GLM with Logit and Probit Links"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script reproduces **Examples 2.1, 2.4, and 2.6** from Fahrmeir & Tutz,
demonstrating:

- Binomial logistic regression on the Caesar dataset
- Hypothesis testing using deviance and Pearson statistics
- Interaction effects in GLMs
- Model goodness-of-fit assessment

## Data Setup

We start by loading the Caesar dataset from the Fahrmeir package
and transforming it from contingency table format to individual-level data.
Remember to install the package "Fahrmeir" to get the dataset!
We will only use `yl` as response for now (yes or no infection status).

```{r}
#| message: false
library(conflicted)
library(dplyr)
library(Fahrmeir)
data(caesar)
head(caesar)
```

## Data Transformation

Convert to tibble format for easier manipulation. Group the data by different
combinations of the covariates, then sum the total counts and infected count for
each combination. Finally, convert back to a usual data.frame and eliminate
unnecessary rows (combinations of covariates) with zero cases.

```{r}
caesar_tibble <- as_tibble(caesar)

caesar_grouped <- group_by(caesar_tibble, noplan, factor, antib)

caesar_summary <- summarize(
    caesar_grouped,
    size = sum(w),
    Infect_count = sum(yl * w)
)
caesar <- as.data.frame(caesar_summary)

caesar <- caesar[caesar$size != 0, ]
row.names(caesar) <- seq_len(nrow(caesar))
caesar
```

## Understanding Factor Coding

Look at a description of the structure of caesar and the coding scheme of each
variable of type "factor" using the function `contrasts()`. By default, they
are usually based on dummy coding. Note that they are coded in an exact opposite
way to Example 2.1 on p.30. Redefine the reference levels of these variables
of type "factor", so the coding is exactly like example 2.1 on p.30 in F & T.

```{r}
str(caesar)

contrasts(caesar$noplan)
contrasts(caesar$factor)
contrasts(caesar$antib)

caesar$noplan <- relevel(caesar$noplan, ref = "planned")
caesar$factor <- relevel(caesar$factor, ref = "without")
caesar$antib <- relevel(caesar$antib, ref = "without")

contrasts(caesar$noplan)
contrasts(caesar$factor)
contrasts(caesar$antib)
```

## Main Effects Model

Alternatively, we can force our covariates as integers with values (1 or 0)
exactly as in example 2.1 in the book. Fit the regression with the logit link
(default). The unscaled deviance gives the same scaled deviance for binomial and
poisson data. (Note: The documentation of glm() about what it is reporting for
the deviance may be misleading, because it ignores the fact that deviance only
becomes twice the log-likelihood difference when rescaled by the dispersion.)

```{r}
caesar$noplan <- as.integer(caesar$noplan == "not")
caesar$factor <- as.integer(caesar$factor == "risk factors")
caesar$antib <- as.integer(caesar$antib == "antibiotics")

caesar_glm <- glm(
    Infect_count / size ~ antib + factor + noplan,
    data = caesar,
    weights = size,
    family = binomial
)

caesar_glm
summary(caesar_glm)
deviance(caesar_glm)

anova(caesar_glm, test = "Chisq")
anova(caesar_glm, test = "F")
```

## Model Diagnostics

Extract residuals of different types. The working residuals are equivalent to
`caesar_glm$residuals`. Perform Pearson and deviance tests for model
goodness-of-fit. Compare fitted and realized proportions as on the top of p.52.

```{r}
resid(caesar_glm, type = "deviance")
resid(caesar_glm, type = "pearson")
resid(caesar_glm, type = "working")
resid(caesar_glm, type = "response")
sum(resid(caesar_glm, type = "deviance")^2)

1 - pchisq(caesar_glm$deviance, df = 3)
1 - pchisq(sum(resid(caesar_glm, type = "pearson")^2), df = 3)

round(caesar$Infect_count / caesar$size, digits = 2)
round(caesar_glm$fitted.values, digits = 2)
```

## Alternative Link Function: Probit

We can also fit with a probit link as was done on p. 31 (towards the end of
Ex 2.1). The coefficient estimates should be the same as those reported by the
book.

```{r}
caesar_glm_probit <- glm(
    Infect_count / size ~ antib + factor + noplan,
    data = caesar,
    weights = size,
    family = binomial(link = "probit")
)
summary(caesar_glm_probit)$coeff
```

## Interaction Effects

Now we play with the interaction effects like in Ex 2.4.

### Adding noplan:antib Interaction

Test the interaction between noplan (planned/emergency surgery) and antib
(antibiotic use). We get the same residual deviances as reported in the book.
The noplan:factor interaction is insignificant.

```{r}
caesar_glm_noplan_antib <- glm(
    Infect_count / size ~ factor + noplan * antib,
    data = caesar,
    weights = size,
    family = binomial
)

anova(caesar_glm_noplan_antib)

noplan_antib_interaction_delta_deviance <- anova(
    caesar_glm_noplan_antib
)$Deviance[5]
1 - pchisq(noplan_antib_interaction_delta_deviance, df = 1)
```

### Adding noplan:factor Interaction

Fit model with noplan:factor interaction. We will see some enormous estimates
and standard errors for some coefficients. This model is accepted at usual level
of significance. Compare fitted and realized proportions as on p.53.

```{r}
caesar_glm_noplan_factor <- glm(
    Infect_count / size ~ noplan + factor + antib + noplan:factor,
    data = caesar,
    weights = size,
    family = binomial
)

# eqiuvalently, you can do
# caesar_glm_noplan_factor <- glm(Infect_count/size ~  antib + noplan*factor,
#                    data=caesar , weight=size, family=binomial)

summary(caesar_glm_noplan_factor)

1 - pchisq(sum(resid(caesar_glm_noplan_factor, type = "pearson")^2), df = 2)
1 - pchisq(sum(resid(caesar_glm_noplan_factor, type = "deviance")^2), df = 2)

round(caesar$Infect_count / caesar$size, 2)
round(caesar_glm_noplan_factor$fitted.values, 2)
```
## Handling Separation Issues

The book recommended we should add fictitious data to the two zero
Infect_count cells to avoid existence of MLE issue. Add 0.5 infection and 1
total case to problematic cell. We do not reject the new model with
noplan_factor interaction at usual levels such as 0.1 or 0.05.

```{r}
caesar
caesar_augment <- caesar
caesar_augment$Infect_count[3] <- 0.5
caesar_augment$size[3] <- caesar_augment$size[3] + 1

caesar_augment

caesar_glm_noplan_factor_augment <- glm(
    Infect_count / size ~ noplan + factor + antib + noplan:factor,
    data = caesar_augment,
    weights = size,
    family = binomial
)

summary(caesar_glm_noplan_factor_augment)

round(
    rbind(
        caesar_augment$Infect_count / caesar_augment$size,
        caesar_glm_noplan_factor_augment$fitted
    ),
    digits = 2
)

1 -
    pchisq(
        sum(resid(caesar_glm_noplan_factor_augment, type = "pearson")^2),
        df = 2
    )
1 -
    pchisq(
        sum(resid(caesar_glm_noplan_factor_augment, type = "deviance")^2),
        df = 2
    )
```

## Key Takeaways

1. **Data transformation**: Converting contingency tables to analysis format requires careful aggregation
2. **Contrast coding**: Choose reference levels strategically for interpretability
3. **Model comparison**: Use deviance and Pearson statistics for goodness-of-fit
4. **Interaction testing**: Compare nested models to test interaction significance
5. **Link functions**: Both logit and probit are commonly used; results often similar

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling Based on Generalized Linear Models*
- [R documentation on `glm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html)
