---
title: "Cell Differentiation Study"
subtitle: "Poisson GLM with Interaction Effects"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script demonstrates:

- **Poisson regression** for count data (cell differentiation)
- Testing **synergistic vs. additive effects** of two treatment factors
- Model diagnostics using **deviance and Pearson statistics**
- **Dispersion estimation** for overdispersion assessment

The study investigates whether TNF (Tumor Necrosis Factor) and IFN (Interferon) agents stimulate cell differentiation **synergistically or independently**.

## Package Setup and Data

```{r}
#| message: false
library(Fahrmeir)
library(gee)      # For robust quasi-likelihood inference
library(MASS)     # For negative binomial modeling

# Load the cells dataset
data(cells)
head(cells)
str(cells)
```

The dataset contains:
- **TNF**: TNF treatment (yes/no)
- **IFN**: IFN treatment (yes/no)
- **y**: Count of differentiated cells

## Exploratory Analysis

Let's look at the structure and summary of the data:

```{r}
cells
summary(cells)

# Summary statistics by treatment group
with(cells, by(y, list(TNF, IFN), summary))
```

## Poisson GLM: Main Effects + Interaction

We fit a Poisson GLM with the interaction term to test for synergistic effects:

```{r}
# Fit Poisson GLM
glm_poisson <- glm(
  formula = y ~ TNF * IFN,
  family = poisson,
  data = cells
)

summary(glm_poisson)
```

### Model Interpretation

The terms in the model are:
- **Intercept**: Baseline log count (reference group: no TNF, no IFN)
- **TNF**: Main effect of TNF treatment
- **IFN**: Main effect of IFN treatment
- **TNF:IFN**: Interaction effect (synergy test)

Extract and display coefficients:

```{r}
coeff_table <- round(summary(glm_poisson)$coefficients, 4)
print(coeff_table)

# Exponentiate to get rate ratios
rate_ratios <- data.frame(
  Coefficient = names(glm_poisson$coefficients),
  "Rate Ratio" = round(exp(glm_poisson$coefficients), 4)
)
print(rate_ratios)
```

## Model Diagnostics

### Deviance Analysis

```{r}
cat("Model Deviance:", round(glm_poisson$deviance, 3), "\n")
cat("Degrees of Freedom:", glm_poisson$df.residual, "\n")

# ANOVA table
anova(glm_poisson, test = "Chisq")
```

### Goodness-of-Fit Testing

Two key statistics for assessing model fit in Poisson regression:

**1. Deviance Test:**
```{r}
p_value_deviance <- 1 - pchisq(glm_poisson$deviance, df = glm_poisson$df.residual)
cat("Deviance:", round(glm_poisson$deviance, 3), "\n")
cat("p-value:", round(p_value_deviance, 4), "\n")
```

**2. Pearson Chi-Square Test:**
```{r}
# Calculate Pearson residuals
pearson_residuals <- resid(glm_poisson, type = "pearson")
pearson_stat <- sum(pearson_residuals^2)

p_value_pearson <- 1 - pchisq(pearson_stat, df = glm_poisson$df.residual)
cat("Pearson Chi-Square:", round(pearson_stat, 3), "\n")
cat("p-value:", round(p_value_pearson, 4), "\n")
```

### Residual Diagnostics

```{r}
# Plot residuals
par(mfrow = c(2, 2))

# Deviance residuals vs fitted
plot(glm_poisson$fitted.values, resid(glm_poisson, type = "deviance"),
     main = "Deviance Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Deviance Residuals")
abline(h = 0, col = "red", lty = 2)

# Pearson residuals vs fitted
plot(glm_poisson$fitted.values, pearson_residuals,
     main = "Pearson Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Pearson Residuals")
abline(h = 0, col = "red", lty = 2)

# Q-Q plot
qqnorm(rstandard(glm_poisson))
qqline(rstandard(glm_poisson))

# Scale-location plot
plot(glm_poisson$fitted.values, sqrt(abs(pearson_residuals)),
     main = "Scale-Location",
     xlab = "Fitted Values", ylab = "√|Pearson Residuals|")

par(mfrow = c(1, 1))
```

## Overdispersion Assessment

Poisson regression assumes the mean equals the variance. When this fails, we have **overdispersion**.

```{r}
# Estimated dispersion parameter
dispersion_pearson <- pearson_stat / glm_poisson$df.residual
dispersion_deviance <- glm_poisson$deviance / glm_poisson$df.residual

cat("Dispersion (Pearson):", round(dispersion_pearson, 4), "\n")
cat("Dispersion (Deviance):", round(dispersion_deviance, 4), "\n")

if (dispersion_pearson > 1) {
  cat("\nOverdispersion detected!\n")
  cat("Consider using quasi-Poisson or negative binomial model.\n")
}
```

## Covariance Matrices

### Poisson Model: Scaled vs Unscaled

Under the Poisson assumption (dispersion = 1), these should be identical:

```{r}
cat("Scaled covariance matrix:\n")
print(round(summary(glm_poisson)$cov.scaled, 4))

cat("\nUnscaled covariance matrix:\n")
print(round(summary(glm_poisson)$cov.unscaled, 4))

cat("\nThese are identical under Poisson assumption (phi=1)\n")
```

## Quasi-Poisson Model

When overdispersion is present, use quasi-Poisson to estimate the scale:

```{r}
# Fit quasi-Poisson model
glm_quasipoisson <- glm(
  formula = y ~ TNF * IFN,
  family = quasipoisson,
  data = cells
)

summary(glm_quasipoisson)
```

### Comparing Covariance Matrices

```{r}
cat("Scaled covariance (quasi-Poisson):\n")
scaled_cov <- summary(glm_quasipoisson)$cov.scaled
print(round(scaled_cov, 4))

cat("\nUnscaled covariance (quasi-Poisson):\n")
unscaled_cov <- summary(glm_quasipoisson)$cov.unscaled
print(round(unscaled_cov, 4))

cat("\nRelationship: cov.scaled = cov.unscaled * dispersion\n")
cat("Dispersion from quasi-Poisson:", round(summary(glm_quasipoisson)$dispersion, 4), "\n")

# Verify relationship
scaled_computed <- unscaled_cov * summary(glm_quasipoisson)$dispersion
cat("\nVerification (max difference):\n")
print(max(abs(scaled_cov - scaled_computed)))
```

### Alternative Dispersion Estimates

```{r}
# Method 1: From quasi-Poisson summary
disp_glm <- summary(glm_quasipoisson)$dispersion
cat("GLM dispersion estimate:", round(disp_glm, 4), "\n")

# Method 2: Using working residuals and weights
disp_working <- with(glm_quasipoisson, sum(weights * residuals^2) / df.residual)
cat("Dispersion from working residuals:", round(disp_working, 4), "\n")

cat("\nNote: The dispersion in quasi-Poisson uses working residuals, not Pearson residuals\n")
```

## Example 2.7: Variance vs Mean and GEE

Plot the empirical variance against mean by TNF level:

```{r}
library(dplyr)

# Compute summary statistics by TNF level
cells_summary <- cells |>
  group_by(TNF) |>
  summarize(
    est_mean = mean(y),
    # Note: var() computes unbiased estimate; multiply by (n-1)/n for biased
    est_var = var(y) * 3 / 4,
    .groups = 'drop'
  ) |>
  as.data.frame()

print(cells_summary)

# Plot variance against mean
plot(cells_summary$est_mean, cells_summary$est_var,
     xlab = "Estimated Mean",
     ylab = "Estimated Variance",
     main = "Variance vs Mean Relationship",
     pch = 16, cex = 1.5)

# Add reference line for Poisson (variance = mean)
x_range <- range(cells_summary$est_mean)
abline(0, 1, lty = 2, col = "red", lwd = 2)
legend("topleft", legend = "Poisson (var = mean)", lty = 2, col = "red")
```

## GEE with Linear Variance Structure

Use GEE with linear variance function (variance ~ mean):

```{r}
library(gee)

# GEE with linear variance structure
gee_linear <- gee(
  formula = y ~ TNF * IFN,
  id = 1:nrow(cells),
  family = quasi(variance = "mu", link = "log"),
  data = cells
)

summary(gee_linear)

cat("\nGEE Coefficients (linear variance):\n")
print(round(summary(gee_linear)$coeff, 4))

cat("\nDispersion from GEE:\n")
print(round(summary(gee_linear)$scale, 4))
```

### Robust vs Naive Standard Errors

```{r}
cat("Robust z-values (sandwich variance):\n")
robust_z <- gee_linear$coeff / sqrt(diag(gee_linear$robust.variance))
robust_p <- round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
print(robust_p)

cat("\nNaive z-values (model-based variance):\n")
naive_z <- gee_linear$coeff / sqrt(diag(gee_linear$naive.variance))
naive_p <- round(2 * pnorm(abs(naive_z), lower.tail = FALSE), 3)
print(naive_p)
```

## GEE with Quadratic Variance Structure

Fit GEE with quadratic variance function (variance ~ mean²):

```{r}
# GEE with quadratic variance structure
gee_quadratic <- gee(
  formula = y ~ TNF * IFN,
  id = 1:nrow(cells),
  family = quasi(variance = "mu^2", link = "log"),
  data = cells
)

summary(gee_quadratic)

cat("\nRobust p-values (quadratic variance):\n")
robust_z_quad <- gee_quadratic$coeff / sqrt(diag(gee_quadratic$robust.variance))
robust_p_quad <- round(2 * pnorm(abs(robust_z_quad), lower.tail = FALSE), 3)
print(robust_p_quad)

cat("\nNote: These robust p-values should match Table 2.7 in F&T\n")
```

## Comparison with geepack

```{r}
library(geepack)

# geepack with scale estimated
gee_geepack_est <- geepack::geeglm(
  formula = y ~ TNF * IFN,
  id = 1:nrow(cells),
  family = poisson(link = "log"),
  data = cells,
  scale.fix = FALSE
)

# geepack with scale fixed to 1
gee_geepack_fix <- geepack::geeglm(
  formula = y ~ TNF * IFN,
  id = 1:nrow(cells),
  family = poisson(link = "log"),
  data = cells,
  scale.fix = TRUE
)

cat("geepack coefficients (scale estimated):\n")
print(round(gee_geepack_est$coefficients, 4))

cat("\ngeepack coefficients (scale fixed = 1):\n")
print(round(gee_geepack_fix$coefficients, 4))
```

## Comparison with Fitted Values

```{r}
fitted_comparison <- data.frame(
  Observation = 1:nrow(cells),
  TNF = cells$TNF,
  IFN = cells$IFN,
  Observed = cells$y,
  Fitted = round(glm_poisson$fitted.values, 2),
  Residual = round(glm_poisson$residuals, 3)
)
print(fitted_comparison)
```

## Key Findings

1. **Model Assessment**: Evaluate whether the Poisson assumption is appropriate via goodness-of-fit tests
2. **Interaction**: Test the TNF:IFN interaction to determine if effects are synergistic
3. **Overdispersion**: If present, consider alternative models (quasi-Poisson, negative binomial)
4. **Interpretation**: Rate ratios provide intuitive interpretation of treatment effects

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling Based on Generalized Linear Models*
- [Poisson Regression in R](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html)
