---
title: "Cell Differentiation Study"
subtitle: "Poisson GLM with Interaction Effects"
date: today
---

## Overview

This script demonstrates:

- **Poisson regression** for count data (cell differentiation)
- Testing **synergistic vs. additive effects** of two treatment factors
- Model diagnostics using **deviance and Pearson statistics**
- **Dispersion estimation** for overdispersion assessment

The study investigates whether TNF (Tumor Necrosis Factor) and IFN (Interferon) agents stimulate cell differentiation **synergistically or independently**.

## Package Setup and Data

```{r}
library(conflicted)
library(dplyr)
library(Fahrmeir)
library(gee)
library(geepack)
library(MASS)

data(cells)
cells
```

## Example 2.3

Aims to study whether the two agents stimulate cell differentiation synergistically or independently. Just fit the Poisson GLM as in Ex 2.3.

```{r}
glm_poisson <- glm(formula = y ~ TNF * IFN, family = poisson, data = cells)

summary(glm_poisson)

summary(glm_poisson)$coeff

round(summary(glm_poisson)$coeff, digits = 4)

anova(glm_poisson)
round(glm_poisson$coeff, digits = 4)
glm_poisson$deviance
```

## Covariance Matrices of Beta

These two are the same, which makes sense under the exponential family assumption is true.

```{r}
summary(glm_poisson)$cov.scaled
summary(glm_poisson)$cov.unscaled
```

## Covariance Matrices of Beta

These two are the same, which makes sense under the exponential family assumption is true.

```{r}
summary(glm_poisson)$cov.scaled
summary(glm_poisson)$cov.unscaled
```

## Example 2.6

Estimated Pearson statistic, same as in Eg 2.6. Both reject the model, but we should take these results with a grain of salt.

```{r}
pearson_chisq_stat <-
    sum((glm_poisson$y - glm_poisson$fitted.values)^2 / glm_poisson$fitted.values)

1 - pchisq(glm_poisson$deviance, df = 12)
1 - pchisq(pearson_chisq_stat, df = 12)
```

The estimated dispersion in Example 2.6, very close to the book.

```{r}
est_disperson <- pearson_chisq_stat / glm_poisson$df.residual
est_disperson
```

Perform quasi "model-based" (also known as "naive") approach assuming the variance structure is correctly specified:

```{r}
glm_quasi_poisson <- glm(
    formula = y ~ TNF * IFN,
    family = quasipoisson,
    data = cells
)

summary(glm_quasi_poisson)

summary(glm_quasi_poisson)$coefficients
anova(glm_quasi_poisson)
resid(glm_quasi_poisson)
resid(glm_quasi_poisson, type = "deviance")
resid(glm_quasi_poisson, type = "working")
resid(glm_quasi_poisson, type = "pearson")
```

These two are different now by the scale/dispersion.

```{r}
summary(glm_quasi_poisson)$cov.scaled
summary(glm_quasi_poisson)$cov.unscaled

summary(glm_poisson)$cov.unscaled * summary(glm_quasi_poisson)$dispersion
```

**WARNING**: the dispersion estimate provided by glm() actually leverages the "working residuals" and the IWLS weights! So it is not the same as our standard estimate based on Pearson residuals! Note that the weights here are same as `glm_poisson$weights`, the weights by assuming phi = 1 under the full likelihood model.

```{r}
summary(glm_quasi_poisson)$dispersion

with(glm_quasi_poisson, sum(weights * residuals^2)) /
    glm_quasi_poisson$df.residual
```

**WARNING**: the dispersion estimate provided by glm() actually leverages the "working residuals" and the IWLS weights! So it is not the same as our standard estimate based on Pearson residuals! Note that the weights here are same as `glm_poisson$weights`, the weights by assuming phi = 1 under the full likelihood model.

```{r}
summary(glm_quasi_poisson)$dispersion

with(glm_quasi_poisson, sum(weights * residuals^2)) /
    glm_quasi_poisson$df.residual
```

## Example 2.7

Plotting variance against mean by TNF. Note that `var()` computes the unbiased variance estimate by default; to reproduce the results on p.59 of F&T, we have to use the biased variance estimate.

```{r}
cells_tbl <- as_tibble(cells)
cells_tbl_grouped_tnf <- group_by(cells_tbl, TNF)
cells_tbl_summary <- summarize(
    cells_tbl_grouped_tnf,
    est_var = var(y) * 3 / 4,
    est_mean = mean(y)
)
cells_summary <- as.data.frame(cells_tbl_summary)
cells_summary
plot(
    cells_summary$est_mean,
    cells_summary$est_var,
    xlab = "mean",
    ylab = "variance"
)
```

Use the gee function for the "robust inference" in this example instead. This should get the second column of table 2.7 corresponding to the linear working variance structure. The dispersion provided by gee is the same as our own estimated dispersion based on Pearson residuals!

```{r}
gee_mu <- gee(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells)),
    family = quasi(variance = "mu", link = "log"),
    data = cells
)

summary(gee_mu)

summary(gee_mu)$coeff
```

The following should be the robust z-values, i.e. estimates rescaled by sandwich variance estimate.

```{r}
robust_z <- gee_mu$coeff / sqrt(diag((gee_mu)$robust.variance))
round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
```

The following should be the estimates rescaled by model-based variance estimate.

```{r}
naivez <- gee_mu$coeff / sqrt(diag((gee_mu)$naive.variance))
round(2 * pnorm(abs(naivez), lower.tail = FALSE), 3)
```

Try to fit with geeglm from the geepack instead.

```{r}
gee_mu_geepack <- geeglm(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells)),
    family = poisson(link = "log"),
    data = cells,
    scale.fix = FALSE
)

gee_mu_geepack_fix <- geeglm(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells)),
    family = poisson(link = "log"),
    data = cells,
    scale.fix = TRUE
)
```

I think the second column in Table 2.7 is not correctly printed; they have printed the naive p-values, which should have been the robust p-values.

Now the third column of table 2.7 corresponding to the purely quadratic working variance structure. The robust p-values match the ones in F&T, but not the naive ones, so all the p-values reported in parentheses of Table 2.7 in F&T should have been robust p-values.

```{r}
glm_gee_mu2 <- gee(
    formula = y ~ TNF * IFN,
    id = seq_len(nrow(cells)),
    family = quasi(variance = "mu^2", link = "log"),
    data = cells
)

summary(glm_gee_mu2)$scale
robust_z <- glm_gee_mu2$coeff / sqrt(diag(glm_gee_mu2$robust.variance))
round(2 * pnorm(abs(robust_z), lower.tail = FALSE), 3)
naivez <- glm_gee_mu2$coeff / sqrt(diag(glm_gee_mu2$naive.variance))
round(2 * pnorm(abs(naivez), lower.tail = FALSE), 3)
```
