---
title: "Example 3.1 & 3.11: Caesar Dataset Analysis"
subtitle: "Multinomial Logistic Regression and Category-Specific Effects"
date: today
---

## Overview

This document mirrors the R script for Examples 3.1 and 3.11 (Fahrmeir & Tutz), covering:
- Multinomial logistic regression on three infection categories (noInf, Inf1, Inf2)
- Fits via `nnet::multinom()` (BFGS) and `VGAM::vglm()` (Fisher scoring)
- A constrained parallel-coefficients model (Example 3.11)
- Additional fits with `glmnet` and alternative wide-format data

## Packages and Data

Load all required packages and the prepared Caesar dataset.

```{r}
#| message: false
library(conflicted)
library(dplyr)
library(nnet)
library(VGAM)
library(glmnet)
library(Fahrmeir)
library(tidyr)

caes_dat <- read.csv("data/Example3-1Caes.csv")
head(caes_dat)
str(caes_dat)
```

### Alternative Data Preparation (from Fahrmeir Package)

Alternatively, one can manipulate the dataset directly from the Fahrmeir package using `dplyr`. The code below:

1. Creates a tibble dataframe from the caesar data
2. Groups by covariate combinations (noplan, factor, antib)
3. Summarizes to compute total counts and category-specific counts
4. Eliminates rows with zero cases
5. Converts factor levels to match the textbook coding

```{r}
#| eval: false
caesar_df <- as_tibble(caesar)
caesar_grouped <- group_by(caesar_df, noplan, factor, antib)
caesar_summary <- summarize(
  caesar_grouped,
  size = sum(w),
  Noinf = w[which(y == 3)],
  Inf1 = w[which(y == 1)],
  Inf2 = w[which(y == 2)]
)
caesar <- as.data.frame(caesar_summary)

caesar <- caesar[which(caesar$size != 0), ]
rownames(caesar) <- seq_len(nrow(caesar))

caesar$factor <- as.integer(caesar$factor == "risk factors")
caesar$noplan <- as.integer(caesar$noplan == "not")
caesar$antib <- as.integer(caesar$antib == "antibiotics")
```

## Example 3.1: Multinom Fit (nnet)

Fit the multinomial model with the response matrix (columns 2:4) and covariates `NoPlan`, `Antib`, and `RiskF`.

```{r}
?multinom
caes_nnet <- multinom(
    as.matrix(caes_dat[, 2:4]) ~ NoPlan + Antib + RiskF,
    data = caes_dat
)

caes_nnet
summary(caes_nnet)
summary(caes_nnet)$coeff

attributes(caes_nnet)
caes_nnet$deviance
```

## glmnet Comparison

Fit the same model with `glmnet` (multinomial family) and reshape coefficients to match `multinom` output.

```{r}
caes_glmnet <- glmnet(
    x = cbind(as.matrix(caes_dat[, 5:7])),
    y = as.matrix(caes_dat[, 2:4]),
    family = "multinomial",
    lambda = c(0)
)

caes_glmnet$beta

beta_matrix <- cbind(
    caes_glmnet$beta[[2]] - caes_glmnet$beta[[1]],
    caes_glmnet$beta[[3]] - caes_glmnet$beta[[1]]
)

beta_matrix <- rbind(
    caes_glmnet$a0[-1] - caes_glmnet$a0[1],
    beta_matrix
)

t(as.matrix(beta_matrix))
```

## Alternative: Original Fahrmeir Data

Fit using the `caesar` data from the Fahrmeir package, setting preferred reference levels before modeling.

```{r}
data(caesar)
caesar
?caesar
as.factor(caesar$y)
caesar$y <- relevel(caesar$y, ref = "3")

as.factor(caesar$noplan)
caesar$noplan <- relevel(caesar$noplan, ref = "planned")

as.factor(caesar$antib)
caesar$antib <- relevel(caesar$antib, ref = "without")

as.factor(caesar$factor)
caesar$factor <- relevel(caesar$factor, ref = "without")

caes_nnet2 <- multinom(y ~ noplan + antib + factor, data = caesar, weights = w)

summary(caes_nnet2)
caes_nnet2$deviance
```

## Exploring `caes_nnet`

Inspect fitted probabilities, residuals, grouped proportions, and predictions for a specific covariate combination.

```{r}
caes_nnet$fitted
caes_nnet$residuals

caes_nnet_grp <- caes_dat[, 2:4]
caes_nnet_grp[, 1] <- caes_dat[, 2] / caes_dat[, 1]
caes_nnet_grp[, 2] <- caes_dat[, 3] / caes_dat[, 1]
caes_nnet_grp[, 3] <- caes_dat[, 4] / caes_dat[, 1]
caes_nnet_grp - caes_nnet$fitted

caes_nnet$deviance
predict(caes_nnet, data.frame(NoPlan = 1, Antib = 1, RiskF = 0), type = "probs")
```{r}

ratios <- exp(summary(caes_nnet)$coeff %*% c(1, 1, 1, 0))
prob_no_inf <- 1 / (sum(ratios) + 1)

prob_no_inf
predict(
    caes_nnet,
    data.frame(NoPlan = 1, Antib = 1, RiskF = 0),
    type = "probs",
    se.fit = TRUE
)
```

## Example 3.1: VGAM Fit

Fit the multinomial model with Fisher scoring and compare deviance/coefficients with `nnet`.

```{r}
caes_vgam <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = caes_dat,
    family = multinomial(refLevel = "noInf")
)

summary(caes_vgam)
coefvlm(caes_vgam, matrix.out = TRUE)
summary(vglm(
    formula = cbind(noInf, Inf2, Inf1) ~ NoPlan + Antib + RiskF,
    data = caes_dat,
    family = multinomial(refLevel = "noInf")
))

fittedvlm(caes_vgam)
summary(caes_nnet)$coeff
```

## Example 3.11: Parallel Coefficients

Constrain `NoPlan` and `RiskF` to be equal across categories using the `parallel` argument.

```{r}
caes_vgam_parallel <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = caes_dat,
    family = multinomial(refLevel = "noInf", parallel = TRUE ~ NoPlan + RiskF - 1)
)

coefvlm(caes_vgam_parallel, matrix.out = TRUE)
```

## VGAM on Filtered Fahrmeir Data

Filter out zero-count rows, reset reference levels, and fit with `vglm`; compare criteria and log-likelihoods.

```{r}
caesar_no_zero_w <- dplyr::filter(caesar, w > 0)
levels(caesar_no_zero_w$noplan)
levels(caesar_no_zero_w$antib)
levels(caesar_no_zero_w$factor)

caesar_no_zero_w$noplan <- relevel(caesar_no_zero_w$noplan, ref = "planned")
caesar_no_zero_w$antib <- relevel(caesar_no_zero_w$antib, ref = "without")
caesar_no_zero_w$factor <- relevel(caesar_no_zero_w$factor, ref = "without")

caes_vgam2 <- vglm(
    formula = y ~ noplan + antib + factor,
    data = caesar_no_zero_w,
    weights = w,
    family = VGAM::multinomial(refLevel = "3")
)

summary(caes_vgam2)
coefvlm(caes_vgam2, matrix.out = TRUE)

attributes(caes_vgam)$criterion
attributes(caes_vgam2)$criterion
logLik(caes_vgam)
logLik(caes_vgam2)
```

## Long-Form Data (Wide-to-Long)

Reshape the prepared data to long format and fit the multinomial model with weights; coefficients should match earlier fits.

```{r}
caesar_new <- pivot_longer(
    data = caes_dat[, -1],
    cols = noInf:Inf2,
    names_to = "inf_type",
    values_to = "w"
)
caesar_new$inf_type <- factor(
    caesar_new$inf_type,
    levels = unique(caesar_new$inf_type)
)
# Extract coefficient matrix

caes_nnet_tibble <- multinom(
    inf_type ~ NoPlan + Antib + RiskF,
    data = caesar_new,
    weights = w
)
summary(caes_nnet_tibble)$coeff
summary(caes_nnet2)$coeff
```

## Notes

- `multinom()` uses BFGS (quasi-Newton); `vglm()` uses Fisher scoring.
- Reference levels are set explicitly to match textbook coding.
- Deviance values differ across implementations depending on how the saturated model is handled.
- Robust/naive p-values in Table 2.7 of F&T should correspond to robust estimates.
coef_matrix <- summary(Caes_nnet)$coeff
cat("Coefficient Matrix (matches textbook Example 3.1, p.109):\n")
print(round(coef_matrix, 4))

# Model deviance (note: this is -2 * loglikelihood, unadjusted)
cat("\nModel Deviance:\n")
print(Caes_nnet$deviance)


## Key Learning Points

1. **Multinomial regression**: For outcomes with >2 unordered categories
2. **Reference category**: Coefficients are relative to the reference (first) category
3. **Interpretation**: Coefficients represent log-odds ratios vs reference
4. **Optimization methods**: BFGS (nnet) vs Fisher scoring (VGAM) give similar results
5. **Constrained models**: Test whether coefficients are equal across categories (parallel assumption)
6. **Data formats**: Wide format (response matrix) vs long format (weights)

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling Based on Generalized Linear Models* (2nd ed.), p. 109-114
- Yee, T. W. (2015). *Vector Generalized Linear and Additive Models: With an Implementation in R*
- [nnet package documentation](https://cran.r-project.org/package=nnet)
- [VGAM package documentation](https://cran.r-project.org/package=VGAM)
