---
title: "Example 3.1 & 3.11: Caesar Dataset Analysis"
subtitle: "Multinomial Logistic Regression and Category-Specific Effects"
date: today
format:
  html:
    code-fold: false
---

## Overview

This example reproduces **Example 3.1 and 3.11** from Fahrmeir & Tutz, demonstrating:

- **Multinomial logistic regression** for categorical outcomes with >2 categories
- Modeling three infection categories: no infection, single infection, multiple infections
- Comparing estimation methods: `nnet::multinom()` and `VGAM::vglm()`
- **Constrained coefficients** (Example 3.11) with parallel assumption testing

## Data Loading and Exploration

```{r}
#| message: false
library(nnet)
library(VGAM)
library(dplyr)
library(tidyr)

# Load the pre-formatted Caesar dataset from CSV
Caes.dat <- read.csv("data/Example3-1Caes.csv")
head(Caes.dat)
is.data.frame(Caes.dat)
str(Caes.dat)
```

### Understanding the Data

The dataset contains surgical infection data with:
- **size**: Total number of patients per combination
- **noInf**: Patients with no infection (reference category)
- **Inf1**: Patients with exactly one infection
- **Inf2**: Patients with two infections
- **NoPlan**: Surgery planning (0=planned, 1=not planned)
- **Antib**: Antibiotic use (0=no, 1=yes)
- **RiskF**: Risk factors (0=no, 1=yes)

## Example 3.1: Multinomial Regression with `nnet`

The `multinom()` function fits multinomial logistic regression using BFGS optimization (quasi-Newton method).

### Fitting the Model

```{r}
# Fit multinomial model
# The response is the 2:4 columns (noInf, Inf1, Inf2)
# First column (noInf) is automatically treated as reference category
Caes_nnet <- nnet::multinom(as.matrix(Caes.dat[, 2:4]) ~ NoPlan + Antib + RiskF,
    data = Caes.dat
)

# Display model summary
Caes_nnet
summary(Caes_nnet)
```

### Extracting Coefficients

```{r}
# Extract coefficient matrix
coef_matrix <- summary(Caes_nnet)$coeff
cat("Coefficient Matrix (matches textbook Example 3.1, p.109):\n")
print(round(coef_matrix, 4))

# Transpose for easier reading
cat("\nTransposed coefficients:\n")
print(t(coef_matrix))

# Model deviance (note: this is -2 * loglikelihood, unadjusted)
cat("\nModel Deviance:\n")
print(Caes_nnet$deviance)
```

### Model Predictions and Fitted Values

```{r}
# Get fitted probabilities for each observation
fitted_probs <- Caes_nnet$fitted
cat("Fitted probabilities (first few rows):\n")
print(round(head(fitted_probs), 4))

# Get residuals (observed - fitted proportions)
residuals_nnet <- Caes_nnet$residuals
cat("\nResiduals (first few rows):\n")
print(round(head(residuals_nnet), 4))
```

### Manual Verification of Residuals

```{r}
# Manually compute proportions from the data
Caes_proportions <- Caes.dat[, 2:4]
Caes_proportions[, 1] <- Caes.dat[, 2] / Caes.dat[, 1] # noInf proportion
Caes_proportions[, 2] <- Caes.dat[, 3] / Caes.dat[, 1] # Inf1 proportion
Caes_proportions[, 3] <- Caes.dat[, 4] / Caes.dat[, 1] # Inf2 proportion

cat("Manual calculation of residuals:\n")
print(round(Caes_proportions - Caes_nnet$fitted, 4))
```

### Predictions for Specific Covariate Combinations

```{r}
# Predict probabilities for a specific case
# NoPlan=1 (not planned), Antib=1 (yes), RiskF=0 (no)
pred_case <- data.frame(NoPlan = 1, Antib = 1, RiskF = 0)

cat("Predicted probabilities for NoPlan=1, Antib=1, RiskF=0:\n")
pred_probs <- predict(Caes_nnet, pred_case, type = "probs")
print(pred_probs)

# Predicted class
cat("\nPredicted infection category:\n")
print(predict(Caes_nnet, pred_case, type = "class"))
```

### Manual Probability Calculation

```{r}
# Manually verify the probability calculation
# Using multinomial logistic regression: P(category j) = exp(η_j) / (1 + sum(exp(η_i)))
# where η_i are the linear predictors relative to reference category

coefs_case <- coef_matrix %*% c(1, 1, 1, 0) # [intercept, NoPlan, Antib, RiskF]
cat("Linear predictors for this case:\n")
print(coefs_case)

# Calculate probabilities
ratios <- exp(coefs_case) # Relative risks vs reference
prob_reference <- 1 / (sum(ratios) + 1) # P(no infection)
prob_other <- ratios * prob_reference # P(Inf1), P(Inf2)

cat("\nManually calculated probabilities:\n")
cat("P(noInf):", prob_reference, "\n")
cat("P(Inf1):", prob_other[1], "\n")
cat("P(Inf2):", prob_other[2], "\n")
```

## Alternative Approach: VGAM with Fisher Scoring

The VGAM package uses Fisher scoring (different optimization) and provides additional flexibility.

### Basic VGAM Model

```{r}
# Fit using VGAM with multinomial family
# Note: Need to specify reference level
Caes_vgam <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = Caes.dat,
    family = multinomial(refLevel = "noInf")
)

summary(Caes_vgam)
```

### Extracting VGAM Coefficients

```{r}
# Extract coefficients in matrix form
coef_vgam <- coefvlm(Caes_vgam, matrix.out = TRUE)
cat("VGAM Coefficients (in matrix form):\n")
print(round(coef_vgam, 4))

# Compare with nnet results
cat("\nComparison: nnet vs VGAM\n")
cat("Both methods should give similar results (up to numerical differences)\n")
cat("nnet coefficients:\n")
print(round(coef_matrix, 4))
```

### VGAM Fitted Probabilities

```{r}
# Extract fitted values
fitted_vgam <- fittedvlm(Caes_vgam)
cat("VGAM fitted probabilities (first few rows):\n")
print(round(head(fitted_vgam), 4))
```

## Example 3.11: Parallel Coefficients (Constrained Model)

Example 3.11 tests whether some coefficients can be constrained to be equal across outcome categories (proportional odds-like assumption).

### Constrained VGAM Model

```{r}
# Fit model with parallel constraint on NoPlan and RiskF
# This means these coefficients are the same for both Inf1 and Inf2 categories
Caes_vgam_parallel <- vglm(
    formula = cbind(noInf, Inf1, Inf2) ~ NoPlan + Antib + RiskF,
    data = Caes.dat,
    family = multinomial(
        refLevel = "noInf",
        parallel = TRUE ~ NoPlan + RiskF - 1
    )
)

summary(Caes_vgam_parallel)
```

### Compare Constrained Coefficients

```{r}
# Extract parallel model coefficients
coef_parallel <- coefvlm(Caes_vgam_parallel, matrix.out = TRUE)
cat("Parallel model coefficients (NoPlan, RiskF constrained):\n")
print(round(coef_parallel, 4))

cat("\nUnconstrained model coefficients (for comparison):\n")
print(round(coef_vgam, 4))

# Notice: Parallel model has single NoPlan and RiskF coefficients
# while Antib has separate coefficients for each category
```

### Likelihood Ratio Test

```{r}
# Test the parallel constraint hypothesis
# H0: NoPlan and RiskF effects are equal across categories
# H1: Effects differ across categories

ll_unconstrained <- logLik(Caes_vgam)[1]
ll_parallel <- logLik(Caes_vgam_parallel)[1]

lr_stat <- 2 * (ll_unconstrained - ll_parallel)
df_diff <- length(coef(Caes_vgam)) - length(coef(Caes_vgam_parallel))
p_value <- 1 - pchisq(lr_stat, df = df_diff)

cat("Likelihood Ratio Test:\n")
cat("H0: Parallel assumption holds for NoPlan, RiskF\n")
cat("LR statistic:", round(lr_stat, 4), "\n")
cat("df:", df_diff, "\n")
cat("p-value:", round(p_value, 4), "\n")

if (p_value > 0.05) {
    cat("\nConclusion: Parallel assumption is reasonable (p >0.05)\n")
} else {
    cat("\nConclusion: Parallel assumption rejected (p <0.05)\n")
}
```

## Alternative Data Format: Long Form

The data can also be reshaped to long form for alternative modeling approaches:

```{r}
# Reshape from wide to long format
caesar_long <- pivot_longer(
    data = Caes.dat[, -1],
    cols = noInf:Inf2,
    names_to = "inf_type",
    values_to = "w"
)

# Convert to factor with proper reference level
caesar_long$inf_type <- factor(caesar_long$inf_type,
    levels = unique(caesar_long$inf_type)
)

cat("Data in long format (first 12 rows):\n")
print(head(caesar_long, 12))

# Fit model using long format
Caes_nnet_long <- multinom(inf_type ~ NoPlan + Antib + RiskF,
    data = caesar_long,
    weights = w
)

cat("\nCoefficients from long format (should match original):\n")
print(round(summary(Caes_nnet_long)$coeff, 4))
```

## Key Learning Points

1. **Multinomial regression**: For outcomes with >2 unordered categories
2. **Reference category**: Coefficients are relative to the reference (first) category
3. **Interpretation**: Coefficients represent log-odds ratios vs reference
4. **Optimization methods**: BFGS (nnet) vs Fisher scoring (VGAM) give similar results
5. **Constrained models**: Test whether coefficients are equal across categories (parallel assumption)
6. **Data formats**: Wide format (response matrix) vs long format (weights)

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling Based on Generalized Linear Models* (2nd ed.), p. 109-114
- Yee, T. W. (2015). *Vector Generalized Linear and Additive Models: With an Implementation in R*
- [nnet package documentation](https://cran.r-project.org/package=nnet)
- [VGAM package documentation](https://cran.r-project.org/package=VGAM)
