---
title: "Example 3.2 & 3.4: BTR Analysis"
subtitle: "Ordinal Regression on Breathing Test Results"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script reproduces **Examples 3.2 and 3.4** from Fahrmeir & Tutz, analyzing **breathing test results (BTR)** with ordinal outcomes.

Topics covered:
- **Ordinal logistic regression** with `polr()` function
- Model fitting with weighted counts
- Contrast coding and reference levels
- Interpretation of coefficients and proportional odds assumption

## Data Loading and Setup

```{r}
#| message: false
library(MASS)

# Load the BTR dataset from CSV
# Note: Ordinal outcome with factors: Positive, Borderline, Negative
BTR.dat <- read.csv("data/Example3-2BTR.csv",
    colClasses = c("factor", "factor", "factor", "numeric")
)

str(BTR.dat)
is.data.frame(BTR.dat)
BTR.dat
```

### Preparing the Data

```{r}
# Explore and modify the class of the data variables
is.factor(BTR.dat$BTR)
is.ordered(BTR.dat$BTR)

# Set BTR to be ordinal (ordered categorical)
BTR.dat$BTR <- as.ordered(BTR.dat$BTR)

is.factor(BTR.dat$Age)
is.factor(BTR.dat$Smoking)
is.factor(BTR.dat$Freq)
```

## Contrast Coding
The default is dummy coding with `contr.treatment`. Let's examine the contrasts:

```{r}
# Display current contrast settings
getOption(x = "contrasts")

contrasts(BTR.dat$Smoking)
contrasts(BTR.dat$Age)
```

# Fit cumulative logistic model with Age, Smoking, and their interaction
# Note: polr() computes coefficient signs opposite to Fahrmeir & Tutz formula
BTR.logit <- polr(BTR ~ Age + Smoking + Age:Smoking,
    data = BTR.dat,
    weights = Freq,
    Hess = TRUE,
    method = "logistic"
)

summary(BTR.logit)
```

### Accessing Model Components

```{r}
# Access model components
names(BTR.logit)

# Linear predictors
BTR.logit$lp

# Threshold parameters (intercepts/cutpoints)
cat("Thresholds:\n")
print(BTR.logit$zeta)
```

### Manual Coefficient Verification

```{r}
# Create design matrix manually for verification
Z <- cbind(
    (BTR.dat$Age == "40to59"),
    (BTR.dat$Smoking == "2Former"),
    (BTR.dat$Smoking == "3Current"),
    (BTR.dat$Age == "40to59") * (BTR.dat$Smoking == "2Former"),
    (BTR.dat$Age == "40to59") * (BTR.dat$Smoking == "3Current")
)

# Hand-computed linear predictors should match
cat("Manual eta calculation (should match BTR.logit$lp):\n")
print(Z %*% BTR.logit$coeff)

cat("\nActual linear predictors:\n")
print(BTR.logit$lp)
```

## Model Interpretation

Extract and interpret coefficients:

```{r}
# Coefficient table
cat("Coefficients:\n")
print(summary(BTR.logit)$coefficients)

# Odds ratios
cat("\nOdds Ratios (exp(coefficients)):\n")
print(round(exp(BTR.logit$coeff), 4))
```

## Predicted Probabilities

```{r}
# Predict for specific combinations
new_data <- data.frame(
    Age = c("20to39", "40to59"),
    Smoking = c("1Never", "2Former")
)

# Get predicted probabilities for each class
predictions <- predict(BTR.logit, new_data, type = "probs")
cat("Predicted probabilities:\n")
print(round(predictions, 4))
```

## Model Diagnostics

```{r}
# Check fitted values
cat("Fitted values (first few):\n")
print(head(BTR.logit$fitted.values, 10))

# Deviance
cat("\nModel Deviance:\n")
print(BTR.logit$deviance)

# Residuals
cat("\nResiduals (first few):\n")
print(head(residuals(BTR.logit), 10))
```

## Proportional Odds Assumption

The proportional odds model assumes coefficients are the same across all cutpoints. We can test this assumption using the Brant test or comparing with non-proportional models.

```{r}
# Loglikelihood
cat("Log-likelihood:\n")
print(logLik(BTR.logit))
```

## Key Learning Points

1. **Ordinal outcomes**: BTR with three ordered categories (Positive > Borderline > Negative)
2. **Proportional odds model**: Assumes constant coefficients across all cutpoints
3. **Weights parameter**: Handles aggregated count data properly
4. **Coefficient interpretation**: Signs differ from textbook convention (check documentation)
5. **Thresholds**: Represent the intercepts for each cutpoint in the latent variable model

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Ripley, B. D. (2002). Modern applied statistics with S (4th ed.) - MASS package
- [polr() documentation](https://www.rdocumentation.org/packages/MASS/versions/7.3-51.6/topics/polr)
