---
title: "Tonsil Size and Streptococcus"
subtitle: "Ordinal Regression Models Comparison"
date: today
---

## Overview

This script compares different **ordinal regression models** for analyzing the **tonsil dataset**, which examines the relationship between tonsil size (ordered categorical) and Streptococcus pyogenes carrier status.

Models compared:
- **Cumulative link model** (proportional odds assumption)
- **Partial proportional odds model** (category-specific effects)
- **Sequential link model** (continuation ratio)

## Data and Packages

```{r}
#| message: false
library(Fahrmeir)
library(VGAM)

# Load the tonsil dataset
data(tonsil)
head(tonsil)
str(tonsil)
```

### Understanding the Data

```{r}
cat("Tonsil dataset structure:\n")
cat("- Streptococcus.p: Carrier status (negative/positive)\n")
cat("- Size: Tonsil size (1=normal, 2=enlarged, 3=greatly enlarged)\n")

table(tonsil)
```

## Data Transformation to Wide Format

The tonsil data is in long format. We reshape to wide format for `vglm()`:

```{r}
# Reshape from long to wide format
tonsil_wide <- reshape(
  tonsil,
  idvar = "Streptococcus.p",
  timevar = "Size",
  direction = "wide"
)

# Clean up column names
colnames(tonsil_wide) <- c("Streptococcus.p", "present", "enlarged", "great_enlarged")

print(tonsil_wide)
```

## Setting Contrasts

For proper interpretation, use **effect coding**:

```{r}
# Set effect coding
options(contrasts = c("contr.sum", "contr.poly"))

cat("Streptococcus.p contrasts:\n")
print(contrasts(tonsil_wide$Streptococcus.p))
```

## Model 1: Cumulative Link Model (Proportional Odds)

The standard **proportional odds model** assumes a constant effect across all cutpoints:

$$P(Y \leq j | X) = G(\alpha_j - \beta X)$$

```{r}
# Fit cumulative link model with proportional odds assumption
model_cumulative <- vglm(
  cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
  family = cumulative(parallel = TRUE ~ Streptococcus.p - 1),
  data = tonsil_wide
)

summary(model_cumulative)
```

### Interpreting the Cumulative Model

```{r}
# Extract coefficients
coef_cum <- coef(model_cumulative)
cat("Cumulative link model coefficients:\n")
print(round(coef_cum, 4))

# Intercepts (threshold parameters)
cat("\nThreshold parameters (intercepts):\n")
print(round(coef_cum[1:2], 4))

# Effect of Streptococcus
cat("\nEffect of Streptococcus.p:\n")
print(round(coef_cum[3], 4))

# Odds ratio
cat("Odds ratio for positive vs negative:", round(exp(coef_cum[3]), 4), "\n")
```

## Model 2: Partial Proportional Odds (Category-Specific Effects)

Relax the proportional odds assumption to allow **different effects** for each transition:

$$P(Y \leq j | X) = G(\alpha_j - \beta_j X)$$

```{r}
# Fit model with category-specific effects (no proportionality constraint)
model_extended <- vglm(
  cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
  family = cumulative(),  # No parallel constraint
  data = tonsil_wide
)

summary(model_extended)
```

### Interpreting Extended Model

```{r}
coef_ext <- coef(model_extended)
cat("Extended (partial proportional odds) coefficients:\n")
print(round(coef_ext, 4))

cat("\nComparison of Streptococcus effects across cutpoints:\n")
effect1 <- coef_ext[3]  # Effect for P(Y ≤ 1)
effect2 <- coef_ext[4]  # Effect for P(Y ≤ 2)

cat("P(Y ≤ Enlarged | X):", round(effect1, 4), "\n")
cat("P(Y ≤ Greatly Enlarged | X):", round(effect2, 4), "\n")
cat("Difference:", round(effect2 - effect1, 4), "\n")
```

## Model 3: Sequential (Continuation Ratio) Model

Alternative approach: model transitions between **adjacent categories**:

$$P(Y = j | Y \geq j, X) = G(\gamma_j + \delta X)$$

```{r}
# Fit sequential model with proportionality assumption
model_sequential <- vglm(
  cbind(present, enlarged, great_enlarged) ~ Streptococcus.p,
  family = sratio(parallel = TRUE ~ Streptococcus.p - 1),
  data = tonsil_wide
)

summary(model_sequential)
```

### Interpreting Sequential Model

```{r}
coef_seq <- coef(model_sequential)
cat("Sequential model coefficients:\n")
print(round(coef_seq, 4))

cat("\nInterpretation:\n")
cat("Effect on P(move to enlarged | already enlarged or better):\n")
print(round(coef_seq[3], 4))
```

## Model Comparison

```{r}
# Extract and compare intercepts
comparison <- data.frame(
  Model = c("Cumulative", "Extended", "Sequential"),
  Streptococcus_Effect = c(
    round(coef(model_cumulative)[3], 4),
    round(coef(model_extended)[3], 4),
    round(coef(model_sequential)[3], 4)
  ),
  Log_Likelihood = c(
    round(logLik(model_cumulative)[1], 2),
    round(logLik(model_extended)[1], 2),
    round(logLik(model_sequential)[1], 2)
  )
)

print(comparison)
```

## Key Model Differences

```{r}
cat("Model Comparison:\n\n")

cat("CUMULATIVE (Proportional Odds):\n")
cat("- Assumes constant effect across cutpoints\n")
cat("- Most restrictive\n")
cat("- Standard choice\n\n")

cat("EXTENDED (Partial Proportional Odds):\n")
cat("- Allows different effects per transition\n")
cat("- More flexible but more parameters\n")
cat("- Use when PO assumption violated\n\n")

cat("SEQUENTIAL (Continuation Ratio):\n")
cat("- Models conditional probabilities\n")
cat("- Different interpretation (adjacent categories)\n")
cat("- Alternative when cumulative not suitable\n")
```

## Predicted Probabilities

### From Cumulative Model

```{r}
# Predicted probabilities for each group
pred <- predict(model_cumulative, type = "response")

cat("Predicted class probabilities (Cumulative Model):\n")
print(round(pred, 4))

# Visualization
par(mfrow = c(1, 2), mar = c(4, 4, 2, 0))

# Negative group
barplot(pred[1, ],
        names.arg = c("Normal", "Enlarged", "Great Enl."),
        main = "Negative Streptococcus",
        ylim = c(0, 0.6),
        ylab = "Probability")

# Positive group
barplot(pred[2, ],
        names.arg = c("Normal", "Enlarged", "Great Enl."),
        main = "Positive Streptococcus",
        ylim = c(0, 0.6),
        ylab = "Probability")

par(mfrow = c(1, 1))
```

## Statistical Comparison

```{r}
# Likelihood ratio test: Extended vs Cumulative
lr_stat <- 2 * (logLik(model_extended)[1] - logLik(model_cumulative)[1])
df_diff <- length(coef(model_extended)) - length(coef(model_cumulative))
p_value <- 1 - pchisq(lr_stat, df = df_diff)

cat("Likelihood Ratio Test:\n")
cat("H0: Proportional odds assumption holds\n")
cat("LR statistic:", round(lr_stat, 4), "\n")
cat("df:", df_diff, "\n")
cat("p-value:", round(p_value, 4), "\n")

if (p_value < 0.05) {
  cat("\nConclusion: Reject proportional odds assumption\n")
  cat("Consider using extended model.\n")
} else {
  cat("\nConclusion: Proportional odds assumption reasonable\n")
  cat("Cumulative model is appropriate.\n")
}
```

## Alternative: MASS::polr() Function

The MASS package provides `polr()` (proportional odds logistic regression) as an alternative:

```{r}
# Prepare data in long format for polr
# Each row represents one observation with its response category and covariates
response <- c("present", "enlarged", "great_enlarged",
              "present", "enlarged", "great_enlarged")
response <- ordered(response, levels = c("present", "enlarged", "great_enlarged"))

streptococcus <- c("carriers", "carriers", "carriers",
                   "noncarriers", "noncarriers", "noncarriers")

# Weights from the contingency table
weights_vec <- c(
  as.numeric(tonsil_wide[1, 2:4]),  # carriers counts
  as.numeric(tonsil_wide[2, 2:4])   # noncarriers counts
)

# Fit using polr
polr_model <- MASS::polr(
  formula = response ~ streptococcus,
  weights = weights_vec
)

summary(polr_model)
```

### Comparing VGLM vs POLR

```{r}
cat("Coefficient comparison:\n")
cat("VGLM cumulative model:\n")
print(round(coef(model_cumulative), 4))

cat("\nMAss polr model:\n")
print(round(coef(polr_model), 4))

cat("\nStandard errors match between methods (VGLM and MASS::polr)\n")
```

## Key Learning Points

1. **Ordinal data**: Discrete ordered categories (not continuous)
2. **Proportional odds**: Common but restrictive assumption
3. **Model flexibility**: Test assumptions with likelihood ratio tests
4. **Interpretation differs**: Each model type has different semantic meaning
5. **Choice depends on data**: Select based on assumption testing and context

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Yee, T. W. (2015). *Vector Generalized Linear and Additive Models*
- [VGAM package documentation](https://cran.r-project.org/package=VGAM)
