---
title: "Ordinal Regression: Happiness Data"
subtitle: "Cumulative Link Models for Ordered Outcomes"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script demonstrates **ordinal (cumulative link) regression** on the **happiness dataset**, analyzing **ordered categorical responses**.

Topics covered:
- Cumulative link models (proportional odds model)
- Category-specific effects via `vglm()`
- Data reshaping and aggregation
- Effect coding and contrast matrices

## Data Loading and Exploration

```{r}
#| message: false
library(Fahrmeir)
library(VGAM)
library(dplyr)

# Load the happiness dataset
data(happy)
head(happy)
str(happy)
```

The happy dataset contains:
- **Sex**: Gender (Males/Females)
- **School**: Education level (<12, 12, 13-16, >16)
- **Rep.happiness**: Reported happiness (ordered: Not happy, Pretty happy, Very happy)
- **n**: Frequency counts

## Part 1: School Education Levels by Sex

### Data Aggregation

```{r}
# Reshape data: aggregate school education by sex
happy_by_sex <- happy %>%
  group_by(Sex) %>%
  summarize(
    school_0to12 = sum(n[School == '<12']),
    school_12 = sum(n[School == '12']),
    school_13to16 = sum(n[School == '13-16']),
    school_over16 = sum(n[School == '>16']),
    .groups = 'drop'
  ) %>%
  as.data.frame()

head(happy_by_sex)
```

### Setting Contrasts

For reproduction of textbook results, we use **effect coding**:

```{r}
# Check default contrasts
contrasts(happy_by_sex$Sex)

# Set effect coding
options(contrasts = c("contr.sum", "contr.poly"))
contrasts(happy_by_sex$Sex)

# Adjust to match textbook (female as reference)
contrasts(happy_by_sex$Sex) <- matrix(c(-1, 1), 2, 1)
contrasts(happy_by_sex$Sex)
```

### Cumulative Link Model - Part 1

Fit a proportional odds model for educational attainment:

```{r}
# Create response matrix (school education levels)
response_1 <- cbind(
  happy_by_sex$school_0to12,
  happy_by_sex$school_12,
  happy_by_sex$school_13to16,
  happy_by_sex$school_over16
)

# Fit cumulative link model
model_1 <- vglm(
  response_1 ~ Sex,
  family = cumulative(),
  data = happy_by_sex
)

summary(model_1)
```

### Extracting and Interpreting Coefficients

```{r}
# Get coefficients
coef_1 <- coef(model_1)
print(round(coef_1, 4))

# Predict probabilities
pred_1 <- predict(model_1, type = "response")
head(pred_1)
```

## Part 2: Happiness Levels by Education

### Data Aggregation

```{r}
# Reshape data: aggregate happiness by education level
happy_by_school <- happy %>%
  group_by(School) %>%
  summarize(
    not_happy = sum(n[Rep.happiness == 'Not to happy']),
    pretty_happy = sum(n[Rep.happiness == 'Pretty happy']),
    very_happy = sum(n[Rep.happiness == 'Very happy']),
    .groups = 'drop'
  ) %>%
  as.data.frame()

head(happy_by_school)
```

### Cumulative Link Model - Part 2

```{r}
# Reset to standard dummy coding for education
options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(happy_by_school$School)

# Create response matrix
response_2 <- cbind(
  happy_by_school$not_happy,
  happy_by_school$pretty_happy,
  happy_by_school$very_happy
)

# Fit model
model_2 <- vglm(
  response_2 ~ School,
  family = cumulative(link = "logitlink"),
  data = happy_by_school
)

summary(model_2)
```

### Interpreting Category-Specific Effects

```{r}
# Extract coefficients
coef_2 <- coef(model_2)
cat("Raw coefficients:\n")
print(round(coef_2, 4))

# Reconstruct textbook coefficients
# For each education level, add intercepts
intercepts <- coef_2[1:2]
education_effects <- coef_2[3:8]

# Recreate full coefficient table
full_coef <- c(intercepts, intercepts + education_effects[1:2],
               intercepts + education_effects[3:4],
               intercepts + education_effects[5:6])

cat("\nRecovered coefficients (textbook format):\n")
print(round(full_coef, 4))
```

## Log-Likelihood and Model Fit

Compute the deviance for the combined model:

```{r}
# Individual model log-likelihoods
ll_1 <- logLik(model_1)
ll_2 <- logLik(model_2)

cat("Log-likelihood (Education part):", round(ll_1[1], 4), "\n")
cat("Log-likelihood (Happiness part):", round(ll_2[1], 4), "\n")
cat("Combined GLM log-likelihood:", round(ll_1[1] + ll_2[1], 4), "\n")
```

### Computing Saturated Model Likelihood

For a saturated model with multinomial distribution, the log-likelihood is:

```{r}
# Separate data by sex groups
happy_males <- happy[happy$Sex == "Males", ]
happy_females <- happy[happy$Sex == "Females", ]

# Saturated model likelihood: each observation is its own category
ll_sat_males <- dmultinom(
  happy_males$n,
  prob = happy_males$n / sum(happy_males$n),
  log = TRUE
)

ll_sat_females <- dmultinom(
  happy_females$n,
  prob = happy_females$n / sum(happy_females$n),
  log = TRUE
)

ll_saturated <- ll_sat_males + ll_sat_females

cat("Saturated model log-likelihood:\n")
cat("  Males:", round(ll_sat_males, 4), "\n")
cat("  Females:", round(ll_sat_females, 4), "\n")
cat("  Total:", round(ll_saturated, 4), "\n")
```

### Deviance Calculation

```{r}
# Deviance: 2 * (saturated LL - GLM LL)
glm_ll <- ll_1[1] + ll_2[1]
deviance <- 2 * (ll_saturated - glm_ll)

cat("\nDeviance Calculation:\n")
cat("Saturated log-likelihood:", round(ll_saturated, 4), "\n")
cat("GLM log-likelihood:", round(glm_ll, 4), "\n")
cat("Deviance = 2 * (", round(ll_saturated, 4), " - ", round(glm_ll, 4), ")\n")
cat("Deviance:", round(deviance, 4), "\n")
cat("(Should match Table 3.11 in F&T: 13.27)\n")
```

### Alternative: Using logLik with sum=FALSE

```{r}
# Individual group contributions to log-likelihood
ll_by_group_1 <- logLik(model_1, sum = FALSE)
ll_by_group_2 <- logLik(model_2, sum = FALSE)

cat("\nLog-likelihood by group (Part 1):\n")
print(ll_by_group_1)

cat("\nLog-likelihood by group (Part 2):\n")
print(ll_by_group_2)
```

## Key Learning Points

1. **Ordinal regression**: Designed for ordered categorical outcomes
2. **Proportional odds assumption**: Constant effects across cutpoints
3. **Category-specific effects**: VGAM allows different slopes per category
4. **Effect coding**: Essential for textbook reproducibility
5. **Data reshaping**: Converting long to wide format for multiple responses

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Yee, T. W. (2015). *Vector Generalized Linear and Additive Models*
- [VGAM package documentation](https://cran.r-project.org/package=VGAM)
