---
title: "Ohio Respiratory Data: GEE Analysis"
subtitle: "Generalized Estimating Equations with Repeated Measures"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script implements **Generalized Estimating Equations (GEE)** to analyze the **Ohio respiratory dataset**, which contains **repeated measurements** of respiratory outcomes from children.

Key topics:
- Handling correlated (non-independent) observations
- Different working correlation structures
- Marginal (population-average) inference
- Robust standard errors

## Data Preparation

```{r}
#| message: false
library(gee)
library(geepack)

# Load and explore
data(ohio)
head(ohio, 20)
str(ohio)
```

### Understanding the Data

```{r}
cat("Number of unique subjects:", length(unique(ohio$id)), "\n")
cat("Measurements per subject:\n")
print(table(table(ohio$id)))
cat("Age coding:\n")
cat("  -2 = 7 years\n")
cat("  -1 = 8 years\n")
cat("   0 = 9 years\n")
cat("   1 = 10 years\n")
```

## Variable Coding

Convert variables to factors with appropriate contrasts:

```{r}
# Convert to factors
ohio$age <- as.factor(ohio$age)
ohio$smoke <- as.factor(ohio$smoke)

# Check default contrasts
cat("Default age contrasts:\n")
print(contrasts(ohio$age))
cat("\nDefault smoke contrasts:\n")
print(contrasts(ohio$smoke))

# Set effect coding (for Fahrmeir & Tutz reproducibility)
options(contrasts = c("contr.sum", "contr.poly"))

cat("\nEffect coding for age:\n")
print(contrasts(ohio$age))
cat("\nEffect coding for smoke:\n")
print(contrasts(ohio$smoke))

# Adjust smoke contrast manually
contrasts(ohio$smoke) <- matrix(c(-1, 1), 2, 1)
cat("\nAdjusted smoke contrasts:\n")
print(contrasts(ohio$smoke))
```

## GEE with Independence Structure

Start with independence (no within-subject correlation) as baseline:

```{r}
gee_indep <- gee(
  formula = resp ~ age * smoke,
  data = ohio,
  family = binomial,
  id = id,
  corstr = "independence"
)

summary(gee_indep)$coefficients
```

## GEE with Exchangeable Structure

Assume constant correlation among repeated measures:

```{r}
gee_exch <- gee(
  formula = resp ~ age * smoke,
  data = ohio,
  family = binomial,
  id = id,
  corstr = "exchangeable"
)

summary(gee_exch)$coefficients
```

## GEE with Unstructured Correlation

Allow unique correlations for each pair of time points:

```{r}
gee_unstruct <- gee(
  formula = resp ~ age * smoke,
  data = ohio,
  family = binomial,
  id = id,
  corstr = "unstructured"
)

summary(gee_unstruct)$coefficients
```

## Robust Variance Estimation

GEE provides **robust standard errors** (sandwich estimator):

```{r}
cat("Robust variance matrix (independence structure, first 4x4):\n")
print(round(gee_indep$robust.variance[1:4, 1:4], 4))

cat("\nRobust variance matrix (exchangeable structure, first 4x4):\n")
print(round(gee_exch$robust.variance[1:4, 1:4], 4))
```

## Comparison of Results

```{r}
results <- data.frame(
  Parameter = rownames(summary(gee_indep)$coefficients),
  Indep_Coef = round(gee_indep$coefficients, 4),
  Indep_SE = round(summary(gee_indep)$coefficients[, "Robust S.E."], 4),
  Exch_Coef = round(gee_exch$coefficients, 4),
  Exch_SE = round(summary(gee_exch)$coefficients[, "Robust S.E."], 4),
  Unstruct_Coef = round(gee_unstruct$coefficients, 4),
  Unstruct_SE = round(summary(gee_unstruct)$coefficients[, "Robust S.E."], 4)
)

print(results)
```

## Scale/Dispersion Parameter

Compute the scale parameter using Pearson residuals and method of moments:

```{r}
# Extract Pearson residuals from independence model
pearson_resid <- (ohio$resp - gee_indep$fitted.values) /
                   sqrt(gee_indep$fitted.values * (1 - gee_indep$fitted.values))

# Method of moments dispersion estimator
scale_by_hand <- sum(pearson_resid^2) / (gee_indep$nobs - length(gee_indep$coefficients))

cat("Scale parameter (method of moments):", round(scale_by_hand, 4), "\n")
cat("Scale from summary():", round(summary(gee_indep)$scale, 4), "\n")
```

### Understanding Naive vs Robust Standard Errors

```{r}
# Fit with scale fixed to 1
gee_scale_fixed <- gee(
  formula = resp ~ age * smoke,
  data = ohio,
  family = binomial,
  id = id,
  corstr = "independence",
  scale.fix = TRUE
)

# Verify relationship: Naive SE * sqrt(scale) = Model SE
naive_se_scaled <- summary(gee_scale_fixed)$coefficients[, "Naive S.E."] *
                     sqrt(summary(gee_indep)$scale)

cat("Naive SE (scale.fix=TRUE) * sqrt(scale):\n")
print(round(naive_se_scaled, 4))

cat("\nRobust SE (scale estimated):\n")
print(round(summary(gee_indep)$coefficients[, "Robust S.E."], 4))

cat("\nDifference (should be ~0):\n")
print(round(naive_se_scaled - summary(gee_indep)$coefficients[, "Robust S.E."], 6))
```

## Working Correlation Matrices

Examine the estimated within-subject correlation structures:

```{r}
cat("Independence Structure (identity matrix):\n")
print(round(gee_indep$working.correlation[1:4, 1:4], 4))

cat("\nExchangeable Structure:\n")
print(round(gee_exch$working.correlation[1:4, 1:4], 4))

cat("\nUnstructured (first 4x4):\n")
print(round(gee_unstruct$working.correlation[1:4, 1:4], 4))
```

### Calculating Beta for Age 10

For age=10 (not in the design matrix), compute combined coefficient:

```{r}
# With effect coding, age=10 is minus the sum of the other three age coefficients
age_coefs <- summary(gee_indep)$coefficients[c("age1", "age2", "age3"), 1]
beta_age_10 <- -sum(age_coefs)

cat("Age coefficient components:\n")
print(round(age_coefs, 4))

cat("\nBeta for age=10 (combined):\n")
print(round(beta_age_10, 4))
```

### Standard Error for Combined Effect

```{r}
# Robust SE for combined age=10 coefficient (using contrast matrix)
contrast_matrix <- matrix(rep(1, 3), nrow=1)
variance_subset <- gee_indep$robust.variance[c("age1", "age2", "age3"),
                                             c("age1", "age2", "age3")]

robust_se_age_10 <- sqrt(contrast_matrix %*% variance_subset %*% t(contrast_matrix))

cat("Robust SE for age=10 coefficient:", round(robust_se_age_10[1,1], 4), "\n")
cat("(Matches Fahrmeir & Tutz p.273)\n")
```

## Table 6.9: Models Without Interactions

Now fit the simpler main effects models:

```{r}
# Independence structure
gee_indep_main <- gee(
  formula = resp ~ age + smoke,
  data = ohio,
  family = binomial,
  id = id,
  corstr = "independence"
)

# Exchangeable structure
gee_exch_main <- gee(
  formula = resp ~ age + smoke,
  data = ohio,
  family = binomial,
  id = id,
  corstr = "exchangeable"
)

# Unstructured
gee_unstruct_main <- gee(
  formula = resp ~ age + smoke,
  data = ohio,
  family = binomial,
  id = id,
  corstr = "unstructured"
)

cat("Table 6.9: Main Effects Models\n\n")

cat("Independence Correlation:\n")
print(summary(gee_indep_main)$coefficients)

cat("\nExchangeable Correlation:\n")
print(summary(gee_exch_main)$coefficients)

cat("\nUnstructured Correlation:\n")
print(summary(gee_unstruct_main)$coefficients)
```

### Comparing Robust Variances

```{r}
cat("Robust variance matrices differ by structure:\n\n")

cat("Independence robust variance (first 3x3):\n")
print(round(gee_indep_main$robust.variance[1:3, 1:3], 4))

cat("\nExchangeable robust variance (first 3x3):\n")
print(round(gee_exch_main$robust.variance[1:3, 1:3], 4))

cat("\nUnstructured robust variance (first 3x3):\n")
print(round(gee_unstruct_main$robust.variance[1:3, 1:3], 4))

cat("\nNote: Unlike with interactions, robust variances ARE different across correlation structures\n")
```

## Key Insights from Models

```{r}
cat("GEE Advantages:\n")
cat("1. Handles repeated measures/clustered data\n")
cat("2. Provides robust (sandwich) standard errors\n")
cat("3. Robust to correlation misspecification\n")
cat("4. Population-average (marginal) interpretation\n")
cat("5. No distributional assumptions needed\n")
cat("6. Valid under weaker assumptions than random effects models\n")
```

## When to Use Each Structure

```{r}
cat("Correlation Structure Selection:\n\n")

cat("Independence:\n")
cat("- When observations truly independent\n")
cat("- Baseline/comparison model\n\n")

cat("Exchangeable:\n")
cat("- All pairs equally correlated\n")
cat("- Most common in practice\n")
cat("- Good for clustered or repeated measures\n\n")

cat("Unstructured:\n")
cat("- Maximum flexibility\n")
cat("- Requires estimation of many parameters\n")
cat("- Use when pattern complex or unclear\n\n")

cat("Other structures (AR-1, etc):\n")
cat("- Time-ordered data\n")
cat("- Spatial/longitudinal patterns\n")
```

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Liang, K. Y., & Zeger, S. L. (1986). Longitudinal data analysis using generalized linear models. *Biometrika*.
- Halekoh, U., HĂ¸jsgaard, S., & Yan, J. (2006). The R package geepack. *Journal of Statistical Software*.
