---
title: "GEE Analysis: CTSIB Balance Test Data"
subtitle: "Generalized Estimating Equations for Repeated Measures"
date: today
---

## Overview

This script reproduces **simulations and GEE analysis** on the **CTSIB (Computer Test of Sensory Integration and Balance)** dataset from Faraway (2016, p.292).

Topics covered:
- GEE methodology for balance/stability outcomes
- Binary response modeling for repeated measurements
- Correlation structure selection
- Robust vs model-based standard errors
- Working with the geepack package

## Data Setup

```{r}
#| message: false
library(gee)
library(geepack)
library(faraway)

# Set default contrasts
options(contrasts = c("contr.treatment", "contr.poly"))

# Load and explore the CTSIB dataset
data(ctsib)
?ctsib  # read the description of the data
head(ctsib, 20)  # take a look at the data
str(ctsib)
```

### Creating Binary Response

```{r}
# Create binary response: 1 = completely stable, 0 = otherwise
ctsib$stable <- ifelse(ctsib$CTSIB == 1, 1, 0)
```

The CTSIB dataset contains measurements on balance/stability for subjects under different conditions.

## GEE with Exchangeable Correlation

Fit GEE model with exchangeable (constant) correlation structure and fixed scale:

```{r}
# Fit GEE with scale fixed at 1
gg <- gee(stable ~ Sex + Age + Height + Weight + Surface + Vision,
          id = Subject,
          family = binomial,
          data = ctsib,
          corstr = "exchangeable",
          scale.fix = TRUE)

cat("GEE Coefficients (scale.fix = TRUE):\n")
print(round(gg$coefficients, 4))
```

## GEE with Estimated Scale

Now allow the scale/dispersion to be estimated:

```{r}
# Fit GEE with scale estimated
gg.scale <- gee(stable ~ Sex + Age + Height + Weight + Surface + Vision,
                id = Subject,
                family = binomial,
                data = ctsib,
                corstr = "exchangeable",
                scale.fix = FALSE)

cat("\nGEE Coefficients (scale.fix = FALSE):\n")
print(gg.scale$coefficients)

cat("\nEstimated Scale:\n")
print(gg.scale$scale)
```

## Comparing Scale-Fixed vs Estimated

```{r}
# Compare coefficients
comparison <- data.frame(
  Variable = names(gg$coefficients),
  Scale_Fixed = round(gg$coefficients, 5),
  Scale_Estimated = round(gg.scale$coefficients, 5)
)
print(comparison)

cat("\nScale parameter difference:\n")
cat("When scale.fix=TRUE: scale = 1\n")
cat("When scale.fix=FALSE: scale =", round(gg.scale$scale, 4), "\n")
```

## Understanding the GEE Approach

The GEE method:
1. Estimates regression coefficients from marginal model
2. Provides robust standard errors via sandwich estimator
3. Handles correlated observations within subjects
4. Does not require specifying correlation structure correctly (though efficiency may vary)

```{r}
cat("GEE Results Summary (scale.fix=TRUE):\n")
cat("Number of coefficients:", length(gg$coefficients), "\n")
cat("Exchangeable correlation structure\n")
```

## Alternative: geepack Package

The geepack package provides `geeglm()` as an alternative interface to GEE. Note: geepack uses a different dispersion estimator based on the original Yan and Fine paper.

```{r}
# Uncomment to use geepack:geeglm
# mod_geepack <- geeglm(stable ~ Sex + Age + Height + Weight + Surface + Vision,
#                       id = Subject,
#                       corstr = "exchangeable",
#                       scale.fix = TRUE,
#                       data = ctsib,
#                       family = binomial)
#
# print(mod_geepack$coefficients)
# summary(mod_geepack)
# summary(mod_geepack)$scale.value
# summary(mod_geepack)$dispersion
```

## Key Insights

1. **Scale parameter**: Reflects overdispersion/underdispersion in the data
2. **Robust variance**: The sandwich estimator is more reliable than model-based variance
3. **Correlation structure**: Different structures may give similar coefficients but different standard errors
4. **Independence assumption**: Often too restrictive for repeated measures data

## Key Learning Points

1. **GEE advantages**: Provides population-average (marginal) effects, robust to correlation misspecification
2. **Working correlation**: Used for efficiency but doesn't need to be correct for valid inference
3. **Sandwich variance**: Makes inference valid even with model misspecification
4. **Scale parameter**: Important for understanding data overdispersion

## References

- Faraway, J. J. (2016). *Extending the Linear Model with R* (2nd ed.), p.292
- Liang, K. Y., & Zeger, S. L. (1986). Longitudinal data analysis using generalized linear models
- Yan, J. & Fine, J. (2004). The R package geepack: GEE estimation of generalized estimating equations
- [gee package documentation](https://cran.r-project.org/package=gee)
- [geepack package documentation](https://cran.r-project.org/package=geepack)
