---
title: "GLM with Mixed Effects"
subtitle: "Binomial GLMM on Ohio Respiratory Data"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script demonstrates **Generalized Linear Mixed Models (GLMM)** for handling **subject-specific random effects** in the Ohio respiratory data.

Methods covered:
- Laplace approximation via `glmer()`
- Adaptive Gauss-Hermite quadrature (AGHQ) for improved accuracy
- Penalized Quasi-Likelihood (PQL) via `glmmPQL()`
- Random effects extraction and interpretation

## Comparing GLMM Estimation Methods

Different packages and methods provide different accuracy-computation tradeoffs:

```{r}
#| message: false
library(lme4)
library(geepack)
library(MASS)

# Load and prepare data
data(ohio)
ohio$age <- as.factor(ohio$age)
ohio$smoke <- as.factor(ohio$smoke)

# Set effect coding
options(contrasts = c("contr.sum", "contr.poly"))
contrasts(ohio$smoke) <- matrix(c(-1, 1), 2, 1)
```

## GLMM via Laplace Approximation

The default `glmer()` approach uses the **Laplace approximation**:

```{r}
glmm_laplace <- glmer(
  resp ~ age * smoke + (1 | id),
  data = ohio,
  family = binomial,
  control = glmerControl(optimizer = "Nelder_Mead")
)

summary(glmm_laplace)
```

## GLMM via Adaptive Gauss-Hermite Quadrature

More accurate numerical integration with `nAGQ`:

```{r}
# AGHQ with 10 quadrature points
glmm_aghq10 <- glmer(
  resp ~ age * smoke + (1 | id),
  data = ohio,
  family = binomial,
  control = glmerControl(optimizer = "Nelder_Mead"),
  nAGQ = 10
)

summary(glmm_aghq10)
```

### Comparison: Laplace vs AGHQ

```{r}
comparison <- data.frame(
  Parameter = names(fixef(glmm_laplace)),
  Laplace = round(fixef(glmm_laplace), 4),
  AGHQ_10 = round(fixef(glmm_aghq10), 4)
)
print(comparison)

cat("\nRandom effect SD:\n")
cat("Laplace:", round(attr(VarCorr(glmm_laplace)$id, "stddev"), 4), "\n")
cat("AGHQ_10:", round(attr(VarCorr(glmm_aghq10)$id, "stddev"), 4), "\n")
```

## Random Effects

Extract and examine the estimated random effects (conditional modes):

```{r}
# Get random effects for each subject
ranef_laplace <- ranef(glmm_laplace)$id
ranef_aghq <- ranef(glmm_aghq10)$id

head(ranef_laplace)
head(ranef_aghq)

# Summary statistics
cat("Laplace - Random Effects Summary:\n")
print(summary(ranef_laplace[, 1]))

cat("\nAGHQ - Random Effects Summary:\n")
print(summary(ranef_aghq[, 1]))
```

## Alternative: Penalized Quasi-Likelihood

`glmmPQL()` from MASS provides a faster (though less accurate) alternative:

```{r}
glmm_pql <- glmmPQL(
  resp ~ age * smoke,
  random = ~ 1 | id,
  family = binomial,
  data = ohio
)

summary(glmm_pql)
```

## Comparing Estimation Methods

```{r}
# Extract coefficients from all methods
coefs_comparison <- data.frame(
  Method = c("Laplace", "AGHQ_10", "PQL"),
  Intercept = c(
    fixef(glmm_laplace)[1],
    fixef(glmm_aghq10)[1],
    fixef(glmm_pql)[1]
  ),
  Age_Effect = c(
    fixef(glmm_laplace)[2],
    fixef(glmm_aghq10)[2],
    fixef(glmm_pql)[2]
  )
)

print(round(coefs_comparison, 4))
```

## Wine Bitterness: Ordinal Mixed Models

Now we demonstrate **ordinal regression with random effects** on the wine tasting dataset:

### Data Setup

```{r}
library(ordinal)

# Load wine dataset
head(wine)
str(wine)

# Set contrasts
options(contrasts = c("contr.treatment", "contr.poly"))

# Verify class and contrasts
cat("Contrast for contact:\n")
print(contrasts(wine$contact))

cat("\nContrast for temp:\n")
print(contrasts(wine$temp))

# Adjust contrasts as in raw script
contrasts(wine$contact) <- c(0, 1)
contrasts(wine$temp) <- c(1, 0)

cat("\nAdjusted contact contrasts:\n")
print(contrasts(wine$contact))

cat("Adjusted temp contrasts:\n")
print(contrasts(wine$temp))
```

### Fixed Effects Ordinal Model: polr()

Fit using MASS::polr() (proportional odds logistic regression):

```{r}
# Fit fixed effects ordinal model
wine_fit_polr <- MASS::polr(
  rating ~ temp + contact,
  data = wine,
  Hess = TRUE,
  method = "logistic"
)

summary(wine_fit_polr)
```

### Fixed Effects Ordinal Model: clm()

Alternative from the ordinal package:

```{r}
# Fit using ordinal::clm()
wine_fit_clm <- clm(
  rating ~ temp + contact,
  data = wine
)

summary(wine_fit_clm)

cat("\nExtracted coefficients:\n")
print(wine_fit_clm$coefficients)
```

### Comparing polr() vs clm()

```{r}
# Compare coefficients
comparison_ordinal <- data.frame(
  Parameter = names(coef(wine_fit_polr)),
  polr_coef = round(coef(wine_fit_polr), 4),
  clm_coef = round(coef(wine_fit_clm), 4)
)

print(comparison_ordinal)

# Compare log-likelihoods
cat("Log-likelihood (polr):", round(logLik(wine_fit_polr), 4), "\n")
cat("Log-likelihood (clm):", round(logLik(wine_fit_clm), 4), "\n")
```

### Random Effects: clmm()

Fit ordinal mixed model with judge random effects:

```{r}
# Fit ordinal mixed model with random judge effect
wine_fit_clmm <- clmm(
  rating ~ temp + contact + (1 | judge),
  data = wine,
  nAGQ = 10
)

summary(wine_fit_clmm)
```

### Comparing Fixed vs Mixed Effects

```{r}
# Extract and compare coefficients
cat("Fixed effects (polr):\n")
print(round(coef(wine_fit_polr), 4))

cat("\nMixed effects (clmm):\n")
print(round(clmm$beta, 4))

# Get fitted values
fitted_clmm <- wine_fit_clmm$fitted.values
cat("\nFitted probabilities (first 10):\n")
print(round(head(fitted_clmm, 10), 3))
```

### Random Effects for Judges

```{r}
# Extract random judge effects
ranef_judges <- ranef(wine_fit_clmm)$judge

cat("Random effects for each judge:\n")
print(round(ranef_judges, 4))

# Interpretation
cat("\nInterpretation:\n")
cat("Positive value: Judge tends to give higher ratings\n")
cat("Negative value: Judge tends to give lower ratings\n")
```

## Model Features

**Fixed effects interpretation:**
- Subject-specific (conditional) estimates
- Different from population-average (marginal) effects in GEE

**Random intercept variance:**
- Between-subject heterogeneity
- Accounts for repeated measurements

**Trade-offs:**
- **Laplace**: Fast, reasonable accuracy
- **AGHQ**: Slower, higher accuracy
- **PQL**: Fastest, lower accuracy

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Faraway, J. J. (2016). *Extending the Linear Model with R* (2nd ed.)
- [lme4 documentation](https://cran.r-project.org/package=lme4)
