---
title: "GLM with Mixed Effects"
subtitle: "Binomial GLMM on Ohio Respiratory Data and Ordinal Mixed Models"
date: today
---

## Overview

This script demonstrates **Generalized Linear Mixed Models (GLMM)** for handling **subject-specific random effects** in longitudinal and ordinal data.

Methods covered:
- Laplace approximation via `glmer()` (lme4)
- Adaptive Gauss-Hermite quadrature (AGHQ) for improved accuracy
- Penalized Quasi-Likelihood (PQL) via `glmmPQL()` (MASS)
- Ordinal mixed models via `clmm()` (ordinal)

### Available GLMM Software Comparison

Many R packages exist for fitting GLMMs, using different estimation methods:

- **MASS::glmmPQL**: Penalized quasi-likelihood
- **lme4::glmer**: Laplace approximation and adaptive Gauss-Hermite quadrature (AGHQ)
- **MCMCglmm**: Markov chain Monte Carlo
- **glmmML**: AGHQ
- **glmmADMB**: Laplace
- **glmm** (from Jim Lindsey's repeated package): AGHQ
- **ordinal::clmm**: For ordinal outcomes with random effects
- **gamlss.mx**, **ASREML-R**, **sabreR**, **mcemGLM**: Other specialized approaches

We'll focus on the most commonly used: `glmer()`, `glmmPQL()`, and `clmm()`.

## Ohio Respiratory Data Setup

Load required packages and prepare the Ohio respiratory dataset:

```{r}
#| message: false
library(conflicted)
library(lme4)
library(geepack)
library(ordinal)
library(MASS)
```

Load and examine the data structure:

```{r}
head(ohio)

# Check current variable classes
class(ohio$resp)
class(ohio$age)
class(ohio$smoke)
```

Convert age and smoke to factors, and set up effect coding to match Fahrmeir & Tutz:

```{r}
ohio$age <- as.factor(ohio$age)
ohio$smoke <- as.factor(ohio$smoke)

# Check default contrasts
contrasts(ohio$age)
contrasts(ohio$smoke)
```

Set effect coding (Fahrmeir & Tutz use this parameterization):

```{r}
options(contrasts = c("contr.sum", "contr.poly"))

contrasts(ohio$age)
contrasts(ohio$smoke)

# Adjust smoke contrast to match book
contrasts(ohio$smoke) <- matrix(c(-1, 1), 2, 1)
contrasts(ohio$smoke)
```

## GLMM via Laplace Approximation

The default `glmer()` approach uses the **Laplace approximation**, which is fast and reasonably accurate. The Nelder-Mead optimizer is recommended by Faraway (2016):

```{r}
glmer_fit <- glmer(
    resp ~ age * smoke + (1 | id),
    data = ohio,
    family = binomial,
    control = glmerControl(optimizer = "Nelder_Mead")
)

summary(glmer_fit)
```

## GLMM via Adaptive Gauss-Hermite Quadrature

More accurate numerical integration with `nAGQ`. Using 10 quadrature points closely reproduces the numbers in Table 7.2 of Fahrmeir & Tutz:

```{r}
glmer_fit_n_agq_10 <- glmer(
    resp ~ age * smoke + (1 | id),
    data = ohio,
    family = binomial,
    control = glmerControl(optimizer = "Nelder_Mead"),
    nAGQ = 10
)

summary(glmer_fit_n_agq_10)
```

"Estimate" the random effects, which are conditional modes as indicated by the documentation of `ranef()`:

```{r}
head(ranef(glmer_fit))
```

## Alternative: Penalized Quasi-Likelihood

This is an alternative (and faster) method for fitting GLMM that doesn't require numerical integration. See Chapter 13 in Faraway (2016) for more details. This approach is based on penalized quasi-likelihood:

```{r}
glmm_pql_fit <- glmmPQL(
    resp ~ age * smoke,
    random = ~ 1 | id,
    family = binomial,
    data = ohio
)

summary(glmm_pql_fit)
```

This should use the version of `ranef()` from the nlme package:

```{r}
head(ranef(glmm_pql_fit))
```

## Wine Bitterness: Ordinal Mixed Models

Now we demonstrate **ordinal regression with random effects** on the wine tasting dataset from the ordinal package.

### Data Setup and Contrast Coding

```{r}
head(wine)

# Check variable classes
class(wine$bottle)
class(wine$rating)
class(wine$judge)
class(wine$temp)
class(wine$contact)
```

Reset contrasts to treatment coding:

```{r}
options(contrasts = c("contr.treatment", "contr.poly"))

# Check default contrasts
contrasts(wine$contact)
contrasts(wine$temp)
```

Adjust contrasts to specific coding scheme:

```{r}
contrasts(wine$contact) <- c(0, 1)
contrasts(wine$temp) <- c(1, 0)

contrasts(wine$contact)
contrasts(wine$temp)
```

### Fixed Effects Ordinal Model: polr()

Fit using `MASS::polr()` for comparison with the ordinal package:

```{r}
wine_fixed_fit_polr <- polr(
    rating ~ temp + contact,
    data = wine,
    Hess = TRUE,
    method = "logistic"
)

summary(wine_fixed_fit_polr)
```

### Fixed Effects Ordinal Model: clm()

Alternative fixed-effects fit using the ordinal package:

```{r}
wine_fixed_fit_clm <- clm(rating ~ temp + contact, data = wine)

summary(wine_fixed_fit_clm)
wine_fixed_fit_clm$coefficients
```

### Comparing with Fahrmeir & Tutz Book Values

The book reports thresholds computed via specific spacing. Let's check:

```{r}
book_coeff <- c(
    -5.289,
    -5.289 + exp(0.974),
    -5.289 + exp(0.974) + exp(0.739),
    -5.289 + exp(0.974) + exp(0.739) + exp(0.426)
)
book_coeff
```

::: {.callout-note}
The reported log-likelihood is close to Table 7.4 in Fahrmeir & Tutz, but the estimated coefficients are not exactly the same. This may be due to:

- Different parameterizations or contrast coding
- Software implementation differences
- Rounding or convergence tolerances

Using the Fahrmeir package dataset directly doesn't fully resolve the discrepancy either.
:::

### Random Effects: clmm()

Fit ordinal mixed model with judge random effects using 10 AGHQ quadrature points:

```{r}
wine_mixed_fit <- clmm(
    rating ~ temp + contact + (1 | judge),
    data = wine,
    nAGQ = 10
)

summary(wine_mixed_fit)
```

### Results Interpretation

The mixed model results are consistent in the sense that the fixed-effect coefficients (beta's) have larger magnitudes when accounting for judge-specific random effects.

```{r}
# Fitted values
head(wine_mixed_fit$fitted.values)

# Contrast structure
wine_mixed_fit$contrasts
```

Extract judge random effects (note: `ranef()` here comes from the ordinal package):

```{r}
ranef(wine_mixed_fit)
```

## Key Learning Points

1. **Subject-specific vs. population-average**: GLMM coefficients are conditional (subject-specific), differing from GEE marginal effects
2. **Estimation methods**:
   - **Laplace**: Fast, reasonable accuracy
   - **AGHQ**: Slower, higher accuracy (more quadrature points = better approximation)
   - **PQL**: Fastest, lower accuracy (avoids integration entirely)
3. **Random effects**: Represent between-subject heterogeneity; extracted as conditional modes
4. **Ordinal mixed models**: Use `clmm()` for cumulative link models with random effects
5. **Contrast coding**: Effect coding vs. dummy coding changes coefficient interpretation but not model fit

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling Based on Generalized Linear Models* (2nd ed.)
- Faraway, J. J. (2016). *Extending the Linear Model with R* (2nd ed.), Chapter 13
- Venables, W. N., & Ripley, B. D. (2002). Modern applied statistics with S (4th ed.)
- [lme4 documentation](https://cran.r-project.org/package=lme4)
- [ordinal package documentation](https://cran.r-project.org/package=ordinal)
