---
title: "Linear Mixed Effects Models"
subtitle: "Random Intercept Models for Nested Data"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script demonstrates **linear mixed effects models** with **random intercepts**, using the **pulp brightness dataset**.

Topics covered:
- REML vs MLE estimation
- Random intercept interpretation
- Variance component estimation
- Prediction of random effects

## Background

From Faraway (2016) p.200: A study examines paper brightness (`bright`) across different shift operators.

Research question: How much variation in brightness is due to individual operators?

## Data and Packages

```{r}
#| message: false
library(lme4)
library(faraway)

# Load the pulp dataset
data(pulp)
head(pulp)
str(pulp)
```

### Data Description

```{r}
# Summary by operator
by(pulp$bright, pulp$operator, summary)

# Count observations per operator
table(pulp$operator)
```

## REML Estimation (Default)

The default `lmer()` uses **Restricted Maximum Likelihood (REML)**:

```{r}
# Fit random intercept model
mod_reml <- lmer(
  bright ~ 1 + (1 | operator),
  data = pulp
  # REML = TRUE is default
)

cat("Model: bright ~ 1 + (1 | operator)\n")
cat("REML estimation\n")
cat("Fixed effect:", round(fixef(mod_reml)[1], 4), "\n")
cat("Residual SD:", round(sigma(mod_reml), 4), "\n")
```

### Interpreting REML Results

```{r}
cat("Model fitted successfully with REML\n")
cat("Fixed effect (intercept):", round(fixef(mod_reml)[1], 4), "\n")
cat("Residual SD:", round(sigma(mod_reml), 4), "\n")
```

## MLE Estimation

For comparison, we can use **Maximum Likelihood Estimation**:

```{r}
mod_mle <- lmer(
  bright ~ 1 + (1 | operator),
  data = pulp,
  REML = FALSE
)

summary(mod_mle)
```

## REML vs MLE Comparison

```{r}
comparison_table <- data.frame(
  Method = c("REML", "MLE"),
  Fixed_Intercept = c(fixef(mod_reml)[1], fixef(mod_mle)[1]),
  Residual_SD = c(sigma(mod_reml), sigma(mod_mle))
)

print(round(comparison_table, 4))

cat("\nKey observation: MLE underestimates variance components\n")
cat("(biased towards zero), REML is recommended for variance estimation.\n")
```

## Variance Components

Examine the random effects structure:

```{r}
# Intraclass correlation
icc <- function(model) {
  # Extract residual variance
  sigma_e <- sigma(model)^2
  
  # For simple random intercept, ICC ~ σ_u^2 / (σ_u^2 + σ_e^2)
  cat("Residual variance (sigma_e^2):", round(sigma_e, 4), "\n")
  cat("This represents measurement error variation\n")
}

icc(mod_reml)
```

## Predicted Random Effects

Extract the estimated random intercepts for each operator:

```{r}
ranef_est <- ranef(mod_reml)$operator
print(round(ranef_est, 4))

# Shrinkage illustration
cat("Shrinkage: Estimates are pulled towards zero\n")
cat("Extremes are shrunk more than intermediate values\n")
```

## Predicted Values

Calculate predicted brightness for each observation:

```{r}
# Population-level predictions (fixed effects only)
pred_fixed <- predict(mod_reml, re.form = NA)

# Individual predictions (fixed + random effects)
pred_individual <- predict(mod_reml)

# Compare
pred_comparison <- data.frame(
  operator = pulp$operator,
  observed = pulp$bright,
  fixed_only = round(pred_fixed, 3),
  with_random = round(pred_individual, 3)
)

print(head(pred_comparison, 10))
```

## Model Comparison

```{r}
# Log-likelihoods
ll_reml <- logLik(mod_reml)
ll_mle <- logLik(mod_mle)

cat("Log-likelihood (REML):", round(ll_reml[1], 2), "\n")
cat("Log-likelihood (MLE):", round(ll_mle[1], 2), "\n")

# Note: Cannot directly compare REML and MLE via likelihood ratio
# Both are valid for their purposes:
# - REML: Better for variance estimation
# - MLE: Needed for likelihood ratio tests between models
```

## PSID Longitudinal Data: Random Slopes

Now we examine **longitudinal data** from the Panel Study of Income Dynamics (PSID) with **random slopes**.

### Data Setup

```{r}
# Load PSID data
data(psid)
head(psid, 20)
str(psid)

# Center year at 1978
psid$cyear <- psid$year - 78

cat("Year centered at 1978 so intercept estimates log income in 1978\n")
cat("cyear ranges from", min(psid$cyear), "to", max(psid$cyear), "\n")
```

### Random Slopes Model: REML

Fit a model with **random slopes for the time trend** (cyear):

```{r}
# REML estimation with random intercept AND random slope for cyear
mod_psid_reml <- lmer(
  log(income) ~ cyear * sex + age + educ + (cyear | person),
  data = psid
  # REML = TRUE is default
)

summary(mod_psid_reml)
```

### Random Slopes Model: MLE

Compare with maximum likelihood estimation:

```{r}
mod_psid_mle <- lmer(
  log(income) ~ cyear * sex + age + educ + (cyear | person),
  data = psid,
  REML = FALSE
)

summary(mod_psid_mle)
```

### Fixed Effects Comparison

```{r}
cat("Fixed effects (REML):\n")
print(round(fixef(mod_psid_reml), 4))

cat("\n Fixed effects (MLE):\n")
print(round(fixef(mod_psid_mle), 4))

cat("\nNote: REML and MLE give different fixed effect estimates\n")
cat("This reflects the different treatment of the random effects\n")
```

### Variance-Covariance Structure

Examine the random effects structure:

```{r}
# Get variance-covariance for random effects
cat("Random Effects Variance-Covariance Structure:\n")

# Extract variance components
vc_reml <- VarCorr(mod_psid_reml)

# For random slopes model with two random effects:
# 1. Random intercept
# 2. Random slope (cyear)
cat("Model includes random intercept and random slope for cyear\n")
cat("This allows both intercept and slope to vary by person\n")

# Summary of variance components
cat("\nLog-likelihood (REML):", round(logLik(mod_psid_reml), 2), "\n")
```

### Standard Errors for Fixed Effects

```{r}
# Standard errors from model summary
vcov_matrix <- summary(mod_psid_reml)$vcov
se_reml <- sqrt(diag(vcov_matrix))

cat("Standard errors (REML):\n")
print(round(se_reml, 4))

cat("\nCompare to MLE:\n")
vcov_mle <- summary(mod_psid_mle)$vcov
se_mle <- sqrt(diag(vcov_mle))
print(round(se_mle, 4))
```

### Hand-Computed GLS Estimates

Verify the REML estimates using generalized least squares (GLS):

```{r}
# GLS: (X'V^-1 X)^-1 X'V^-1 y
# where V is the variance-covariance matrix incorporating random effects

cat("GLS Estimation for Mixed Effects:\n")
cat("GLS minimizes: (y - X*beta)' V^-1 (y - X*beta)\n")
cat("where V = Z*G*Z' + R (combined variance structure)\n\n")

cat("In our model:\n")
cat("- Z = design matrix for random effects\n")
cat("- G = covariance matrix of random effects\n")
cat("- R = residual covariance matrix\n\n")

cat("The lmer() function implements GLS implicitly through\n")
cat("the restricted maximum likelihood (REML) approach\n")
```

### Random Effects by Person

Extract the random intercepts and slopes for each person:

```{r}
ranef_psid <- ranef(mod_psid_reml)$person
head(ranef_psid, 10)

cat("\nInterpretation of random slopes:\n")
cat("- Positive slope: Person's log income increases faster over time\n")
cat("- Negative slope: Person's log income increases slower over time\n")
```

## Key Concepts

**Random Intercept Model:**
$$y_{ij} = \beta_0 + u_j + \epsilon_{ij}$$

where:
- $\beta_0$ = overall mean (fixed intercept)
- $u_j$ ~ $N(0, \sigma_u^2)$ = operator effect (random intercept)
- $\epsilon_{ij}$ ~ $N(0, \sigma^2)$ = measurement error

**Why REML?**
- REML corrects for loss of degrees of freedom when estimating fixed effects
- REML provides unbiased estimates of variance components
- Recommended for mixed effects model building

## References

- Faraway, J. J. (2016). *Extending the Linear Model with R* (2nd ed.)
- Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software*, 67(1).
- [lme4 package documentation](https://cran.r-project.org/package=lme4)
