---
title: "Survival Analysis"
subtitle: "Kaplan-Meier Estimation and Weibull Models"
date: today
format:
  html:
    code-fold: false
---

## Overview

This script demonstrates **survival analysis** techniques including:

- **Kaplan-Meier (KM) estimator** for survival functions
- Handling **censored data**
- **Weibull distribution** suitability testing
- Comparison of survival curves between groups

## Background

The dataset contains breast cancer survival data with:
- **time**: Follow-up time (in months or years)
- **status**: Event indicator (0=censored, 1=event/death)
- **group**: Staining status (P=positive, N=negative)

## Data Loading

```{r}
#| message: false
library(survival)

# Load the breast cancer survival dataset
bcancer <- read.csv("data/survival.csv")

head(bcancer)
tail(bcancer)
```

### Understanding Censoring

```{r}
str(bcancer)
summary(bcancer)

# Check censoring rates
cat("Event indicator (status):\n")
table(bcancer$status)

cat("\nCensoring rate:",
    round(100 * sum(bcancer$status == 0) / nrow(bcancer), 1), "%\n")

cat("\nEvents by group:\n")
table(bcancer$status, bcancer$group)
```

## Kaplan-Meier Estimation

### Fitting the KM Estimator

```{r}
# Fit KM estimator stratified by group
km_fit <- survfit(Surv(time, status) ~ group, data = bcancer)

summary(km_fit)
```

### Interpreting KM Output

```{r}
cat("Survival object details:\n")
print(km_fit)

cat("\nSummary by group:\n")
summary(km_fit)

# Alternative: Direct summary table
km_summary <- summary(km_fit)
cat("\nNumber of observations by group:\n")
print(table(bcancer$group))
```

## Survival Curves

### Basic KM Plot

```{r}
# Set graphical parameters for better visualization
orig_par <- par(mar = c(4, 4, 2, 1), las = 1)

# Plot KM curves
plot(km_fit,
     lwd = 2,
     cex.axis = 1.2,
     xlab = "Time",
     ylab = "Survival Probability",
     main = "Kaplan-Meier Survival Curves by Staining Group"
)

# Add legend
legend("bottomleft",
       legend = names(km_fit$strata),
       lty = c(1, 1),
       col = c(1, 2),
       bty = "n")

par(orig_par)  # Restore original parameters
```

### Cumulative Survival Summary

```{r}
# Extract survival probabilities at specific timepoints
survprob_summary <- summary(km_fit)$surv
time_summary <- summary(km_fit)$time

cat("Survival probabilities over time:\n")
print(cbind(Time = time_summary, Survival = round(survprob_summary, 3)))
```

## Weibull Model Assessment

### Checking Weibull Suitability

The Weibull distribution is suitable if the log-log survival plot is approximately linear:

$$\log(-\log(S(t))) = \alpha + \beta \log(t)$$

```{r}
# Prepare data for Weibull assessment
negative_group <- bcancer$group == "N"
positive_group <- bcancer$group == "P"

# Set up plot
opar <- par(mar = c(4, 4, 2, 0), las = 1)

plot(log(km_fit$time), log(-log(km_fit$surv)),
     type = "n",
     xlab = "Log(Time)",
     ylab = "Log(-Log(Survival))",
     main = "Weibull Assessment"
)

# Add lines for each group
lines(log(km_fit$time[negative_group]),
      log(-log(km_fit$surv[negative_group])),
      col = "darkgreen",
      type = "b",
      pch = 16,
      lwd = 2)

lines(log(km_fit$time[positive_group]),
      log(-log(km_fit$surv[positive_group])),
      col = "purple",
      type = "b",
      pch = 17,
      lwd = 2)

# Add legend
legend("topleft",
       legend = c("Negative (N)", "Positive (P)"),
       col = c("darkgreen", "purple"),
       pch = c(16, 17),
       lty = 1,
       bty = "n")

par(opar)
```

### Interpreting Weibull Plot

```{r}
cat("Weibull Model Assessment:\n")
cat("- If plot is approximately LINEAR: Weibull is appropriate\n")
cat("- If plot is CURVED: May need other distributions (lognormal, etc)\n")
cat("- PARALLEL lines: Suggest proportional hazards assumption\n")
```

## Survival Statistics

### Median Survival Times

```{r}
cat("Median survival times by group:\n")
print(summary(km_fit)$table[, "median"])

# Survival at specific timepoints
cat("\n\nSurvival at 50% follow-up time:\n")
```

## Proportional Hazards Assumption: Cloglog Plot

The complementary log-log (cloglog) plot tests the **proportional hazards assumption**:
- Parallel lines in cloglog space indicate proportional hazards
- Intersecting or crossing lines suggest violation

```{r}
# Cloglog plot for PH assumption
opar <- par(mar = c(4, 4, 2, 1), las = 1)

plot(km_fit,
     fun = "cloglog",
     lty = 1:2,
     col = c("Red", "Blue"),
     xlim = c(5, 300),
     xlab = "Time (log scale)",
     ylab = "Log(-Log(Survival))",
     main = "Cloglog Plot: Testing Proportional Hazards"
)

legend("topleft",
       legend = c("Negative Stain", "Positive Stain"),
       lty = 1:2,
       col = c("Red", "Blue"),
       bty = "n")

par(opar)

cat("Proportional Hazards Assessment:\n")
cat("- PARALLEL lines: PH assumption satisfied\n")
cat("- CROSSING/INTERSECTING lines: PH assumption violated\n")
```

## Exponential Regression

### Fitting Exponential Survival Model

The exponential distribution has constant hazard rate:

$$h(t) = \lambda, \quad S(t) = \exp(-\lambda t)$$

```{r}
# Fit exponential regression model
exp_model <- survreg(
  Surv(time, status) ~ group,
  data = bcancer,
  dist = "exponential"
)

summary(exp_model)
```

### Wald Test for Group Effect

```{r}
# Extract coefficient and variance
beta_group <- exp_model$coefficients[2]
var_beta <- exp_model$var[2, 2]

# Wald test statistic
wald_stat <- beta_group / sqrt(var_beta)
p_value_wald <- 2 * pnorm(-abs(wald_stat))

cat("Wald Test for Group Effect:\n")
cat("Coefficient:", round(beta_group, 4), "\n")
cat("Std. Error:", round(sqrt(var_beta), 4), "\n")
cat("Z-statistic:", round(wald_stat, 4), "\n")
cat("p-value:", round(p_value_wald, 4), "\n")
```

### Likelihood Ratio Test

```{r}
# LR test: compare log-likelihoods
ll_full <- exp_model$loglik[2]  # Model with group
ll_intercept <- exp_model$loglik[1]  # Intercept only

lr_stat <- -2 * (ll_intercept - ll_full)
p_value_lr <- pchisq(lr_stat, df = 1, lower.tail = FALSE)

cat("\nLikelihood Ratio Test:\n")
cat("Log-likelihood (intercept only):", round(ll_intercept, 4), "\n")
cat("Log-likelihood (with group):", round(ll_full, 4), "\n")
cat("LR statistic:", round(lr_stat, 4), "\n")
cat("p-value:", round(p_value_lr, 4), "\n")
```

### Interpreting Exponential Model

```{r}
# Exponential rate parameters
intercept_coef <- exp_model$coefficients[1]
group_coef <- exp_model$coefficients[2]

# Lambda (rate) for each group
lambda_neg <- exp(-intercept_coef)
lambda_pos <- exp(-(intercept_coef + group_coef))

cat("Rate Parameters (λ):\n")
cat("Negative group (N):", round(lambda_neg, 6), "\n")
cat("Positive group (P):", round(lambda_pos, 6), "\n")
cat("Ratio (P/N):", round(lambda_pos / lambda_neg, 4), "\n")
```

### Verification: MLE Formula

For exponential distribution, the MLE of λ is:
$$\hat{\lambda} = \frac{\text{# events}}{\text{total time at risk}}$$

```{r}
# Compute MLEs by hand for each group
events_neg <- sum(bcancer$status[bcancer$group == "N"])
time_neg <- sum(bcancer$time[bcancer$group == "N"])
lambda_neg_manual <- events_neg / time_neg

events_pos <- sum(bcancer$status[bcancer$group == "P"])
time_pos <- sum(bcancer$time[bcancer$group == "P"])
lambda_pos_manual <- events_pos / time_pos

cat("Manual MLE Calculation:\n")
cat("Negative: ", events_neg, "events /", time_neg, "time =", round(lambda_neg_manual, 6), "\n")
cat("Positive: ", events_pos, "events /", time_pos, "time =", round(lambda_pos_manual, 6), "\n")

cat("\nComparison with survreg():\n")
cat("Negative - survreg:", round(lambda_neg, 6), ", manual:", round(lambda_neg_manual, 6), "\n")
cat("Positive - survreg:", round(lambda_pos, 6), ", manual:", round(lambda_pos_manual, 6), "\n")
```

### Confidence Intervals for Rate Parameters

Using the delta method:

```{r}
# Delta method for exp(-β)
beta <- exp_model$coefficients[2]
var_beta_full <- exp_model$var[2, 2]

# Variance of exp(-β)
var_lambda_pos <- (exp(-beta))^2 * var_beta_full

# CI using delta method
ci_lower_delta <- exp(-beta) - qnorm(0.975) * sqrt(var_lambda_pos)
ci_upper_delta <- exp(-beta) + qnorm(0.975) * sqrt(var_lambda_pos)

cat("Delta Method CI for Positive Group Rate:\n")
cat("Point estimate:", round(exp(-beta), 6), "\n")
cat("95% CI: (", round(ci_lower_delta, 6), ",", round(ci_upper_delta, 6), ")\n")

# Alternative: Using transformation (monotone function)
beta_coef <- exp_model$coefficients[2]
var_beta_coef <- exp_model$var[2, 2]

ci_lower_trans <- exp(-(beta_coef + qnorm(0.975) * sqrt(var_beta_coef)))
ci_upper_trans <- exp(-(beta_coef - qnorm(0.975) * sqrt(var_beta_coef)))

cat("\nTransformation Method CI:\n")
cat("95% CI: (", round(ci_lower_trans, 6), ",", round(ci_upper_trans, 6), ")\n")

cat("\nNote: Transformation method avoids negative CI bounds\n")
```

## Key Concepts

**Kaplan-Meier Estimator:**
- Non-parametric estimate of survival function
- Handles censored data properly
- Step function (decreases at event times only)

**Censoring:**
- Right censoring: Follow-up ended before event
- Event indicator (status) identifies censored vs uncensored

**Weibull Assumption:**
- Parametric model for survival times
- Constant hazard ratio (proportional hazards)
- Log-linear relationship between log time and log-log survival

## Formal Testing

```{r}
# Log-rank test for survival difference
logrank_test <- survdiff(Surv(time, status) ~ group, data = bcancer)
print(logrank_test)
```

## References

- Fahrmeir, L. & Tutz, G. (2001). *Multivariate Statistical Modelling*
- Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data*
- [R survival package documentation](https://cran.r-project.org/package=survival)
