---
title: "Survival Analysis"
subtitle: "Breast Cancer Data: Kaplan-Meier, Exponential and Weibull Models"
date: today
---

## Overview

This script demonstrates survival analysis techniques on breast cancer data.

## Data Loading

```{r}
library(conflicted)
library(survival)

bcancer <- read.csv("data/survival.csv")

head(bcancer)
tail(bcancer)
```

Status is a "non-censored" indicator:
- 0 if the person is right censored
- 1 if it is uncensored

Group is for staining status: positive (P) or negative (N) staining.

## Kaplan-Meier Estimation

Fit KM estimator for the survival function:

```{r}
cancer.km <- survfit(Surv(time, status) ~ group, data = bcancer)

cancer.km
summary(cancer.km)
```

## Graphical Parameters

Set up graphical parameters:

```{r}
par()$las

orig_par <- par(mar = c(3, 3, 1, 1), las = 1)
par()$las
orig_par$las
orig_par$mar
par()$las
par()$mar
```

`las` = 0 by default. `las` = 1 adjusts to horizontal axis labels, even for the y-axis. This will assign new graphical parameter settings, but also store the original graphical parameters to be retrieved later.

Plot KM survival curves for the two groups:

```{r}
plot(cancer.km, lwd = 2, cex.axis = 1.5)
par(orig_par)
```

## Weibull Suitability Check

Check Weibull suitability:

```{r}
opar <- par(mar = c(4, 4, 0, 0), las = 1)
negative <- bcancer$group == "N"
positive <- bcancer$group == "P"

plot(
    log(cancer.km$time),
    log(-log(cancer.km$surv)),
    type = "n",
    xlab = "Log(time)",
    ylab = "Log(-Log(S(time)))"
)

lines(
    log(cancer.km$time[negative]),
    log(-log(cancer.km$surv[negative])),
    col = "darkgreen",
    type = "b"
)

lines(
    log(cancer.km$time[positive]),
    log(-log(cancer.km$surv[positive])),
    col = "purple",
    type = "b"
)

par(opar)
par()$las
par()$mar
```

## Proportional Hazards Assumption

Check PH assumption with cloglog plot:

```{r}
opar <- par(mar = c(2, 2, 0, 0), las = 1)

plot(
    cancer.km,
    fun = "cloglog",
    lty = 1:2,
    col = c("Red", "Blue"),
    xlim = c(5, 300)
)

legend(
    "topleft",
    legend = c("Negative Stain", "Positive Stain"),
    lty = 1:2,
    col = c("Red", "Blue"),
    bty = "n"
)

par(opar)
```

## Exponential Regression

```{r}
cancer_exp <- survreg(
    Surv(time, status) ~ group,
    data = bcancer,
    dist = "exponential"
)

summary(cancer_exp)
```

### Wald Test

```{r}
wald_stat <- cancer_exp$coeff[2] / sqrt(cancer_exp$var[2, 2])
2 * pnorm(-abs(wald_stat))
```

p-value

### Likelihood Ratio Test

```{r}
lr_stat <- -2 * (cancer_exp$loglik[1] - cancer_exp$loglik[2])
pchisq(q = lr_stat, df = 1, lower.tail = FALSE)
```

p-value

### Rate Parameters

```{r}
exp(-cancer_exp$coef[1])

exp(-sum(cancer_exp$coef))
```

### Verification: MLE Formula

For exponential distribution, MLE of lambda is # events / total time at risk:

```{r}
sum(bcancer$status[bcancer$group == "N"]) /
    sum(bcancer$time[bcancer$group == "N"])

sum(bcancer$status[bcancer$group == "P"]) /
    sum(bcancer$time[bcancer$group == "P"])
```

### Delta Method Confidence Interval

```{r}
beta <- cancer_exp$coeff[2]
var_beta <- (cancer_exp$var)[2, 2]
var_delta <- (exp(-beta))^2 * var_beta

exp(-beta) + qnorm(0.975) * sqrt(var_delta)
exp(-beta) - qnorm(0.975) * sqrt(var_delta)
```

The lower bound doesn't cross to the negative number region, which is lucky.

Alternative using monotone transformation:

```{r}
exp(-(beta - qnorm(0.975) * sqrt(var_beta)))
exp(-(beta + qnorm(0.975) * sqrt(var_beta)))
```

This guarantees that the lower bound doesn't cross to the negative number region.

## Weibull Regression

```{r}
cancer_wbl <- survreg(
    Surv(time, status) ~ group,
    data = bcancer,
    dist = "weibull"
)

summary(cancer_wbl)
cancer_wbl$coef
cancer_wbl$scale
cancer_wbl$var
```

### Rate and Shape Parameters

```{r}
lambda_n <- exp(-cancer_wbl$coef[1])
lambda_p <- exp(-sum(cancer_wbl$coef))

alpha <- 1 / cancer_wbl$scale

s_n_100 <- exp(-(lambda_n * 100)^alpha)
s_p_100 <- exp(-(lambda_p * 100)^alpha)

lambda_n
lambda_p
alpha
s_n_100
s_p_100
```

### Hazard Ratio

```{r}
hazard_ratio <- (lambda_p / lambda_n)^alpha
hazard_ratio
```

### Wald Test for Shape Parameter

```{r}
log(cancer_wbl$scale) / sqrt(cancer_wbl$var[3, 3])

2 * pnorm(-abs(log(cancer_wbl$scale) / sqrt(cancer_wbl$var[3, 3])))
```

p-value

### Likelihood Ratio Test

```{r}
2 * (cancer_wbl$loglik[2] - cancer_exp$loglik[2])

qchisq(.95, df = 1)

pchisq(
    2 * (cancer_wbl$loglik[2] - cancer_exp$loglik[2]),
    df = 1,
    lower.tail = FALSE
)
```
